### 2016-10-30 TexArxiv/clgm.tex
added TexArxiv folder

\include{clgm_def}
\begin{document}

\title{Rao-Blackwellised Two-filter algorithm for Conditionally Linear and Gaussian models}
\date{}

\author{Ngoc Minh Nguyen\footnote{LTCI, CNRS and T\'el\'ecom ParisTech, 46 rue Barrault 75634 Paris Cedex 13, France.}, Sylvain {L}e {C}orff\footnote{Laboratoire de Math\'ematiques d'Orsay, Univ. Paris-Sud, CNRS, Universit\'e Paris-Saclay, 91405 Orsay, France.} and Eric Moulines\footnote{Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique, France.}}

\lhead{Ngoc Minh Nguyen et al.}
\rhead{Conditionally Linear and Gaussian models}

\maketitle

\begin{abstract}
Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for conditionally linear and Gaussian models. The most widely used Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{kim:1994,lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the Forward Backward decomposition of the marginal smoothing distribution. In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps without any approximation in both forward and backward passes. In this paper, these algorithms are compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance. The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield. Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.

%In this paper, commodity markets are described using a two factor model based on the spot price and a convenience yield representing the benefit or cost required for holding the physical commodity. It is assumed that the market structure has different regimes which depend on existing stocks and contracts or other externalities. These different regimes are modeled using a finite state space unobserved Markov chain, each regime corresponding to an  equilibrium behavior of the spot price and the convenience yield. At each time step, it is assumed that noisy observations of future contracts are available which leads to represent commodity markets dynamics using a conditionally linear and Gaussian state space model. We propose a new algorithm to approximate the conditional distributions of the states (spot, convenience yield and regime indicator) given a set of observations. These approximations are obtained combining Sequential Monte Carlo methods and Kalman smoothing techniques  and may be used to detect significant regime switching times and to estimate model parameters with an Expectation Maximization based algorithm. The performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.
\end{abstract}

\section{Introduction} \label{sec:intro}
%Modeling commodity prices is a crucial step to valuate contingent claims related to energy markets and to optimize storage or extraction strategies. In \cite{gibson:schwartz:1990,schwartz:1997}, the authors proposed a model where the spot price of a commodity$(S_t, t \geq 0)$ depends on a second factor $(\delta_t, t \geq 0)$, referred to as the instantaneous convenience yield. This factor plays the role of dividends in equity markets and models the benefit of holding the physical commodity or the storage and maintenance costs required to keep the commodity. In this model, this convenience yield is described as an Ornstein-Uhlenbeck process:
%\begin{align*}
%\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma S_{t} \rmd W_{t}^{1} \eqsp,\\
%\rmd \delta_{t} & = \kappa(\alpha-\delta_{t})\rmd t+\eta \rmd W_{t}^{2} \eqsp, \quad \rmd \langle W_t^1, W_t^2\rangle = \rho \rmd t\eqsp,
%\end{align*}
%where the parameter $\left(r, \sigma, \kappa, \alpha, \eta, \rho \right)$ are constant and $( (W_t^1,W_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$. A similar model may be written under the risk-neutral probability $\mathbb{Q}$ with an obvious transformation of the parameters. This convenience yield based model has been extended in \cite{casassus:colin:dufresne:2005,liu:tang:2011}. In this model, the convenience yield is meant to revert to its equilibrium value $\alpha$ at speed $\kappa$. 
%
%
%This model appear to be too restrictive as energy markets are not likely to revert to a single equilibrium value. This assumption is relaxed using Markov switching models to allow several possible regimes for the spot price and the convenience yield. The regime switching autoregressive model introduced in \cite{hamilton:1988,hamilton:1989} for the term structure of interest rates has been generalized by \cite{kim:1994} to general state-space models. This Markov switching model also extends the dynamic linear model of \cite{shumway:stoffer:1991} by using a time dependent switching process and introducing regime changes in the observation model. Regime switching models based on an unobserved Markov chain have been widely used to represent commodity prices, see for instance \cite{fong:see:2002} for crude oil data or \cite{mount:ning:cai:2006} for electricity prices. See also \cite{naifar:dohaiman:2013} for a two-regime model (crisis regime and non-crisis regime) to obtain the evidence of a significant correlation between crude oil prices and short-term interest rate during the financial crisis. In \cite{chiarella:clewlow:kang:2009}, a two-factor model for the gas forward price is introduced to model where the parameters of the volatility function may change depending on the market structure. In order to capture gas price dynamics, \cite{chen:forsyth:2010} proposed a single factor model for the crude oil spot price using a mean reverting process with two equilibrium levels. Following \cite{almansour:2016}, the spot price and convenience yield are described in this paper as:
%\begin{align*}
%\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd W_{t}^{1} \eqsp,\\
%\rmd\delta_{t} & = \kappa_{a_t}(\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
%\end{align*}
%where $(a_t)_{t\ge 0}$ is a finite state space Markov process. 
%
%This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield. A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).

%Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_k= (\ln S_k, \delta_k)$ (the sampling period is taken to be 1) is modeled as a 
State space models are bivariate stochastic processes $\{(Y_k,Z_k)\}_{k\ge 1}$ where the state sequence $\{Z_k\}_{k\ge 1}$ is a Markov chain which is only partially observed through the sequence $\{Y_k\}_{k\ge 1}$. Conditionally on the state sequence $\{Z_k\}_{k\ge 1}$ the observations are independent and for all $\ell\ge 1$ the conditional distribution of $Y_{\ell}$ given $\{Z_k\}_{k\ge 1}$ depends on $Z_{\ell}$ only. These models are used in a large variety of disciplines such as econometrics, target tracking, biology, signal processing, see \cite{delmoral:2013} and the references therein. In general state space models, bayesian filtering and smoothing problems, i.e. computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks required for instance  to perform maximum likelihood inference. Filtering refers to the estimation of the distributions of $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time k, while smoothing is the estimation of the distribution of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$. These distributions are in particular crucial to infer a sequence of states given the observations in real world online learning applications, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications... In the case where the state and observation models are linear and Gaussian, the filtering and smoothing problems may be solved optimally using the Kalman filter or the Rauch-Tung-Striebel smoother of \cite{kalman:1960} and \cite{rauch:striebel:tung:1965}. Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters or Gaussian sum filters...

This paper focuses on a special class of state space models, known as Conditionally Linear and Gaussian Models (CLGM), of the form:
\begin{equation}
\label{eq:model:state}
Z_k = d_{a_{k}} + T_{a_{k}}Z_{k-1}+H_{a_{k}}\varepsilon_k\eqsp,
\end{equation}
where
\begin{enumerate}[-]
\item $(\varepsilon_k)_{k\ge 0}$ is a sequence of independent and identically distributed (i.i.d.) standard Gaussian variables,
\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in $\{1,\ldots,J\}$.  $Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
\item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
\item $Z_1$ is a Gaussian random vector of size $\dimz$ with mean $\mu_1$ and variance $\Sigma_1$ and is independent of $(\varepsilon_k)_{k\ge 0}$. 
\end{enumerate}
At each time step, the observation $Y_k$ is given by:
\begin{equation}
\label{eq:model:obs}
Y_k = c_{a_k} + B_{a_k}Z_k + G_{a_k}\eta_k\eqsp,
\end{equation}
where
\begin{enumerate}[-]
\item $(\eta_k)_{k\ge 0}$ is a i.i.d. sequence of standard Gaussian variables, independent of $(\varepsilon_k)_{k\ge 0}$ and $Z_1$.
\item $(G_j)_{1\le j \le J}$ are $\dimy \times \dimy$ positive-definite matrices and $(B_j)_{1\le j \le J}$ are $\dimy \times \dimz$ matrices.
\end{enumerate}
These switching models based on an unobserved Markov chain have been widely used in target tracking, underwater navigation or in finance to represent for instance commodity prices which are likely to go through regimes shifts, see for instance \cite{fong:see:2002} for crude oil data or \cite{mount:ning:cai:2006} for electricity prices. In this paper, the regime switching approach is applied to Chicago Mercantile Exchange crude oil data (see Section~\ref{sec:numerical:experiments}).

An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian. To exploit this specific structure using Rao Blackwellisation techniques, an alternative approach to linearization techniques to solve filtering and smoothing problems is to use Sequential Monte Carlo (SMC) methods. These methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}. By using the tractable substructure of CLGM \cite{chen:liu:2000,doucet:godsill:andrieu:2000,doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005} introduced several SMC methods combined with Kalman filters to address the filtering problem. These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case. However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables. Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}. However, only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps. Instead of using a structural approximation of the model such as in \cite{kim:1994}, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions. In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored. The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass. \cite{briers:doucet:maskell:2010} proposed another Rao Blackwellised smoother which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.  This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques. In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$. In this paper, we propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables. This explicit marginalization provides an additional Rao Blackwellisation step and produce more robust approximations.

The paper is organized as follows. The proposed Monte Carlo algorithm to estimate the marginal smoothing distributions of the hidden states is presented in Section~\ref{sec:RaoBlackwell} along with the other widely used Rao Blackwellised smoothing algorithms. The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data. Then, the parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.  
The performance of this algorithm is assessed using simulated data and CME crude oil (WTI) data. Intermediate computations and explicit approximations provided by the algorithm are given in the Appendix.

\section{Rao Blackwellised smoothing algorithms}
\label{sec:RaoBlackwell}
This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$. For all $m\,\times\,m$ positive-definite matrix $A$ and  all $z\in\mathbb{R}^{m}$ define
\[
\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp.
\]
Let $m_{}(a_{i},z_{i-1};z_i)$ be the probability density of the conditional distribution of $Z_i$ given $(a_{i},Z_{i-1})$ and $g_{}(a_{i},z_{i};y_i)$ be the probability density of the conditional distribution of $Y_i$ given $(a_{i},Z_{i})$:
\begin{align}
\label{eq:definition-m}
m_{}(a_{i},z_{i-1};z_i) & \eqdef \left|2\pi\barH_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_i}}{z_i -d_{a_{i}} -T_{a_{i}}z_{i-1}}^2\right\}\eqsp,\\
\label{eq:definition-g}
g_{}(a_{i},z_{i};y_i) & \eqdef \left|2\pi\barG_{a_i}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barG_{a_i}}{y_i - c_{a_i} - B_{a_i}z_{i}}^2\right\}\eqsp,
\end{align}
where
\[
\barG_j \eqdef G_jG'_j\;,\; \barH_{j} \eqdef H_jH'_j\eqsp.
\]
%Maximum likelihood estimation of the parameters using the Expectation Maximization algorithm or gradient ascent algorithms requires to compute the posterior distributions of the states $a_{i-1:i},z_{i-1:i}$  given the observations $Y_{1:n}$, written $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$, which cannot be computed in closed-form (the dependency of $\phi^n_{i}$ on $Y_{1:n}$ is dropped for better clarity). SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$. 
All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distribution and share the same forward filter detailed in Section~\ref{sec:forward}.

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  Labels trajectories at time $i-1$ are defined by:
\[
a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
\]
where
\[
\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
\]
Similarly, means and variances trajectories are defined by: 
\[
\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
\]
The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
$\mu_i^k$ and $P_i^k$ are updated according to:
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

%The conditional distribution of $Y_1$ given $a^1_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^j$ may be written:
%\begin{align*}
%\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
%& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
%\end{align*}
%\begin{algorithm}
%\DontPrintSemicolon
%\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
%\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
%{\bf Initialization}\;
%\For{$k \gets 1$ \textbf{to} $N$} {
%Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
%Compute $\mu_1^k$ and $P_1^k$:
%\begin{align*}
%K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
%\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
%P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
%\end{align*}
%Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
%}
%{\bf Iterations}\;
%\For{$i \gets 2$ \textbf{to} $n$} {
%Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
%\For{$k \gets 1$ \textbf{to} $N$} {
%For all $j \in \{1,\ldots, J\}$ compute
%\[
%\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
%\]
%Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
%Update $\mu_i^k$ and $P_i^k$:
%\begin{align*}
%\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
%P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
%\end{align*}
%Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
%}
%}

%The recursive updates of $(a^k_{1:i})_{1\le k \le N}$, $(\omega^k_{i})_{1\le k \le N}$,  $\mu^k_{i}$ and $P^k_{i}$ are  displayed in Algorithm~\ref{algo:forward}.
%In the initialization step, the first weight is given, for all $1\le k \le N$, by $\omega^k_1 = p(y_1|a^k_{1})$. By \eqref{eq:model:obs}, the conditional distribution of $Y_1$ given $a_1$ is a Gaussian distribution with mean $c_{a_1}+B_{a_1}\mu_1$ and variance $B_{a_1}\Sigma_1B'_{a_1} + \barG_{a_1}$. This yields:
%\[
%\omega^k_{1} \propto \exp\left\{-\frac{1}{2}\normMat{B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}}{Y_1-c_{a^k_1}-B_{a_1^k}\mu_1}^2\right\} \eqsp,
%\]

%By Algorithm~\ref{algo:forward}, $p^N_{}(a_{i},z_{i}|y_{1:i})$ is given by:
%\begin{equation*}
%p^N_{}(a_{i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{i}}(a_{i})\omega^k_{i}|2\pi P^k_{i}|^{-1/2}\times \exp\left\{-\frac{1}{2}\normMat{P_{i}^k}{z_{i}-\mu^k_{i}}^2\right\}\eqsp.
%\end{equation*}

\subsection{Path-space smoother of \cite{kitagawa:1996}}
As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step. In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
\[
p^{\mathsf{Kit}}_N(a_{1:n}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
\]
In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
\[
p^{\mathsf{Kit}}_N(a_{i}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
\]
This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps.

\subsection{Algorithm based on the structural approximation of \cite{kim:1994}}
Another method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{kim:1994} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$. Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available. Then, the approximation of \cite{kim:1994} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
\[
p(a_i|a_{i+1},y_{1:i}) \propto Q(a_i,a_{i+1})p(a_i|y_{1:i})\eqsp.
\]
Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$. $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$. Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$. If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the indicator variables is:
\[
p^{\mathsf{Kim}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
\]

\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm. The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
\[
p(a_{1:n}|y_{1:n}) = p(a_{1:i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp,
\]
where
\[
p(a_{1:i}|a_{i+1:n},y_{1:n}) \propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})p(a_{1:i}|y_{1:i})\eqsp.
\]
The second factor in the last equation may be approximated using the trajectories $(a^k_{1:i})_{1\le k \le N}$ and importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter. Therefore, $p(a_{1:i}|a_{i+1:n},y_{1:n})$ may be approximated by:
\[
p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a^k_{1:i})\eqsp,
\]
with $\tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$. Then, a trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$ may be sampled following these steps:
\begin{enumerate}[-]
\item Set $\tilde{a}_n= a_n^k$ with probability $\omega_n^k$, $1\le k \le N$.
\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probability $\tilde{\omega}_i^k$, $1\le k \le N$.
\end{enumerate}
This algorithm requires to compute the quantity  $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$. As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using backward Kalman filtering techniques from time $n$ to time $i$. However, this has to be done for each trajectory $a^k_{1:i}$, $1\le k \le N$, which leads to an algorithm with a prohibitive computational complexity. \cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a procedure computationally less intensive by noting that:
\begin{equation}
\label{eq:FFBS:incremental}
p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) = \int p(y_{i+1:n},a_{i+1:n}|z_i,a^k_{i})p(z_{i}|a^k_{1:i},y_{1:i})\rmd z_i\eqsp,
\end{equation}
which is the two-filter decomposition of the smoothing distribution. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
\[
p(y_{i+1:n},a_{i+1:n}|z_i,a_{i}) \propto Q(a_i,a_{i+1}) \exp\left\{-(z_i'\Omega_iz_i-2\lambda_i'z_i)/2\right\}\eqsp,
\]
where the proportionality is with respect to $(a_i,z_i)$ and
\[
p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\hat{\Omega}_iz_i-2\hat{\lambda}_i'z_i)/2\right\}\eqsp,
\]
where the proportionality is with respect to $z_i$. 
\begin{align*}
\hat{\Omega}_n &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
\hat{\lambda}_n &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
\end{align*}
Then, for $1\le i\le n-1$, define $m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\hat{\Omega}_{i+1}H_{a_{i+1}} + I$. Then,
\begin{align*}
\Omega_i &= T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\hat{\Omega}_{i+1}T_{a_{i+1}}\eqsp,\\
\lambda_i &=T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
\end{align*}
As $p(y_{i:n},a_{i+1:n}|z_i,a_{i}) = p(y_{i}|z_i,a_{i})p(y_{i+1:n},a_{i+1:n}|z_i,a_{i})$,
\begin{align*}
\hat{\Omega}_i &= \Omega_i + B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
\hat{\lambda}_i & = \lambda_i + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
\end{align*}
Then, by \eqref{eq:FFBS:incremental},
\[
p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})\propto Q(a_i^k,a_{i+1})|\Lambda_i|^{-1/2}\exp\left\{\eta_i/2\right\}\eqsp,
\]
where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$. These backward simulation steps are summarized in Algorithm~\ref{algo:FFBS}.

\subsection{Rao Blackwellised Two-filter Smoother}
Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}. Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods. For $3\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
\begin{equation}
\label{eq:dens:bwd_fwd}
%\phi^n_{i}(a_{i-1:i},z_{i-1:i})  =  \sum_{a_{i-2}}\sum_{a_{i+1}} \int_{z_{i-2}} \int_{z_{i+1}} \psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
\phi^n_{i}(a_{i-1:i},z_{i-1:i})  =  \sum_{a_{i-2}}\sum_{a_{i+1}} \int_{z_{i-2}} \int_{z_{i+1}} \psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
\end{equation}
where $\psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-2$ and $i+1$.
It may be decomposed as follows:
\begin{multline*}
\psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1}) \propto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})p_{}(a_{i-2},z_{i-2}|y_{1:i-2})Q(a_{i-2},a_{i-1})\\
\times \left\{\prod_{k=i}^{i+1} m_{}(a_{k-1},z_{k-2};z_{k-1})g_{}(a_{k-1},z_{k-1};y_{k-1})Q(a_{k-1},a_k)\right\}m_{}(a_{i+1},z_{i};z_{i+1})\eqsp,
\end{multline*}
where the proportionality is with respect to $(a_{i-2:i+1},z_{i-2:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ may be written as
\begin{multline}
\label{eq:smoothing}
\phi^n_{i}(a_{i-1:i},z_{i-1:i})  \propto r_{i}(a_{i-1},z_{i-1},y_{1:i-2})m_{}(a_{i},z_{i-1};z_i)g_{}(a_{i-1},z_{i-1};y_{i-1})Q(a_{i-1},a_i)\\
\times g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
\end{multline}
where $m$ and $g$ are defined in \eqref{eq:definition-m} and \eqref{eq:definition-g},  $Q$ is the transition matrix of the
Markov chain $\{a_k, k \geq 0\}$ and
\begin{align}
r_{i}(a_{i-1},z_{i-1},y_{1:i-2}) &= p_{}(a_{i-1},z_{i-1}|y_{1:i-2})\eqsp,\\
&=\sum_{a_{i-2}}\int_{z_{i-2}}p_{}(a_{i-2},z_{i-2}|y_{1:i-2})m_{}(a_{i-1},z_{i-2};z_{i-1})Q(a_{i-2},a_{i-1}) \rmd z_{i-2}\eqsp,\label{eq:smooth:r}\\
t_{i}(a_i,z_i,y_{i:n}) &= \sum_{a_{i+1}}\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp.\label{eq:smooth:t}
\end{align}
In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of states $a_{1:i-2}^k$ associated with importance weights $\omega_{i-2}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-2},z_{i-2}|y_{1:i-2})$. 
Plugging this approximation into \eqref{eq:smooth:r} provides a non normalized approximation $r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ of $r_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ integrating over all possible choices of $(a_{i-2},z_{i-2})$:
\[
r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2}) = \sum_{a_{i-2}}\int_{z_{i-2}}p^N_{}(a_{i-2},z_{i-2}|y_{1:i-2})m_{}(a_{i-1},z_{i-2};z_{i-1})Q(a_{i-2},a_{i-1}) \rmd z_{i-2}\eqsp.
\]
Define
\begin{align*}
\Upsilon^k_{i-2}(a_{i-1}) &\eqdef (P^k_{i-2})^{-1} + T'_{a_{i-1}}\barH_{a_{i-1}}^{-1}T_{a_{i-1}}\eqsp,\\
%\upsilon^k_{i-2}(a_{i-1})&\eqdef (P^k_{i-2})^{-1}\mu^k_{i-2} +  T'_{a_{i-1}}\barH_{a_{i-1}}^{-1}(z_{i-1}-d_{a_{i-1}})\eqsp,\\
\upsilon^k_{i-2}(a_{i-1})&\eqdef \barH_{a_{i-1}}^{-1}T_{a_{i-1}}(\Upsilon^k_{i-2})^{-1}(a_{i-1})(P^k_{i-2})^{-1}\mu^k_{i-2}\eqsp,\\
\Omega^k_{i-2}(a_{i-1}) &= \left(\barH_{a_{i-1}}^{-1} - \barH_{a_{i-1}}^{-1}T_{a_{i-1}}(\Upsilon^k_{i-2})^{-1}(a_{i-1})T'_{a_{i-1}}\barH_{a_{i-1}}^{-1}\right)^{-1}\eqsp,\\
r_{i-2}^k(a_{i-1}) &= \upsilon^k_{i-2}(a_{i-1})+(\Omega^k_{i-2}(a_{i-1}))^{-1}d_{a_{i-1}}\eqsp,\\
\omega_{\mathsf{f},i}^k(a_{i-1}) &\eqdef \omega^k_{i-2}Q(a^k_{i-2},a_{i-1})\left|P^k_{i-2}\right|^{-1/2}\left|\Upsilon^k_{i-2}(a_{i-1})\right|^{-1/2} \left|2\pi\barH_{a_{i-1}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{d_{a_{i-1}}}^2\right\}\\
&\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-2}^k}{\mu^k_{i-2}}^2+\frac{1}{2}\normMat{P_{i-2}^k\Upsilon^k_{i-2}(a_{i-1})P_{i-2}^k}{\mu^k_{i-2}}^2 - (\upsilon^k_{i-2})'(a_{i-1})d_{a_{i-1}} \right\}\eqsp.
\end{align*}
Then,
\begin{equation}
\label{eq:rN}
r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i-1})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{z_{i-1}}^2+z'_{i-1}r_{i-2}^k(a_{i-1})\right\}\eqsp.
%\times\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{z_{i-1}-d_{a_{i-1}}}^2\right\}\eqsp.%q(a^k_{i-2},a_{i-1})\omega^k_{i-2}\left|P^k_{i-2}\right|^{-1/2}\left|\Upsilon^k_{i-2}(a_{i-1})\right|^{-1/2}\exp\left\{ -\frac{1}{2}\normMat{P_{i-2}^k}{\mu^k_{i-2}}^2\right\}\\
%\times \left|2\pi\barH_{a_{i-1}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_{i-1}}}{z_{i-1}-d_{a_{i-1}}}^2\right\}\exp\left\{\frac{1}{2}(\upsilon^k_{i-2})'(a_{i-1})(\Upsilon^k_{i-2})^{-1}(a_{i-1})\upsilon^k_{i-2}(a_{i-1})\right\}\eqsp.
\end{equation}

The backward pass described below produces a sequence of states $\tilde{a}_{i+1:n}^{\ell}$ associated with importance weights $\tilde{\omega}_{i+1}^{\ell}$, $1\le \ell\le N$ which are used to approximate $p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$. Plugging this approximation into \eqref{eq:smooth:t} provides a non normalized approximation $t^N_{i}(a_i,z_i,y_{i:n})$ of $t_{i}(a_i,z_i,y_{i:n})$ integrating over all possible choices $(a_{i+1},z_{i+1})$.

These steps are then combined to form the normalized SMC approximation $\phi^{n,N}_{i}(a_{i-1:i},z_{i-1:i})$ of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ using \eqref{eq:smoothing}. As $r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ and $t^N_{i}(a_i,z_i,y_{i:n})$ are not normalized, the normalization of $\phi^{n,N}_{i}(a_{i-1:i},z_{i-1:i})$ is obtained by integrating over the states $a_{i-1:i},z_{i-1:i}$, when $r_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ and $t_{i}(a_i,z_i,y_{i:n})$ are replaced by $r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ and $t^N_{i}(a_i,z_i,y_{i:n})$ in \eqref{eq:smoothing}. %These combination and normalization steps are given in Appendix~\ref{sec:approx:smooth}.
Our procedure allows to construct sequence of regimes with non-degenerated importance weights in the combination step. This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ over the states at times $i-2$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged. This method often leads to importance weights which are close to be numerically degenerated.

%\subsection{Backward filter}
%\label{sec:backward}
%This section displays a SMC algorithm to approximate $t_{i}(a_i,z_i,y_{i:n})$ using backward recursions. 
As the function $(a_{i+1},z_{i+1})\mapsto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$ is not a probability density function it is not possible to use SMC samples to approximate it. The backward filter relies on artificial densities to introduce a surrogate target density function which may be approximated using SMC methods. Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:
\[
\tilde{p}_{n}(a_n,z_n,y_n) \eqdef \gamma^{}_{n}(a_n,z_n)g_{}(a_n,z_n;y_n)\eqsp,\quad \tilde{p}_{n}(y_n)\eqdef \sum_{a_n=1}^J\int \gamma^{}_{n}(a_n,z_n)g_{}(a_n,z_n;y_n)\rmd z_n\eqsp,
\]
and, for all $1\le i\le n-1$,
\begin{align*}
\tilde{p}_{i}(a_{i:n},z_{i:n},y_{i:n}) &\eqdef \gamma^{}_{i}(a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)\eqsp,\\
\tilde{p}_{i}(y_{i:n})&\eqdef \sum_{a_{i:n}=1}^J\int \gamma^{}_{i}(a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)\rmd z_{i:n}\eqsp.
\end{align*}
By Lemma~\ref{eq:technical:twofilter}, these probability densities may be used to approximate the quantities $p_{}(y_{i:n}|a_{i},z_{i})$, $1\le i \le n$, in \eqref{eq:smooth:t}.
\begin{lemma}
\label{eq:technical:twofilter}
For all $1\le i \le n-1$,
\begin{align}
\tilde{p}_{i}(a_i,z_i|y_{i:n}) &= p_{}(y_{i:n}|a_i,z_i)\gamma^{}_{i}(a_i,z_i)/\tilde{p}_{i}(y_{i:n})\eqsp,\label{eq:p:y1:n}\\
\tilde{p}_{i}(a_{i},z_{i}|y_{i:n}) &= \gamma^{}_i(a_i,z_i)\sum_{a_{i+1:n}=1}^J\frac{\tilde{p}_{i}(a_{i:n}|y_{i:n})p_{}(y_{i:n}|a_{i:n},z_{i})}{\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'}\label{eq:p:ai:zi}\eqsp.
\end{align}
\end{lemma}
\begin{proof}
The proof is postponed to Appendix~\ref{sec:app:proofs}.
\end{proof}
The SMC approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$, described in Algorithm~\ref{algo:backward},  produces sequence of states $\tilde{a}^{\ell}_{i:n}$ associated with importance weights $\tilde{\omega}^{\ell}_i$ for $1\le \ell\le N$ to define a sample based approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$ of the form $\tilde{p}^N_{}(a_{i:n}|y_{i:n}) \eqdef \sum_{\ell=1}^N \tilde{\omega}^{\ell}_i \delta_{\tilde{a}^{\ell}_{i:n}}(a_{i:n})$. By \eqref{eq:p:y1:n} and \eqref{eq:p:ai:zi}, we get
\begin{align*}
p_{}(y_{i:n}|a_i,z_i)
&= \frac{\tilde{p}_{i}(y_{i:n})  \tilde{p}_{i}(a_i,z_i|y_{i:n})}{\gamma^{}_{i}(a_i,z_i)} \\
&= \tilde{p}_{i}(y_{i:n})  \sum_{a_{i+1:n}=1}^J\frac{\tilde{p}_{i}(a_{i:n}|y_{i:n})p_{}(y_{i:n}|a_{i:n},z_{i})}{\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'}\label{eq:p:ai:zi}\eqsp,
\end{align*}
which suggests the following particle approximation $p^N_{}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$:
\[
p^N_{}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_{i}(y_{i:n}) \sum_{\ell=1}^N\frac{\tilde{\omega}^{\ell}_i p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z_{i})}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}\delta_{\tilde{a}^{\ell}_{i}}(a_{i})\eqsp.
\]
The conditional likelihood of the observations given the sequence of states $p_{}(y_{i:n}|a_{i:n},z_{i})$ can be computed explicitly using a Gaussian backward smoother; these computations are summarized in Lemma~\ref{lem:py}.
\begin{lemma}
\label{lem:py}
For all $1\le i \le n$,
\begin{equation}
\label{eq:py}
p_{}(y_{i:n}|a_{i:n},z_i) =  \exp\left\{-\frac{1}{2}\tilde{c}_{i}(a_{i:n}) - \frac{1}{2}\normMat{\tilde{P}_{i}(a_{i:n})}{z_i}^2 + z'_i\tilde{\nu}_{i}(a_{i:n})\right\}\eqsp,
\end{equation}
where
\begin{align}
\tilde{c}_{n}(a_{n}) &= \dimy\log(2\pi) + \log \left|\barG_{a_n}\right| + \normMat{\barG_{a_n}}{y_n-c_{a_n}}^2\eqsp,\label{eq:cn}\\
\tilde{P}_{n}^{-1}(a_{n}) &=B'_{a_n}\barG_{a_n}^{-1}B_{a_n}\eqsp,\label{eq:Pn}\\
\tilde{\nu}_{n}(a_{n})&=B'_{a_n}\barG_{a_n}^{-1}(y_n-c_{a_n})\label{eq:nun}
\end{align}
and, for all $1\le i\le n-1$,
\begin{align}
\tilde{c}_{i}(a_{i:n}) &= \tilde{c}_{i|i+1}(a_{i:n}) + \dimy\log(2\pi)  + \log|\barG_{a_i}| + \normMat{\barG_{a_i}}{y_i-c_{a_i}}^2\eqsp,\label{eq:ck}\\
\tilde{P}_{i}^{-1}(a_{i:n}) &= \tilde{P}_{i|i+1}^{-1}(a_{i:n}) + B'_{a_i}\barG_{a_i}^{-1}B_{a_i}\eqsp,\label{eq:Pk}\\
\tilde{\nu}_{i}(a_{i:n})&=\tilde{\nu}_{i|i+1}(a_{i:n}) + B'_{a_i}\barG_{a_i}^{-1}(y_{i}-c_{a_{i}})\eqsp,\label{eq:nuk}
\end{align}
with
\begin{align*}
\Delta_{i}(a_{i+1:n}) &= \left(I_{\dimz} + H'_{a_{i+1}}\tilde{P}_{i+1}^{-1}(a_{i+1:n})H_{a_{i+1}}\right)^{-1}\eqsp,\\
\tilde{r}_{i|i+1}(a_{i+1:n})&= \tilde{\nu}_{i+1}(a_{i+1:n}) + \barH_{a_{i+1}}^{-1}d_{a_{i+1}}\eqsp,\\
\tilde{c}_{i|i+1}(a_{i+1:n}) &= \tilde{c}_{i+1}(a_{i+1:n}) + \log|\barH_{a_{i+1}}| + d'_{a_{i+1}}\barH_{a_{i+1}}^{-1}d_{a_{i+1}}-\log|H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}|\\
&\hspace{6cm} - \tilde{r}'_{i|i+1}(a_{i+1:n})H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\tilde{r}_{i|i+1}(a_{i+1:n})\eqsp,\\
\tilde{P}_{i|i+1}^{-1}(a_{i+1:n}) &=T'_{a_{i+1}}\left(I_{\dimz}-\barH_{a_{i+1}}^{-1}H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\right)\barH_{a_{i+1}}^{-1}T_{a_{i+1}}\eqsp,\\
\tilde{\nu}_{i|i+1}(a_{i+1:n})&=T'_{a_{i+1}}\barH_{a_{i+1}}^{-1}\left[-d_{a_{i+1}}+H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\left(\tilde{\nu}_{i+1}(a_{i+1:n})+\barH_{a_{i+1}}^{-1}d_{a_{i+1}}\right)\right]\eqsp.
\end{align*}
\end{lemma}
\begin{proof}
The proof is postponed to Appendix~\ref{sec:app:proofs}.
\end{proof}
By Lemma~\ref{lem:py}, the SMC approximation $p_{}^N(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$ is then given by
\begin{equation}
\label{eq:pN:yi:n}
p_{}^N(y_{i:n}|a_{i},z_{i}) = \tilde{p}_i(y_{i:n}) \sum_{\ell=1}^N  \frac{\delta_{\tilde{a}^{\ell}_i}(a_i)\tilde{\omega}_{i}^{\ell}}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'} \exp\left\{- \frac{1}{2}\normMat{\tilde{P}^{\ell}_{i}}{z_i}^2 + z'_i\tilde{\nu}^{\ell}_{i} - \frac{1}{2} \tilde{c}^{\ell}_i \right\}\eqsp,
\end{equation}
where $(\tilde{P}^{\ell}_{i})^{-1} \eqdef \tilde{P}_{i}^{-1}(\tilde{a}^{\ell}_{i:n})$, $\tilde{\nu}^{\ell}_{i}\eqdef \tilde{\nu}_{i}(\tilde{a}^{\ell}_{i:n})$, $\tilde{c}^{\ell}_{i} \eqdef \tilde{c}^{\ell}_{i}(\tilde{a}^{\ell}_{i:n})$. Define
\begin{align*}
\Delta^{\ell}_{i} &\eqdef \left(I + H'_{\tilde{a}^\ell_{i+1}}(\tilde{P}^{\ell}_{i+1})^{-1}H_{\tilde{a}^\ell_{i+1}}\right)^{-1}\eqsp,\\
\delta^{\ell}_{i}&\eqdef\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}(d_{\tilde{a}^\ell_{i+1}}+T_{\tilde{a}^\ell_{i+1}}z_i)\eqsp.
\end{align*}
Then, by \eqref{eq:smooth:t}, the SMC approximation $t^N_{i}(a_i,z_i,y_{i:n})$ of $t_{i}(a_i,z_i,y_{i:n})$ is given by:
\begin{align}
t^N_{i}(a_i,z_i,y_{i:n}) &=\sum_{a_{i+1}=1}^J\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p^N_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp,\nonumber\\
&=  \tilde{p}_{i+1}(y_{i+1:n}) \sum_{\ell=1}^N C_i^{-1}(\tilde{a}^{\ell}_{i+1:n}) Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_i H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2} \eqsp\nonumber \\
&\hspace{4.3cm} \times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{\tilde{a}^{\ell}_{i+1}}}{d_{\tilde{a}^{\ell}_{i+1}}+T_{\tilde{a}^{\ell}_{i+1}}z_i}^2\right\} \eqsp,\\
&= \sum_{\ell=1}^N\tilde{\omega}_{\mathsf{b},i}^\ell(a_i)\exp\left\{-\frac{1}{2}\normMat{\tilde{S}_{i+1}^\ell}{z_i}+z_i'\tilde{s}_{i+1}^\ell\right\}\eqsp,\label{eq:tN}
\end{align}
where
\begin{align*}
C_i(\tilde{a}^{\ell}_{i+1:n}) &\eqdef \exp \left\{\tilde{c}^{\ell}_{i+1}/2\right\}  \int_{z_{i+1}} \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1}, z)\tilde{p}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z) \rmd z\eqsp,\\
\tilde{\omega}_{\mathsf{b},i}^\ell(a_i) &= \tilde{p}_{i+1}(y_{i+1:n})C_i(\tilde{a}^{\ell}_{i+1:n})^{-1}Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_i H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2}\\
&\times\exp\{-d'_{\tilde{a}^\ell_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}}/2\}\exp\{(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}}(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})/2\} \eqsp,\\
(\tilde{S}_{i+1}^\ell)^{-1} &= T'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}(T_{\tilde{a}^{\ell}_{i+1}}-H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}T_{\tilde{a}^{\ell}_{i+1}})\eqsp,\\
\tilde{s}_{i+1}^\ell &=T'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}(H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}}(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})-d_{\tilde{a}^\ell_{i+1}})\eqsp.
\end{align*}
In the experiments (see section~\ref{sec:exp}), $\gamma^{}_i(a_i,z_i)$ is set as a mixture of Gaussian distributions with parameters specified below. Note that for such a choice,
 the integral $\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'$ may be computed explicitly; see Lemma~\ref{lem:integral:gammap}. This choice differs from \cite{briers:doucet:maskell:2010} where it is advocated to set $\gamma^{}_i$ as the product of two independent densities $\gamma_i^{a}(a_i)$ and $\gamma_i^{z}(z_i)$.
 The accuracy of the algorithm relies heavily on a proper tuning of this artificial density: it is shown in Section~\ref{sec:exp} that fitting a mixture improves the performance of the algorithm. Let $\phi_{\mu,\Sigma}$ be the probability density function of a $\dimz$ dimensional Gaussian random variable with mean $\mu$ and variance matrix $\Sigma$.

\begin{lemma}
\label{lem:integral:gammap}
For all $1\le i\le n$,
\begin{multline*}
\int \phi_{\mu_i,\Sigma_i}(z_i)p_{}(y_{i:n}|a_{i:n},z_i)\rmd z_{i} = \exp\left\{-\frac{1}{2}\log |\Sigma_i| - \frac{1}{2}\mu_i'\Sigma^{-1}_i\mu_i\right\}\\
\times\exp\left\{-\frac{1}{2}\tilde{c}_{i}(a_{i:n})+\frac{1}{2}\log|\tilde{\Omega}_{i}(a_{i:n})| +\frac{1}{2}\tilde{z}'_i(a_{i:n})\tilde{\Omega}_{i}(a_{i:n})\tilde{z}_i(a_{i:n})\right\}\eqsp,
\end{multline*}
where
\[
\tilde{\Omega}_{i}(a_{i:n})\eqdef \left(\Sigma_i^{-1} + \tilde{P}_{i}^{-1}(a_{i:n})\right)^{-1}\quad\mbox{and}\quad
\tilde{z}_i(a_{i:n})\eqdef \Sigma_i^{-1}\mu_i+\tilde{\nu}_{i}(a_{i:n})
\]
and where $c_{i}$, $\tilde{P}_{i}$ and $\nu_{i}$ are given in Lemma~\ref{lem:py}.
\end{lemma}
\begin{proof}
The proof is postponed to Appendix~\ref{sec:app:proofs}.
\end{proof}

\section{Numerical experiments}
\label{sec:numerical:experiments}
This section highlights the performance of the Rao Blackwellised Two-Filter approximation (RBTF) of the marginal smoothing distributions using both simulated data and a real data set. This approximation is compared to the one provided by \cite{briers:doucet:maskell:2010}.  We also describe numerically the impact of the artificial law choice on the accuracy and stability of the algorithms. 
\subsection{Choice of artificial distributions}
\label{sec:artificial:distribution}
As noted in \cite{briers:doucet:maskell:2010}, the backward filter performance highly depends on the sequence $(\gamma_i)_{0\le i\le n}$. A poor choice may imply degeneracy of the weighted samples produced by the backward pass and lead to  a poor estimation of the posterior probabilities of the state variables given the observations.
%In contrast of the forward pass, the initial law of state variables are not totally observed in almost cases but we have at least an idea to guess its distribution. But it is not the case for the distribution at terminal time. Because it depends on the evolution of state variables in the past.
%We propose three artificial laws. %In the next section, we compared the performance of our proposition in backward filter in comparaison to the one of \cite{briers:doucet:maskell:2010} and also the performance of these three artificial distributions.
%The first and the third one are our propositions and the second one is the one proposed by \cite{briers:doucet:maskell:2010} from our exchange.
\begin{enumerate}[-]
\item %According to \cite{briers:doucet:maskell:2010}, a sensible choice for $\gamma_i$ is given by the prior distribution of the states $(a_i,z_i)$.
%In our case, an easy to implement solution to approximate this
A first choice for $\gamma_i$ is given by an estimation of the prior distribution of the states $(a_i,z_i)$ obtained by simulating $P$ independent chains $\left(a^{p}_{1:n}, z^{p}_{1:n} \right)_{1\le p \le P}$ using \eqref{eq:model:state}. %and \eqref{eq:model:obs}.
Then, for all $1\le i \le n$, $\gamma^{}_i$ is defined as
\begin{align*}
\gamma^{}_i (a_i, z_i) &= \sum_{p=1}^{P} m_{}(a^{p}_{i-1},z^p_{i-1}; z_{i}) Q(a^{p}_{i-1},a_i)/P\eqsp,\\
&= (2\pi)^{-\dimz/2}\sum_{p=1}^{P} Q(a^{p}_{i-1},a_i) \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^{p}_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^{p}_{i-1}}}{z_i - d_{a^p_{i-1}} -T z^{p}_{i-1}}^2\right\}/P \eqsp.
\end{align*}
%\paragraph{Last time step}
%Define the terminal proposal distribution as $\tilde{Q}_n(\tilde{a}_n^{\ell}) \propto \int_{z_n} \gamma^{}_n(\tilde{a}_n^{\ell},z_n) g_{}(\tilde{a}_n^{\ell}, z_n, y_n) \rmd z_n$. By \eqref{eq:model:obs},
%\begin{multline*}
%\tilde{Q}_n(\tilde{a}_n^{\ell})  \propto \sum_{p=1}^{P} Q(a^{p}_{n-1},\tilde{a}_n^{\ell}) \exp\left\{ -\frac{1}{2} \log \left| \tilde{\Lambda}_p \right| +\frac{1}{2}\normMat{\tilde{\Lambda}_p}{\tilde{\beta}_p}^2 - \frac{1}{2}\normMat{\barG}{y_n-c_{\tilde{a}_n^{\ell}}} \right\} \\
%\times \exp\left\{ -\frac{1}{2} \log|\barH_{a^p_{n-1}}| -\frac{1}{2}\normMat{\barH_{a^p_{n-1}}}{d_{a^p_{n-1}} + T z^p_{n-1}}\right\} \eqsp.
%\end{multline*}
%where
%\[
%\tilde{\Lambda}_p \eqdef \barH^{-1}_{a^p_{n-1}} + B'\barG^{-1} B \quad \mbox{and} \quad \tilde{\beta}_p \eqdef \barH^{-1}_{a^p_{n-1}} \left( d_{a^p_{n-1}} + T z^p_{n-1} \right) + B'\barG^{-1}(y_n-c_{a^l_{n}})\eqsp.
%\]
%For this choice of terminal proposal distribution, the corresponding particles are uniformly weighted i.e. for all $1 \le \ell \le N$, $\tilde{\omega}^{\ell}_n = 1/N$.
%\paragraph{Backward iterations}
%For $i=n-1$ to $i=n$, the proposal distribution $\tilde{q}_i$ is defined as
%\begin{align*}
%\tilde{Q}_i (\tilde a^{\ell}_{i+1:n}, \tilde a^{\ell}_i) &\propto \frac{Q(\tilde a^{\ell}_{i},\tilde a^{\ell}_{i+1}) \int_{z_i} \gamma^{}_i(\tilde a^{\ell}_{i},z_i) p_{}(y_{i:n}|\tilde a^{\ell}_{i:n},z_i) \rmd z_i }{\int_{z_{i+1}} \gamma^{}_{i+1}(\tilde a^{\ell}_{i+1},z_{i+1}) p_{}(y_{i+1:n}|\tilde a^{\ell}_{i+1:n},z_{i+1}) \rmd z_{i+1}}\eqsp,
%\end{align*}
%with, by Lemma~\ref{lem:py},
%\begin{align*}
%\int_{z_i} \gamma^{}_i(a_{i},z_i) p_{}(y_{i:n}|a_{i:n},z_i) dz_i &= \sum_{p=1}^{P} \int_{z_i} Q(a^{p}_{i-1}, a_{i}) m_{}(a^p_{i-1}, z^{p}_{i-1}; z_i) p_{}(y_{i:n}|a_{i:n},z_i) \rmd z_i/P \eqsp,\\
%&\hspace{-4.2cm} \propto \sum_{p=1}^{P} q(a^{p}_{i-1},a_i) \exp \left\{ -\frac{1}{2} \log\left| \barH_{a^p_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^p_{i-1}}}{d_{a^p_{i-1}} + T z^p_{i-1}}^2 - \frac{1}{2} c_{i}(a_{i:n}) \right\}\\
%&\hspace{-3.2cm} \times \int \exp \left\{-\frac{1}{2} z'_{i} \left( \barH^{-1}_{a^p_{i-1}} + \tilde{P}^{-1}_{i}(a_{i:n})  \right) z_i + z'_i \left( \barH^{-1}_{a^p_{i-1}} \left( d_{a^p_{i-1}} + T z^p_{i-1} \right) + \nu_{i}(a_{i:n}) \right)  \right\} \rmd z_i \eqsp, \\
%&\hspace{-3.2cm} = \sum_{p=1}^{P} Q(a^{p}_{i-1},a_i) \left| \barH_{a^p_{i-1}} \right|^{-1/2}\exp \left\{-\frac{1}{2}\normMat{\barH_{a^p_{i-1}}}{d_{a^p_{i-1}} + T z^p_{i-1}}^2 - \frac{1}{2} c_{i}(a_{i:n}) \right\} \eqsp\\
%&\hspace{-2.2cm} \times \left|R^p_{i-1}(a_{i:n})\right|^{1/2}\exp \left\{\frac{1}{2} \delta'_{i-1}(a^p_{i-1},a_{i:n}) R^p_{i-1}(a_{i:n}) \delta_{i-1}(a^p_{i-1},a_{i:n})  \right\}/P \eqsp,
%\end{align*}
%where
%\begin{align*}
%R^p_{i-1}(a_{i:n}) \eqdef \left( \barH^{-1}_{a^p_{i-1}} + \tilde{P}^{-1}_{i}(a_{i:n}) \right)^{-1} &= H_{a^p_{i-1}} \Delta_{i-1}(a^p_{i-1},a_{i:n})H'_{a^p_{i-1}}\eqsp,\\
%\delta_{i-1}(a^p_{i-1:n}) &\eqdef \left( \barH^{-1}_{a^p_{i-1}} \left( d_{a^p_{i-1}} + T z^p_{i-1} \right) + \nu_{i}(a_{i:n}) \right)\eqsp.
%\end{align*}
%For this choice of proposal distribution, the corresponding particles are uniformly weighted i.e. for all $1 \le \ell \le N$, $\tilde{\omega}^{\ell}_i = 1/N$.

%\paragraph{Application for the approximation of $t_{i}$}
%Then,
%\begin{multline}
%\label{eq:tN}
%t^N_{i}(a_i,z_i,y_{i:n}) =  \tilde{p}_{i+1}(y_{i+1:n})g(a_i,z_i,y_i) \sum_{\ell=1}^N C^{-1}_i(\tilde a^{\ell}_{i+1:n}) Q(a_i, \tilde a^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{a_i}|^{-1/2} |H_{a_i} \Delta^{\ell}_i H'_{a_i}|^{1/2} \eqsp \\
%\times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{a_i}\Delta^l_i H'_{a_i} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{a_i}}{d_{a_i}+Tz_i}\right\} \eqsp,
%\end{multline}
%where
%\begin{multline*}
%C_i(\tilde a^{\ell}_{i+1:n})  =  \sum_{p=1}^{P} Q(a^{p}_{i},\tilde a^{\ell}_{i+1}) \exp \left\{ -\frac{1}{2} \log\left| \barH_{a^p_{i}} \right| -\frac{1}{2}\normMat{\barH_{a^p_{i}}}{d_{a^p_{i}} + T z^p_{i}}^2\right\} \\
%\times \exp \left\{\frac{1}{2} \log \left|H_{a^p_{i}} \Delta^{\ell}_{i}(a^p_i) H'_{a^p_{i}}\right| + \frac{1}{2} (\delta^{\ell}_{i}(a_i^p) )'  H_{a^p_{i}} \Delta^{\ell}_{i}(a^p_i)H'_{a^p_{i}} \delta^{\ell}_{i}(a_i^p)  \right\}/P \eqsp.
%\end{multline*}
This approach may lead to poor results as it does not take into account the observations received up to time $i$. Its performance depends heavily on the number of simulations $P$ and a large value of $P$ increases the computational cost of the algorithm.
%It can be not very stable because of lack of the information from observations obtained by the forward pass. This default is corrected by the second artificial law.

%\item Another choice proposed by one of the authors of \cite{briers:doucet:maskell:2010} is the distribution of the states at time $i$ obtained from the forward pass $(\omega^{k}_i, \mu^k_{i}, P^k_{i})^{N}_{k=1}$. For all $1\le i \le n$, $\gamma^{}_i$ is defined as
%\begin{align*}
%\gamma^{}_i (a_i, z_i) &= \sum_{k=1}^{N} \delta_{a_i}(a^k_i) \omega^{k}_i \exp\left\{ -\frac{1}{2} \log \left|\bar{P}^k_{i} \right| -\frac{1}{2}\normMat{\bar{P}^k_{i}}{z_i - \mu^k_{i} }^2\right\} \eqsp.
%\end{align*}
%This approach uses the information brought by the observations and reduces the computational cost as it does not require to sample a large number of sequences of states only to compute $\gamma_i$.
%But by our experiments, using directly and exactly the forward pass information can be not a perfect choice. Since forward filter doesn't exprime partly but not exactly the distribution of state variable for given observations. The aim of backward pass is to complement the lack of forward pass. So the backward pass needs to be estimated more flexibly.

\item $\gamma_i$ may also be computed as a combination of the two previous approaches:
%At time step $i$, we store $(\mu^k_{i-1}), k=1,\ldots,N$ obtained during the forward pass of the previous step $i-1$ weighted with $\omega^k_{i-1}, k=1,\ldots,N$ and take the variance $\barH_{a^k_{i-1}}, k=1,\ldots,N$ naturally obtained from \eqref{eq:model:state}. Formally,
\begin{align*}
\gamma^{}_i (a_i, z_i) &= \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
\end{align*}
This approach is more flexible since it matches freely the forward distribution with the next time step using \eqref{eq:model:state}.
\end{enumerate}

\subsection{Simulated data}
We consider the two-factor model \eqref{eq:model:state} and \eqref{eq:model:obs} with $J=2$. The performance of the algorithms are analyzed in two different settings with $n=1500$ observations. In both settings the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data. The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data, see Section~\ref{sec:exp}.  It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$). The initial state variance is high to emphasize the fact that there is no prior precise knowledge on the state value.

\paragraph{Experiment 1:}
\[
\pi(1)=\pi(2) = 1/2\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
\]
\[
Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T = \begin{pmatrix}1 & -0.02\\ 0 & 0.96\end{pmatrix}\;\;\barH_1 = \begin{pmatrix}0.25 & 0.01\\ 0.01 & 0.01\end{pmatrix}\;\;\barH_2 = \begin{pmatrix}0.3 & 0\\ 0 & 0.3\end{pmatrix}\eqsp,
\]
\[
B = \begin{pmatrix}1 & 1 & 1 & 1\\ 0 & 0.5 & 0.5 & 0.1 \end{pmatrix}'\;\; \barG = \mathrm{diag}\left(0.1,0.1,0.1,0.1\right)\;\; \mu_1 = \begin{pmatrix}1 \\ 0\end{pmatrix}\;\; \Sigma_1 = \begin{pmatrix} 5 & 0 \\ 0  & 5\end{pmatrix}\eqsp.
\]
The approximations of the smoothing distributions are computed with $N=10$. RBTF and the algorithm of \cite{briers:doucet:maskell:2010} are compared to the approximations provided by RBTF with $1000$ particles. 
%Figure~\ref{fig:exp1:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.

\paragraph{Experiment 2:}
In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$. 
%and $B$ becomes:
%\[
%B = \begin{pmatrix}1 & 0 & 0.5 & 0 \\ 0 & 1 & 0.5 & 0\end{pmatrix}\eqsp.
%\]
%It becomes a situation more difficult because of the noisy level. 
The smoothing distributions are estimated with $N=15$ particles. Both methods are compared to the approximations provided by RBTF with $1000$ particles. 
%Figure~\ref{fig:exp2:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.
 %Figure~\ref{fig:exp1:label} and %\ref{fig:exp2:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.

\medskip

Figure~\ref{fig:exp:artificial} displays the estimated posterior distributions of the labels $(a_k)_{1\le k\le 500}$ given by RBTF for the first $500$ time steps with the  artificial distributions introduced in Section~\ref{sec:artificial:distribution}. In experiment~1, the three artificial laws behave similarly except that the first choice is unable to detect the last regime switching for more than 50 time steps. 
%In experiment 2, the second method is completely unstable which indicates that the third artificial leads to the best performance. Therefore, the third artificial law is used in the following numerical experiments.

\begin{figure}
\centering
\includegraphics[scale=.5]{Artificial_Compare_Exp1_1_500.eps}
\includegraphics[scale=.5]{Artificial_Compare_Exp2_1_500.eps}
\caption{estimated probabilities by RBTF in experiment 1 [top] and experiment 2 [bottom] with the two artificial laws: yellow (the first one) and red (the third one).}
\label{fig:exp:artificial}
\end{figure}

Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms. In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).

\begin{figure}
\centering
\includegraphics[scale=.5]{N10_1_500.eps}
\includegraphics[scale=.5]{N10_501_1000.eps}
\includegraphics[scale=.5]{N10_1001_1500.eps}
\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true labels (squared black).}
\label{fig:exp1:label}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.5]{N15_1_500.eps}
\includegraphics[scale=.5]{N15_501_1000.eps}
\includegraphics[scale=.5]{N15_1001_1500.eps}
\caption{Experiment 2 - estimated probabilites by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
\label{fig:exp2:label}
\end{figure}

\subsection{Application to CME crude oil (WTI)}
\label{sec:exp}

\subsubsection*{State dynamics}
As explained in Section~\ref{sec:intro}, the spot price and convenience yield are described in this paper as:
\begin{align*}
\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd W_{t}^{1} \eqsp,\\
\rmd\delta_{t} & = \kappa_{a_t}(\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
\end{align*}
We assume that $J=2$, each regime characterizing the slope of the future curve. A backwardation regime (resp. contango regime) is defined as a negative slope (resp. a positive slope). Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$. It is assumed that each factor (spot, convenience yield) has its own constant market price of risk, $\lambda^S$ for $S_t$ and $\lambda^{\delta}$ for $\delta_t$. The historical dynamics of the Gibson-Schwartz model is then described as follows:
\begin{align*}
\rmd S_{t} & = (\mu -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t} \rmd \widetilde{W}_{t}^{1} \eqsp,\\
\rmd \delta_{t} & = \kappa_{a_t}(\widetilde{\alpha}_{a_t}-\delta_{t})\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd \langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho \rmd t\eqsp,
\end{align*}
where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $. 
The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:
\[
Z_i = d_{a_{i-1}} + T_{a_{i-1}} Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
\]
%where (setting $\Delta t_i = t_i-t_{i-1}=1$):
%\[
%d_{j} \eqdef
%\begin{pmatrix} \left[r- \alpha_{j} - \sigma^2_{j}/2 \right]  + \alpha_{j}[1-\rme^{-\kappa}]/\kappa\\
%\alpha_{j} [1-\rme^{-\kappa}] \end{pmatrix}\eqsp, \;\; T \eqdef
%\begin{pmatrix} 1 & -[1-\rme^{-\kappa}]/\kappa \\ 0 & \rme^{-\kappa} \end{pmatrix} \eqsp,
%\]
%\begin{align*}
%\barH_{j}(1,1) &= \sigma^2_{j} + \eta^2_{j}\left\{1 + (1-\rme^{-2\kappa})/(2\kappa) - 2(1-\rme^{-\kappa})/\kappa\right\}/\kappa^2
%- 2\rho_{j} \eta_{j} \sigma_{j}\left\{ 1 - (1-\rme^{-\kappa})/\kappa \right\}/\kappa\eqsp, \\
%\barH_{j}(1,2) & = \left(\rho_{j} \eta_{j} \sigma_{j}-\eta^2_{j}/\kappa\right)\left(1-\rme^{-\kappa} \right)/\kappa + \eta^2_{j}\left(1-\rme^{-2\kappa} \right)/(2\kappa^2)\eqsp,\\
%\barH_{j}(2,1) &= \barH_{j}(1,2)\eqsp,\quad \barH_{j}(2,2) = \eta^2_{j}\left( 1-\rme^{-2\kappa} \right)/(2\kappa)\eqsp.
%\end{align*}
where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\Delta t_i = t_i-t_{i-1}$):
\begin{align*}
d_{a_{i-1}} &\eqdef
\begin{pmatrix} \left[\mu- \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}}\\
\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa_{a_{i-1}} \Delta t_i}] \end{pmatrix}\eqsp,\\
T_{a_{i-1}} &\eqdef
\begin{pmatrix} 1 & -[1-e^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}} \\ 0 & e^{-\kappa_{a_{i-1}} \Delta t_i} \end{pmatrix} \eqsp,\\
\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta t_i + \eta^2_{a_{i-1}}\left\{\Delta t_i + (1-e^{-2\kappa_{a_{i-1}} \Delta t_i})/(2\kappa_{a_{i-1}}) - 2(1-e^{-\kappa \Delta t_i})/\kappa_{a_{i-1}}\right\}/\kappa_{a_{i-1}}^2 \\
&\hspace{5.2cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-e^{-\kappa_{a_{i-1}} \Delta t_i})/\kappa_{a_{i-1}} \right\}/\kappa\eqsp, \\
\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa_{a_{i-1}}\right)\left(1-e^{-\kappa_{a_{i-1}}\Delta t_i} \right)/\kappa_{a_{i-1}} \\
&\hspace{5.2cm}+ \eta^2_{a_{i-1}}\left(1-e^{-2\kappa_{a_{i-1}} \Delta t_i} \right)/(2\kappa_{a_{i-1}}^2)\eqsp,\\
\barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp,\\
\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-e^{-2\kappa_{a_{i-1}}\Delta t_i} \right)/(2\kappa_{a_{i-1}})\eqsp.
\end{align*}
%Remarks que the equations \eqref{eq:modele} under the pricing measure $\mathbb{Q}$ becomes:
%\begin{equation}
%\begin{cases}
%\rmd S_{t} & = (r -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd \widetilde{W}_{t}^{1} \eqsp,\\
%\rmd\delta_{t} & = \left[\kappa(\alpha_{a_t} -\delta_{t}) - \lambda\right]\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd\langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho_{a_t} \rmd t\eqsp.
%\end{cases}
%\end{equation}
%where the constant $\lambda$ denotes the market price of convenience yield risk and $\widetilde{W}^1, \widetilde{W}^2$ are two $\mathbb{Q}-$Brownian motions. By introducing a new mean-reversion level for the convenience yeild process under $\mathbb{Q}$ 
%\[
%\widetilde{\alpha} = \alpha - \lambda/\kappa 
%\]
%we obtain the dynamics:
%\[
%\rmd\delta_{t} = \kappa(\widetilde{\alpha}_{a_t} -\delta_{t}) \rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2}
%\]
%The behaviors of $Z_i$ under $\mathbb{Q}$ is $Z_i = \widetilde{d}_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i 
%$ with 
%\[
%\widetilde{d}_{a_{i-1}} \eqdef
%\begin{pmatrix} \left[r - \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa \Delta t_i}]/\kappa\\
%\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa \Delta t_i}] \end{pmatrix}\eqsp,
%\]
%The transition matrix $Q$ is assumed to be the same under the pricing measure. 
\subsubsection*{Observation model}
The observations are Wednesday future contracts of the West Texas Intermediate crude oil (WTI) traded in the Chicago Mercantile Exchange (CME) from $11$ January 1995 to $13$ November 2013. The contracts are numbered $F_1, F_2, \ldots, F_{36}$ where $F_1$ (or front month) is the earliest delivery future contract, $F_2$ is the second earliest delivery future contract and so on. Among these $36$ contracts, the four future contracts: $F_1, F_4, F_6, F_{13}$ are used since their trading volumes and their impacts on the Term Structures are the most important ($F_1$ is the most liquid contract, $F_{13}$ characterizes the gap between prices over a one year period, $F_4$ and $F_6$ are intermediate future contracts that are mostly traded). As in \cite{almansour:2016}, we consider that each  future contract has a fixed time to maturity: $F_1, F_4, F_6, F_{13}$ have time to  maturity $4$, $16$, $26$, $56$ weeks.  Our time series contains $n =975$ weekly data with $534$ in backwardation and $441$ in contango (the backwardation effect is more frequent with crude oil data). At each time $t_i = i\Delta t$, with $\Delta t = 0.0192$, the observations of the $\dimy = 4$ future prices are $Y_i \eqdef (\ln(F^{(market)}_{i\Delta t,m_1}), \ldots, \ln(F^{(market)}_{i\Delta t,m_{\dimy}}))'$, where $F_{t_i,m}$ is the future price at $t_i$ for a maturity  $m$ weeks.
%In addition, the backwardation is more likely to stay in backwardation than switch to contango and the same holds for the contango regime.

\medskip

Following \cite{almansour:2016}, it is assumed that $\kappa_1 = \kappa_2 = \kappa$ (and therefore $T_1=T_2=T$) to obtain a closed form solution for $F_{t_i,m}$ which may be written: 
\[
F_{t_i,m} \eqdef \exp\left( \mathsf{A}_{m}(a_{i}) + \mathsf{B}_{m} Z_{i}\right)\eqsp,
\]
where $\mathsf{B}_0	= \begin{pmatrix}1 & 0 \end{pmatrix}$ and  $\mathsf{B}_m    = \mathsf{B}_{m-1} T$ so that  $ \mathsf{B}_m= \begin{pmatrix} 1 & -\left(1-\rme^{-\kappa m \Delta t}\right)/\kappa \end{pmatrix}$ and
%$B_\tau$ is the $1 \times 2$ matrix given by $B_{\tau} \eqdef  \begin{pmatrix} 1 &  -(1-e^{-\kappa \tau})/\kappa \end{pmatrix}$ and
%for all $1\le j\le J$, $A_0(j) \eqdef 0$. Define $F_{i,m} \eqdef F_{i\Delta t, m \Delta t }$ and $\mathcal{F}_i$ the $\sigma$-field generated by $(a_k)_{0\le k \le i}$ and $(Z_k)_{0\le k \le i}$. By definition of future prices and the assumption that there is no regime switching between $t_i$ and $t_{i+1}$, we have:
%%\begin{align*}
%%A_{t,\tau} &\eqdef r \tau - \int^{\tau}_0 \alpha_{a_{t+s}} \left(1- e^{-\kappa(\tau-s)} \right)\rmd s + (1/\kappa^2)\int^{\tau}_0 \eta^2_{a_{t+s}} \left(1-e^{-\kappa (\tau-s)}\right)^2 \rmd s \\
%%&\hspace{7cm}  - (2/\kappa)\int^{\tau}_0 \rho_{a_{t+s}} \eta_{a_{t+s}}\sigma_{a_{t+s}} \left( 1 -e^{-\kappa \left(\tau-s\right)} \right)\rmd s \eqsp,\\
%%B_{\tau} &\eqdef  \begin{pmatrix} 1 &  -(1-e^{-\kappa \tau})/\kappa \end{pmatrix}\eqsp.
%%\end{align*}
%\begin{align*}
%F_{i,m} &= \exp (A_m(a_i) + B_m Z_i)\eqsp,\\
%&=\mathbb{E}\left[\exp (A_{m-1}(a_{i+1}) + B_{m-1} Z_{i+1})\middle|\mathcal{F}_i\right]\eqsp,\\
%&= \mathbb{E}\left[\sum^J_{k=1} 1_{a_{i+1}=k}\exp (A_{m-1}(k) + B_{m-1} Z_{i+1})\middle|\mathcal{F}_i\right] \eqsp,\\
%&= \mathbb{E}\left[\sum^J_{k=1} 1_{a_{i+1}=k}\exp \left(A_{m-1}(k) + B_{m-1} ( d_{a_i} + T Z_i + H_{a_i} \epsilon_{i+1} ) \right)\middle|\mathcal{F}_i\right] \eqsp,\\
%&= \sum^J_{k=1} q(a_i,k) \exp \left(A_{m-1}(k) + B_{m-1} d_{a_i} + B_{m-1} \barH_{a_i} B'_{m-1}/2 + B_{m-1} T Z_i \right)\eqsp.
%\end{align*}
%From the above equality, we obtain,
for all $1\le j\le J$, $\mathsf{A}_0(j) = 0$,  and
\begin{equation*}
\mathsf{A}_m(j) = \ln \left(\sum^J_{k=1} Q(j,k) \exp(\mathsf{A}_{m-1}(k)) \right) + \mathsf{B}_{m-1} d_j + \frac{1}{2} \mathsf{B}_{m-1} \barH_j B'_{m-1}\eqsp.
\end{equation*}
Therefore, the observations of the  logfuture prices are given, for all $1\le i\le n$, by: 
\[
Y_i = c_{a_i} + B Z_i + G \eta_i\eqsp,
\] 
where $\xi_i$ is a standard multivariate Gaussian random variable and:
\[
c_{j}' = \begin{pmatrix} \mathsf{A}_{m_1}(j) & \ldots & \mathsf{A}_{m_{\dimy}}(j) \end{pmatrix}\eqsp,
\;\;\;
B' = \begin{pmatrix}\mathsf{B}_{m_1}' &  \ldots & \mathsf{B}_{m_{\dimy}}' \end{pmatrix}\eqsp, \;\;\;
G = \mathrm{diag}(g_1, \ldots , g_{d} )\eqsp.
\]
The model is then driven by the parameter:
\[
\theta \eqdef \{\pi, Q, \mu_1, \Sigma_1, \kappa, (\alpha_j)_{1\leq j\leq J}, (\sigma_j)_{1\leq j\leq J}, (\eta_j)_{1\leq j\leq J}, (\rho_j)_{1\leq j\leq J}, \mu, (\lambda_j)_{1\leq j\leq J}, (g_{\ell})_{1\leq \ell\leq d} \}\eqsp.
\]
The aim of this section is to estimate $\theta$ and the posterior probabilities $\mathbb{P}(a_k=j|Y_{1:n})$, $1\le k \le n$, $1\le j \le J$. Given the observations $Y_{1:n}$, the EM algorithm introduced in \cite{dempster:laird:rubin:1977} maximizes the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ defined by
\begin{equation*}
%\label{eq:loglik}
\ell_{\theta}^{n}(Y_{1:n}) \eqdef \log\left(\sum_{a_1=1}^J\ldots\sum_{a_n=1}^J\int p_{\theta}(a_{1:n},z_{1:n},Y_{1:n})\,\rmd z_{1:n}\right)\eqsp,
\end{equation*}
where the complete data likelihood $p_{\theta}$ is given by
\begin{equation*}
%\label{eq:completelik}
p_{\theta}(a_{1:n},z_{1:n},Y_{1:n}) \eqdef \pi(a_1)\phi_{\mu_1,\Sigma_1}(z_1)g_{\theta}(a_1,z_1;Y_1)\prod^{n}_{i=2}Q(a_{i-1},a_i)m_{\theta}\left(a_{i-1},z_{i-1};z_i\right)g_{\theta}(a_i,,z_i;Y_i)\eqsp.
\end{equation*}
Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{1:n}\right]$  the conditional expectation given $Y_{1:n}$ when the parameter value is $\theta$. The EM algorithm iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
\begin{enumerate}%[label=\roman*)]
	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p})\eqdef \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{1:n},Y_{1:n})\middle|Y_{1:n}\right]$ ;
	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
\end{enumerate}
%\paragraph{E-step}
%By \eqref{eq:completelik}, the intermediate quantity $Q(\theta,\theta_p)$ may be written (up to some additive constant which does not depend on $\theta$) $Q(\theta,\theta_p) = I^{n}_{1,\theta_p} + J^{n}_{1,\theta_p} + \sum_{i=2}^nI^{n}_{i,\theta_p} + \sum_{i=2}^nJ^{n}_{i,\theta_p}$,
%where
%\begin{align*}
% I^{n}_{1,\theta_p} &\eqdef \mathbb{E}_{\theta_p}\left[\log\phi_{\mu_1,\Sigma_1}(Z_1)\middle|Y_{1:n}\right]\eqsp,\\
% &= - \frac{1}{2}\log|\Sigma_1| -\frac{1}{2}\mathrm{Tr}\left(\Sigma_1^{-1}\mathbb{E}_{\theta_p}\left[Z_1Z'_1\middle|Y_{1:n}\right]\right) +\mu'_1\Sigma_1^{-1}\mathbb{E}_{\theta_p}\left[Z_1\middle|Y_{1:n}\right]- \frac{1}{2}\mu'_1\Sigma_1^{-1}\mu_1\eqsp,\\
% J^{n}_{1,\theta_p} &\eqdef \mathbb{E}_{\theta_p}\left[\log g_{\theta}(a_1,z_1;Y_1)\middle|Y_{1:n}\right]\eqsp,\\
% &= -\frac{1}{2}\log|\barG|-\frac{1}{2}\sum_{j=1}^J\mathrm{Tr}\left(\barG^{-1}\mathbb{E}_{\theta_p}\left[\left(Y_1-c_{j}-BZ_1\right)\left(Y_1-c_{j}-BZ_1\right)'\1_{a_1=j}\middle|Y_{1:n}\right]\right)
%\end{align*}
%and, for all $2\le i\le n$,
%\begin{align*}
% I^{n}_{i,\theta_p} &\eqdef\mathbb{E}_{\theta_p}\left[\log m_{\theta}\left(a_{i-1},z_{i-1};z_i\right)\middle|Y_{1:n}\right]\eqsp,\\
% &= -\frac{1}{2}\sum_{j=1}^J\log |\barH_{j}|\mathbb{E}_{\theta_p}\left[\1_{a_{i-1}=j}\middle|Y_{1:n}\right]\\
% &\hspace{.4cm}-\frac{1}{2}\sum_{j=1}^J \mathrm{Tr}\left(\barH^{-1}_j\mathbb{E}_{\theta_p}\left[\left(Z_i-d_{j}-TZ_{i-1}\right)\left(Z_i-d_{j}-TZ_{i-1}\right)'\1_{a_{i-1}=j}\middle|Y_{1:n}\right]\right)\eqsp.
% \end{align*}
% \begin{align*}
% J^{n}_{i,\theta_p} &\eqdef\mathbb{E}_{\theta_p}\left[\log g_{\theta}(a_i,z_i;Y_i)\middle|Y_{1:n}\right]\eqsp,\\
% &= -\frac{1}{2}\log|\barG|-\frac{1}{2}\sum_{j=1}^J\mathrm{Tr}\left(\barG^{-1}\mathbb{E}_{\theta_p}\left[\left(Y_i-c_{j}-BZ_i\right)\left(Y_i-c_{j}-BZ_i\right)'\1_{a_i=j}\middle|Y_{1:n}\right]\right)\eqsp.
%\end{align*}
All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our proposed algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
%\paragraph{M-step}
As the function $\theta\mapsto Q^N(\theta,\theta_p)$ cannot be maximized analytically, the M-step is performed numerically using the Covariance Matrix Adaptation Evolution Strategy (CMA-ES) introduced in \cite{hansen:ostermeier:2001}. This derivative-free optimization procedure is known to perform well in complex multimodal optimization settings, see e.g. \cite{hansen:kern:2004}. \\

%In details, among $510$ backwardation there are $485$ data whose regime staying in this regime, the same remarks on $417$ among $441$ contango doesn't change its regime.

%The future expiration date usually expires on the third business day prior to the $25-th$ calendar day of the month preceeding the delivery month
%\footnotetext{For WTI crude oil, trading in the current delivery month shall cease on the third business day prior to the $25$th calendar day of the month preceding the delivery month. If the $25$th calendar day of the month is a non-business day, trading shall cease on the third business dat prior to the last business day preceding the $25$th calendar day. See http://www.cmegroup.com for more details}

%Contrary to \cite{almansour:2014}, we calculate the exact time to maturity for each day and each future contract. For example, on $19-05-2010$ the time to maturity of the front contract $F_1$ is only $1$ day (the earliest expiration date is $20-05-2010$) but it is considered with a time to maturity $1$ month in \cite{almansour:2014}. Our calculus takes into account exactly the real time to maturity.

\subsubsection*{Numerical results}
\label{sec:crudeoil}
%Following \cite{carmona:ludkovski:2004}, under the risk neutral probability $\mathbb{Q}$, we assume that the spot price $S_t$ of the Chicago Mercantile Exchange (CME) crude oil and its associated convenience yield $\delta_t$ are solutions to the SDE:
%\begin{align*}
%\rmd S_t &= \left(\tilde{\mu}_{a_{t}} - \delta_t\right) S_t \rmd t + \sigma_{a_t} S_t \rmd \tilde{W}^1_t \eqsp,\\
%\rmd \delta_t &= \kappa \left( \tilde{\alpha}_{a_t} - \delta_t \right) \rmd t + \eta_{a_t} \rmd \tilde{W}^2_t \eqsp,
%\end{align*}
%where $\tilde{W}^1$ and $\tilde{W}^2$ are two correlated Brownian motions under the historical probability $\mathbb{P}$.
%
%In this case,  $\delta_t$ is an Orstein-Uhlenbeck process.
%
%Under the risk-neutral probability $\mathbb{Q}$, this dynamic becomes:
%\begin{align*}
%\rmd S_t &= \left(r - \delta_t\right)S_t \rmd t + \sigma_{a_t} S_t \rmd W^1_t \eqsp,\\
%\rmd \delta_t &= \left[\kappa \left(\alpha_{a_t} - \delta_t \right) - \lambda_{a_t} \right] \rmd t + \eta_{a_t} \rmd  W^2_t \eqsp,
%\end{align*}
%Define $\alpha_{a_t} \eqdef \tilde{\alpha}_{a_t} - \lambda_{a_t}/\kappa$. Therefore, the dynamic of $Z_t = (\ln(S_t), \delta_t)$ under the risk-neutral mesure is:
%\begin{align*}
%\rmd \ln(S_t) &= \left(r - \delta_t- \sigma^2_{a_t}/2 \right) dt + \sigma_{a_t} \rmd W^1_t \eqsp,\\
%\rmd \delta_t &= \kappa\left( \alpha_{a_t} - \delta_t\right)dt + \eta_{a_t} \rmd W^2_t \eqsp,
%\end{align*}
%where $(a_t)_{t\ge 0}$ is a Markov process taking values in  $\{1,\ldots,J\}$ indicating the market regime and $W^1$ and $W^2$ are Brownian motions under $\mathbb{Q}$ such that $\rmd<W^1, W^2>_t = \rho_{a_t}\rmd t$. The interest rate $r$ is constant and $\lambda_{a_t}$ is a risk associated with the convenience yield dynamic when the market regime is $a_t$. In the following, it is assumed that between two consecutive observations at times $t_{i-1}$ and $t_{i}$, the market regime is fixed so that the solutions to the previous SDE may be computed explicitly if $a_t$ is set to $a_{i-1} = a_{t_{i-1}}$. Therefore, it is assumed that the regime indicators are given at times $(t_i)_{i\ge 1}$ by a discrete time Markov chain $(a_{i})_{i\ge 1}$ and do not change between $t_{i-1}$ and $t_{i}$.
%The explicit integration of \eqref{eq:modele} between $t_{i-1}$ and $t_i$ under the approximation that there is no regime change between these two successive time instants provides the following values of the parameter of the discrete time model:
%for $Z_i \eqdef (\ln (S_{t_i}),\delta_{t_i})$:
%\[
%Z_i = d_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
%\]
%with $\varepsilon_i$ a standard $2$-dimensional Gaussian random variable and parameters defined as

%For WTI crude oil, trading in the current delivery month shall cease on the third business day prior to the $25$th calendar day of the month preceding the delivery month. If the $25$th calendar day of the month is a non-business day, trading shall cease on the third business dat prior to the last business day preceding the $25$th calendar day. In the event that the official exchange holiday schedule changes subsequent to the listing of a crude oil futures, the originally listed expiration date shall remain in effect. In the event that the originally listed expiration day is declared a holiday, expiration will move to the business day immediately prior. For example, the $25$th Jan $2014$ is sunday, the business day immediately prior is $23$th Jan (friday) to which the $3$rd business day prior is $20th Jan$. Therefore the maturity dates of $Feb 2014$ Contract is $20th Jan 2014$.
%For the CME WTI crude oil, the future prices are informed each trading day for each contract. Based on the expiration date, we may obtain an exact time to maturity for these futures. In our case, we are only interested in certain constant times to maturity: $4W$, $3M$, $6M$, $1Y$, $1,5Y$, $2Y$ and $3Y$. We chose the corresponding future contracts numbers: $1, 3, 7, 13, 19, 25$ and $37$ since this series is near the series $4W, 3M, 6M, 1Y, 1,5Y, 2Y, 3Y$. Futures prices for such fixed time to maturity may be estimated from the existing futures by smoothing splines or linear interpolation method.  For our weekly (every Wednesday) future prices data, we used the linear interpolation from $1995$-$01$-$01$ to $2013$-$05$-$01$. The period $1995$-$01$-$01$ to $2013$-$05$-$01$ used for CME WTI Crude Oil future prices is the same as the one for the data used in \cite{almansour:2014} so that we may use the same weekly transition matrix $Q$: $Q(1,1) = 0.9807$ and $Q(2,2) = 0.9734$ (the authors used the New York Mercantile Exchange (NYMEX) WTI crude oil). The initials values for the EM algorithm are given below:

The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi(1) = \pi(2) = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime. The other parameters are initialized as shown in the following table:
\begin{figure}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
		$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
		\hline
		5.0 & 0.1 & -0.05 & 0.4 & 0.4 & 0.5 & 0.5 & 0.75 & 0.65 & 0.0 & 0.0 & 0.0 & 0.1 & 0.1 & 0.1 & 0.1
	\end{tabular}
	\caption{Initial values for the EM algorithm.}
\end{figure}\\
The number of particles is set to $N = 100$, $\Delta t = 1/52$ and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}. The interest rate is $r = 0.0296$ as in \cite{almansour:2016}. The initial guess for the mean and variance of the initial state are
\[
\mu_1 = \begin{pmatrix} \ln F^{(market)}_{1,4} & r-\cfrac{\ln F^{(market)}_{1,16}-\ln F^{(market)}_{1,4}}{(16-4)\Delta t} \end{pmatrix}\quad \mbox{and}\quad \Sigma_1 = \begin{pmatrix} 0.05 & 0 \\ 0 & 0.05 \end{pmatrix}\eqsp.
\]
The CMA-ES algorithm is used with an initial standard deviation for the parameters $\sigma_{cmaes} = 0.005$, a number of selected search points $\mu_{cmaes} = 20$ and a population size $\lambda_{cmaes} = 100$. The algorithm is stopped after $10000$ iterations.
%and its bounded conditions are \\
%\begin{tabular}{cccccccccccccc}
%PAR & $\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
%\hline
%INF & 0.01 & -10.0 & -10.0 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 1e-10 & 1e-10 & 1e-10 & 1e-10 \\
%SUP & 1e10 & 10.0 & 10.0 & 1e10 & 1e10 & 1e10 & 1e10 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 \\
%\end{tabular}
In Gibson-Schwartz model \cite{gibson:schwartz:1990}, a stronger backwardation effect implies a greater value for $\alpha$ for the same values of the other parameters. For the CME WTI Crude Oil, backwardation effect is more frequent than contango effect so that $\alpha_1$ should be greater than $\alpha_2$. Therefore, this condition is imposed for all simulations in the CMA-ES algorithm. The results after $2500$ iterations of the EM algorithm are given in Figure~\ref{fig:CMEfinalestimates}.
\begin{table}[h!]
	\centering
	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c||c|clc|c|c||}
		Parameter&$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
		\hline
		Value&2.6378 & 0.0889 & -0.0281 & 0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
		\hline
		Std. Dev& 0.1999 & 0.004248 & 0.00149 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052   \\
	\end{tabular}
	\vspace{1.cm}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|||}
		Parameter&$\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
		\hline
		Value&0.1712 & 0.2422	& -0.1731 & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
		\hline
		Std. Dev& 0.004147 & 0.03243 & 0.01973 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4
	\end{tabular}
	\caption{Final estimates after 3500 iterations.}
	\label{fig:CMEfinalestimates}
\end{table}

%Remarks that the changement of $Q$ influences basically on the model parameters. A bad value of $Q$ can lead the parameters not to follow a good order in fonction of regimes. To prevent this trouble, our technique is firstly to fix the initial transition matrix $Q$ for the $100$ first iterations. That aids the model parameters to stabilize in a good order. More clearly in our experiment, fixing $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ up to 100 iterations drives CMAES to give a good order $\sigma_1 \geq \sigma_2$ $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$. When we no longer fix $Q$ after $100$ iterations, such the parameters in return conduct $\{Q(1,1), \; Q(2,2)\}$ to be in the good order.

As expected, we obtain $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$ at convergence of the EM algorithm. Moreover, $Q(1,1) > Q(2,2)$ is corresponding to the prediction that we did from the data description. From the results, $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$ indicates the first regime (backwardation) characterized by a higher value in both volatility and equilibrium level of convenience yield, and the second regime (contango) characterized by a lower value in both volatility and equilibrium convenience yield level. It verifies the theory of storage that the volatility of the commodity spot price is high when the inventory is low, and the convenience yield is all the higher as inventory is low.
Interestingly, we see that $\lambda^{\delta}_1 \geq \lambda^{\delta}_2$ \footnote{\cite{almansour:2016} obtain the same result in the opposite order, because of their definition of the historical mean-reverting value} and the two historical mean-reverting values are $\widetilde{\alpha}_1 = 0.0073$ and $\widetilde{\alpha}_2 = 0.0305$. This underlines that, while the first regime of convenience yield under $\mathbb{P}$ makes spot price be more likely to increase (with a smaller historical mean-reverting), the term structure of future prices has a tendency to be in backwardation. 
	This also shows opposite behaviors of historical prices evolution and term structure future prices. Intuitively, Figure~\ref{fig:LogPriceTermStructure:data} compares the evolution of future $1M$ (the nearest contracts) to the term structure observed from CME WTI crude oil. The term structure is measured by the difference of future $13M$ and future $1M$ to avoid the eventual saisonality. The figure shows that it is not necessary to have an inverse relationship between the evolution of price and term structure. But when a significant drop in price occurs, the term structure measured as $F_{13} - F_1$ increases (i.e. in contango).
\begin{figure}
\centering
\includegraphics[scale=.6]{LogPrice_TermStructure.eps}
\caption{Log-price (red line) and slope of future curves (blue line)}
\label{fig:LogPriceTermStructure:data}
\end{figure}
%The first regime (in backwardation) is identified by a higher mean-reversion of convenience yield $\alpha$ and a higher volatility in states variables ($\sigma$ and $\eta$). 
%And vice versa the second regime is characterized by a lower value of $\alpha, \sigma, \eta$.
%\begin{verbatim}
%TODO: Add reference: Explain why $rho$ must be high (at least positive).
%TODO: Explain why the noises $g1$, $g4$ are greater than the intermediate contracts
%\end{verbatim}

The correlation between the spot price and the convenience yield is positive and high in both two regimes. The positive correlation verifies one of the fundamental principes in commodity market as in \cite{gibson:schwartz:1990}. That explains when the spot prices are expected to inscrease, the convenience yield are also expected to inscrease. In addition, the higher the correlation is, the more the slope of future curve descreases in function of maturity. It descreases even further in the first regime (in backwardation): the higher volatility higher convenience yield with the higher correlation.

Figure~\ref{fig:posterior:data} and~\ref{fig:posterior:data:logprice} displays the the estimated posterior probabilities of the regime labels and the observed future slope. When the future curve is in backwardation (resp. contango), the model is expected to be in the first regime (resp. second regime), except for the period where the slope of the future curve is too small and in the period from December $2008$ to April $2009$ (beginning of the crisis).

%After convergence, the value obtained for $\kappa$ is slightly greater than the one obtained in \cite{almansour:2014} but we also obtain slightly lower values for $\eta_1$ and $\eta_2$. The result also shows that the future price of maturity $4W$ is such that $g_4$ is greater than all the other $g_i$'s. It may be explained by the fact that there is no observation of spot contract which is likely to impact mostly the front future contract.

%\begin{figure}
%\centering
%\includegraphics[scale=.38]{alpha_M.png}\\
%\includegraphics[scale=.38]{sigma_M.png}\\
%\includegraphics[scale=.38]{eta_M.png}
%\caption{Estimation of $\alpha$, $\sigma$ and $\eta$.}
%\end{figure}

\begin{figure}
	\centering
	\includegraphics[scale=.6]{Posterior_Data.eps}
	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
	\label{fig:posterior:data}
\end{figure}

\begin{figure}
	\centering
	\includegraphics[scale=.75]{Posterior_Data_LogPrice.eps}
	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
	\label{fig:posterior:data:logprice}
\end{figure}

%\paragraph{Comparaison of estimated future prices with the observations}
%It is interesting to assess the accuracy of the estimated parameters obtained with the EM algorithm. These estimated parameters are used to estimate the future prices and compare them with the observed ones for each maturity. The results of this estimation procedure is displayed in Figure~\ref{fig:density:4W3M} and Figure~\ref{fig:density:6M1Y}.

%\begin{figure*}[htb]
%	\centering
%		\begin{tabular}{||c|c|c|c|c||}
%		Residuals & $F_{4W}$ & $F_{3M}$ & $F_{6M}$ & $F_{1Y}$\\
%		\hline
%		mean & -0.002654812 & 0.001178231 & 0.002130222 & 0.002391526 \\
%		std & 0.01461169 & 0.00575261 & 0.01039791 & 0.01796323
%		\end{tabular}
%	\caption{Table of differences between estimated and observed future prices.}
%	\label{tab:tableTest}
%\end{figure*}
%Mean value of the residuals between the estimated and observed log-future prices are close to $0$ with a small variance. The standard deviation of futures $4W$ has an important value in comparison to futures $3M$ or $6M$ which goes along with the remark for the estimation of $G$ in the equation \eqref{eq:model:obs} by the EM algorithm.

%\FloatBarrier

%Our algorithm is calibrated using future prices of maturities $4W, 3M, 6M$ and $1Y$ but may be used to estimate future prices of maturities $8W, 9M, 2Y, 3Y$. These estimations may be compared to the observations given in the market. Densities of the estimated prices for these others maturities $8W, 9M, 2Y, 3Y$ are shown in Figure~\ref{fig:densothers}.
%\begin{figure}
%\centering
%\includegraphics[scale=.4]{Density_Estim_Obs_OUTSIDE.png}
%\caption{Densities of estimated and observed log-future prices $8W, 9M, 2Y, 3Y$.}
%\label{fig:densothers}
%\end{figure}
%At first glance, mean values and standard deviations of the difference between observed and estimated future prices are greater than in the previous case since these maturities are not used to calibrate the model. Nevertheless, these errors remain close to $0$. In addition, standard deviations of futures $2Y$ and $3Y$ are greater than the ones of $8W$ and $9M$, it is reasonable since $2Y$ and $3Y$ futures are outside of the interval of maturities used to calibrate the model.
%\FloatBarrier

\section{Conclusions}
In this paper, we propose an extension of the two-filter algorithm to estimate the conditional distribution of the hidden states given the observations in a Conditionally Linear and Gaussian model (CLGM). This extension improves the performance of the usual Rao Blackwellised two-filter algorithm using few particles by marginalizing explicitly parts of the paths before combining the forward and backward filters. The paper analyzes the behavior of the Term Structure in a crude oil market by discretizing a Gibson-Schwartz model with regime switchings. Numerical results highlight that the estimated posterior probabilities of the regimes are closely related to backwardation and contango effects in crude oil market.

\appendix

\section{Technical lemmas}%~\ref{eq:technical:twofilter}-\ref{lem:integral:gammap}
\begin{lemma}
\label{lem:integratedSDE}
Let $(X_t,\delta_t)_{t\ge0}$ be solutions to the following SDE:
\begin{align*}
\rmd X_t &= \left(\mu - \delta_t- \sigma^2/2 \right) dt + \sigma d W^1_t \eqsp,\\
\rmd \delta_t &= \kappa\left( \alpha - \delta_t\right)dt + \eta dW^2_t \eqsp,
\end{align*}
$(W_t^1)_{t\ge 0}$ and $(W_t^2)_{t\ge 0}$ are standard Brownian motions such that $\rmd \langle W_t^1,W_t^2 \rangle = \rho\rmd t$.
Then, for all $t\ge 0$ and $h>0$,
\[
\begin{pmatrix} X_{t+h} \\ \delta_{t+h}\end{pmatrix} = d_h + T_h\begin{pmatrix} X_{t} \\ \delta_{t}\end{pmatrix} + H_h\varepsilon\eqsp,
\]
where $\varepsilon$ is a standard 2-dimensional Gaussian random variable and (with $\barH_h\eqdef H_h'H_h$),
\[
d \eqdef
\begin{pmatrix} \left[\mu- \alpha - \sigma^2/2 \right] h + \alpha[1-e^{-\kappa h}]/\kappa\\
\alpha [1-e^{-\kappa h}] \end{pmatrix}\eqsp, \;\; T_h \eqdef
\begin{pmatrix} 1 & -[1-e^{-\kappa h}]/\kappa \\ 0 & e^{-\kappa h} \end{pmatrix} \eqsp,
\]
\begin{align*}
\barH_h(1,1) & \eqdef \sigma^2 h+ \eta^2\left\{h + (1-e^{-2\kappa h})/(2\kappa) - 2(1-e^{-\kappa h})/\kappa  \right\}/\kappa^2 \\
&\hspace{7cm} -2\rho\eta\sigma\left\{h - (1-e^{-\kappa h})/\kappa \right\}/\kappa\eqsp, \\
\barH_h(1,2) & \eqdef \left(\rho \eta \sigma-\eta^2/\kappa\right)\left(1-e^{-\kappa h} \right)/\kappa + \eta^2\left(1-e^{-2\kappa h} \right)/(2\kappa^2)\eqsp,\\
\barH_h(2,1) & \eqdef \barH_h(1,2)\eqsp,\\
\barH_h(2,2) & \eqdef \eta^2\left( 1-e^{-2\kappa h} \right)/(2\kappa)\eqsp.
\end{align*}
\end{lemma}

\begin{proof}
For all $t\ge 0$,
\[
X_t = X_0 + (\mu-\sigma^2/2)t - \int_0^t\delta_s\rmd s +\sigma W_t^1
\]
and, as $(\delta_t)_{0\le t\le T}$ is an Ornstein-Uhlenbeck process,
\[
\delta_t = \delta_0\rme^{-\kappa t} + \alpha(1-\rme^{-\kappa t}) + \int_0^t \eta \rme^{\kappa(s-t)}\rmd W^2_s\eqsp.
\]
Then,
\begin{align*}
\int_0^t\delta_s\rmd s &= (\delta_0-\alpha)(1-\rme^{-\kappa t})/\kappa + \alpha t + \eta\int_0^t\int_0^s\rme^{\kappa(u-s)}\rmd W^2_u\rmd s\eqsp,\\
&=(\delta_0-\alpha)(1-\rme^{-\kappa t})/\kappa + \alpha t + (\eta/\kappa) \int_0^t (1-\rme^{-\kappa (t-s)}) \rmd W^2_s\eqsp.
\end{align*}
Defining $\tilde{W}^1_t \eqdef - (\eta/\kappa) \int_0^t (1-\rme^{-\kappa (t-s)}) \rmd W^2_s+\sigma W_t^1$ and $\tilde{W}_t^2\eqdef \int_0^t \eta \rme^{\kappa(s-t)}\rmd W^2_s$, this yields:
\begin{align*}
X_t &= X_0 +  (\mu-\sigma^2/2)t + (\alpha-\delta_0)(1-\rme^{-\kappa t})/\kappa - \alpha t + \tilde{W}_t^1\eqsp,\\
\delta_t &= \delta_0\rme^{-\kappa t} + \alpha(1-\rme^{-\kappa t}) + \tilde{W}^2_t\eqsp.
\end{align*}
The proof is concluded upon noting that $\tilde{W}^1_t$ and $\tilde{W}^2_t$ are centered Gaussian random variables such that:
\begin{itemize}
\item $\!\!\mathrm{Var}\left[\tilde{W}^1_t\right] = \sigma^2  t + \eta^2\left\{ t + (1-e^{-2\kappa t})/(2\kappa) - 2(1-e^{-\kappa  t})/\kappa\right\}/\kappa^2 - 2\rho \eta \sigma\left\{ t - (1-e^{-\kappa  t})/\kappa \right\}/\kappa\eqsp,$
\item $\!\!\mathrm{Var}\left[\tilde{W}^2_t\right] = \eta^2(1-\rme^{-2\kappa t})/(2\kappa)\eqsp,$
\item $\!\!\mathrm{Cov}\left[\tilde{W}^1_t,\tilde{W}^2_t\right] = \left(\rho \eta\sigma-\eta^2/\kappa\right)\left(1-e^{-\kappa t} \right)/\kappa + \eta^2 \left(1-e^{-2\kappa t} \right)/(2\kappa^2)\eqsp.$
\end{itemize}
\end{proof}

\label{sec:app:proofs}
Lemmas~\ref{eq:technical:twofilter}, \ref{lem:py} and \ref{lem:integral:gammap} are close to \cite[Proposition 5, Proposition 6]{briers:doucet:maskell:2010}. The proofs are detailed in this appendix for completeness.
\begin{proof}[Proof of Lemma~\ref{eq:technical:twofilter}]
For all $1\le i \le n-1$,
\begin{align*}
p_{}(y_{i:n}|a_i,z_i) &= \sum_{a_{i+1:n}}\int p_{}(y_{i:n},a_{i+1:n},z_{i+1:n}|a_i,z_i)\rmd z_{i+1:n}\eqsp,\\
&=  \sum_{a_{i+1:n}}\int p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})\rmd z_{i+1:n}\eqsp,\\
&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\sum_{a_{i+1:n}}\int \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})\rmd z_{i+1:n}\eqsp,\\
&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\sum_{a_{i+1:n}}\int \tilde{p}_{i}(a_{i:n},z_{i:n}|y_{i:n})\rmd z_{i+1:n}\eqsp,\\
&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\tilde{p}_{i}(a_{i},z_{i}|y_{i:n})\eqsp,
\end{align*}
which concludes the proof of \eqref{eq:p:y1:n}. To prove \eqref{eq:p:ai:zi} write,
\begin{align*}
\tilde{p}_{i}(a_{i:n},z_{i}|y_{i:n}) &= \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\int p_{}(y_{i:n}|a_{i:n},z_{i:n})p_{}(a_{i+1:n},z_{i+1:n}|a_{i},z_{i})\rmd z_{i+1:n}\eqsp,\\
&= \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\int \frac{p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(z_{i+1:n}|y_{i:n},a_{i:n},z_i)}{p_{}(z_{i+1:n}|a_{i:n},z_i)}p_{}(a_{i+1:n},z_{i+1:n}|a_{i},z_{i})\rmd z_{i+1:n}\eqsp,\\
&=\frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(a_{i+1:n}|a_i)\eqsp.
\end{align*}
Therefore,
\[
\tilde{p}_{i}(a_{i},z_{i}|y_{i:n}) = \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\sum_{a_{i+1:n}}p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(a_{i+1:n}|a_i)
\]
and the proof is completed upon noting that
\[
\tilde{p}_{i}(a_{i:n}|y_{i:n}) = \frac{p_{}(a_{i+1:n}|a_i)}{\tilde{p}_{i}(y_{i:n})} \int \gamma^{}_{i}(a_i,z)p_{}(y_{i:n}|a_{i:n},z)\rmd z\eqsp.
\]
\end{proof}

\begin{proof}[Proof of Lemma~\ref{lem:py}]
The result is proved by backward induction. \eqref{eq:cn}, \eqref{eq:Pn} and \eqref{eq:nun} follow directly from \eqref{eq:model:obs}. Assume that for a given $1\le i\le n-1$, $p(y_{i+1:n}|a_{i+1:n},z_{i+1})$ is given by \eqref{eq:py}. Write
\[
p(y_{i:n}|a_{i:n},z_i) = \int p(z_{i+1}|z_i,a_{i+1})p(y_i|z_i,a_i)p(y_{i+1:n}|a_{i+1:n},z_{i+1})\rmd z_{i+1}\eqsp,
\]
with
\begin{align*}
p(z_{i+1}|z_i,a_{i+1}) &= \exp\left\{-\frac{\dimz}{2}\log(2\pi)-\frac{1}{2}\log|\barH_{a_{i+1}}|-\frac{1}{2}\normMat{\barH_{a_{i+1}}}{z_{i+1}-d_{a_{i+1}}-T_{a_{i+1}}z_i}^2\right\}\eqsp,\\
p(y_i|z_i,a_i)& = \exp\left\{-\frac{\dimy}{2}\log(2\pi)-\frac{1}{2}\log|\barG_{a_i}|-\frac{1}{2}\normMat{\barG_{a_i}}{y_{i}-c_{a_{i}}-B_{a_i}z_i}^2\right\}\eqsp,\\
p(y_{i+1:n}|a_{i+1:n},z_{i+1})& = \exp\left\{-\frac{1}{2}c_{i+1}(a_{i+1:n}) - \frac{1}{2}\normMat{\tilde{P}_{i+1}(a_{i+1:n})}{z_{i+1}}^2 + z'_{i+1}\tilde{\nu}_{i+1}(a_{i+1:n})\right\}\eqsp.
\end{align*}
Let $\Delta_{i}$ and  $\delta_{i}$ be given by:
\begin{align*}
\Delta_{i}(a_{i+1:n}) &\eqdef \left(I_{\dimz} + H'_{a_{i+1}}\tilde{P}_{i+1}^{-1}(a_{i+1:n})H_{a_{i+1}}\right)^{-1}\eqsp,\\
\delta_{i}(a_{i+1:n}) &\eqdef \nu_{i+1}(a_{i+1:n}) + \barH_{a_{i+1}}^{-1}(d_{a_{i+1}}+T_{a_{i+1}}z_i)\eqsp.
\end{align*}
Then, $\barH_{a_{i+1}}^{-1} + \tilde{P}_{i+1}^{-1}(a_{i+1:n}) = \left(H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\right)^{-1}$ and \eqref{eq:ck}, \eqref{eq:Pk} and \eqref{eq:nuk} follows from
\begin{multline*}
\int \exp\left\{-\frac{1}{2}\normMat{H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}}{z_{i+1}}^2 + z'_{i+1}\delta_{i}(a_{i+1:n})\right\}\rmd z_{i+1}\\
 =\exp\left\{\frac{1}{2}\log(2\pi) + \frac{1}{2}\log|H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}|+\frac{1}{2}\delta'_{i}(a_{i+1:n})'H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\delta_{i}(a_{i+1:n})\right\}\eqsp.
\end{multline*}
\end{proof}

\begin{proof}[Proof of Lemma~\ref{lem:integral:gammap}]
By Lemma~\ref{lem:py},
\begin{multline*}
\phi_{\mu_i,\Sigma_i}(z_i)p(y_{i:n}|a_{i:n},z_i) = \exp\left\{-\frac{\dimz}{2}\log(2\pi) -\frac{1}{2}\log|\Sigma_i|-\frac{1}{2}\normMat{\Sigma_i}{z_i-\mu_i}^2\right\}\\
\times\exp\left\{-\frac{1}{2}c_{i}(a_{i:n})-\frac{1}{2}\normMat{\tilde{P}_{i}(a_{i:n})}{z_i}^2+z'_i\nu_{i}(a_{i:n})\right\}\eqsp.
\end{multline*}
The proof is completed noting that
\begin{multline*}
\int \exp\left\{-\frac{1}{2}z'_i\tilde{\Omega}_{i}^{-1}(a_{i:n})z_i + z'_i(\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n}))\right\}\rmd z_{i}\\
= \exp\left\{\frac{\dimz}{2}\log(2\pi) + \frac{1}{2}\log|\tilde{\Omega}_{i}(a_{i:n})| +\frac{1}{2}\left[\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n})\right]'\tilde{\Omega}_{i}(a_{i:n})\left[\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n})\right]\right\}\eqsp.
\end{multline*}
\end{proof}

\clearpage
\newpage

\section{Algorithms}
The forward filter is described in Algorithm~\ref{algo:forward}. The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^j$ may be written:
\begin{align*}
\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
\end{align*}
\begin{algorithm}
\DontPrintSemicolon
\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
{\bf Initialization}\;
\For{$k \gets 1$ \textbf{to} $N$} {
Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
Compute $\mu_1^k$ and $P_1^k$:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
}
{\bf Iterations}\;
\For{$i \gets 2$ \textbf{to} $n$} {
Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
\For{$k \gets 1$ \textbf{to} $N$} {
For all $j \in \{1,\ldots, J\}$ compute
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
\]
Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
Update $\mu_i^k$ and $P_i^k$:
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
}
}
\caption{Forward filter, see \cite{chen:liu:2000}}
\label{algo:forward}
\end{algorithm}

\begin{algorithm}
\DontPrintSemicolon
\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
{\bf Initialization}\;
Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
{\bf Iterations}\;
\For{$i \gets n-1$ \textbf{to} $1$} {
\For{$k \gets 1$ \textbf{to} $N$} {
Compute
\[
\omega^k_{i|n} \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})\eqsp.
\]
}
Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$.
}
\caption{Smoothing algorithm based on the approximation of \cite{kim:1994}}
\label{algo:kim}
\end{algorithm}

\begin{algorithm}
\DontPrintSemicolon
\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
{\bf Initialization}\;
Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
Compute $\hat{\Omega}_n$ and $\hat{\lambda}_n$:
\[
\hat{\Omega}_n = B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}B_{\tilde{a}_n}\quad\mbox{and}\quad \hat{\lambda}_n =B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}(y_n-c_{\tilde{a}_n})\eqsp.
\]
{\bf Iterations}\;
\For{$i \gets n-1$ \textbf{to} $1$} {
Compute 
\[
m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{\tilde{a}_{i+1}}\quad\mbox{and}\quad M_{i+1} = H_{\tilde{a}_{i+1}}'\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}} + I\eqsp.
\]
Set
\begin{align*}
\Omega_i &= T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})\hat{\Omega}_{i+1}T_{\tilde{a}_{i+1}}\eqsp,\\
\lambda_i &=T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})m_{i+1}\eqsp.
\end{align*}
\For{$k \gets 1$ \textbf{to} $N$} {
Compute
\[
\Lambda^k_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I \quad\mbox{and}\quad \eta^k_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}\eqsp.
\]
Set
\[
\tilde{\omega}_{i|n}^k \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})|\Lambda^k_i|^{-1/2}\exp\left\{\eta^k_i/2\right\}\eqsp.
\]
}
Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$ and
\[
\hat{\Omega}_i =  \Omega_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}B_{\tilde{a}_i} \quad\mbox{and}\quad \hat{\lambda}_i = \lambda_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}(y_i-c_{\tilde{a}_i})\eqsp.
\]
}
\caption{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
\label{algo:FFBS}
\end{algorithm}

\begin{algorithm}[b]
\DontPrintSemicolon
\KwIn{Artificial densities $(\gamma^{}_i)_{1\le i\le n}$ and proposal densities $(\tilde{q}_i)_{1\le i \le n}$.}
\KwOut{A set of weighted trajectories $(\tilde{a}^j_{i:n})_{1\le j\le N}$ with importance weights $(\tilde{\omega}^j_i)_{1\le j\le N}$, $1\le i \le n$.}
{\bf Initialization}\;
For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_n\sim\tilde{q}_n(\cdot)$.\;
For $1\le \ell \le N$, set:
\[
\tilde{\omega}^{\ell}_n \propto \frac{\int\gamma^{}_n(\tilde{a}^{\ell}_n,z')g_{}(\tilde{a}^{\ell}_n,z';y_n)\rmd z'}{\tilde{q}_n(\tilde{a}^{\ell}_n)}\eqsp.
\]
{\bf Iterations}\;
\For{$i \gets n-1$ \textbf{to} $1$} {
Resample the set $(\tilde{a}^{\ell}_{i+1:n})_{1\le j\le N}$ using the normalized weights $(\tilde{\omega}^{\ell}_{i+1})_{1\le j \le N}$.\;
For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_{i}\sim\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot)$.\;
For $1\le \ell \le N$, set:
\[
\tilde{\omega}^{\ell}_i \propto \frac{q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
\]
}
\caption{Backward information filter, see \cite[Section~4.1.1]{briers:doucet:maskell:2010}}
\label{algo:backward}
\end{algorithm}

\bibliographystyle{alpha}

\bibliography{./CLGM_bib}
\end{document}

### 2016-10-31 TexArxiv/clgm.tex
no comment

@@
 \]
+The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^{j,k}$ may be written:
+\[
+\gamma_i^{j,k} =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp,
+\]
+where 
+\begin{align*}
+\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
+P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp.
+\end{align*}
 $\mu_i^k$ and $P_i^k$ are updated according to:
 \begin{align*}
-\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
-P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
 K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
@@
 
-%The conditional distribution of $Y_1$ given $a^1_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^j$ may be written:
-%\begin{align*}
-%\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
-%& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
-%\end{align*}
-%\begin{algorithm}
-%\DontPrintSemicolon
-%\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
-%\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
-%{\bf Initialization}\;
-%\For{$k \gets 1$ \textbf{to} $N$} {
-%Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
-%Compute $\mu_1^k$ and $P_1^k$:
-%\begin{align*}
-%K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
-%\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
-%P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
-%\end{align*}
-%Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
-%}
-%{\bf Iterations}\;
-%\For{$i \gets 2$ \textbf{to} $n$} {
-%Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
-%\For{$k \gets 1$ \textbf{to} $N$} {
-%For all $j \in \{1,\ldots, J\}$ compute
-%\[
-%\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
-%\]
-%Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
-%Update $\mu_i^k$ and $P_i^k$:
-%\begin{align*}
-%\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
-%P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
-%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
-%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
-%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
-%\end{align*}
-%Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
-%}
-%}
-
-%The recursive updates of $(a^k_{1:i})_{1\le k \le N}$, $(\omega^k_{i})_{1\le k \le N}$,  $\mu^k_{i}$ and $P^k_{i}$ are  displayed in Algorithm~\ref{algo:forward}.
-%In the initialization step, the first weight is given, for all $1\le k \le N$, by $\omega^k_1 = p(y_1|a^k_{1})$.
-By \eqref{eq:model:obs}, the conditional distribution of $Y_1$ given $a_1$ is a Gaussian distribution with mean $c_{a_1}+B_{a_1}\mu_1$ and variance $B_{a_1}\Sigma_1B'_{a_1} + \barG_{a_1}$.
-This yields:
-%\[
-%\omega^k_{1} \propto \exp\left\{-\frac{1}{2}\normMat{B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}}{Y_1-c_{a^k_1}-B_{a_1^k}\mu_1}^2\right\} \eqsp,
-%\]
-
-%By Algorithm~\ref{algo:forward}, $p^N_{}(a_{i},z_{i}|y_{1:i})$ is given by:
-%\begin{equation*}
-%p^N_{}(a_{i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{i}}(a_{i})\omega^k_{i}|2\pi P^k_{i}|^{-1/2}\times \exp\left\{-\frac{1}{2}\normMat{P_{i}^k}{z_{i}-\mu^k_{i}}^2\right\}\eqsp.
-%\end{equation*}
-
 \subsection{Path-space smoother of \cite{kitagawa:1996}}
@@
 This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps.
+See Section~\ref{sec:numerical:experiments} for a numerical illustration.
 
@@
 Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available.
+Then, note that, for all $1\le i\le n$,
+\[
+p(a_{i:n}|y_{1:n}) = p(a_{i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp.
+\]
 Then, the approximation of \cite{kim:1994} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
@@
 \]
-Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$.
-$\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$.
-Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$.
+Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$:
+\begin{enumerate}[-]
+\item $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$. 
+\item Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$. 
+\end{enumerate}
 If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the indicator variables is:
@@
 \subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
-\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm.
+\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}.
 The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
@@
 \begin{enumerate}[-]
-\item Set $\tilde{a}_n= a_n^k$ with probability $\omega_n^k$, $1\le k \le N$.
-\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probability $\tilde{\omega}_i^k$, $1\le k \le N$.
+\item Set $\tilde{a}_n= a_n^k$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$.
+\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probabilities proportional to $(\tilde{\omega}_i^k)_{1\le k \le N}$.
 \end{enumerate}
 This algorithm requires to compute the quantity  $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$.
-As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using backward Kalman filtering techniques from time $n$ to time $i$.
-However, this has to be done for each trajectory $a^k_{1:i}$, $1\le k \le N$, which leads to an algorithm with a prohibitive computational complexity.
+As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using backward Kalman filtering techniques.
+However, this has to be done for each trajectory $(a^k_{1:i})_{1\le k \le N}$, which leads to an algorithm with a prohibitive computational complexity.
 \cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a procedure computationally less intensive by noting that:
@@
 \end{equation}
-which is the two-filter decomposition of the smoothing distribution. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
+which is similar to the two-filter decomposition of the smoothing distribution. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
 \[
@@
 \]
-where the proportionality is with respect to $z_i$. 
+where the proportionality is with respect to $z_i$. These quantities may be computed recursively backward in time with:
 \begin{align*}
@@
 \end{align*}
-Then, for $1\le i\le n-1$, define $m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\hat{\Omega}_{i+1}H_{a_{i+1}} + I$. Then,
+Then, for $1\le i\le n-1$, define $m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\hat{\Omega}_{i+1}H_{a_{i+1}} + I$ and write
 \begin{align*}
@@
 \]
-where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$.
-These backward simulation steps are summarized in Algorithm~\ref{algo:FFBS}.
+where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$ where $P_i^k = \Gamma_i^k(\Gamma_i^k)'$.
+If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{lindsten:bunch:sarkka:schon:godsill:2015} of the joint smoothing distribution of the indicator variables is:
+\[
+p^{\mathsf{Lbscg}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
+\]
 
@@
 Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}.
-Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods.
+Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations).
+%Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods.
 For $3\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
@@
 %\phi^n_{i}(a_{i-1:i},z_{i-1:i})  =  \sum_{a_{i-2}}\sum_{a_{i+1}} \int_{z_{i-2}} \int_{z_{i+1}} \psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
-\phi^n_{i}(a_{i-1:i},z_{i-1:i})  =  \sum_{a_{i-2}}\sum_{a_{i+1}} \int_{z_{i-2}} \int_{z_{i+1}} \psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
+\phi^n_{i}(a_{i},z_{i})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
 \end{equation}
-where $\psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-2$ and $i+1$.
+where $\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-1$ and $i+1$.
 It may be decomposed as follows:
 \begin{multline*}
-\psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1}) \propto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})p_{}(a_{i-2},z_{i-2}|y_{1:i-2})Q(a_{i-2},a_{i-1})\\
-\times \left\{\prod_{k=i}^{i+1} m_{}(a_{k-1},z_{k-2};z_{k-1})g_{}(a_{k-1},z_{k-1};y_{k-1})Q(a_{k-1},a_k)\right\}m_{}(a_{i+1},z_{i};z_{i+1})\eqsp,
+\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1}) \propto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})p_{}(a_{i-1},z_{i-1}|y_{1:i-1})Q(a_{i-1},a_{i})m_{}(a_{i},z_{i-1};z_{i})\\
+\times g_{}(a_{i},z_{i};y_{i})Q(a_{i},a_{i+1})m_{}(a_{i+1},z_{i};z_{i+1})\eqsp,
 \end{multline*}
-where the proportionality is with respect to $(a_{i-2:i+1},z_{i-2:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ may be written as
-\begin{multline}
+where the proportionality is with respect to $(a_{i-1:i+1},z_{i-1:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ may be written as
+\begin{equation}
 \label{eq:smoothing}
-\phi^n_{i}(a_{i-1:i},z_{i-1:i})  \propto r_{i}(a_{i-1},z_{i-1},y_{1:i-2})m_{}(a_{i},z_{i-1};z_i)g_{}(a_{i-1},z_{i-1};y_{i-1})Q(a_{i-1},a_i)\\
-\times g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
-\end{multline}
-where $m$ and $g$ are defined in \eqref{eq:definition-m} and \eqref{eq:definition-g},  $Q$ is the transition matrix of the
-Markov chain $\{a_k, k \geq 0\}$ and
+\phi^n_{i}(a_{i},z_{i})  \propto r_{i}(a_{i},z_{i},y_{1:i-1})g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
+\end{equation}
+where $m$ and $g$ are defined in \eqref{eq:definition-m} and \eqref{eq:definition-g} and
 \begin{align}
-r_{i}(a_{i-1},z_{i-1},y_{1:i-2}) &= p_{}(a_{i-1},z_{i-1}|y_{1:i-2})\eqsp,\\
-&=\sum_{a_{i-2}}\int_{z_{i-2}}p_{}(a_{i-2},z_{i-2}|y_{1:i-2})m_{}(a_{i-1},z_{i-2};z_{i-1})Q(a_{i-2},a_{i-1}) \rmd z_{i-2}\eqsp,\label{eq:smooth:r}\\
-t_{i}(a_i,z_i,y_{i:n}) &= \sum_{a_{i+1}}\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp.\label{eq:smooth:t}
+r_{i}(a_{i},z_{i},y_{1:i-1}) &= p_{}(a_{i},z_{i}|y_{1:i-1})\eqsp,\\
+&=\sum_{a_{i-1}}\int_{z_{i-1}}p_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-1}\eqsp,\label{eq:smooth:r}\\
+t_{i}(a_i,z_i,y_{i+1:n}) &= \sum_{a_{i+1}}\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp.\label{eq:smooth:t}
 \end{align}
-In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of states $a_{1:i-2}^k$ associated with importance weights $\omega_{i-2}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-2},z_{i-2}|y_{1:i-2})$.
-Plugging this approximation into \eqref{eq:smooth:r} provides a non normalized approximation $r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ of $r_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ integrating over all possible choices of $(a_{i-2},z_{i-2})$:
-\[
-r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2}) = \sum_{a_{i-2}}\int_{z_{i-2}}p^N_{}(a_{i-2},z_{i-2}|y_{1:i-2})m_{}(a_{i-1},z_{i-2};z_{i-1})Q(a_{i-2},a_{i-1}) \rmd z_{i-2}\eqsp.
+In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of indicator variables $a_{1:i-1}^k$ associated with importance weights $\omega_{i-1}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-1},z_{i-1}|y_{1:i-1})$.
+Plugging this approximation into \eqref{eq:smooth:r} provides a non normalized approximation $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ of $r_{i}(a_{i},z_{i},y_{1:i-1})$ integrating over all possible choices of $(a_{i-1},z_{i-1})$:
+\[
+r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{a_{i-1}}\int_{z_{i-1}}p^N_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-2}\eqsp.
 \]
@@
 \begin{align*}
-\Upsilon^k_{i-2}(a_{i-1}) &\eqdef (P^k_{i-2})^{-1} + T'_{a_{i-1}}\barH_{a_{i-1}}^{-1}T_{a_{i-1}}\eqsp,\\
-%\upsilon^k_{i-2}(a_{i-1})&\eqdef (P^k_{i-2})^{-1}\mu^k_{i-2} +  T'_{a_{i-1}}\barH_{a_{i-1}}^{-1}(z_{i-1}-d_{a_{i-1}})\eqsp,\\
-\upsilon^k_{i-2}(a_{i-1})&\eqdef \barH_{a_{i-1}}^{-1}T_{a_{i-1}}(\Upsilon^k_{i-2})^{-1}(a_{i-1})(P^k_{i-2})^{-1}\mu^k_{i-2}\eqsp,\\
-\Omega^k_{i-2}(a_{i-1}) &= \left(\barH_{a_{i-1}}^{-1} - \barH_{a_{i-1}}^{-1}T_{a_{i-1}}(\Upsilon^k_{i-2})^{-1}(a_{i-1})T'_{a_{i-1}}\barH_{a_{i-1}}^{-1}\right)^{-1}\eqsp,\\
-r_{i-2}^k(a_{i-1}) &= \upsilon^k_{i-2}(a_{i-1})+(\Omega^k_{i-2}(a_{i-1}))^{-1}d_{a_{i-1}}\eqsp,\\
-\omega_{\mathsf{f},i}^k(a_{i-1}) &\eqdef \omega^k_{i-2}Q(a^k_{i-2},a_{i-1})\left|P^k_{i-2}\right|^{-1/2}\left|\Upsilon^k_{i-2}(a_{i-1})\right|^{-1/2} \left|2\pi\barH_{a_{i-1}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{d_{a_{i-1}}}^2\right\}\\
-&\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-2}^k}{\mu^k_{i-2}}^2+\frac{1}{2}\normMat{P_{i-2}^k\Upsilon^k_{i-2}(a_{i-1})P_{i-2}^k}{\mu^k_{i-2}}^2 - (\upsilon^k_{i-2})'(a_{i-1})d_{a_{i-1}} \right\}\eqsp.
+\Upsilon^k_{i-1}(a_{i}) &\eqdef (P^k_{i-1})^{-1} + T'_{a_{i}}\barH_{a_{i}}^{-1}T_{a_{i}}\eqsp,\\
+\upsilon^k_{i-1}(a_{i})&\eqdef \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})(P^k_{i-1})^{-1}\mu^k_{i-1}\eqsp,\\
+\Omega^k_{i-1}(a_{i}) &= \left(\barH_{a_{i}}^{-1} - \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})T'_{a_{i}}\barH_{a_{i}}^{-1}\right)^{-1}\eqsp,\\
+r_{i-1}^k(a_{i}) &= \upsilon^k_{i-1}(a_{i})+(\Omega^k_{i-1}(a_{i}))^{-1}d_{a_{i}}\eqsp,\\
+\omega_{\mathsf{f},i}^k(a_{i}) &\eqdef \omega^k_{i-1}Q(a^k_{i-1},a_{i})\left|P^k_{i-1}\right|^{-1/2}\left|\Upsilon^k_{i-1}(a_{i})\right|^{-1/2} \left|2\pi\barH_{a_{i}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{d_{a_{i}}}^2\right\}\\
+&\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-1}^k}{\mu^k_{i-1}}^2+\frac{1}{2}\normMat{P_{i-1}^k\Upsilon^k_{i-1}(a_{i})P_{i-1}^k}{\mu^k_{i-1}}^2 - (\upsilon^k_{i-1})'(a_{i})d_{a_{i}} \right\}\eqsp.
 \end{align*}
@@
 \label{eq:rN}
-r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i-1})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{z_{i-1}}^2+z'_{i-1}r_{i-2}^k(a_{i-1})\right\}\eqsp.
+r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
 %\times\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{z_{i-1}-d_{a_{i-1}}}^2\right\}\eqsp.%q(a^k_{i-2},a_{i-1})\omega^k_{i-2}\left|P^k_{i-2}\right|^{-1/2}\left|\Upsilon^k_{i-2}(a_{i-1})\right|^{-1/2}\exp\left\{ -\frac{1}{2}\normMat{P_{i-2}^k}{\mu^k_{i-2}}^2\right\}\\
@@
 The backward pass described below produces a sequence of states $\tilde{a}_{i+1:n}^{\ell}$ associated with importance weights $\tilde{\omega}_{i+1}^{\ell}$, $1\le \ell\le N$ which are used to approximate $p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$.
-Plugging this approximation into \eqref{eq:smooth:t} provides a non normalized approximation $t^N_{i}(a_i,z_i,y_{i:n})$ of $t_{i}(a_i,z_i,y_{i:n})$ integrating over all possible choices $(a_{i+1},z_{i+1})$.
-
-These steps are then combined to form the normalized SMC approximation $\phi^{n,N}_{i}(a_{i-1:i},z_{i-1:i})$ of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ using \eqref{eq:smoothing}.
-As $r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ and $t^N_{i}(a_i,z_i,y_{i:n})$ are not normalized, the normalization of $\phi^{n,N}_{i}(a_{i-1:i},z_{i-1:i})$ is obtained by integrating over the states $a_{i-1:i},z_{i-1:i}$, when $r_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ and $t_{i}(a_i,z_i,y_{i:n})$ are replaced by $r^N_{i}(a_{i-1},z_{i-1},y_{1:i-2})$ and $t^N_{i}(a_i,z_i,y_{i:n})$ in \eqref{eq:smoothing}.
+Plugging this approximation into \eqref{eq:smooth:t} provides a non normalized approximation $t^N_{i}(a_i,z_i,y_{i+1:n})$ of $t_{i}(a_i,z_i,y_{i+1:n})$ integrating over all possible choices $(a_{i+1},z_{i+1})$.
+
+These steps are then combined to form a non normalized SMC approximation of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ using \eqref{eq:smoothing}.
+As $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ are not normalized, the normalization of the SMC approximation of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
 %These combination and normalization steps are given in Appendix~\ref{sec:approx:smooth}.
 Our procedure allows to construct sequence of regimes with non-degenerated importance weights in the combination step.
-This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ over the states at times $i-2$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged.
+This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $\phi^n_{i}(a_{i},z_{i})$ over the states at times $i-1$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged.
 This method often leads to importance weights which are close to be numerically degenerated.
@@
 %This section displays a SMC algorithm to approximate $t_{i}(a_i,z_i,y_{i:n})$ using backward recursions. 
-As the function $(a_{i+1},z_{i+1})\mapsto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$ is not a probability density function it is not possible to use SMC samples to approximate it.
+As the function $(a_{i+1},z_{i+1})\mapsto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$ is not a probability density function, it is not possible to use SMC samples to approximate it.
 The backward filter relies on artificial densities to introduce a surrogate target density function which may be approximated using SMC methods.
@@
 
-\bibliography{./CLGM_bib}
+\bibliography{./clgm_bib}
 \end{document}

### 2016-11-01 TexArxiv/clgm.tex
minor modifictations in the numerical experiments

@@
 \begin{abstract}
+This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces.
 Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for conditionally linear and Gaussian models.
@@
 In general state space models, bayesian filtering and smoothing problems, i.e.
-computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks required for instance  to perform maximum likelihood inference.
+the computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks required for instance  to perform maximum likelihood inference.
 Filtering refers to the estimation of the distributions of $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time k, while smoothing is the estimation of the distribution of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$.
@@
 However, only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps.
-Instead of using a structural approximation of the model such as in \cite{kim:1994}, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
+Instead of using a structural approximation of the model such as in \cite{kim:1994}, \cite{fong:godsill:doucet:west:2002,lindsten:schon:2011} proposed forward-backward smoothers which require to sample the hidden linear states in the backward pass.
+\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
 In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
@@
 In this paper, we propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables.
-This explicit marginalization provides an additional Rao Blackwellisation step and produce more robust approximations.
+This explicit marginalization provides an additional Rao Blackwellisation step and produces more robust approximations.
 
 The paper is organized as follows.
-The proposed Monte Carlo algorithm to estimate the marginal smoothing distributions of the hidden states is presented in Section~\ref{sec:RaoBlackwell} along with the other widely used Rao Blackwellised smoothing algorithms.
+An overview of the proposed Monte Carlo algorithms to estimate the smoothing distributions of the hidden states is presented in Section~\ref{sec:RaoBlackwell}.
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
-Then, the parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.
-The performance of this algorithm is assessed using simulated data and CME crude oil (WTI) data.
+Then, a new model is proposed to explain regime switchings for commodity markets.
+The parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.
+The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data.
 Intermediate computations and explicit approximations provided by the algorithm are given in the Appendix.
@@
 SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$.
-All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distribution and share the same forward filter detailed in Section~\ref{sec:forward}.
+All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter detailed in Section~\ref{sec:forward}.
 
@@
 \]
-In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter.
+In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter as follows.
 \paragraph{Initialization}
@@
 
-\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
-\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}.
+\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
+\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}.
 The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
@@
 
-\subsection{Rao Blackwellised Two-filter Smoother}
+\subsection{Rao Blackwellised Two-filter Smoother (RBTF)}
 Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}.
 Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations).
+\cite{briers:doucet:maskell:2010} introduced the following decomposition of the smoothing distributions for all $2\le i \le n$:
+\begin{equation}
+\label{eq:dens:bwd_fwd_doucet}
+p(a_{i},z_{i}|y_{1:n})  \propto  p_{}(a_{i},z_{i}|y_{1:i-1})p(y_{i:n}|a_i,z_i)\eqsp.
+\end{equation}
+The first term in the right hand side may be approximated using the forward filter, however, as the function $(a_{i},z_{i})\mapsto p_{}(y_{i:n}|a_{i},z_{i})$ is not a probability density function, it is not possible to use SMC samples to approximate it.
+The backward filter uses artificial densities to introduce a surrogate target density function which may be approximated recursively using SMC methods.
+Then, the forward and backward weighted samples are combined using \eqref{eq:dens:bwd_fwd_doucet} to approximate $p(a_{i},z_{i}|y_{1:n})$.
+%Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:
 %Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods.
-For $3\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
+For $2\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
 \begin{equation}
@@
 %\phi^n_{i}(a_{i-1:i},z_{i-1:i})  =  \sum_{a_{i-2}}\sum_{a_{i+1}} \int_{z_{i-2}} \int_{z_{i+1}} \psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
-\phi^n_{i}(a_{i},z_{i})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
+p(a_{i},z_{i}|y_{1:n})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp,
 \end{equation}
 where $\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-1$ and $i+1$.
-It may be decomposed as follows:
+Note that the EM algorithm requires the approximation of $p(a_{i-1},z_{i-1},a_{i-1},z_{i-1}|y_{1:n})$ in the E-step, this may be obtained following the same steps by marginalizing explicitly the linear states at time $i-2$ and $i+1$, see Section~\ref{sec:numerical:experiments}.
+Intermediate computations follow the same steps as  for the approximation of $p(a_{i},z_{i}|y_{1:n})$.
+$\psi^n_{i}$ may be decomposed as follows:
 \begin{multline*}
@@
 \end{multline*}
-where the proportionality is with respect to $(a_{i-1:i+1},z_{i-1:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ may be written as
+where the proportionality is with respect to $(a_{i-1:i+1},z_{i-1:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $p(a_{i},z_{i}|y_{1:n}) $ may be written as
 \begin{equation}
 \label{eq:smoothing}
-\phi^n_{i}(a_{i},z_{i})  \propto r_{i}(a_{i},z_{i},y_{1:i-1})g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
+p(a_{i},z_{i}|y_{1:n})   \propto r_{i}(a_{i},z_{i},y_{1:i-1})g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
 \end{equation}
@@
 \begin{align*}
-\Upsilon^k_{i-1}(a_{i}) &\eqdef (P^k_{i-1})^{-1} + T'_{a_{i}}\barH_{a_{i}}^{-1}T_{a_{i}}\eqsp,\\
-\upsilon^k_{i-1}(a_{i})&\eqdef \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})(P^k_{i-1})^{-1}\mu^k_{i-1}\eqsp,\\
+\Upsilon^k_{i-1}(a_{i}) &= (P^k_{i-1})^{-1} + T'_{a_{i}}\barH_{a_{i}}^{-1}T_{a_{i}}\eqsp,\\
+\upsilon^k_{i-1}(a_{i})&= \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})(P^k_{i-1})^{-1}\mu^k_{i-1}\eqsp,\\
 \Omega^k_{i-1}(a_{i}) &= \left(\barH_{a_{i}}^{-1} - \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})T'_{a_{i}}\barH_{a_{i}}^{-1}\right)^{-1}\eqsp,\\
 r_{i-1}^k(a_{i}) &= \upsilon^k_{i-1}(a_{i})+(\Omega^k_{i-1}(a_{i}))^{-1}d_{a_{i}}\eqsp,\\
-\omega_{\mathsf{f},i}^k(a_{i}) &\eqdef \omega^k_{i-1}Q(a^k_{i-1},a_{i})\left|P^k_{i-1}\right|^{-1/2}\left|\Upsilon^k_{i-1}(a_{i})\right|^{-1/2} \left|2\pi\barH_{a_{i}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{d_{a_{i}}}^2\right\}\\
+\omega_{\mathsf{f},i}^k(a_{i}) &= \omega^k_{i-1}Q(a^k_{i-1},a_{i})\left|P^k_{i-1}\right|^{-1/2}\left|\Upsilon^k_{i-1}(a_{i})\right|^{-1/2} \left|2\pi\barH_{a_{i}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{d_{a_{i}}}^2\right\}\\
 &\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-1}^k}{\mu^k_{i-1}}^2+\frac{1}{2}\normMat{P_{i-1}^k\Upsilon^k_{i-1}(a_{i})P_{i-1}^k}{\mu^k_{i-1}}^2 - (\upsilon^k_{i-1})'(a_{i})d_{a_{i}} \right\}\eqsp.
@@
 
-These steps are then combined to form a non normalized SMC approximation of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ using \eqref{eq:smoothing}.
-As $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ are not normalized, the normalization of the SMC approximation of $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
+These steps are then combined to form a non normalized SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ using \eqref{eq:smoothing}.
+As $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ are not normalized, the normalization of the SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
 %These combination and normalization steps are given in Appendix~\ref{sec:approx:smooth}.
 Our procedure allows to construct sequence of regimes with non-degenerated importance weights in the combination step.
-This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $\phi^n_{i}(a_{i},z_{i})$ over the states at times $i-1$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged.
+This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $p(a_{i},z_{i}|y_{1:n})$ over the states at times $i-1$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged.
 This method often leads to importance weights which are close to be numerically degenerated.
@@
 As the function $(a_{i+1},z_{i+1})\mapsto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$ is not a probability density function, it is not possible to use SMC samples to approximate it.
-The backward filter relies on artificial densities to introduce a surrogate target density function which may be approximated using SMC methods.
+%The backward filter relies on artificial densities to introduce a surrogate target density function which may be approximated using SMC methods.
 Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:
@@
 \end{proof}
-The SMC approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$, described in Algorithm~\ref{algo:backward},  produces sequence of states $\tilde{a}^{\ell}_{i:n}$ associated with importance weights $\tilde{\omega}^{\ell}_i$ for $1\le \ell\le N$ to define a sample based approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$ of the form $\tilde{p}^N_{}(a_{i:n}|y_{i:n}) \eqdef \sum_{\ell=1}^N \tilde{\omega}^{\ell}_i \delta_{\tilde{a}^{\ell}_{i:n}}(a_{i:n})$.
+A set of weighted trajectories $(\tilde{a}^\ell_{i:n})_{1\le \ell\le N}$ with importance weights $(\tilde{\omega}^\ell_i)_{1\le \ell\le N}$, $1\le i \le n$, may be sampled recursively backward in time to produce a SMC approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$.
+\begin{enumerate}[-]
+\item For $1\le \ell \le N$, sample $\tilde{a}^{j}_n\sim\tilde{q}_n(\cdot)$ and set:
+\[
+\tilde{\omega}^{\ell}_n \propto \frac{\int\gamma^{}_n(\tilde{a}^{\ell}_n,z')g_{}(\tilde{a}^{\ell}_n,z';y_n)\rmd z'}{\tilde{q}_n(\tilde{a}^{\ell}_n)}\eqsp.
+\]
+\item For all $1\le i \le n-1$, resample the set $(\tilde{a}^{\ell}_{i+1:n})_{1\le j\le N}$ using the normalized weights $(\tilde{\omega}^{\ell}_{i+1})_{1\le j \le N}$.
+Then, for $1\le \ell \le N$, sample $\tilde{a}^{j}_{i}\sim\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot)$  and set:
+\[
+\tilde{\omega}^{\ell}_i \propto \frac{q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
+\]
+\end{enumerate}
+To obtain uniformly weighted samples at each time step, in the numerical experiments we use:
+\[
+\tilde{q}_n(\cdot) = \int\gamma^{}_n(\cdot,z')g_{}(\cdot,z';y_n)\rmd z'\quad\mbox{and}\quad \tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot) = \frac{q(\cdot,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\cdot,z')p_{}(y_{i:n}|(\cdot,\tilde{a}^{\ell}_{i+1:n}),z')\rmd z'}{\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
+\]
 By \eqref{eq:p:y1:n} and \eqref{eq:p:ai:zi}, we get
@@
 p_{}(y_{i:n}|a_i,z_i)
-&= \frac{\tilde{p}_{i}(y_{i:n})  \tilde{p}_{i}(a_i,z_i|y_{i:n})}{\gamma^{}_{i}(a_i,z_i)} \\
+&= \frac{\tilde{p}_{i}(y_{i:n})  \tilde{p}_{i}(a_i,z_i|y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\eqsp, \\
 &= \tilde{p}_{i}(y_{i:n})  \sum_{a_{i+1:n}=1}^J\frac{\tilde{p}_{i}(a_{i:n}|y_{i:n})p_{}(y_{i:n}|a_{i:n},z_{i})}{\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'}\label{eq:p:ai:zi}\eqsp,
 \end{align*}
-which suggests the following particle approximation $p^N_{}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$:
-\[
-p^N_{}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_{i}(y_{i:n}) \sum_{\ell=1}^N\frac{\tilde{\omega}^{\ell}_i p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z_{i})}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}\delta_{\tilde{a}^{\ell}_{i}}(a_{i})\eqsp.
+which suggests the following particle approximation $p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$:
+\[
+p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_{i}(y_{i:n}) \sum_{\ell=1}^N\frac{\tilde{\omega}^{\ell}_i p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z_{i})}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}\delta_{\tilde{a}^{\ell}_{i}}(a_{i})\eqsp.
 \]
@@
 \end{proof}
-By Lemma~\ref{lem:py}, the SMC approximation $p_{}^N(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$ is then given by
+By Lemma~\ref{lem:py}, the SMC approximation $p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$ is then given by
 \begin{equation}
 \label{eq:pN:yi:n}
-p_{}^N(y_{i:n}|a_{i},z_{i}) = \tilde{p}_i(y_{i:n}) \sum_{\ell=1}^N  \frac{\delta_{\tilde{a}^{\ell}_i}(a_i)\tilde{\omega}_{i}^{\ell}}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'} \exp\left\{- \frac{1}{2}\normMat{\tilde{P}^{\ell}_{i}}{z_i}^2 + z'_i\tilde{\nu}^{\ell}_{i} - \frac{1}{2} \tilde{c}^{\ell}_i \right\}\eqsp,
+p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_i(y_{i:n}) \sum_{\ell=1}^N  \frac{\delta_{\tilde{a}^{\ell}_i}(a_i)\tilde{\omega}_{i}^{\ell}}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'} \exp\left\{- \frac{1}{2}\normMat{\tilde{P}^{\ell}_{i}}{z_i}^2 + z'_i\tilde{\nu}^{\ell}_{i} - \frac{1}{2} \tilde{c}^{\ell}_i \right\}\eqsp,
 \end{equation}
@@
 \end{align*}
-Then, by \eqref{eq:smooth:t}, the SMC approximation $t^N_{i}(a_i,z_i,y_{i:n})$ of $t_{i}(a_i,z_i,y_{i:n})$ is given by:
+Then, by \eqref{eq:smooth:t}, the SMC approximation $t^N_{i}(a_i,z_i,y_{i+1:n})$ of $t_{i}(a_i,z_i,y_{i+1:n})$ is given by:
 \begin{align}
-t^N_{i}(a_i,z_i,y_{i:n}) &=\sum_{a_{i+1}=1}^J\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p^N_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp,\nonumber\\
+t^N_{i}(a_i,z_i,y_{i+1:n}) &=\sum_{a_{i+1}=1}^J\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p^N_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp,\nonumber\\
 &=  \tilde{p}_{i+1}(y_{i+1:n}) \sum_{\ell=1}^N C_i^{-1}(\tilde{a}^{\ell}_{i+1:n}) Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_i H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2} \eqsp\nonumber \\
 &\hspace{4.3cm} \times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{\tilde{a}^{\ell}_{i+1}}}{d_{\tilde{a}^{\ell}_{i+1}}+T_{\tilde{a}^{\ell}_{i+1}}z_i}^2\right\} \eqsp,\\
-&= \sum_{\ell=1}^N\tilde{\omega}_{\mathsf{b},i}^\ell(a_i)\exp\left\{-\frac{1}{2}\normMat{\tilde{S}_{i+1}^\ell}{z_i}+z_i'\tilde{s}_{i+1}^\ell\right\}\eqsp,\label{eq:tN}
+&= \sum_{\ell=1}^N\tilde{\omega}_{\mathsf{b},i}^\ell(a_i)\exp\left\{-\frac{1}{2}\normMat{\tilde{S}_{i+1}^\ell}{z_i}^2+z_i'\tilde{s}_{i+1}^\ell\right\}\eqsp,\label{eq:tN}
 \end{align}
@@
 \end{align*}
-In the experiments (see section~\ref{sec:exp}), $\gamma^{}_i(a_i,z_i)$ is set as a mixture of Gaussian distributions with parameters specified below. Note that for such a choice,
+In the experiments (see section~\ref{sec:exp}), $\gamma^{}_i(a_i,z_i)$ is set as a mixture of Gaussian distributions. Note that for such a choice,
  the integral $\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'$ may be computed explicitly; see Lemma~\ref{lem:integral:gammap}.
 This choice differs from \cite{briers:doucet:maskell:2010} where it is advocated to set $\gamma^{}_i$ as the product of two independent densities $\gamma_i^{a}(a_i)$ and $\gamma_i^{z}(z_i)$.
- The accuracy of the algorithm relies heavily on a proper tuning of this artificial density: it is shown in Section~\ref{sec:exp} that fitting a mixture improves the performance of the algorithm.
-Let $\phi_{\mu,\Sigma}$ be the probability density function of a $\dimz$ dimensional Gaussian random variable with mean $\mu$ and variance matrix $\Sigma$.
+The accuracy of the algorithm relies heavily on a proper tuning of this artificial density.%: it is shown in Section~\ref{sec:exp} that fitting a mixture improves the performance of the algorithm.
 
@@
 \end{multline*}
-where
+where $\phi_{\mu,\Sigma}$ is the probability density function of a $\dimz$ dimensional Gaussian random variable with mean $\mu$ and variance matrix $\Sigma$ and
 \[
@@
 \label{sec:numerical:experiments}
-This section highlights the performance of the Rao Blackwellised Two-Filter approximation (RBTF) of the marginal smoothing distributions using both simulated data and a real data set.
-This approximation is compared to the one provided by \cite{briers:doucet:maskell:2010}.
-We also describe numerically the impact of the artificial law choice on the accuracy and stability of the algorithms.
-\subsection{Choice of artificial distributions}
-\label{sec:artificial:distribution}
-As noted in \cite{briers:doucet:maskell:2010}, the backward filter performance highly depends on the sequence $(\gamma_i)_{0\le i\le n}$.
-A poor choice may imply degeneracy of the weighted samples produced by the backward pass and lead to  a poor estimation of the posterior probabilities of the state variables given the observations.
-%In contrast of the forward pass, the initial law of state variables are not totally observed in almost cases but we have at least an idea to guess its distribution.
-But it is not the case for the distribution at terminal time.
-Because it depends on the evolution of state variables in the past.
-%We propose three artificial laws.
-%In the next section, we compared the performance of our proposition in backward filter in comparaison to the one of \cite{briers:doucet:maskell:2010} and also the performance of these three artificial distributions.
-%The first and the third one are our propositions and the second one is the one proposed by \cite{briers:doucet:maskell:2010} from our exchange.
-\begin{enumerate}[-]
-\item %According to \cite{briers:doucet:maskell:2010}, a sensible choice for $\gamma_i$ is given by the prior distribution of the states $(a_i,z_i)$.
-%In our case, an easy to implement solution to approximate this
-A first choice for $\gamma_i$ is given by an estimation of the prior distribution of the states $(a_i,z_i)$ obtained by simulating $P$ independent chains $\left(a^{p}_{1:n}, z^{p}_{1:n} \right)_{1\le p \le P}$ using \eqref{eq:model:state}.
-%and \eqref{eq:model:obs}.
-Then, for all $1\le i \le n$, $\gamma^{}_i$ is defined as
-\begin{align*}
-\gamma^{}_i (a_i, z_i) &= \sum_{p=1}^{P} m_{}(a^{p}_{i-1},z^p_{i-1}; z_{i}) Q(a^{p}_{i-1},a_i)/P\eqsp,\\
-&= (2\pi)^{-\dimz/2}\sum_{p=1}^{P} Q(a^{p}_{i-1},a_i) \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^{p}_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^{p}_{i-1}}}{z_i - d_{a^p_{i-1}} -T z^{p}_{i-1}}^2\right\}/P \eqsp.
-\end{align*}
-%\paragraph{Last time step}
-%Define the terminal proposal distribution as $\tilde{Q}_n(\tilde{a}_n^{\ell}) \propto \int_{z_n} \gamma^{}_n(\tilde{a}_n^{\ell},z_n) g_{}(\tilde{a}_n^{\ell}, z_n, y_n) \rmd z_n$.
-By \eqref{eq:model:obs},
-%\begin{multline*}
-%\tilde{Q}_n(\tilde{a}_n^{\ell})  \propto \sum_{p=1}^{P} Q(a^{p}_{n-1},\tilde{a}_n^{\ell}) \exp\left\{ -\frac{1}{2} \log \left| \tilde{\Lambda}_p \right| +\frac{1}{2}\normMat{\tilde{\Lambda}_p}{\tilde{\beta}_p}^2 - \frac{1}{2}\normMat{\barG}{y_n-c_{\tilde{a}_n^{\ell}}} \right\} \\
-%\times \exp\left\{ -\frac{1}{2} \log|\barH_{a^p_{n-1}}| -\frac{1}{2}\normMat{\barH_{a^p_{n-1}}}{d_{a^p_{n-1}} + T z^p_{n-1}}\right\} \eqsp.
-%\end{multline*}
-%where
-%\[
-%\tilde{\Lambda}_p \eqdef \barH^{-1}_{a^p_{n-1}} + B'\barG^{-1} B \quad \mbox{and} \quad \tilde{\beta}_p \eqdef \barH^{-1}_{a^p_{n-1}} \left( d_{a^p_{n-1}} + T z^p_{n-1} \right) + B'\barG^{-1}(y_n-c_{a^l_{n}})\eqsp.
-%\]
-%For this choice of terminal proposal distribution, the corresponding particles are uniformly weighted i.e. for all $1 \le \ell \le N$, $\tilde{\omega}^{\ell}_n = 1/N$.
-%\paragraph{Backward iterations}
-%For $i=n-1$ to $i=n$, the proposal distribution $\tilde{q}_i$ is defined as
-%\begin{align*}
-%\tilde{Q}_i (\tilde a^{\ell}_{i+1:n}, \tilde a^{\ell}_i) &\propto \frac{Q(\tilde a^{\ell}_{i},\tilde a^{\ell}_{i+1}) \int_{z_i} \gamma^{}_i(\tilde a^{\ell}_{i},z_i) p_{}(y_{i:n}|\tilde a^{\ell}_{i:n},z_i) \rmd z_i }{\int_{z_{i+1}} \gamma^{}_{i+1}(\tilde a^{\ell}_{i+1},z_{i+1}) p_{}(y_{i+1:n}|\tilde a^{\ell}_{i+1:n},z_{i+1}) \rmd z_{i+1}}\eqsp,
-%\end{align*}
-%with, by Lemma~\ref{lem:py},
-%\begin{align*}
-%\int_{z_i} \gamma^{}_i(a_{i},z_i) p_{}(y_{i:n}|a_{i:n},z_i) dz_i &= \sum_{p=1}^{P} \int_{z_i} Q(a^{p}_{i-1}, a_{i}) m_{}(a^p_{i-1}, z^{p}_{i-1}; z_i) p_{}(y_{i:n}|a_{i:n},z_i) \rmd z_i/P \eqsp,\\
-%&\hspace{-4.2cm} \propto \sum_{p=1}^{P} q(a^{p}_{i-1},a_i) \exp \left\{ -\frac{1}{2} \log\left| \barH_{a^p_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^p_{i-1}}}{d_{a^p_{i-1}} + T z^p_{i-1}}^2 - \frac{1}{2} c_{i}(a_{i:n}) \right\}\\
-%&\hspace{-3.2cm} \times \int \exp \left\{-\frac{1}{2} z'_{i} \left( \barH^{-1}_{a^p_{i-1}} + \tilde{P}^{-1}_{i}(a_{i:n})  \right) z_i + z'_i \left( \barH^{-1}_{a^p_{i-1}} \left( d_{a^p_{i-1}} + T z^p_{i-1} \right) + \nu_{i}(a_{i:n}) \right)  \right\} \rmd z_i \eqsp, \\
-%&\hspace{-3.2cm} = \sum_{p=1}^{P} Q(a^{p}_{i-1},a_i) \left| \barH_{a^p_{i-1}} \right|^{-1/2}\exp \left\{-\frac{1}{2}\normMat{\barH_{a^p_{i-1}}}{d_{a^p_{i-1}} + T z^p_{i-1}}^2 - \frac{1}{2} c_{i}(a_{i:n}) \right\} \eqsp\\
-%&\hspace{-2.2cm} \times \left|R^p_{i-1}(a_{i:n})\right|^{1/2}\exp \left\{\frac{1}{2} \delta'_{i-1}(a^p_{i-1},a_{i:n}) R^p_{i-1}(a_{i:n}) \delta_{i-1}(a^p_{i-1},a_{i:n})  \right\}/P \eqsp,
-%\end{align*}
-%where
-%\begin{align*}
-%R^p_{i-1}(a_{i:n}) \eqdef \left( \barH^{-1}_{a^p_{i-1}} + \tilde{P}^{-1}_{i}(a_{i:n}) \right)^{-1} &= H_{a^p_{i-1}} \Delta_{i-1}(a^p_{i-1},a_{i:n})H'_{a^p_{i-1}}\eqsp,\\
-%\delta_{i-1}(a^p_{i-1:n}) &\eqdef \left( \barH^{-1}_{a^p_{i-1}} \left( d_{a^p_{i-1}} + T z^p_{i-1} \right) + \nu_{i}(a_{i:n}) \right)\eqsp.
-%\end{align*}
-%For this choice of proposal distribution, the corresponding particles are uniformly weighted i.e. for all $1 \le \ell \le N$, $\tilde{\omega}^{\ell}_i = 1/N$.
-
-%\paragraph{Application for the approximation of $t_{i}$}
-%Then,
-%\begin{multline}
-%\label{eq:tN}
-%t^N_{i}(a_i,z_i,y_{i:n}) =  \tilde{p}_{i+1}(y_{i+1:n})g(a_i,z_i,y_i) \sum_{\ell=1}^N C^{-1}_i(\tilde a^{\ell}_{i+1:n}) Q(a_i, \tilde a^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{a_i}|^{-1/2} |H_{a_i} \Delta^{\ell}_i H'_{a_i}|^{1/2} \eqsp \\
-%\times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{a_i}\Delta^l_i H'_{a_i} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{a_i}}{d_{a_i}+Tz_i}\right\} \eqsp,
-%\end{multline}
-%where
-%\begin{multline*}
-%C_i(\tilde a^{\ell}_{i+1:n})  =  \sum_{p=1}^{P} Q(a^{p}_{i},\tilde a^{\ell}_{i+1}) \exp \left\{ -\frac{1}{2} \log\left| \barH_{a^p_{i}} \right| -\frac{1}{2}\normMat{\barH_{a^p_{i}}}{d_{a^p_{i}} + T z^p_{i}}^2\right\} \\
-%\times \exp \left\{\frac{1}{2} \log \left|H_{a^p_{i}} \Delta^{\ell}_{i}(a^p_i) H'_{a^p_{i}}\right| + \frac{1}{2} (\delta^{\ell}_{i}(a_i^p) )'  H_{a^p_{i}} \Delta^{\ell}_{i}(a^p_i)H'_{a^p_{i}} \delta^{\ell}_{i}(a_i^p)  \right\}/P \eqsp.
-%\end{multline*}
-This approach may lead to poor results as it does not take into account the observations received up to time $i$.
-Its performance depends heavily on the number of simulations $P$ and a large value of $P$ increases the computational cost of the algorithm.
-%It can be not very stable because of lack of the information from observations obtained by the forward pass. This default is corrected by the second artificial law.
-
-%\item Another choice proposed by one of the authors of \cite{briers:doucet:maskell:2010} is the distribution of the states at time $i$ obtained from the forward pass $(\omega^{k}_i, \mu^k_{i}, P^k_{i})^{N}_{k=1}$.
-For all $1\le i \le n$, $\gamma^{}_i$ is defined as
-%\begin{align*}
-%\gamma^{}_i (a_i, z_i) &= \sum_{k=1}^{N} \delta_{a_i}(a^k_i) \omega^{k}_i \exp\left\{ -\frac{1}{2} \log \left|\bar{P}^k_{i} \right| -\frac{1}{2}\normMat{\bar{P}^k_{i}}{z_i - \mu^k_{i} }^2\right\} \eqsp.
-%\end{align*}
-%This approach uses the information brought by the observations and reduces the computational cost as it does not require to sample a large number of sequences of states only to compute $\gamma_i$.
-%But by our experiments, using directly and exactly the forward pass information can be not a perfect choice.
-Since forward filter doesn't exprime partly but not exactly the distribution of state variable for given observations.
-The aim of backward pass is to complement the lack of forward pass.
-So the backward pass needs to be estimated more flexibly.
-
-\item $\gamma_i$ may also be computed as a combination of the two previous approaches:
-%At time step $i$, we store $(\mu^k_{i-1}), k=1,\ldots,N$ obtained during the forward pass of the previous step $i-1$ weighted with $\omega^k_{i-1}, k=1,\ldots,N$ and take the variance $\barH_{a^k_{i-1}}, k=1,\ldots,N$ naturally obtained from \eqref{eq:model:state}.
-Formally,
-\begin{align*}
-\gamma^{}_i (a_i, z_i) &= \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
-\end{align*}
-This approach is more flexible since it matches freely the forward distribution with the next time step using \eqref{eq:model:state}.
-\end{enumerate}
+This section highlights the performance of the Rao Blackwellised Two-Filter approximation (RBTF) of the marginal smoothing distributions in comparison to the other Rao Blackwellised smoothers introduced in Section~\ref{sec:RaoBlackwell}.
 
@@
 The performance of the algorithms are analyzed in two different settings with $n=1500$ observations.
-In both settings the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data.
-The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data, see Section~\ref{sec:exp}.
+In both settings the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}.
+The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.
 It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$).
@@
 
-Figure~\ref{fig:exp:artificial} displays the estimated posterior distributions of the labels $(a_k)_{1\le k\le 500}$ given by RBTF for the first $500$ time steps with the  artificial distributions introduced in Section~\ref{sec:artificial:distribution}.
-In experiment~1, the three artificial laws behave similarly except that the first choice is unable to detect the last regime switching for more than 50 time steps.
+%Figure~\ref{fig:exp:artificial} displays the estimated posterior distributions of the labels $(a_k)_{1\le k\le 500}$ given by RBTF for the first $500$ time steps with the  artificial distributions:
+
+%In experiment~1, the three artificial laws behave similarly except that the first choice is unable to detect the last regime switching for more than 50 time steps. 
 %In experiment 2, the second method is completely unstable which indicates that the third artificial leads to the best performance.
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=.5]{Artificial_Compare_Exp1_1_500.eps}
-\includegraphics[scale=.5]{Artificial_Compare_Exp2_1_500.eps}
-\caption{estimated probabilities by RBTF in experiment 1 [top] and experiment 2 [bottom] with the two artificial laws: yellow (the first one) and red (the third one).}
-\label{fig:exp:artificial}
-\end{figure}
-
-Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms.
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.5]{Artificial_Compare_Exp1_1_500.eps}
+%\includegraphics[scale=.5]{Artificial_Compare_Exp2_1_500.eps}
+%\caption{estimated probabilities by RBTF in experiment 1 [top] and experiment 2 [bottom] with the two artificial laws: yellow (the first one) and red (the third one).}
+%\label{fig:exp:artificial}
+%\end{figure}
+
+Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms using:
+\[
+\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
+\]
 In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
@@
 \subsubsection*{State dynamics}
-As explained in Section~\ref{sec:intro}, the spot price and convenience yield are described in this paper as:
+Modeling commodity prices is a crucial step to valuate contingent claims related to energy markets and to optimize storage or extraction strategies.
+In \cite{gibson:schwartz:1990,schwartz:1997}, the authors proposed a model where the spot price of a commodity$(S_t, t \geq 0)$ depends on a second factor $(\delta_t, t \geq 0)$, referred to as the instantaneous convenience yield.
+This factor plays the role of dividends in equity markets and models the benefit of holding the physical commodity or the storage and maintenance costs required to keep the commodity.
+In this model, this convenience yield is described as an Ornstein-Uhlenbeck process:
+\begin{align*}
+\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma S_{t} \rmd W_{t}^{1} \eqsp,\\
+\rmd \delta_{t} & = \kappa(\alpha-\delta_{t})\rmd t+\eta \rmd W_{t}^{2} \eqsp, \quad \rmd \langle W_t^1, W_t^2\rangle = \rho \rmd t\eqsp,
+\end{align*}
+where the parameter $\left(r, \sigma, \kappa, \alpha, \eta, \rho \right)$ are constant and $( (W_t^1,W_t^2), t \geq 0)$ are standard Brownian motions.
+This model appear to be too restrictive as energy markets are not likely to revert to a single equilibrium value.
+This assumption is relaxed using Markov switching models to allow several possible regimes for the spot price and the convenience yield.
+Following \cite{almansour:2016}, the spot price and convenience yield are described in this paper as:
 \begin{align*}
@@
 \end{align*}
-We assume that $J=2$, each regime characterizing the slope of the future curve.
-A backwardation regime (resp.
-contango regime) is defined as a negative slope (resp.
-a positive slope).
+where $(a_t)_{t\ge 0}$ is a finite state space Markov process. 
+%We assume that $J=2$, each regime characterizing the slope of the future curve. A backwardation regime (resp. contango regime) is defined as a negative slope (resp. a positive slope). 
 Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$.
@@
 where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $.
+This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield.
+A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).
+Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a Conditionally Linear Gaussian Model.
 The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:
@@
 %\label{eq:completelik}
-p_{\theta}(a_{1:n},z_{1:n},Y_{1:n}) \eqdef \pi(a_1)\phi_{\mu_1,\Sigma_1}(z_1)g_{\theta}(a_1,z_1;Y_1)\prod^{n}_{i=2}Q(a_{i-1},a_i)m_{\theta}\left(a_{i-1},z_{i-1};z_i\right)g_{\theta}(a_i,,z_i;Y_i)\eqsp.
+p_{\theta}(a_{1:n},z_{1:n},Y_{1:n}) \eqdef \pi(a_1)\phi_{\mu_1,\Sigma_1}(z_1)g_{\theta}(a_1,z_1;Y_1)\prod^{n}_{i=2}Q(a_{i-1},a_i)m_{\theta}\left(a_{i},z_{i-1};z_i\right)g_{\theta}(a_i,,z_i;Y_i)\eqsp,
 \end{equation*}
+where $\phi_{\mu_1,\Sigma_1}$ is the Gaussian probability density function with mean $\mu_1$ and variance $\Sigma_1$.
 Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{1:n}\right]$  the conditional expectation given $Y_{1:n}$ when the parameter value is $\theta$.
@@
 %\end{align*}
-All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our proposed algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
+All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our RBTF algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
 %\paragraph{M-step}
@@
 \cite{hansen:kern:2004}.
-\\
 
@@
 \end{figure}\\
-The number of particles is set to $N = 100$, $\Delta t = 1/52$ and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}.
-The interest rate is $r = 0.0296$ as in \cite{almansour:2016}.
-The initial guess for the mean and variance of the initial state are
+The number of particles is set to $N = 100$, $\Delta t = 1/52$. % and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}. 
+The interest rate is $r = 0.0296$ as in \cite{almansour:2016}. The initial guess for the mean and variance of the initial state are
 \[
@@
 
-\section{Conclusions}
-In this paper, we propose an extension of the two-filter algorithm to estimate the conditional distribution of the hidden states given the observations in a Conditionally Linear and Gaussian model (CLGM).
-This extension improves the performance of the usual Rao Blackwellised two-filter algorithm using few particles by marginalizing explicitly parts of the paths before combining the forward and backward filters.
-The paper analyzes the behavior of the Term Structure in a crude oil market by discretizing a Gibson-Schwartz model with regime switchings.
-Numerical results highlight that the estimated posterior probabilities of the regimes are closely related to backwardation and contango effects in crude oil market.
-
 \appendix
@@
 
-\clearpage
-\newpage
-
-\section{Algorithms}
-The forward filter is described in Algorithm~\ref{algo:forward}.
+%\clearpage
+%\newpage
+%
+%\section{Algorithms}
+%The forward filter is described in Algorithm~\ref{algo:forward}.
 The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^j$ may be written:
-\begin{align*}
-\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
-& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
-\end{align*}
-\begin{algorithm}
-\DontPrintSemicolon
-\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
-\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
-{\bf Initialization}\;
-\For{$k \gets 1$ \textbf{to} $N$} {
-Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
-Compute $\mu_1^k$ and $P_1^k$:
-\begin{align*}
-K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
-\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
-P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
-\end{align*}
-Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
-}
-{\bf Iterations}\;
-\For{$i \gets 2$ \textbf{to} $n$} {
-Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
-\For{$k \gets 1$ \textbf{to} $N$} {
-For all $j \in \{1,\ldots, J\}$ compute
-\[
-\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
-\]
-Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
-Update $\mu_i^k$ and $P_i^k$:
-\begin{align*}
-\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
-P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
-K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
-\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
-P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
-\end{align*}
-Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
-}
-}
-\caption{Forward filter, see \cite{chen:liu:2000}}
-\label{algo:forward}
-\end{algorithm}
-
-\begin{algorithm}
-\DontPrintSemicolon
-\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
-\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
-{\bf Initialization}\;
-Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
-{\bf Iterations}\;
-\For{$i \gets n-1$ \textbf{to} $1$} {
-\For{$k \gets 1$ \textbf{to} $N$} {
-Compute
-\[
-\omega^k_{i|n} \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})\eqsp.
-\]
-}
-Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$.
-}
-\caption{Smoothing algorithm based on the approximation of \cite{kim:1994}}
-\label{algo:kim}
-\end{algorithm}
-
-\begin{algorithm}
-\DontPrintSemicolon
-\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
-\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
-{\bf Initialization}\;
-Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
-Compute $\hat{\Omega}_n$ and $\hat{\lambda}_n$:
-\[
-\hat{\Omega}_n = B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}B_{\tilde{a}_n}\quad\mbox{and}\quad \hat{\lambda}_n =B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}(y_n-c_{\tilde{a}_n})\eqsp.
-\]
-{\bf Iterations}\;
-\For{$i \gets n-1$ \textbf{to} $1$} {
-Compute 
-\[
-m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{\tilde{a}_{i+1}}\quad\mbox{and}\quad M_{i+1} = H_{\tilde{a}_{i+1}}'\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}} + I\eqsp.
-\]
-Set
-\begin{align*}
-\Omega_i &= T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})\hat{\Omega}_{i+1}T_{\tilde{a}_{i+1}}\eqsp,\\
-\lambda_i &=T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})m_{i+1}\eqsp.
-\end{align*}
-\For{$k \gets 1$ \textbf{to} $N$} {
-Compute
-\[
-\Lambda^k_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I \quad\mbox{and}\quad \eta^k_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}\eqsp.
-\]
-Set
-\[
-\tilde{\omega}_{i|n}^k \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})|\Lambda^k_i|^{-1/2}\exp\left\{\eta^k_i/2\right\}\eqsp.
-\]
-}
-Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$ and
-\[
-\hat{\Omega}_i =  \Omega_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}B_{\tilde{a}_i} \quad\mbox{and}\quad \hat{\lambda}_i = \lambda_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}(y_i-c_{\tilde{a}_i})\eqsp.
-\]
-}
-\caption{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
-\label{algo:FFBS}
-\end{algorithm}
-
-\begin{algorithm}[b]
-\DontPrintSemicolon
-\KwIn{Artificial densities $(\gamma^{}_i)_{1\le i\le n}$ and proposal densities $(\tilde{q}_i)_{1\le i \le n}$.}
-\KwOut{A set of weighted trajectories $(\tilde{a}^j_{i:n})_{1\le j\le N}$ with importance weights $(\tilde{\omega}^j_i)_{1\le j\le N}$, $1\le i \le n$.}
-{\bf Initialization}\;
-For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_n\sim\tilde{q}_n(\cdot)$.\;
-For $1\le \ell \le N$, set:
-\[
-\tilde{\omega}^{\ell}_n \propto \frac{\int\gamma^{}_n(\tilde{a}^{\ell}_n,z')g_{}(\tilde{a}^{\ell}_n,z';y_n)\rmd z'}{\tilde{q}_n(\tilde{a}^{\ell}_n)}\eqsp.
-\]
-{\bf Iterations}\;
-\For{$i \gets n-1$ \textbf{to} $1$} {
-Resample the set $(\tilde{a}^{\ell}_{i+1:n})_{1\le j\le N}$ using the normalized weights $(\tilde{\omega}^{\ell}_{i+1})_{1\le j \le N}$.\;
-For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_{i}\sim\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot)$.\;
-For $1\le \ell \le N$, set:
-\[
-\tilde{\omega}^{\ell}_i \propto \frac{q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
-\]
-}
-\caption{Backward information filter, see \cite[Section~4.1.1]{briers:doucet:maskell:2010}}
-\label{algo:backward}
-\end{algorithm}
+%\begin{align*}
+%\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
+%& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
+%\end{align*}
+%\begin{algorithm}
+%\DontPrintSemicolon
+%\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
+%\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
+%{\bf Initialization}\;
+%\For{$k \gets 1$ \textbf{to} $N$} {
+%Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
+%Compute $\mu_1^k$ and $P_1^k$:
+%\begin{align*}
+%K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
+%\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
+%P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
+%\end{align*}
+%Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
+%}
+%{\bf Iterations}\;
+%\For{$i \gets 2$ \textbf{to} $n$} {
+%Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
+%\For{$k \gets 1$ \textbf{to} $N$} {
+%For all $j \in \{1,\ldots, J\}$ compute
+%\[
+%\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
+%\]
+%Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
+%Update $\mu_i^k$ and $P_i^k$:
+%\begin{align*}
+%\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
+%P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
+%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
+%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
+%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
+%\end{align*}
+%Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
+%}
+%}
+%\caption{Forward filter, see \cite{chen:liu:2000}}
+%\label{algo:forward}
+%\end{algorithm}
+%
+%\begin{algorithm}
+%\DontPrintSemicolon
+%\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
+%\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
+%{\bf Initialization}\;
+%Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
+%{\bf Iterations}\;
+%\For{$i \gets n-1$ \textbf{to} $1$} {
+%\For{$k \gets 1$ \textbf{to} $N$} {
+%Compute
+%\[
+%\omega^k_{i|n} \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})\eqsp.
+%\]
+%}
+%Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$.
+%}
+%\caption{Smoothing algorithm based on the approximation of \cite{kim:1994}}
+%\label{algo:kim}
+%\end{algorithm}
+%
+%
+%\begin{algorithm}
+%\DontPrintSemicolon
+%\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
+%\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
+%{\bf Initialization}\;
+%Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
+%Compute $\hat{\Omega}_n$ and $\hat{\lambda}_n$:
+%\[
+%\hat{\Omega}_n = B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}B_{\tilde{a}_n}\quad\mbox{and}\quad \hat{\lambda}_n =B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}(y_n-c_{\tilde{a}_n})\eqsp.
+%\]
+%{\bf Iterations}\;
+%\For{$i \gets n-1$ \textbf{to} $1$} {
+%Compute 
+%\[
+%m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{\tilde{a}_{i+1}}\quad\mbox{and}\quad M_{i+1} = H_{\tilde{a}_{i+1}}'\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}} + I\eqsp.
+%\]
+%Set
+%\begin{align*}
+%\Omega_i &= T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})\hat{\Omega}_{i+1}T_{\tilde{a}_{i+1}}\eqsp,\\
+%\lambda_i &=T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})m_{i+1}\eqsp.
+%\end{align*}
+%\For{$k \gets 1$ \textbf{to} $N$} {
+%Compute
+%\[
+%\Lambda^k_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I \quad\mbox{and}\quad \eta^k_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}\eqsp.
+%\]
+%Set
+%\[
+%\tilde{\omega}_{i|n}^k \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})|\Lambda^k_i|^{-1/2}\exp\left\{\eta^k_i/2\right\}\eqsp.
+%\]
+%}
+%Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$ and
+%\[
+%\hat{\Omega}_i =  \Omega_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}B_{\tilde{a}_i} \quad\mbox{and}\quad \hat{\lambda}_i = \lambda_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}(y_i-c_{\tilde{a}_i})\eqsp.
+%\]
+%}
+%\caption{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
+%\label{algo:FFBS}
+%\end{algorithm}
+%
+%
+%\begin{algorithm}[b]
+%\DontPrintSemicolon
+%\KwIn{Artificial densities $(\gamma^{}_i)_{1\le i\le n}$ and proposal densities $(\tilde{q}_i)_{1\le i \le n}$.}
+%\KwOut{A set of weighted trajectories $(\tilde{a}^j_{i:n})_{1\le j\le N}$ with importance weights $(\tilde{\omega}^j_i)_{1\le j\le N}$, $1\le i \le n$.}
+%{\bf Initialization}\;
+%For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_n\sim\tilde{q}_n(\cdot)$.\;
+%For $1\le \ell \le N$, set:
+%\[
+%\tilde{\omega}^{\ell}_n \propto \frac{\int\gamma^{}_n(\tilde{a}^{\ell}_n,z')g_{}(\tilde{a}^{\ell}_n,z';y_n)\rmd z'}{\tilde{q}_n(\tilde{a}^{\ell}_n)}\eqsp.
+%\]
+%{\bf Iterations}\;
+%\For{$i \gets n-1$ \textbf{to} $1$} {
+%Resample the set $(\tilde{a}^{\ell}_{i+1:n})_{1\le j\le N}$ using the normalized weights $(\tilde{\omega}^{\ell}_{i+1})_{1\le j \le N}$.\;
+%For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_{i}\sim\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot)$.\;
+%For $1\le \ell \le N$, set:
+%\[
+%\tilde{\omega}^{\ell}_i \propto \frac{q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
+%\]
+%}
+%\caption{Backward information filter, see \cite[Section~4.1.1]{briers:doucet:maskell:2010}}
+%\label{algo:backward}
+%\end{algorithm}
 

### 2016-11-04 TexArxiv/clgm.tex
Changes in forward filter
Revised introduction

@@
 This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces.
-Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for conditionally linear and Gaussian models.
+Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models.
 The most widely used Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{kim:1994,lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the Forward Backward decomposition of the marginal smoothing distribution.
@@
 Conditionally on the state sequence $\{Z_k\}_{k\ge 1}$ the observations are independent and for all $\ell\ge 1$ the conditional distribution of $Y_{\ell}$ given $\{Z_k\}_{k\ge 1}$ depends on $Z_{\ell}$ only.
-These models are used in a large variety of disciplines such as econometrics, target tracking, biology, signal processing, see \cite{delmoral:2013} and the references therein.
+These models are used in a large variety of disciplines such as financial econometrics, biology, signal processing, see \cite{delmoral:2013} and the references therein.
 In general state space models, bayesian filtering and smoothing problems, i.e.
-the computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks required for instance  to perform maximum likelihood inference.
-Filtering refers to the estimation of the distributions of $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time k, while smoothing is the estimation of the distribution of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$.
-These distributions are in particular crucial to infer a sequence of states given the observations in real world online learning applications, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications...
-In the case where the state and observation models are linear and Gaussian, the filtering and smoothing problems may be solved optimally using the Kalman filter or the Rauch-Tung-Striebel smoother of \cite{kalman:1960} and \cite{rauch:striebel:tung:1965}.
-Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters or Gaussian sum filters...
-
-This paper focuses on a special class of state space models, known as Conditionally Linear and Gaussian Models (CLGM), of the form:
+the computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks.
+Filtering refers to the estimation of the distributions of the hidden state $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time $k$, while fixed-interval smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$.
+These distributions are in particular crucial to infer a sequence of states given the observations, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications...
+In the case where the state and observation models are linear and Gaussian, the filtering is solved explicitly using the Kalman filter \cite{kalman:1960} while the smoothing problem is solved using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier smoother \cite{}.
+%Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters, Gaussian sum filters or Se...
+This paper focuses on  Conditionally Linear and Gaussian Models (CLGM) given for $k\ge 1$ by:
 \begin{equation}
@@
 \begin{enumerate}[-]
-\item $(\varepsilon_k)_{k\ge 0}$ is a sequence of independent and identically distributed (i.i.d.) standard Gaussian variables,
-\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in $\{1,\ldots,J\}$.  $Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
+\item $(\varepsilon_k)_{k\ge 1}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance,
+\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$.  %$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
 \item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
-\item $Z_1$ is a Gaussian random vector of size $\dimz$ with mean $\mu_1$ and variance $\Sigma_1$ and is independent of $(\varepsilon_k)_{k\ge 0}$. 
+\item $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ and is independent of $(\varepsilon_k)_{k\ge 1}$. 
 \end{enumerate}
-At each time step, the observation $Y_k$ is given by:
+At each time step $k\ge 0$, the observation $Y_k$ is given by:
 \begin{equation}
@@
 \begin{enumerate}[-]
-\item $(\eta_k)_{k\ge 0}$ is a i.i.d. sequence of standard Gaussian variables, independent of $(\varepsilon_k)_{k\ge 0}$ and $Z_1$.
+\item $(\eta_k)_{k\ge 0}$ is a i.i.d. sequence of $\dimy$-dimensional Gaussian vectors, independent of $(\varepsilon_k)_{k\ge 1}$ and $Z_1$.
 \item $(G_j)_{1\le j \le J}$ are $\dimy \times \dimy$ positive-definite matrices and $(B_j)_{1\le j \le J}$ are $\dimy \times \dimz$ matrices.
 \end{enumerate}
-These switching models based on an unobserved Markov chain have been widely used in target tracking, underwater navigation or in finance to represent for instance commodity prices which are likely to go through regimes shifts, see for instance \cite{fong:see:2002} for crude oil data or \cite{mount:ning:cai:2006} for electricity prices.
+Conditionally Linear and Gaussian models have been widely used, see \cite{sarkka:2013} and the references therein.
 In this paper, the regime switching approach is applied to Chicago Mercantile Exchange crude oil data (see Section~\ref{sec:numerical:experiments}).
@@
 An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian.
-To exploit this specific structure using Rao Blackwellisation techniques, an alternative approach to linearization techniques to solve filtering and smoothing problems is to use Sequential Monte Carlo (SMC) methods.
-These methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}.
-By using the tractable substructure of CLGM \cite{chen:liu:2000,doucet:godsill:andrieu:2000,doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005} introduced several SMC methods combined with Kalman filters to address the filtering problem.
-These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case.
-However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables.
-Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}.
-However, only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps.
-Instead of using a structural approximation of the model such as in \cite{kim:1994}, \cite{fong:godsill:doucet:west:2002,lindsten:schon:2011} proposed forward-backward smoothers which require to sample the hidden linear states in the backward pass.
-\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
-In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
-The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass.
-\cite{briers:doucet:maskell:2010} proposed another Rao Blackwellised smoother which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
+To exploit this specific structure, it has been suggested in \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to combine Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter leading to Rao Blackwellised Monte Carlo filters.
+%These SMC methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}.
+Improvements of these early filtering techniques  have been introduced in \cite{doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005}.
+%These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case.
+%However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables.
+%Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}.
+%,lindsten:schon:2011
+%In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
+The use of Rao Blackwellisation to solve the smoothing problem has been proved to be challenging and has received satisfactory solutions only recently.
+The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it requires to sample the hidden linear states in the backward pass.
+An alternative, based on a structural approximation of the model suggested in \cite{kim:1994}, was introduced by \cite{barber:2006} to handle the backward pass.
+The first fully Rao Blackwellised SMC smoother which does replace the model XXX was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
 This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques.
+More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
+%The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass.
+
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
@@
 In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of indicator variables $a_{1:i-1}^k$ associated with importance weights $\omega_{i-1}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-1},z_{i-1}|y_{1:i-1})$.
-Plugging this approximation into \eqref{eq:smooth:r} provides a non normalized approximation $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ of $r_{i}(a_{i},z_{i},y_{1:i-1})$ integrating over all possible choices of $(a_{i-1},z_{i-1})$:
-\[
-r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{a_{i-1}}\int_{z_{i-1}}p^N_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-2}\eqsp.
+Plugging this approximation into \eqref{eq:smooth:r} provides a normalized approximation $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ of $r_{i}(a_{i},z_{i},y_{1:i-1})$ integrating over all possible choices of $(a_{i-1},z_{i-1})$:
+\[
+r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{a_{i-1}}\int_{z_{i-1}}p^N_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-1}\eqsp.
 \]
 Define
-\begin{align*}
-\Upsilon^k_{i-1}(a_{i}) &= (P^k_{i-1})^{-1} + T'_{a_{i}}\barH_{a_{i}}^{-1}T_{a_{i}}\eqsp,\\
-\upsilon^k_{i-1}(a_{i})&= \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})(P^k_{i-1})^{-1}\mu^k_{i-1}\eqsp,\\
-\Omega^k_{i-1}(a_{i}) &= \left(\barH_{a_{i}}^{-1} - \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})T'_{a_{i}}\barH_{a_{i}}^{-1}\right)^{-1}\eqsp,\\
-r_{i-1}^k(a_{i}) &= \upsilon^k_{i-1}(a_{i})+(\Omega^k_{i-1}(a_{i}))^{-1}d_{a_{i}}\eqsp,\\
-\omega_{\mathsf{f},i}^k(a_{i}) &= \omega^k_{i-1}Q(a^k_{i-1},a_{i})\left|P^k_{i-1}\right|^{-1/2}\left|\Upsilon^k_{i-1}(a_{i})\right|^{-1/2} \left|2\pi\barH_{a_{i}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{d_{a_{i}}}^2\right\}\\
-&\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-1}^k}{\mu^k_{i-1}}^2+\frac{1}{2}\normMat{P_{i-1}^k\Upsilon^k_{i-1}(a_{i})P_{i-1}^k}{\mu^k_{i-1}}^2 - (\upsilon^k_{i-1})'(a_{i})d_{a_{i}} \right\}\eqsp.
+%\begin{align*}
+%\Upsilon^k_{i-1}(a_{i}) &= (P^k_{i-1})^{-1} + T'_{a_{i}}\barH_{a_{i}}^{-1}T_{a_{i}}\eqsp,\\
+%\upsilon^k_{i-1}(a_{i})&= \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})(P^k_{i-1})^{-1}\mu^k_{i-1}\eqsp,\\
+%\Omega^k_{i-1}(a_{i}) &= \left(\barH_{a_{i}}^{-1} - \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})T'_{a_{i}}\barH_{a_{i}}^{-1}\right)^{-1}\eqsp,\\
+%r_{i-1}^k(a_{i}) &= \upsilon^k_{i-1}(a_{i})+(\Omega^k_{i-1}(a_{i}))^{-1}d_{a_{i}}\eqsp,\\
+%\omega_{\mathsf{f},i}^k(a_{i}) &= \omega^k_{i-1}Q(a^k_{i-1},a_{i})\left|P^k_{i-1}\right|^{-1/2}\left|\Upsilon^k_{i-1}(a_{i})\right|^{-1/2} \left|2\pi\barH_{a_{i}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{d_{a_{i}}}^2\right\}\\
+%&\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-1}^k}{\mu^k_{i-1}}^2+\frac{1}{2}\normMat{P_{i-1}^k\Upsilon^k_{i-1}(a_{i})P_{i-1}^k}{\mu^k_{i-1}}^2 - (\upsilon^k_{i-1})'(a_{i})d_{a_{i}} \right\}\eqsp.
+%\end{align*}
+\begin{align*}
+\Omega^k_{i-1}(a_{i}) &= T_{a_{i}}P^k_{i-1}T'_{a_{i}} + \barH_{a_{i}}\eqsp,\;\mu^k_{i-1}(a_{i}) = d_{a_{i}} + T_{a_{i}}\mu^k_{i-1}\eqsp,\;r_{i-1}^k(a_{i}) = (\Omega^k_{i-1}(a_{i}))^{-1}\mu^k_{i-1}(a_{i})\eqsp,\\
+\omega_{\mathsf{f},i}^k(a_{i}) &= \omega^k_{i-1}Q(a^k_{i-1},a_{i}) \left|2\pi \Omega^k_{i-1}(a_{i})\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{\mu^k_{i-1}(a_{i})}^2\right\}\eqsp.
 \end{align*}
@@
 These steps are then combined to form a non normalized SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ using \eqref{eq:smoothing}.
-As $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ are not normalized, the normalization of the SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
+As $t^N_{i}(a_i,z_i,y_{i+1:n})$ is not normalized, the normalization of the SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
 %These combination and normalization steps are given in Appendix~\ref{sec:approx:smooth}.

### 2016-11-07 TexArxiv/clgm.tex
Added rejuvenation for the FFBS

@@
 These distributions are in particular crucial to infer a sequence of states given the observations, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications...
-In the case where the state and observation models are linear and Gaussian, the filtering is solved explicitly using the Kalman filter \cite{kalman:1960} while the smoothing problem is solved using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier smoother \cite{}.
+In the case where the state and observation models are linear and Gaussian, the filtering is solved explicitly using the Kalman filter \cite{kalman:1960} while the smoothing problem is solved using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier smoother \cite{bryson:frazier:1963}.
 %Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters, Gaussian sum filters or Se...
@@
 \begin{enumerate}[-]
-\item $(\varepsilon_k)_{k\ge 1}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance,
+\item $(\varepsilon_k)_{k\ge 1}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance.
 \item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$.  %$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
 \item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
-\item $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ and is independent of $(\varepsilon_k)_{k\ge 1}$. 
+\item $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_k)_{k\ge 1}$. 
 \end{enumerate}
@@
 The use of Rao Blackwellisation to solve the smoothing problem has been proved to be challenging and has received satisfactory solutions only recently.
-The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it requires to sample the hidden linear states in the backward pass.
+The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it required to sample the hidden linear states in the backward pass.
 An alternative, based on a structural approximation of the model suggested in \cite{kim:1994}, was introduced by \cite{barber:2006} to handle the backward pass.
-The first fully Rao Blackwellised SMC smoother which does replace the model XXX was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
+The first fully Rao Blackwellised SMC smoother which does use this strong approximation of the backward pass was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
 This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques.
@@
 
+In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao Blackwellisation steps to produce more robust approximations.
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
-In this paper, we propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables.
-This explicit marginalization provides an additional Rao Blackwellisation step and produces more robust approximations.
+We propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables.
+Similarly, in the backward pass of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}, an indicator variable $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$.
+In this case, particle rejuvenation may be introduced by using the forward weighted samples at time $i-1$ and extending these trajectories at time $i$ with a Kalman filter for all possible values of the indicator variable.
+Then, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$ using an appropriately adapted weight.
 
 The paper is organized as follows.
-An overview of the proposed Monte Carlo algorithms to estimate the smoothing distributions of the hidden states is presented in Section~\ref{sec:RaoBlackwell}.
+An overview of the proposed Monte Carlo algorithms to estimate the smoothing distributions of the hidden states as long as the additional explicit marginalization steps are presented in Section~\ref{sec:RaoBlackwell}.
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
@@
 This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$.
-For all $m\,\times\,m$ positive-definite matrix $A$ and  all $z\in\mathbb{R}^{m}$ define
+For all $m\,\times\,m$ positive-definite matrix $A$ let $|A|$ be the determinant of $A$ and  for all $z\in\mathbb{R}^{m}$ define
 \[
@@
 Then, by \eqref{eq:FFBS:incremental},
-\[
+\begin{equation}
+\label{eq:post:obs}
 p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})\propto Q(a_i^k,a_{i+1})|\Lambda_i|^{-1/2}\exp\left\{\eta_i/2\right\}\eqsp,
-\]
+\end{equation}
 where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$ where $P_i^k = \Gamma_i^k(\Gamma_i^k)'$.
@@
 \]
-
+\paragraph{Additional Rao-Blackwellisation step:}
+an improved version of this FFBS algorithm may be defined for all $2\le i\le n-1$ by writing :
+\begin{align*}
+p(a_{1:i}|a_{i+1:n},y_{1:n}) &\propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})p(a_{1:i}|y_{1:i})\eqsp,\\
+&\propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})\int p(a_{1:i-1},z_{i-1}|y_{1:i-1})Q(a_{i-1},a_i)\\
+&\hspace{6cm}m(a_i,z_{i-1};z_i)g(a_i,z_i;y_i)\rmd z_{i-1:i}
+\end{align*}
+and replacing $p(a_{1:i-1},z_{i-1}|y_{1:i-1})$ in the integral by 
+\[
+p^N_{}(a_{1:i-1},z_{i-1}|y_{1:i-1}) = \sum_{k=1}^N\delta_{a^k_{1:i-1}}(a_{1:i-1})\,\omega^k_{i-1}\,p_{}(z_{i-1}|a^k_{1:i-1},y_{1:i-1})\eqsp.
+\]
+Kalman filtering techniques allow to compute the integral for each trajectory $\{a^k_{1:i-1}\}_{1\le k\le N}$ and each $a_i\in\{1,\ldots,J\}$.
+On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the indicator variables of the filtering pass $\{a_i^k\}_{1\le k \le N}$.
+This means that a Kalman filter must be run for all trajectory $a^k_{1:i-1}$ which may be extended by $a_i\in\{1,\ldots,J\}$ to compute $\Lambda_i$ and $\eta_i$.
+Then, by integrating over all possible paths $a_{1:i-1}$, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$.
 \subsection{Rao Blackwellised Two-filter Smoother (RBTF)}
@@
 %Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods.
-For $2\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
+\paragraph{Additional Rao-Blackwellisation step:}
+for $2\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
 \begin{equation}

### 2016-11-07 TexArxiv/clgm.tex
Changes in Forward filter

@@
 Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models.
-The most widely used Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{kim:1994,lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the Forward Backward decomposition of the marginal smoothing distribution.
-In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps without any approximation in both forward and backward passes.
-In this paper, these algorithms are compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance.
+The algorithm proposed by  \cite{briers:doucet:maskell:2010} extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps.
+Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution% In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
+ and use explicit marginalization of the linear states in both forward and backward passes.
+In this paper, we propose an overview of these algorithms and introduce  compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance.
 The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield.
@@
 An alternative, based on a structural approximation of the model suggested in \cite{kim:1994}, was introduced by \cite{barber:2006} to handle the backward pass.
-The first fully Rao Blackwellised SMC smoother which does use this strong approximation of the backward pass was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
-This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC and Kalman filtering techniques.
+This approximation introduces a systemic bias which might be difficult to control.
+The first fully Rao Blackwellised SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
+This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC for the indicator variables and Kalman filtering techniques for the hidden linear states.
 More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
@@
 
-In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao Blackwellisation steps to produce more robust approximations.
+In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao Blackwellisation steps, extending the ideas of \cite{fa}.
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
@@
 The paper is organized as follows.
-An overview of the proposed Monte Carlo algorithms to estimate the smoothing distributions of the hidden states as long as the additional explicit marginalization steps are presented in Section~\ref{sec:RaoBlackwell}.
+An extensive survey of the proposed fully Rao Blackwellised Monte Carlo smoothers  as long as the additional explicit marginalization steps are presented in Section~\ref{sec:RaoBlackwell}.
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
-Then, a new model is proposed to explain regime switchings for commodity markets.
+In Section~\ref{}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data.
+A detailed derivation of the algorithms is provided in the Appendix.
+
+%Then, a new model is proposed to explain regime switchings for commodity markets.
 The parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.
-The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data.
-Intermediate computations and explicit approximations provided by the algorithm are given in the Appendix.
+%The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data.
 
@@
 This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$.
-For all $m\,\times\,m$ positive-definite matrix $A$ let $|A|$ be the determinant of $A$ and  for all $z\in\mathbb{R}^{m}$ define
+For all $m\,\times\,m$ matrix  let $|A|$ be the determinant of $A$ if $A$ is positive-definite matrix, for all $z\in\mathbb{R}^{m}$ define
 \[
@@
 \]
-In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter as follows.
+In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by using the Kalman filter's update.
 \paragraph{Initialization}
@@
 \paragraph{Iterations}
-For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
-Labels trajectories at time $i-1$ are defined by:
-\[
-a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
+%For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
+The ancestral path of indicator variables at time $i-1$ is defined by:
+%\begin{equation}
+%\label{eq:ancestral:traj}
+%a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
+%\end{equation}
+%where the ancestral variables are defined as follows:
+%\[
+%\mathsf{A}^k_{i} = k \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
+%\]
+%Similarly, means and variances trajectories are defined by: 
+%\[
+%\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
+%\]
+%The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
+For all sampled trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$ and all $1\le j \le J$, define:
+\[
+\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
+\]
+The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ so that:% which means that $\gamma_i^{j,k}$ may be written:
+\[
+\gamma_i^{j,k} \propto  Q(a_{i-1}^k,j)|B_jP^{j,k}_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^{j,k}_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^{j,k}_{i|i-1}}^2\right\} \eqsp,
 \]
 where 
-\[
-\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
-\]
-Similarly, means and variances trajectories are defined by: 
-\[
-\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
-\]
-The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
-\[
-\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
-\]
-The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^{j,k}$ may be written:
-\[
-\gamma_i^{j,k} =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp,
-\]
-where 
-\begin{align*}
-\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
-P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp.
-\end{align*}
-$\mu_i^k$ and $P_i^k$ are updated according to:
-\begin{align*}
-K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
-\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
-P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
-\end{align*}
-The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
-
-\subsection{Path-space smoother of \cite{kitagawa:1996}}
-As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step.
+\begin{align*}
+\mu^{j,k}_{i|i-1} &= d_{j} + T_{j}\mu^k_{i-1}\eqsp,\\
+P^{j,k}_{i|i-1} &= T_{j}P^k_{i-1}T'_{j} + \barH_{j}\eqsp.
+\end{align*}
+
+%$\mu_i^k$ and $P_i^k$ are updated according to:
+%\begin{align*}
+%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
+%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
+%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
+%\end{align*}
+%The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
+
+%\subsection{Path-space smoother of \cite{kitagawa:1996}}
+%As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step.
 In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
-\[
-p^{\mathsf{Kit}}_N(a_{1:n}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
-\]
-In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
-\[
-p^{\mathsf{Kit}}_N(a_{i}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
-\]
-This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps.
+%\[
+%p^{\mathsf{Kit}}_N(a_{1:n}|y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
+%\]
+%In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
+%\[
+%p^{\mathsf{Kit}}_N(a_{i}|y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
+%\]
+%This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps.
 See Section~\ref{sec:numerical:experiments} for a numerical illustration.
 
-\subsection{Algorithm based on the structural approximation of \cite{kim:1994}}
-Another method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{kim:1994} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$.
+\subsection{Algorithm based on the structural approximation of \cite{kim:1994}, \cite{barber:2006}}
+The method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{barber:2006} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$.
 Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available.
@@
 \]
-Then, the approximation of \cite{kim:1994} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
+Then, the approximation of \cite{kim:1994,barber:2006} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
 \[
@@
 \subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
-\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}.
+\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} without using the approximation of \cite{kim:1994}.
 The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
@@
 \]
-The second factor in the last equation may be approximated using the trajectories $(a^k_{1:i})_{1\le k \le N}$ and importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter.
+The second factor in the last equation may be approximated using the ancestral  trajectories $(a^k_{1:i})_{1\le k \le N}$ defined in \eqref{eq:ancestral:traj} and importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter.
 Therefore, $p(a_{1:i}|a_{i+1:n},y_{1:n})$ may be approximated by:
@@
 \item Set $\tilde{a}_n= a_n^k$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$.
-\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probabilities proportional to $(\tilde{\omega}_i^k)_{1\le k \le N}$.
+\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probabilities proportional to $(\tilde{\omega}_{i|n}^k)_{1\le k \le N}$.
 \end{enumerate}
@@
 \end{equation}
-which is similar to the two-filter decomposition of the smoothing distribution. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
-\[
-p(y_{i+1:n},a_{i+1:n}|z_i,a_{i}) \propto Q(a_i,a_{i+1}) \exp\left\{-(z_i'\Omega_iz_i-2\lambda_i'z_i)/2\right\}\eqsp,
+which is similar to the two-filter decomposition of the smoothing distribution, see Section~\ref{sec:two-filer}. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
+\[
+p(y_{i+1:n},a_{i+1:n}|z_i,a_{i}) \propto Q(a_i,a_{i+1}) \exp\left\{-(z_i'\Omega_i(a_{i+1:n})z_i-2\lambda_i'(a_{i+1:n})z_i)/2\right\}\eqsp,
 \]
@@
 \[
-p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\hat{\Omega}_iz_i-2\hat{\lambda}_i'z_i)/2\right\}\eqsp,
+p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\hat{\Omega}_i(a_{i:n})z_i-2\hat{\lambda}'_i(a_{i:n})z_i)/2\right\}\eqsp,
 \]
@@
 \begin{align*}
-\hat{\Omega}_n &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
-\hat{\lambda}_n &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
+\hat{\Omega}_n(a_n) &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
+\hat{\lambda}_n(a_n) &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
 \end{align*}
@@
 \begin{align*}
-\Omega_i &= T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\hat{\Omega}_{i+1}T_{a_{i+1}}\eqsp,\\
-\lambda_i &=T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
+\Omega_i(a_{i+1:n}) &= T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}(a_{i+1:n})H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\hat{\Omega}_{i+1}(a_{i+1:n})T_{a_{i+1}}\eqsp,\\
+\lambda_i(a_{i+1:n}) &=T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}(a_{i+1:n})H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
 \end{align*}
@@
 \begin{align*}
-\hat{\Omega}_i &= \Omega_i + B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
-\hat{\lambda}_i & = \lambda_i + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
+\hat{\Omega}_i(a_{i:n}) &= \Omega_i(a_{i+1:n})+ B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
+\hat{\lambda}_i(a_{i:n}) & = \lambda_i(a_{i+1:n}) + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
 \end{align*}
@@
 \end{equation}
-where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$ where $P_i^k = \Gamma_i^k(\Gamma_i^k)'$.
+where the proportionality is with respect to $a^k_{1:i}$ and 
+\begin{align*}
+\Lambda^k_i(a_{i+1:n})&= (\Gamma_i^k)'\Omega_i(a_{i+1:n})\Gamma_i^k + I\eqsp,\\
+\eta^k_i(a_{i+1:n}) &= \|\mu_i^k\|^2_{\Omega^{-1}_i(a_{i+1:n})} - 2\lambda'_i(a_{i+1:n})\mu_i^k-\|(\Gamma_i^k)'(\lambda_i(a_{i+1:n})-\Omega_i(a_{i+1:n})\mu_i^k)\|^2_{\Lambda_i(a_{i+1:n})}\eqsp,
+\end{align*}
+where $P_i^k = \Gamma_i^k(\Gamma_i^k)'$. Therefore,
+\[
+\tilde{\omega}_{i|n} \propto \omega_i^kQ(a_i^k,a_{i+1})|\Lambda^k_i(a_{i+1:n})|^{-1/2}\exp\left\{\eta^k_i(a_{i+1:n})/2\right\}\eqsp.
+\]
 If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{lindsten:bunch:sarkka:schon:godsill:2015} of the joint smoothing distribution of the indicator variables is:
@@
 \]
-\paragraph{Additional Rao-Blackwellisation step:}
-an improved version of this FFBS algorithm may be defined for all $2\le i\le n-1$ by writing :
+\textbf{Additional particle rejuvenation step}\\
+An improved version of this FFBS algorithm may be defined for all $2\le i\le n-1$ by writing :
 \begin{align*}
@@
 &\propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})\int p(a_{1:i-1},z_{i-1}|y_{1:i-1})Q(a_{i-1},a_i)\\
-&\hspace{6cm}m(a_i,z_{i-1};z_i)g(a_i,z_i;y_i)\rmd z_{i-1:i}
-\end{align*}
-and replacing $p(a_{1:i-1},z_{i-1}|y_{1:i-1})$ in the integral by 
-\[
-p^N_{}(a_{1:i-1},z_{i-1}|y_{1:i-1}) = \sum_{k=1}^N\delta_{a^k_{1:i-1}}(a_{1:i-1})\,\omega^k_{i-1}\,p_{}(z_{i-1}|a^k_{1:i-1},y_{1:i-1})\eqsp.
-\]
-Kalman filtering techniques allow to compute the integral for each trajectory $\{a^k_{1:i-1}\}_{1\le k\le N}$ and each $a_i\in\{1,\ldots,J\}$.
+&\hspace{6cm}m(a_i,z_{i-1};z_i)g(a_i,z_i;y_i)\rmd z_{i-1:i}\eqsp.
+\end{align*}
+Replacing $p(a_{1:i-1},z_{i-1}|y_{1:i-1})$ in the integral by the particle approximation obtained during the forward pass and using Kalman filtering techniques for each trajectory $\{a^k_{1:i-1}\}_{1\le k\le N}$ and each $a_i\in\{1,\ldots,J\}$ yields:
+\[
+\int p^N(a_{1:i-1},z_{i-1}|y_{1:i-1})Q(a_{i-1},a_i)m(a_i,z_{i-1};z_i)g(a_i,z_i;y_i)\rmd z_{i-1:i} \propto \sum_{k=1}^N \omega_{i|i-1}^k(a_i)\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp,
+\]
+where
+\[
+\omega_{i|i-1}^k(a_i) = \omega_{i-1}^k Q(a_{i-1}^k,a_i)|\Sigma^k_{i|i-1}(a_i)|^{-1/2}\mathrm{e}^{-\left(y_i - y^k_{i|i-1}(a_i)\right)'\left(\Sigma^k_{i|i-1}(a_i)\right)^{-1}\left(y_i - y^k_{i|i-1}(a_i)\right)/2}\eqsp,
+\]
+\[
+y^k_{i|i-1}(a_i) = c_{a_i} + B_{a_i}\mu^k_{i-1}\;\mbox{and}\; \Sigma^k_{i|i-1}(a_i) = B_{a_i}P^k_{i-1}B'_{a_i} + \barG_{a_i}\eqsp.
+\]
+%\[
+%p^N_{}(a_{1:i-1},z_{i-1}|y_{1:i-1}) = \sum_{k=1}^N\delta_{a^k_{1:i-1}}(a_{1:i-1})\,\omega^k_{i-1}\,p_{}(z_{i-1}|a^k_{1:i-1},y_{1:i-1})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp.
+%\]
 On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the indicator variables of the filtering pass $\{a_i^k\}_{1\le k \le N}$.
-This means that a Kalman filter must be run for all trajectory $a^k_{1:i-1}$ which may be extended by $a_i\in\{1,\ldots,J\}$ to compute $\Lambda_i$ and $\eta_i$.
-Then, by integrating over all possible paths $a_{1:i-1}$, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$.
+This means that a Kalman filter must be used for each trajectory $a^k_{1:i-1}$ which may be extended by $a_i\in\{1,\ldots,J\}$.
+Denote by $\mu_{i|i-1}^k(a_i)$ and $P_{i|i-1}^k(a_i)$ the mean and covariance matrix of the law of $z_i$ given $(a^k_{1:i-1},a_i)$.
+Then,
+\[
+p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i}) = Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{\eta^k_{i|i-1}(a_{i:n})/2\right\}\eqsp,
+\]
+where the proportionality is with respect to $(a^k_{1:i-1},a_i)$ and 
+\begin{align*}
+\Lambda^k_{i|i-1}(a_{i:n})&= (\Gamma_{i|i-1}^k(a_i))'\Omega_i(a_{i+1:n})\Gamma_{i|i-1}^k(a_i) + I\eqsp,\\
+\eta^k_{i|i-1}(a_{i:n}) &= \|\mu_{i|i-1}^k(a_i)\|^2_{\Omega^{-1}_i(a_{i+1:n})} - 2\lambda'_i(a_{i+1:n})\mu_{i|i-1}^k(a_i)\\
+&\hspace{3cm}-\|(\Gamma_{i|i-1}^k(a_i))'(\lambda_i(a_{i+1:n})-\Omega_i(a_{i+1:n})\mu_{i|i-1}^k(a_i))\|^2_{\Lambda_i(a_{i+1:n})}\eqsp,
+\end{align*}
+where $P_{i|i-1}^k(a_i) = \Gamma_{i|i-1}^k(a_i)(\Gamma_{i|i-1}^k(a_i))'$. $p(a_{1:i}|a_{i+1:n},y_{1:n})$ is then approximated by :
+\[
+p^N(a_{1:i}|a_{i+1:n},y_{1:n})\propto\sum_{k=1}^N \omega_{i|i-1}^k(a_i)Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{\eta^k_{i|i-1}(a_{i:n})/2\right\}\delta_{a^k_{1:i-1}}(a_{1:i-1})
+\]
+By integrating over all possible paths $a_{1:i-1}$, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$.
 \subsection{Rao Blackwellised Two-filter Smoother (RBTF)}
-Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}.
+\label{sec:two-filer}
+%Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}.
 Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations).

### 2016-11-08 TexArxiv/clgm.tex
Changes in Forward filter

@@
 \label{sec:forward}
-The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}.
+The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using a standard Rao Blackwellised  algorithm.
 The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
@@
 In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by using the Kalman filter's update.
-\paragraph{Initialization}
-At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$.
+
+\vspace{.2cm}
+
+\noindent\textbf{Initialization}\\
+At time $i=1$, write, for all $1\le j \le J$, 
+\[
+\mu^j_{1|0} = c_{j}+B_{j}\mu_1\;\;\mbox{and}\;\;P_{1|0}^j =B_{j}\Sigma_1B'_{j} + \barG_{j}\eqsp.
+\]
+$(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to 
+\[
+\pi_j p(a_1=j|y_1) \propto \pi_j |P_{1|0}^j|^{-1/2}\exp\left\{-(y_1-\mu^j_{1|0})'(P_{1|0}^j)^{-1}(y_1-\mu^j_{1|0})/2\right\}\eqsp. 
+\]
 Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
@@
 \end{align*}
-Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$.
-This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$.
-\paragraph{Iterations}
+Each particle particle $a_1^k$ is associated with the importance weight $\omega^k_1 = 1/N$.
+
+\vspace{.2cm}
+
+%This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$.
+\noindent\textbf{Iterations}\\
 %For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
@@
 %The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
-For all sampled trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$ and all $1\le j \le J$, define:
+Several procedures may be used to extend the trajectories $(a^k_{1:i-1})_{1\le k \le N}$ at time $i$.
+For all sampled trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$ and all $1\le j \le J$,  \cite{chen:liu:2000} used the incremental weights:
 \[
@@
 \]
-where 
-\begin{align*}
-\mu^{j,k}_{i|i-1} &= d_{j} + T_{j}\mu^k_{i-1}\eqsp,\\
-P^{j,k}_{i|i-1} &= T_{j}P^k_{i-1}T'_{j} + \barH_{j}\eqsp.
-\end{align*}
+where  $\mu^{j,k}_{i|i-1} = d_{j} + T_{j}\mu^k_{i-1}$ and $P^{j,k}_{i|i-1} = T_{j}P^k_{i-1}T'_{j} + \barH_{j}$.
+In \cite{chen:liu:2000}, for all $1\le k \le N$, an ancestral path is chosen with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
+Then, the new indicator variable $a_i^k$ is sampled in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$.
+A drawback of this method is that only ancestral paths that have been selected using the importance weights $(\omega^k_{i-1})_{1\le k \le N}$ are extended at time $i$.
+Following \cite{barembruch:garivier:moulines:2008}, this may be improved by considering all the offsprings of all ancestral trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$.
+Each ancestral path has $J$ offsprings at time $i$, it is thus necessary to choose a given number of trajectories at time $i$ (for instance $N$) among the $NJ$ possible paths.
 

### 2016-11-08 TexArxiv/clgm.tex
MAJ

@@
 
-\title{Rao-Blackwellised Two-filter algorithm for Conditionally Linear and Gaussian models}
+\title{Rao-Blackwellized  smoother for Conditionally Linear and Gaussian models}
 \date{}
@@
 These distributions are in particular crucial to infer a sequence of states given the observations, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications...
-In the case where the state and observation models are linear and Gaussian, the filtering is solved explicitly using the Kalman filter \cite{kalman:1960} while the smoothing problem is solved using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier smoother \cite{bryson:frazier:1963}.
+In the case where the state and observation models are linear and Gaussian, the filtering can be solved explicitly using the Kalman filter \cite{kalman:1960}.
+Exact solutions of the fixed-horizon smoothing problem can be obtained using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier "two-filter" smoother \cite{bryson:frazier:1963}.
 %Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters, Gaussian sum filters or Se...
@@
 \end{enumerate}
-Conditionally Linear and Gaussian models have been widely used, see \cite{sarkka:2013} and the references therein.
-In this paper, the regime switching approach is applied to Chicago Mercantile Exchange crude oil data (see Section~\ref{sec:numerical:experiments}).
-
-An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian.
+Conditionally Linear and Gaussian models play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
+An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditionally to the sequence of labels, the filtering and the smoothing distribution of the "continuous" states can be solved explicitly.
 To exploit this specific structure, it has been suggested in \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to combine Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter leading to Rao Blackwellised Monte Carlo filters.

### 2016-11-08 TexArxiv/clgm.tex
MAJ

@@
 \end{enumerate}
-Conditionally Linear and Gaussian models play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
-An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditionally to the sequence of labels, the filtering and the smoothing distribution of the "continuous" states can be solved explicitly.
-To exploit this specific structure, it has been suggested in \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to combine Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter leading to Rao Blackwellised Monte Carlo filters.
+Conditionally Linear and Gaussian Models (CLGM) play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
+An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditionally to the sequence of labels, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be solved explicitly.
+To exploit this specific structure, it has been suggested in the pioneering paper of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter to compute the conditional distribution of the 'continuous' states sequence $(Z_j)_{j=1}^n$.
+This is a specific instance of Rao-Blackwellised Monte Carlo filters, often referred to as the \emph{Mixture Kalman Filter}.
 %These SMC methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}.
@@
 %In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
-The use of Rao Blackwellisation to solve the smoothing problem has been proved to be challenging and has received satisfactory solutions only recently.
+
+The use of Rao Blackwellisation to solve the smoothing problem has been proved to be more challenging and has received satisfactory solutions only recently.
+
 The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it required to sample the hidden linear states in the backward pass.
-An alternative, based on a structural approximation of the model suggested in \cite{kim:1994}, was introduced by \cite{barber:2006} to handle the backward pass.
-This approximation introduces a systemic bias which might be difficult to control.
+
+An alternative approach, based on the so-called  'structural approximation' of the model suggested in an early paper by \cite{kim:1994}, was proposed by \cite{barber:2006} to avoid to sample a 'continuous' state in the backward pass.
+This approximation is rather ad-hoc and the resulting smoother is not consistent when the number of particles goes to infinity.
+The inaccuracy  introduced by the approximation might be difficult to control.
+
 The first fully Rao Blackwellised SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
 This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC for the indicator variables and Kalman filtering techniques for the hidden linear states.
+
 More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
@@
 
-In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao Blackwellisation steps, extending the ideas of \cite{fa}.
+In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao-Blackwellisation steps allowing the sample "new" particles in the backward pass.
+This approach may be seen as an extension of the ideas of \cite{fa} for Rao-Blackwellized smoother.
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
@@
 The paper is organized as follows.
-An extensive survey of the proposed fully Rao Blackwellised Monte Carlo smoothers  as long as the additional explicit marginalization steps are presented in Section~\ref{sec:RaoBlackwell}.
+The proposed modification of the fully Rao Blackwellised Monte Carlo smoothers are presented in Section~\ref{sec:RaoBlackwell}.
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
@@
 A detailed derivation of the algorithms is provided in the Appendix.
-
-%Then, a new model is proposed to explain regime switchings for commodity markets.
-The parameters of the proposed model are estimated using a Monte Carlo EM algorithm where the required conditional expectations are replaced by their estimates based on the proposed SMC method.
-%The performance of the extended Rao-Blackwellised two-filter  algorithm is assessed using simulated data and CME crude oil (WTI) data.
 

### 2016-11-09 TexArxiv/clgm.tex
MAJ
New forward filter

@@
 In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao-Blackwellisation steps allowing the sample "new" particles in the backward pass.
-This approach may be seen as an extension of the ideas of \cite{fa} for Rao-Blackwellized smoother.
+This approach may be seen as an extension of the ideas of \cite{fearnhead:clifford:2003} for Rao-Blackwellized smoother.
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
@@
 The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
-\[
+\begin{equation}
+\label{eq:forward:SMC}
 p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
-\]
+\end{equation}
 In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by using the Kalman filter's update.
@@
 Each ancestral path has $J$ offsprings at time $i$, it is thus necessary to choose a given number of trajectories at time $i$ (for instance $N$) among the $NJ$ possible paths.
+Following \cite{fearnhead:clifford:2003,barembruch:garivier:moulines:2008}, this may be improved by considering all the offsprings of all ancestral trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$.
+Each ancestral path has $J$ offsprings at time $i$, it is thus necessary to choose a given number of trajectories at time $i$ (for instance $N$) among the $NJ$ possible paths.
+To obtain the weight associated with each offspring write the following approximation of  $p(a_{1:i}|y_{1:i})$ based on the weighted samples at time $i-1$:
+\begin{align*}
+p^N(a_{1:i}|y_{1:i})&\propto \sum_{k=1}^N\omega^k_{i-1}Q(a^k_{i-1},a_i)p(y_i|a^k_{1:i-1},a_i,y_{1:i-1})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp,\\
+&\propto \sum_{k=1}^N\sum_{j=1}^J\omega^k_{i-1}\gamma_i^{j,k}\delta_{(a^k_{1:i-1},j)}(a_{1:i})\eqsp.
+\end{align*}
+Therefore, each ancestral trajectory of the form  $(a^k_{1:i-1},j)$, $1\le k \le N$, $1\le j\le J$, is associated with the normalized weight $\tilde{\omega}^{j,k}_{i} \propto \omega^k_{i-1}\gamma_i^{j,k}$.
+Several random selection schemes have been proposed to discard some of the possible offsprings to maintain an average number of $N$ particles at each time step.
+Following \cite{barembruch:garivier:moulines:2008}, we might choose between the Kullback-Leibler Optimal Selection (KL-OS) or the Chi-Squared Optimal Selection (CS-OS) to associate a new weight to each of the $NJ$ trajectories.
+If the new weight is 0, then the corresponding particle can be removed.\\
+\textbf{KL-OS:} $\lambda$ is chosen as the solution of :
+\[
+\sum_{k=1}^N\sum_{j=1}^J\mathrm{min}\left(\tilde{\omega}^{j,k}_{i}/\lambda,1\right) = N\eqsp.
+\]
+For all $1\le j \le J$ and $1\le k \le N$, if $\tilde{\omega}^{j,k}_{i}\ge \lambda$ then the new weight $\tilde{\Omega}^{j,k}_{i}$ is $\tilde{\Omega}^{j,k}_{i}=\tilde{\omega}^{j,k}_{i}$ and if $\tilde{\omega}^{j,k}_{i}< \lambda$:
+\[
+\tilde{\Omega}^{j,k}_{i}=
+\left\{
+ \begin{array}{rl}
+  &\lambda \;\mbox{with probability } \tilde{\omega}^{j,k}_{i}/\lambda\eqsp,\\
+&0 \;\mbox{with probability } 1-\tilde{\omega}^{j,k}_{i}/\lambda\eqsp.
+\end{array}
+\right.
+\]
+\textbf{CS-OS:} $\lambda$ is chosen as the solution of :
+\[
+\sum_{k=1}^N\sum_{j=1}^J\mathrm{min}\left(\sqrt{\tilde{\omega}^{j,k}_{i}/\lambda},1\right) = N\eqsp.
+\]
+For all $1\le j \le J$ and $1\le k \le N$, if $\tilde{\omega}^{j,k}_{i}\ge \lambda$ then the new weight $\tilde{\Omega}^{j,k}_{i}$ is $\tilde{\Omega}^{j,k}_{i}=\tilde{\omega}^{j,k}_{i}$ and if $\tilde{\omega}^{j,k}_{i}< \lambda$:
+\[
+\tilde{\Omega}^{j,k}_{i}=
+\left\{
+ \begin{array}{rl}
+  &\sqrt{\tilde{\omega}^{j,k}_{i}\lambda} \;\mbox{with probability } \sqrt{\tilde{\omega}^{j,k}_{i}/\lambda}\eqsp,\\
+&0 \;\mbox{with probability } 1-\sqrt{\tilde{\omega}^{j,k}_{i}/\lambda}\eqsp.
+\end{array}
+\right.
+\]
+Then, in both cases, all particles such that $\tilde{\Omega}^{j,k}_{i} = 0$ are discarded and for all the other trajectories defined as an ancestral path $(a^k_{1:i-1})$ extended by $a^k_i = j$, the new corresponding weight $\omega$ in \eqref{eq:forward:SMC} is given by the normalized weight $\tilde{\Omega}^{j,k}_{i}$.
 
@@
 \centering
-\includegraphics[scale=.5]{N10_1_500.eps}
-\includegraphics[scale=.5]{N10_501_1000.eps}
-\includegraphics[scale=.5]{N10_1001_1500.eps}
+\includegraphics[scale=.5]{N10_1_500}
+\includegraphics[scale=.5]{N10_501_1000}
+\includegraphics[scale=.5]{N10_1001_1500}
 \caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true labels (squared black).}
@@
 \centering
-\includegraphics[scale=.5]{N15_1_500.eps}
-\includegraphics[scale=.5]{N15_501_1000.eps}
-\includegraphics[scale=.5]{N15_1001_1500.eps}
+\includegraphics[scale=.5]{N15_1_500}
+\includegraphics[scale=.5]{N15_501_1000}
+\includegraphics[scale=.5]{N15_1001_1500}
 \caption{Experiment 2 - estimated probabilites by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
@@
 \centering
-\includegraphics[scale=.6]{LogPrice_TermStructure.eps}
+\includegraphics[scale=.6]{LogPrice_TermStructure}
 \caption{Log-price (red line) and slope of future curves (blue line)}
@@
 	\centering
-	\includegraphics[scale=.6]{Posterior_Data.eps}
+	\includegraphics[scale=.6]{Posterior_Data}
 	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
@@
 	\centering
-	\includegraphics[scale=.75]{Posterior_Data_LogPrice.eps}
+	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
 	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}

### 2016-11-09 TexArxiv/clgm.tex
MAJ

--- 2016-11-08 TexArxiv/clgm.tex
@@
 The method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{barber:2006} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$.
-Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available.
+Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$ approximating the filtering distributions are available.
 Then, note that, for all $1\le i\le n$,
@@
 \subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
-\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} without using the approximation of \cite{kim:1994}.
+\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} without using the ad-hoc approximation of \cite{kim:1994}.
 The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
@@
 \]
-The second factor in the last equation may be approximated using the ancestral  trajectories $(a^k_{1:i})_{1\le k \le N}$ defined in \eqref{eq:ancestral:traj} and importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter.
+The second factor in the last equation may be approximated using the ancestral  trajectories $(a^k_{1:i})_{1\le k \le N}$ and the associated importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter.
 Therefore, $p(a_{1:i}|a_{i+1:n},y_{1:n})$ may be approximated by:
 \[
-p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a^k_{1:i})\eqsp,
-\]
-with $\tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$.
+p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a_{1:i})\eqsp, \quad \text{with} \quad \tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) \eqsp.
+\]
 Then, a trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$ may be sampled following these steps:
@@
 This algorithm requires to compute the quantity  $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$.
-As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using backward Kalman filtering techniques.
+As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using  Kalman filtering techniques.
 However, this has to be done for each trajectory $(a^k_{1:i})_{1\le k \le N}$, which leads to an algorithm with a prohibitive computational complexity.
-\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a procedure computationally less intensive by noting that:
+\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a procedure computationally less intensive by conditioning with respect to $z_i$ and then marginalizing with respect to this variable:
 \begin{equation}
@@
 \end{equation}
-which is similar to the two-filter decomposition of the smoothing distribution, see Section~\ref{sec:two-filer}. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
+This is similar to the two-filter decomposition of the smoothing distribution, see Section~\ref{sec:two-filer}. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
 \[

### 2016-11-12 TexArxiv/clgm.tex
Numerical results updated

@@
 \[
-p^{\mathsf{Lbscg}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
+p^{\mathsf{Lbscg}}_{\tilde{N}}(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
 \]
@@
 \[
-y^k_{i|i-1}(a_i) = c_{a_i} + B_{a_i}\mu^k_{i-1}\;\mbox{and}\; \Sigma^k_{i|i-1}(a_i) = B_{a_i}P^k_{i-1}B'_{a_i} + \barG_{a_i}\eqsp.
+y^k_{i|i-1}(a_i) = c_{a_i} + B_{a_i}(d_{a_i}+T_{a_i}\mu^k_{i-1})\;\mbox{and}\; \Sigma^k_{i|i-1}(a_i) = B_{a_i}(T_{a_i}P^k_{i-1}T'_{a_i}+\barH_{a_i})B'_{a_i} + \barG_{a_i}\eqsp.
 \]
@@
 %Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods.
-\paragraph{Additional Rao-Blackwellisation step:}
+\textbf{Additional particle rejuvenation step}\\
 for $2\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
@@
 \label{sec:numerical:experiments}
-This section highlights the performance of the Rao Blackwellised Two-Filter approximation (RBTF) of the marginal smoothing distributions in comparison to the other Rao Blackwellised smoothers introduced in Section~\ref{sec:RaoBlackwell}.
-
 \subsection{Simulated data}
-We consider the two-factor model \eqref{eq:model:state} and \eqref{eq:model:obs} with $J=2$.
-The performance of the algorithms are analyzed in two different settings with $n=1500$ observations.
-In both settings the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}.
-The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.
-It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$).
-The initial state variance is high to emphasize the fact that there is no prior precise knowledge on the state value.
-
+This section highlights the performance of the additional Rao Blackwellisation steps for the Two-Filter and the FFBS approximations of the marginal smoothing distributions in the case where $J=2$.
+In each setting the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}.
+First, the FFBS algorithm is applied to a simple one-dimensional model with:
+\[
+\pi(1)=\pi(2) = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
+\]
+\[
+Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T_1 = T_2 = 1\;\;\barH_1 = \barH_2= 0.05\eqsp,
+\]
+\[
+B_1 = B_2 =1\;\;\barG_1 = 0.3\;\;\barG_2 = 0.1\eqsp.
+\]
+The two FFBS algorithms are used with $N = \tilde{N} = 100$.
+Figure~\ref{fig:ffbs} displays the RMSE values as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm.
+Then, the performance of the two-filter based algorithms are compared in two different settings with $n=1500$ observations.
 \paragraph{Experiment 1:}
@@
 %\end{figure}
-
+The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.
+It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$).
 Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms using:

### 2016-11-13 TexArxiv/clgm.tex
Abstract and Introduction

@@
 The algorithm proposed by  \cite{briers:doucet:maskell:2010} extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps.
-Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution% In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
+Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution % In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
  and use explicit marginalization of the linear states in both forward and backward passes.
-In this paper, we propose an overview of these algorithms and introduce  compared to an extension of the Two-filter based algorithm of \cite{briers:doucet:maskell:2010} with additional explicit Rao Blackwellised steps to improve its performance.
+In this paper, we propose an overview of these algorithms and introduce an additional Rao Blackwellisation step for all procedures which allows to sample new particles in the backward pass to improve the performance of the algorithm.
 The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield.
@@
 \end{enumerate}
-Conditionally Linear and Gaussian Models (CLGM) play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
-An important feature of these models is that, conditionally on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditionally to the sequence of labels, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be solved explicitly.
+CLGM play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
+A crucial feature of these models is that, conditional on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditional on the sequence of labels, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be solved explicitly.
 To exploit this specific structure, it has been suggested in the pioneering paper of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter to compute the conditional distribution of the 'continuous' states sequence $(Z_j)_{j=1}^n$.

### 2016-11-13 TexArxiv/clgm.tex
Numerical experiment

@@
 \subsection{Simulated data}
-This section highlights the performance of the additional Rao Blackwellisation steps for the Two-Filter and the FFBS approximations of the marginal smoothing distributions in the case where $J=2$.
+This section highlights the performance of the additional Rao Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where $J=2$.
 In each setting the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}.
@@
 \]
-The two FFBS algorithms are used with $N = \tilde{N} = 100$.
-Figure~\ref{fig:ffbs} displays the RMSE values as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm.
+The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with $2000$ particles.
+Figure~\ref{fig:ffbs:posterior} illustrates that the algorithm with rejuvenation slightly improves the original algorithm in this simple case.
+\begin{figure}
+\centering
+\includegraphics[scale=.3]{FFBSplot.eps}
+\caption{Estimated posterior probabilities by the two FFBS algorithms.}
+\label{fig:ffbs:posterior}
+\end{figure}
+In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the labels as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).
+
 Then, the performance of the two-filter based algorithms are compared in two different settings with $n=1500$ observations.

### 2016-11-13 TexArxiv/clgm.tex
New experiments

@@
 \]
-The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with $2000$ particles.
+The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles.
 Figure~\ref{fig:ffbs:posterior} illustrates that the algorithm with rejuvenation slightly improves the original algorithm in this simple case.
@@
 \end{figure}
-In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the labels as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).
+Figure~\ref{fig:ffbs:err} displays the mean estimation error over $100$ independent Monte Carlo runs.
+The estimation error is defined as the difference between the the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
+\begin{figure}
+\centering
+\includegraphics[scale=.3]{err_est_ffbs.eps}
+\caption{Posterior probabilities estimation error fo the two FFBS algorithms.}
+\label{fig:ffbs:err}
+\end{figure}
+%In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the labels as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).
 

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
-In Section~\ref{}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data.
+In Section~\ref{sec:crudeoil}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data.
 A detailed derivation of the algorithms is provided in the Appendix.
@@
 
-\section{Numerical experiments}
+\section{Simulated data}
 \label{sec:numerical:experiments}
-\subsection{Simulated data}
-This section highlights the performance of the additional Rao Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where $J=2$.
-In each setting the transition matrix $Q$ is such that the states are prone to stay in the same regime  for many time steps as expected for the WTI crude oil data,  see Section~\ref{sec:exp}.
+This section highlights the performance improvement brought by  the additional Rao-Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$.
+In each setting the transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data;  see Section~\ref{sec:exp}.
 First, the FFBS algorithm is applied to a simple one-dimensional model with:
@@
 \centering
-\includegraphics[scale=.3]{FFBSplot.eps}
+\includegraphics[scale=.3]{FFBSplot}
 \caption{Estimated posterior probabilities by the two FFBS algorithms.}
@@
 \centering
-\includegraphics[scale=.3]{err_est_ffbs.eps}
+\includegraphics[scale=.3]{err_est_ffbs}
 \caption{Posterior probabilities estimation error fo the two FFBS algorithms.}
@@
 \includegraphics[scale=.5]{N15_1001_1500}
-\caption{Experiment 2 - estimated probabilites by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
+\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
 \label{fig:exp2:label}
@@
 
-\subsection{Application to CME crude oil (WTI)}
+\section{Application to CME crude oil (WTI)}
+\label{sec:crudeoil}
 \label{sec:exp}
 
-\subsubsection*{State dynamics}
+\subsection{Model}
 Modeling commodity prices is a crucial step to valuate contingent claims related to energy markets and to optimize storage or extraction strategies.
@@
 %We assume that $J=2$, each regime characterizing the slope of the future curve. A backwardation regime (resp. contango regime) is defined as a negative slope (resp. a positive slope).
-Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$.
+%Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$.
 It is assumed that each factor (spot, convenience yield) has its own constant market price of risk, $\lambda^S$ for $S_t$ and $\lambda^{\delta}$ for $\delta_t$.
 The historical dynamics of the Gibson-Schwartz model is then described as follows:
-\begin{align*}
-\rmd S_{t} & = (\mu -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t} \rmd \widetilde{W}_{t}^{1} \eqsp,\\
-\rmd \delta_{t} & = \kappa_{a_t}(\widetilde{\alpha}_{a_t}-\delta_{t})\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd \langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho \rmd t\eqsp,
-\end{align*}
-where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $.
+%\begin{align*}
+%\rmd S_{t} & = (\mu -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t} \rmd \widetilde{W}_{t}^{1} \eqsp,\\
+%\rmd \delta_{t} & = \kappa_{a_t}(\widetilde{\alpha}_{a_t}-\delta_{t})\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd \langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho \rmd t\eqsp,
+%\end{align*}
+%where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $.
 This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield.
 A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).
-Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a Conditionally Linear Gaussian Model.
+Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a Conditionally Linear Gaussian Model.
 The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:
@@
 d_{a_{i-1}} &\eqdef
-\begin{pmatrix} \left[\mu- \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}}\\
-\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa_{a_{i-1}} \Delta t_i}] \end{pmatrix}\eqsp,\\
+\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \alpha_{a_{i-1}}[1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}}\\
+\alpha_{a_{i-1}} [1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}] \end{pmatrix}\eqsp,\\
 T_{a_{i-1}} &\eqdef
-\begin{pmatrix} 1 & -[1-e^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}} \\ 0 & e^{-\kappa_{a_{i-1}} \Delta t_i} \end{pmatrix} \eqsp,\\
-\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta t_i + \eta^2_{a_{i-1}}\left\{\Delta t_i + (1-e^{-2\kappa_{a_{i-1}} \Delta t_i})/(2\kappa_{a_{i-1}}) - 2(1-e^{-\kappa \Delta t_i})/\kappa_{a_{i-1}}\right\}/\kappa_{a_{i-1}}^2 \\
-&\hspace{5.2cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-e^{-\kappa_{a_{i-1}} \Delta t_i})/\kappa_{a_{i-1}} \right\}/\kappa\eqsp, \\
-\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa_{a_{i-1}}\right)\left(1-e^{-\kappa_{a_{i-1}}\Delta t_i} \right)/\kappa_{a_{i-1}} \\
-&\hspace{5.2cm}+ \eta^2_{a_{i-1}}\left(1-e^{-2\kappa_{a_{i-1}} \Delta t_i} \right)/(2\kappa_{a_{i-1}}^2)\eqsp,\\
+\begin{pmatrix} 1 & -[1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}} \\ 0 & \rme^{-\kappa_{a_{i-1}} \Delta t_i} \end{pmatrix} \eqsp,\\
+\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta t_i + \eta^2_{a_{i-1}}\left\{\Delta t_i + (1-\rme^{-2\kappa_{a_{i-1}} \Delta t_i})/(2\kappa_{a_{i-1}}) - 2(1-\rme^{-\kappa \Delta t_i})/\kappa_{a_{i-1}}\right\}/\kappa_{a_{i-1}}^2 \\
+&\hspace{5.2cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-\rme^{-\kappa_{a_{i-1}} \Delta t_i})/\kappa_{a_{i-1}} \right\}/\kappa\eqsp, \\
+\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa_{a_{i-1}}\right)\left(1-\rme^{-\kappa_{a_{i-1}}\Delta t_i} \right)/\kappa_{a_{i-1}} \\
+&\hspace{5.2cm}+ \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa_{a_{i-1}} \Delta t_i} \right)/(2\kappa_{a_{i-1}}^2)\eqsp,\\
 \barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp,\\
-\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-e^{-2\kappa_{a_{i-1}}\Delta t_i} \right)/(2\kappa_{a_{i-1}})\eqsp.
+\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa_{a_{i-1}}\Delta t_i} \right)/(2\kappa_{a_{i-1}})\eqsp.
 \end{align*}
@@
 %The transition matrix $Q$ is assumed to be the same under the pricing measure.
-\subsubsection*{Observation model}
 The observations are Wednesday future contracts of the West Texas Intermediate crude oil (WTI) traded in the Chicago Mercantile Exchange (CME) from $11$ January 1995 to $13$ November 2013.
@@
 
-\subsubsection*{Numerical results}
-\label{sec:crudeoil}
+\subsection{Numerical results}
 %Following \cite{carmona:ludkovski:2004}, under the risk neutral probability $\mathbb{Q}$, we assume that the spot price $S_t$ of the Chicago Mercantile Exchange (CME) crude oil and its associated convenience yield $\delta_t$ are solutions to the SDE:

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 \rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd W_{t}^{1} \eqsp,\\
-\rmd\delta_{t} & = \kappa_{a_t}(\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
+\rmd\delta_{t} & = \kappa (\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
 \end{align*}
 where $(a_t)_{t\ge 0}$ is a finite state space Markov process.
-%We assume that $J=2$, each regime characterizing the slope of the future curve. A backwardation regime (resp. contango regime) is defined as a negative slope (resp. a positive slope).
+
 %Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$.
@@
 \[
-Z_i = d_{a_{i-1}} + T_{a_{i-1}} Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
+Z_i = d_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
 \]
@@
 %\end{align*}
-where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\Delta t_i = t_i-t_{i-1}$):
+where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\Delta = t_i-t_{i-1}$):
 \begin{align*}
 d_{a_{i-1}} &\eqdef
-\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \alpha_{a_{i-1}}[1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}}\\
-\alpha_{a_{i-1}} [1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}] \end{pmatrix}\eqsp,\\
-T_{a_{i-1}} &\eqdef
-\begin{pmatrix} 1 & -[1-\rme^{-\kappa_{a_{i-1}} \Delta t_i}]/\kappa_{a_{i-1}} \\ 0 & \rme^{-\kappa_{a_{i-1}} \Delta t_i} \end{pmatrix} \eqsp,\\
-\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta t_i + \eta^2_{a_{i-1}}\left\{\Delta t_i + (1-\rme^{-2\kappa_{a_{i-1}} \Delta t_i})/(2\kappa_{a_{i-1}}) - 2(1-\rme^{-\kappa \Delta t_i})/\kappa_{a_{i-1}}\right\}/\kappa_{a_{i-1}}^2 \\
-&\hspace{5.2cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-\rme^{-\kappa_{a_{i-1}} \Delta t_i})/\kappa_{a_{i-1}} \right\}/\kappa\eqsp, \\
-\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa_{a_{i-1}}\right)\left(1-\rme^{-\kappa_{a_{i-1}}\Delta t_i} \right)/\kappa_{a_{i-1}} \\
-&\hspace{5.2cm}+ \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa_{a_{i-1}} \Delta t_i} \right)/(2\kappa_{a_{i-1}}^2)\eqsp,\\
+\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta + \alpha_{a_{i-1}}[1-\rme^{-\kappa \Delta}]/\kappa\\
+\alpha_{a_{i-1}} [1-\rme^{-\kappa \Delta}] \end{pmatrix}\eqsp,\\
+T &\eqdef
+\begin{pmatrix} 1 & -[1-\rme^{-\kappa \Delta}]/\kappa \\ 0 & \rme^{-\kappa \Delta}  \end{pmatrix} \eqsp,\\
+\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta + \eta^2_{a_{i-1}}\left\{\Delta  + (1-\rme^{-2\kappa \Delta})/(2\kappa) - 2(1-\rme^{-\kappa \Delta})/\kappa\right\}/\kappa^2 \\
+&\hspace{4.4cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-\rme^{-\kappa \Delta})/\kappa \right\}/\kappa\eqsp, \\
+\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa \right)\left(1-\rme^{-\kappa \Delta}} \right)/\kappa} \\
+&\hspace{7.8cm}+ \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa \Delta}} \right)/(2\kappa^2)\eqsp,\\
 \barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp,\\
-\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa_{a_{i-1}}\Delta t_i} \right)/(2\kappa_{a_{i-1}})\eqsp.
+\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa \Delta} \right)/(2\kappa)\eqsp.
 \end{align*}
@@
 %In addition, the backwardation is more likely to stay in backwardation than switch to contango and the same holds for the contango regime.
-
-\medskip
-
-Following \cite{almansour:2016}, it is assumed that $\kappa_1 = \kappa_2 = \kappa$ (and therefore $T_1=T_2=T$) to obtain a closed form solution for $F_{t_i,m}$ which may be written:
+A closed form solution for $F_{t_i,m}$  may be written:
 \[

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 \]
-%where (setting $\Delta t_i = t_i-t_{i-1}=1$):
+%where (setting $\tau t_i = t_i-t_{i-1}=1$):
 %\[
@@
 %\end{align*}
-where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\Delta = t_i-t_{i-1}$):
+where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\tau = t_i-t_{i-1}$):
 \begin{align*}
 d_{a_{i-1}} &\eqdef
-\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta + \alpha_{a_{i-1}}[1-\rme^{-\kappa \Delta}]/\kappa\\
-\alpha_{a_{i-1}} [1-\rme^{-\kappa \Delta}] \end{pmatrix}\eqsp,\\
+\begin{pmatrix} \left[\mu- \alpha_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \tau + \alpha_{a_{i-1}}[1-\rme^{-\kappa \tau}]/\kappa\\
+\alpha_{a_{i-1}} [1-\rme^{-\kappa \tau}] \end{pmatrix}\eqsp,\\
 T &\eqdef
-\begin{pmatrix} 1 & -[1-\rme^{-\kappa \Delta}]/\kappa \\ 0 & \rme^{-\kappa \Delta}  \end{pmatrix} \eqsp,\\
-\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \Delta + \eta^2_{a_{i-1}}\left\{\Delta  + (1-\rme^{-2\kappa \Delta})/(2\kappa) - 2(1-\rme^{-\kappa \Delta})/\kappa\right\}/\kappa^2 \\
-&\hspace{4.4cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \Delta t_i - (1-\rme^{-\kappa \Delta})/\kappa \right\}/\kappa\eqsp, \\
-\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa \right)\left(1-\rme^{-\kappa \Delta}} \right)/\kappa} \\
-&\hspace{7.8cm}+ \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa \Delta}} \right)/(2\kappa^2)\eqsp,\\
-\barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp,\\
-\barH_{a_{i-1}}(2,2) &= \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa \Delta} \right)/(2\kappa)\eqsp.
+\begin{pmatrix} 1 & -[1-\rme^{-\kappa \tau}]/\kappa \\ 0 & \rme^{-\kappa \tau}  \end{pmatrix} \eqsp,\\
+\barH_{a_{i-1}}(1,1) &= \sigma^2_{a_{i-1}} \tau + \eta^2_{a_{i-1}}\left\{\tau  + (1-\rme^{-2\kappa \tau})/(2\kappa) - 2(1-\rme^{-\kappa \tau})/\kappa\right\}/\kappa^2 \\
+&\hspace{4.4cm} - 2\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}\left\{ \tau t_i - (1-\rme^{-\kappa \tau})/\kappa \right\}/\kappa\eqsp, \\
+\barH_{a_{i-1}}(1,2) & = \left(\rho_{a_{i-1}} \eta_{a_{i-1}} \sigma_{a_{i-1}}-\eta^2_{a_{i-1}}/\kappa \right)\left(1-\rme^{-\kappa \tau} \right)/\kappa + \eta^2_{a_{i-1}}\left(1-\rme^{-2\kappa \tau} \right)/(2\kappa^2)\eqsp,\\
+\barH_{a_{i-1}}(2,1) &= \barH_{a_{i-1}}(1,2)\eqsp, \quad \barH_{a_{i-1}}(2,2) = \eta^2_{a_{i-1}}\left( 1-\rme^{-2\kappa \tau} \right)/(2\kappa)\eqsp.
 \end{align*}
@@
 %\widetilde{d}_{a_{i-1}} \eqdef
-%\begin{pmatrix} \left[r - \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \Delta t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa \Delta t_i}]/\kappa\\
-%\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa \Delta t_i}] \end{pmatrix}\eqsp,
+%\begin{pmatrix} \left[r - \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \tau t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa \tau t_i}]/\kappa\\
+%\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa \tau t_i}] \end{pmatrix}\eqsp,
 %\]
@@
 Our time series contains $n =975$ weekly data with $534$ in backwardation and $441$ in contango (the backwardation effect is more frequent with crude oil data).
-At each time $t_i = i\Delta t$, with $\Delta t = 0.0192$, the observations of the $\dimy = 4$ future prices are $Y_i \eqdef (\ln(F^{(market)}_{i\Delta t,m_1}), \ldots, \ln(F^{(market)}_{i\Delta t,m_{\dimy}}))'$, where $F_{t_i,m}$ is the future price at $t_i$ for a maturity  $m$ weeks.
+At each time $t_i = i\tau$, with $\tau = 0.0192$, the observations of the $\dimy = 4$ future prices are $Y_i \eqdef (\ln(F^{(market)}_{i\tau t,m_1}), \ldots, \ln(F^{(market)}_{i\tau t,m_{\dimy}}))'$, where $F_{t_i,m}$ is the future price at $t_i$ for a maturity  $m$ weeks.
 %In addition, the backwardation is more likely to stay in backwardation than switch to contango and the same holds for the contango regime.
@@
 \]
-where $\mathsf{B}_0	= \begin{pmatrix}1 & 0 \end{pmatrix}$ and  $\mathsf{B}_m    = \mathsf{B}_{m-1} T$ so that  $ \mathsf{B}_m= \begin{pmatrix} 1 & -\left(1-\rme^{-\kappa m \Delta t}\right)/\kappa \end{pmatrix}$ and
-%$B_\tau$ is the $1 \times 2$ matrix given by $B_{\tau} \eqdef  \begin{pmatrix} 1 &  -(1-e^{-\kappa \tau})/\kappa \end{pmatrix}$ and
-%for all $1\le j\le J$, $A_0(j) \eqdef 0$.
-Define $F_{i,m} \eqdef F_{i\Delta t, m \Delta t }$ and $\mathcal{F}_i$ the $\sigma$-field generated by $(a_k)_{0\le k \le i}$ and $(Z_k)_{0\le k \le i}$.
-By definition of future prices and the assumption that there is no regime switching between $t_i$ and $t_{i+1}$, we have:
-%%\begin{align*}
-%%A_{t,\tau} &\eqdef r \tau - \int^{\tau}_0 \alpha_{a_{t+s}} \left(1- e^{-\kappa(\tau-s)} \right)\rmd s + (1/\kappa^2)\int^{\tau}_0 \eta^2_{a_{t+s}} \left(1-e^{-\kappa (\tau-s)}\right)^2 \rmd s \\
-%%&\hspace{7cm}  - (2/\kappa)\int^{\tau}_0 \rho_{a_{t+s}} \eta_{a_{t+s}}\sigma_{a_{t+s}} \left( 1 -e^{-\kappa \left(\tau-s\right)} \right)\rmd s \eqsp,\\
-%%B_{\tau} &\eqdef  \begin{pmatrix} 1 &  -(1-e^{-\kappa \tau})/\kappa \end{pmatrix}\eqsp.
-%%\end{align*}
-%\begin{align*}
-%F_{i,m} &= \exp (A_m(a_i) + B_m Z_i)\eqsp,\\
-%&=\mathbb{E}\left[\exp (A_{m-1}(a_{i+1}) + B_{m-1} Z_{i+1})\middle|\mathcal{F}_i\right]\eqsp,\\
-%&= \mathbb{E}\left[\sum^J_{k=1} 1_{a_{i+1}=k}\exp (A_{m-1}(k) + B_{m-1} Z_{i+1})\middle|\mathcal{F}_i\right] \eqsp,\\
-%&= \mathbb{E}\left[\sum^J_{k=1} 1_{a_{i+1}=k}\exp \left(A_{m-1}(k) + B_{m-1} ( d_{a_i} + T Z_i + H_{a_i} \epsilon_{i+1} ) \right)\middle|\mathcal{F}_i\right] \eqsp,\\
-%&= \sum^J_{k=1} q(a_i,k) \exp \left(A_{m-1}(k) + B_{m-1} d_{a_i} + B_{m-1} \barH_{a_i} B'_{m-1}/2 + B_{m-1} T Z_i \right)\eqsp.
-%\end{align*}
-%From the above equality, we obtain,
+where $\mathsf{B}_0	= \begin{pmatrix}1 & 0 \end{pmatrix}$ and  $\mathsf{B}_m    = \mathsf{B}_{m-1} T$ so that  $ \mathsf{B}_m= \begin{pmatrix} 1 & -\left(1-\rme^{-\kappa m \tau t}\right)/\kappa \end{pmatrix}$ and
 for all $1\le j\le J$, $\mathsf{A}_0(j) = 0$,  and
@@
 The EM algorithm iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
-\begin{enumerate}%[label=\roman*)]
+\begin{enumerate}
 	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p})\eqdef \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{1:n},Y_{1:n})\middle|Y_{1:n}\right]$ ;
@@
 \end{figure}\\
-The number of particles is set to $N = 100$, $\Delta t = 1/52$. % and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}.
+The number of particles is set to $N = 100$, $\tau  = 1/52$. % and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}.
 The interest rate is $r = 0.0296$ as in \cite{almansour:2016}. The initial guess for the mean and variance of the initial state are
 \[
-\mu_1 = \begin{pmatrix} \ln F^{(market)}_{1,4} & r-\cfrac{\ln F^{(market)}_{1,16}-\ln F^{(market)}_{1,4}}{(16-4)\Delta t} \end{pmatrix}\quad \mbox{and}\quad \Sigma_1 = \begin{pmatrix} 0.05 & 0 \\ 0 & 0.05 \end{pmatrix}\eqsp.
+\mu_1 = \begin{pmatrix} \ln F^{(market)}_{1,4} & r-\cfrac{\ln F^{(market)}_{1,16}-\ln F^{(market)}_{1,4}}{(16-4)\tau} \end{pmatrix}\quad \mbox{and}\quad \Sigma_1 = \begin{pmatrix} 0.05 & 0 \\ 0 & 0.05 \end{pmatrix}\eqsp.
 \]

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 \[
-c_{j}' = \begin{pmatrix} \mathsf{A}_{m_1}(j) & \ldots & \mathsf{A}_{m_{\dimy}}(j) \end{pmatrix}\eqsp,
+c_{j}' = [ \mathsf{A}_{m_1}(j), \dots, \mathsf{A}_{m_{\dimy}}(j) ]\eqsp,
 \;\;\;
-B' = \begin{pmatrix}\mathsf{B}_{m_1}' &  \ldots & \mathsf{B}_{m_{\dimy}}' \end{pmatrix}\eqsp, \;\;\;
-G = \mathrm{diag}(g_1, \ldots , g_{d} )\eqsp.
-\]
-The model is then driven by the parameter:
-\[
-\theta \eqdef \{\pi, Q, \mu_1, \Sigma_1, \kappa, (\alpha_j)_{1\leq j\leq J}, (\sigma_j)_{1\leq j\leq J}, (\eta_j)_{1\leq j\leq J}, (\rho_j)_{1\leq j\leq J}, \mu, (\lambda_j)_{1\leq j\leq J}, (g_{\ell})_{1\leq \ell\leq d} \}\eqsp.
+B' = [\mathsf{B}_{m_1}',  \dots, \mathsf{B}_{m_{\dimy}}']\eqsp, \;\;\;
+G = \mathrm{diag}(g_1, \dots , g_{d} )\eqsp.
+\]
+The model depends of the parameters:
+\[
+\theta \eqdef \{\pi, Q, \mu_1, \Sigma_1, \kappa, (\alpha_j)_{1\leq j\leq J}, (\sigma_j)_{1\leq j\leq J}, (\eta_j)_{1\leq j\leq J}, (\rho_j)_{1\leq j\leq J}, \mu, (\lambda_j)_{1\leq j\leq J}, (g_{\ell})_{1\leq \ell\leq d} \}\eqsp .
 \]
@@
 where $\phi_{\mu_1,\Sigma_1}$ is the Gaussian probability density function with mean $\mu_1$ and variance $\Sigma_1$.
-Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{1:n}\right]$  the conditional expectation given $Y_{1:n}$ when the parameter value is $\theta$.
+Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{1:n}\right]$  the conditional expectation given $Y_{1:n}$ when the parameter value is set to $\theta$.
 The EM algorithm iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
@@
 From the results, $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$ indicates the first regime (backwardation) characterized by a higher value in both volatility and equilibrium level of convenience yield, and the second regime (contango) characterized by a lower value in both volatility and equilibrium convenience yield level.
-It verifies the theory of storage that the volatility of the commodity spot price is high when the inventory is low, and the convenience yield is all the higher as inventory is low.
+This in accordance with the theory of storage that the volatility of the commodity spot price is high when the inventory is low, and the convenience yield is all the higher as inventory is low.
+
 Interestingly, we see that $\lambda^{\delta}_1 \geq \lambda^{\delta}_2$ \footnote{\cite{almansour:2016} obtain the same result in the opposite order, because of their definition of the historical mean-reverting value} and the two historical mean-reverting values are $\widetilde{\alpha}_1 = 0.0073$ and $\widetilde{\alpha}_2 = 0.0305$.

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 \tilde{c}_{i|i+1}(a_{i+1:n}) &= \tilde{c}_{i+1}(a_{i+1:n}) + \log|\barH_{a_{i+1}}| + d'_{a_{i+1}}\barH_{a_{i+1}}^{-1}d_{a_{i+1}}-\log|H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}|\\
-&\hspace{6cm} - \tilde{r}'_{i|i+1}(a_{i+1:n})H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\tilde{r}_{i|i+1}(a_{i+1:n})\eqsp,\\
+&\hspace{4.0cm} - \tilde{r}'_{i|i+1}(a_{i+1:n})H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\tilde{r}_{i|i+1}(a_{i+1:n})\eqsp,\\
 \tilde{P}_{i|i+1}^{-1}(a_{i+1:n}) &=T'_{a_{i+1}}\left(I_{\dimz}-\barH_{a_{i+1}}^{-1}H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\right)\barH_{a_{i+1}}^{-1}T_{a_{i+1}}\eqsp,\\
@@
 &=  \tilde{p}_{i+1}(y_{i+1:n}) \sum_{\ell=1}^N C_i^{-1}(\tilde{a}^{\ell}_{i+1:n}) Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_i H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2} \eqsp\nonumber \\
-&\hspace{4.3cm} \times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{\tilde{a}^{\ell}_{i+1}}}{d_{\tilde{a}^{\ell}_{i+1}}+T_{\tilde{a}^{\ell}_{i+1}}z_i}^2\right\} \eqsp,\\
+&\hspace{2.3cm} \times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{\tilde{a}^{\ell}_{i+1}}}{d_{\tilde{a}^{\ell}_{i+1}}+T_{\tilde{a}^{\ell}_{i+1}}z_i}^2\right\} \eqsp,\\
 &= \sum_{\ell=1}^N\tilde{\omega}_{\mathsf{b},i}^\ell(a_i)\exp\left\{-\frac{1}{2}\normMat{\tilde{S}_{i+1}^\ell}{z_i}^2+z_i'\tilde{s}_{i+1}^\ell\right\}\eqsp,\label{eq:tN}
@@
 
-Interestingly, we see that $\lambda^{\delta}_1 \geq \lambda^{\delta}_2$ \footnote{\cite{almansour:2016} obtain the same result in the opposite order, because of their definition of the historical mean-reverting value} and the two historical mean-reverting values are $\widetilde{\alpha}_1 = 0.0073$ and $\widetilde{\alpha}_2 = 0.0305$.
-This underlines that, while the first regime of convenience yield under $\mathbb{P}$ makes spot price be more likely to increase (with a smaller historical mean-reverting), the term structure of future prices has a tendency to be in backwardation.
-	This also shows opposite behaviors of historical prices evolution and term structure future prices.
-Intuitively, Figure~\ref{fig:LogPriceTermStructure:data} compares the evolution of future $1M$ (the nearest contracts) to the term structure observed from CME WTI crude oil.
-The term structure is measured by the difference of future $13M$ and future $1M$ to avoid the eventual saisonality.
-The figure shows that it is not necessary to have an inverse relationship between the evolution of price and term structure.
-But when a significant drop in price occurs, the term structure measured as $F_{13} - F_1$ increases (i.e.
+%Interestingly, we see that $\lambda^{\delta}_1 \geq \lambda^{\delta}_2$ \footnote{\cite{almansour:2016} obtain the same result in the opposite order, because of their definition of the historical mean-reverting value} and the two historical mean-reverting values are $\widetilde{\alpha}_1 = 0.0073$ and $\widetilde{\alpha}_2 = 0.0305$.
+This underlines that,
+
+Figure~\ref{fig:LogPriceTermStructure:data} compares the evolution of future $1M$ (the nearest contracts) to the \emph{term structure} observed from CME WTI crude oil, defined as the difference of future $13M$ and future $1M$ (to avoid seasonality).
+The figure shows that it is not necessary to have an inverse relationship between the  price of the nearest contract and the term structure.
+But when a significant drop in the price  of the nearest contract occurs, the term structure increases (i.e.
 in contango).
@@
 \centering
-\includegraphics[scale=.6]{LogPrice_TermStructure}
+\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
 \caption{Log-price (red line) and slope of future curves (blue line)}
@@
 The correlation between the spot price and the convenience yield is positive and high in both two regimes.
-The positive correlation verifies one of the fundamental principes in commodity market as in \cite{gibson:schwartz:1990}.
-That explains when the spot prices are expected to inscrease, the convenience yield are also expected to inscrease.
-In addition, the higher the correlation is, the more the slope of future curve descreases in function of maturity.
-It descreases even further in the first regime (in backwardation): the higher volatility higher convenience yield with the higher correlation.
+This is an accordance to what as been observed in most commodity market; see \cite{gibson:schwartz:1990}.
+The slope of future curve decreases in function of maturity.
 
@@
 	\centering
-	\includegraphics[scale=.6]{Posterior_Data}
+	\includegraphics[scale=.75,angle=90]{Posterior_Data}
 	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}

### 2016-11-15 TexArxiv/clgm.tex
MAJ abstract

@@
 This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces.
-Only a few methods have been proposed to reduce Monte Carlo variance of smoothers with Rao Blackwellisation steps for these models.
-The algorithm proposed by  \cite{briers:doucet:maskell:2010} extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps.
-Other Rao Blackwellised Sequential Monte Carlo smoothers introduced in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} are based on the forward backward decomposition of the marginal smoothing distribution % In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
- and use explicit marginalization of the linear states in both forward and backward passes.
-In this paper, we propose an overview of these algorithms and introduce an additional Rao Blackwellisation step for all procedures which allows to sample new particles in the backward pass to improve the performance of the algorithm.
-The procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield.
-Parameters are estimated using  an Expectation Maximization and the performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.
+To reduce Monte Carlo variance of smoothers, it is typical in these models to use Rao-Blackwellisation.
+Particle approximation is used to sample the sequence of the hidden regimes; the gaussian state sequence are then explicitly integrated conditional to the sequence of regimes and observations, using variants of the Kalman filter / smoother.
+The first successful attempt to use Rao-Blackwellisation for smoothing has been presented in  \cite{briers:doucet:maskell:2010} which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filter / smoother.
+More recently, in a series of article, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed to use a  forward backward decomposition of the smoothing distribution, mimicking the Rauch-Tung-Striebel smoother for the regime combined with backward Kalman updates.
+% In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
+In this paper, we provide a survey of these algorithms in a common unifying framework.
+We then investigate the benefit of introducing an additional rejuvenation step  which allows to sample at each time instant new  regime conditional to the forward and backward particles.
+To support our claims, a limited Monte Carlo study is presented.
+These procedures are applied to commodity markets which are described using a  factor model based on the spot price and a convenience yield for crude oil data.
 

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 In this paper, we provide a survey of these algorithms in a common unifying framework.
-We then investigate the benefit of introducing an additional rejuvenation step  which allows to sample at each time instant new  regime conditional to the forward and backward particles.
+We then investigate the benefit of introducing an additional rejuvenation step  which allows to sample at each time instant new  regime conditional to the forward and backward particles approximating the regimes.
 To support our claims, a limited Monte Carlo study is presented.
@@
 At each time step, it is assumed that noisy observations of future contracts are available which leads to represent commodity markets dynamics using a conditionally linear and Gaussian state space model.
-We propose a new algorithm to approximate the conditional distributions of the states (spot, convenience yield and regime indicator) given a set of observations.
+We propose a new algorithm to approximate the conditional distributions of the states (spot, convenience yield and regime) given a set of observations.
 These approximations are obtained combining Sequential Monte Carlo methods and Kalman smoothing techniques  and may be used to detect significant regime switching times and to estimate model parameters with an Expectation Maximization based algorithm.
@@
 Filtering refers to the estimation of the distributions of the hidden state $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time $k$, while fixed-interval smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$.
-These distributions are in particular crucial to infer a sequence of states given the observations, see for instance \cite{thrun:burgard:fox:2005} for Simultaneous Localization and Mapping problems, \cite{barshalom:li:kirubarjan:2001} for target tracking problems or \cite{sarkka:2013,douc:moulines:stoffer:2013} for other applications in engineering, telecommunications...
-In the case where the state and observation models are linear and Gaussian, the filtering can be solved explicitly using the Kalman filter \cite{kalman:1960}.
+
+When the state and observation models are linear and Gaussian, the filtering can be solved explicitly using the Kalman filter \cite{kalman:1960}.
 Exact solutions of the fixed-horizon smoothing problem can be obtained using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier "two-filter" smoother \cite{bryson:frazier:1963}.
@@
 \item $(\varepsilon_k)_{k\ge 1}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance.
-\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$.  %$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
+\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$, called \emph{regimes}  %$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
 \item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
-\item $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_k)_{k\ge 1}$.
+\item \textcolor{red}{j'aurais mis $Z_0$} $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_k)_{k\ge 1}$.
 \end{enumerate}
@@
 CLGM play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
-A crucial feature of these models is that, conditional on a trajectory $(a_1,\ldots,a_n)$ of indicator variables, both the state equation and the observation equation are linear and Gaussian, which implies that conditional on the sequence of labels, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be solved explicitly.
-To exploit this specific structure, it has been suggested in the pioneering paper of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample indicator variables with the Kalman filter to compute the conditional distribution of the 'continuous' states sequence $(Z_j)_{j=1}^n$.
+A crucial feature of these models is that, conditional on the regime sequence $(a_1,\ldots,a_n)$ , both the state equation and the observation equation are linear and Gaussian, which implies that conditional on the sequence of regimes and the observations, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be computed explicitly.
+
+To exploit this specific structure, it has been suggested in the pioneering papers of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample the regimes with the Kalman filter to compute the conditional distribution of the states sequence $(Z_j)_{j=1}^n$ conditional on the regimes and the observations.
 This is a specific instance of Rao-Blackwellised Monte Carlo filters, often referred to as the \emph{Mixture Kalman Filter}.
@@
 The first fully Rao Blackwellised SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
-This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC for the indicator variables and Kalman filtering techniques for the hidden linear states.
+This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC for the regime sequence and Kalman filtering techniques for the hidden linear states.
 
@@
 We propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables.
-Similarly, in the backward pass of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}, an indicator variable $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$.
-In this case, particle rejuvenation may be introduced by using the forward weighted samples at time $i-1$ and extending these trajectories at time $i$ with a Kalman filter for all possible values of the indicator variable.
+Similarly, in the backward pass of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}, a regime $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$.
+In this case, particle rejuvenation may be introduced by using the forward weighted samples at time $i-1$ and extending these trajectories at time $i$ with a Kalman filter for all possible values of the regime.
 Then, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$ using an appropriately adapted weight.
@@
 In \cite{chen:liu:2000}, for all $1\le k \le N$, an ancestral path is chosen with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
-Then, the new indicator variable $a_i^k$ is sampled in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$.
+Then, the new regime $a_i^k$ is sampled in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$.
 A drawback of this method is that only ancestral paths that have been selected using the importance weights $(\omega^k_{i-1})_{1\le k \le N}$ are extended at time $i$.
@@
 \subsection{Algorithm based on the structural approximation of \cite{kim:1994}, \cite{barber:2006}}
-The method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{barber:2006} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$.
+The method to approximate the smoothing distributions of the rzgimes given all the observations proposed by \cite{barber:2006} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$.
 Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$ approximating the filtering distributions are available.
@@
 \end{enumerate}
-If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the indicator variables is:
+If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the regimes is:
 \[
@@
 \subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
-\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} without using the ad-hoc approximation of \cite{kim:1994}.
+\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the regime backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} without using the ad-hoc approximation of \cite{kim:1994}.
 The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
@@
 \]
-If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{lindsten:bunch:sarkka:schon:godsill:2015} of the joint smoothing distribution of the indicator variables is:
+If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{lindsten:bunch:sarkka:schon:godsill:2015} of the joint smoothing distribution of the regime is:
 \[
@@
 %\]
-On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the indicator variables of the filtering pass $\{a_i^k\}_{1\le k \le N}$.
+On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the regime of the filtering pass $\{a_i^k\}_{1\le k \le N}$.
 This means that a Kalman filter must be used for each trajectory $a^k_{1:i-1}$ which may be extended by $a_i\in\{1,\ldots,J\}$.
@@
 %Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}.
-Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations).
+Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  regimes given all the observations).
 \cite{briers:doucet:maskell:2010} introduced the following decomposition of the smoothing distributions for all $2\le i \le n$:
@@
 \end{align}
-In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of indicator variables $a_{1:i-1}^k$ associated with importance weights $\omega_{i-1}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-1},z_{i-1}|y_{1:i-1})$.
+In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of regimes $a_{1:i-1}^k$ associated with importance weights $\omega_{i-1}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-1},z_{i-1}|y_{1:i-1})$.
 Plugging this approximation into \eqref{eq:smooth:r} provides a normalized approximation $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ of $r_{i}(a_{i},z_{i},y_{1:i-1})$ integrating over all possible choices of $(a_{i-1},z_{i-1})$:
@@
 \end{figure}
-%In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the labels as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).
+%In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the regimes as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).
 
@@
 It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$).
-Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the labels given by the algorithms using:
+Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using:
 \[
@@
 \includegraphics[scale=.5]{N10_1001_1500}
-\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true labels (squared black).}
+\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
 \label{fig:exp1:label}
@@
 \includegraphics[scale=.5]{N15_1001_1500}
-\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true labels (squared black).}
+\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
 \label{fig:exp2:label}
@@
 
-Figure~\ref{fig:posterior:data} and~\ref{fig:posterior:data:logprice} displays the the estimated posterior probabilities of the regime labels and the observed future slope.
+Figure~\ref{fig:posterior:data} and~\ref{fig:posterior:data:logprice} displays the the estimated posterior probabilities of the regimes and the observed future slope.
 When the future curve is in backwardation (resp.

### 2016-11-15 TexArxiv/clgm.tex
MAJ intro

@@
 More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
+The update of the smoothing distribution of the regime given the observation shares some striking similarities with the Rauch-Tung-Striebel smoothing procedure, which is at the heart of the FFBS procedure.
+The Rao-Blackwellisation requires to update backward in time the smoothing distribution of the states given the regimes and the observations, which is achieved by using a \emph{à la Kalman} backward update.
 %The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass.
@@
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
-In Section~\ref{sec:crudeoil}, an application of regime switchings for commodity markets is presented using CME crude oil (WTI) data.
+In Section~\ref{sec:crudeoil}, an application to commodity markets is presented; the performance of our procedure is illustrated on crude oil  data.
 A detailed derivation of the algorithms is provided in the Appendix.

### 2016-11-15 TexArxiv/clgm.tex
MAJ introduction
MAJ clgm.Tex

@@
 \end{enumerate}
+\textcolor{red}{après on part de $k=1$}
 At each time step $k\ge 0$, the observation $Y_k$ is given by:
@@
 This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$.
-For all $m\,\times\,m$ matrix  let $|A|$ be the determinant of $A$ if $A$ is positive-definite matrix, for all $z\in\mathbb{R}^{m}$ define
+For all $m\,\times\,m$ matrix  let $|A|$ be the determinant of $A$.
+If $A$ is a positive-definite matrix, for all $z\in\mathbb{R}^{m}$ define
 \[
@@
 SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$.
-All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter detailed in Section~\ref{sec:forward}.
+All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter presented in Section~\ref{sec:forward}.
 
@@
 The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using a standard Rao Blackwellised  algorithm.
-The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
+The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with normalized importance weights $(\omega^k_{i})_{1\le k \le N}$ ($\sum_{k=1}^N \omega^k_i = 1$) used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
 \begin{equation}
@@
 \subsection{Algorithm based on the structural approximation of \cite{kim:1994}, \cite{barber:2006}}
-The method to approximate the smoothing distributions of the rzgimes given all the observations proposed by \cite{barber:2006} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$.
+The method to approximate the smoothing distributions of the regimes given all the observations proposed by \cite{barber:2006} relies on an approximation of $p(a_i|a_{i+1:n},y_{1:n})$, which is hardly justified and may introduce some bias.
 Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$ approximating the filtering distributions are available.

### 2016-11-15 TexArxiv/clgm.tex
MAJ

@@
 \]
-where
+This decomposition is similar to the Rauch-Tung-Striebel decomposition of the filtering distribution.
+The first factor on the right hand side of the previous equation is nevertheless more difficult to handle because it itself relies on all the observations.
+Nevertheless, as noted by \cite{sarkka:bunch:godsill:2012}, this term can be computed recursively by considering the following obvious decomposition:
 \[

### 2016-11-18 TexArxiv/clgm.tex
Last modifications before submission

@@
 
-\title{Rao-Blackwellized  smoother for Conditionally Linear and Gaussian models}
+\title{A survey of Rao-Blackwellized Sequential Monte Carlo smoothers for Conditionally Linear and Gaussian models}
 \date{}
@@
 This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces.
-To reduce Monte Carlo variance of smoothers, it is typical in these models to use Rao-Blackwellisation.
-Particle approximation is used to sample the sequence of the hidden regimes; the gaussian state sequence are then explicitly integrated conditional to the sequence of regimes and observations, using variants of the Kalman filter / smoother.
-The first successful attempt to use Rao-Blackwellisation for smoothing has been presented in  \cite{briers:doucet:maskell:2010} which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filter / smoother.
+To reduce Monte Carlo variance of smoothers, it is typical in these models to use Rao-Blackwellisation: particle approximation is used to sample the sequence of hidden regimes while the Gaussian states are explicitly integrated conditional on the sequence of regimes and observations, using variants of the Kalman filter / smoother.
+The first successful attempt to use Rao-Blackwellisation for smoothing has been presented in  \cite{briers:doucet:maskell:2010} which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filters / smoothers.
 More recently, in a series of article, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed to use a  forward backward decomposition of the smoothing distribution, mimicking the Rauch-Tung-Striebel smoother for the regime combined with backward Kalman updates.
-% In \cite{kim:1994}, a structural approximation of the model is used to simplify the backward pass while \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} use Rao Blackwellisation steps
 In this paper, we provide a survey of these algorithms in a common unifying framework.
-We then investigate the benefit of introducing an additional rejuvenation step  which allows to sample at each time instant new  regime conditional to the forward and backward particles approximating the regimes.
-To support our claims, a limited Monte Carlo study is presented.
-These procedures are applied to commodity markets which are described using a  factor model based on the spot price and a convenience yield for crude oil data.
-
-%In this paper, commodity markets are described using a two factor model based on the spot price and a convenience yield representing the benefit or cost required for holding the physical commodity.
-It is assumed that the market structure has different regimes which depend on existing stocks and contracts or other externalities.
-These different regimes are modeled using a finite state space unobserved Markov chain, each regime corresponding to an  equilibrium behavior of the spot price and the convenience yield.
-At each time step, it is assumed that noisy observations of future contracts are available which leads to represent commodity markets dynamics using a conditionally linear and Gaussian state space model.
-We propose a new algorithm to approximate the conditional distributions of the states (spot, convenience yield and regime) given a set of observations.
-These approximations are obtained combining Sequential Monte Carlo methods and Kalman smoothing techniques  and may be used to detect significant regime switching times and to estimate model parameters with an Expectation Maximization based algorithm.
-The performance of the proposed procedure is assessed with Chicago Mercantile Exchange crude oil data.
+We then investigate the benefit of introducing an additional rejuvenation step  which allows to sample at each time instant new  regimes conditional on the forward and backward particles.
+To support our claims, a Monte Carlo study is presented.
+These procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield for crude oil data.
 \end{abstract}
@@
 \section{Introduction} \label{sec:intro}
-%Modeling commodity prices is a crucial step to valuate contingent claims related to energy markets and to optimize storage or extraction strategies.
-In \cite{gibson:schwartz:1990,schwartz:1997}, the authors proposed a model where the spot price of a commodity$(S_t, t \geq 0)$ depends on a second factor $(\delta_t, t \geq 0)$, referred to as the instantaneous convenience yield.
-This factor plays the role of dividends in equity markets and models the benefit of holding the physical commodity or the storage and maintenance costs required to keep the commodity.
-In this model, this convenience yield is described as an Ornstein-Uhlenbeck process:
-%\begin{align*}
-%\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma S_{t} \rmd W_{t}^{1} \eqsp,\\
-%\rmd \delta_{t} & = \kappa(\alpha-\delta_{t})\rmd t+\eta \rmd W_{t}^{2} \eqsp, \quad \rmd \langle W_t^1, W_t^2\rangle = \rho \rmd t\eqsp,
-%\end{align*}
-%where the parameter $\left(r, \sigma, \kappa, \alpha, \eta, \rho \right)$ are constant and $( (W_t^1,W_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$.
-A similar model may be written under the risk-neutral probability $\mathbb{Q}$ with an obvious transformation of the parameters.
-This convenience yield based model has been extended in \cite{casassus:colin:dufresne:2005,liu:tang:2011}.
-In this model, the convenience yield is meant to revert to its equilibrium value $\alpha$ at speed $\kappa$.
-%
-%
-%This model appear to be too restrictive as energy markets are not likely to revert to a single equilibrium value.
-This assumption is relaxed using Markov switching models to allow several possible regimes for the spot price and the convenience yield.
-The regime switching autoregressive model introduced in \cite{hamilton:1988,hamilton:1989} for the term structure of interest rates has been generalized by \cite{kim:1994} to general state-space models.
-This Markov switching model also extends the dynamic linear model of \cite{shumway:stoffer:1991} by using a time dependent switching process and introducing regime changes in the observation model.
-Regime switching models based on an unobserved Markov chain have been widely used to represent commodity prices, see for instance \cite{fong:see:2002} for crude oil data or \cite{mount:ning:cai:2006} for electricity prices.
-See also \cite{naifar:dohaiman:2013} for a two-regime model (crisis regime and non-crisis regime) to obtain the evidence of a significant correlation between crude oil prices and short-term interest rate during the financial crisis.
-In \cite{chiarella:clewlow:kang:2009}, a two-factor model for the gas forward price is introduced to model where the parameters of the volatility function may change depending on the market structure.
-In order to capture gas price dynamics, \cite{chen:forsyth:2010} proposed a single factor model for the crude oil spot price using a mean reverting process with two equilibrium levels.
-Following \cite{almansour:2016}, the spot price and convenience yield are described in this paper as:
-%\begin{align*}
-%\rmd S_{t} & = (r-\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd W_{t}^{1} \eqsp,\\
-%\rmd\delta_{t} & = \kappa_{a_t}(\alpha_{a_t} -\delta_{t})\rmd t+\eta_{a_t} \rmd W_{t}^{2} \eqsp, \quad \rmd\langle W_t^1, W_t^2\rangle = \rho_{a_t} \rmd t\eqsp,
-%\end{align*}
-%where $(a_t)_{t\ge 0}$ is a finite state space Markov process.
-%
-%This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield.
-A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).
-
-%Assuming that the switching rate between regimes is small compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_k= (\ln S_k, \delta_k)$ (the sampling period is taken to be 1) is modeled as a
 State space models are bivariate stochastic processes $\{(Y_k,Z_k)\}_{k\ge 1}$ where the state sequence $\{Z_k\}_{k\ge 1}$ is a Markov chain which is only partially observed through the sequence $\{Y_k\}_{k\ge 1}$.
@@
 
-When the state and observation models are linear and Gaussian, the filtering can be solved explicitly using the Kalman filter \cite{kalman:1960}.
+When the state and observation models are linear and Gaussian, filtering can be solved explicitly using the Kalman filter \cite{kalman:1960}.
 Exact solutions of the fixed-horizon smoothing problem can be obtained using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier "two-filter" smoother \cite{bryson:frazier:1963}.
-%Many extensions of these algorithms based on local linearization methods have been proposed for nonlinear dynamical systems such as extended Kalman filters, iterated Kalman filters, Gaussian sum filters or Se...
 This paper focuses on  Conditionally Linear and Gaussian Models (CLGM) given for $k\ge 1$ by:
@@
 \item $(\varepsilon_k)_{k\ge 1}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance.
-\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$, called \emph{regimes}  %$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
+\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$, called \emph{regimes}.
+%$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
 \item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
-\item \textcolor{red}{j'aurais mis $Z_0$} $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_k)_{k\ge 1}$.
+\item $Z_0$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_k)_{k\ge 1}$.
 \end{enumerate}
-\textcolor{red}{après on part de $k=1$}
-At each time step $k\ge 0$, the observation $Y_k$ is given by:
+%\textcolor{red}{après on part de $k=1$}
+At each time step $k\ge 1$, the observation $Y_k$ is given by:
 \begin{equation}
@@
 The first fully Rao Blackwellised SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
-This two-filter approach combines a forward and a backward filters which are approximated numerically using SMC for the regime sequence and Kalman filtering techniques for the hidden linear states.
+This two-filter approach combines a forward filter with a backward information filter which are approximated numerically using SMC for the regime sequence and Kalman filtering techniques for the hidden linear states.
 
 More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
-The update of the smoothing distribution of the regime given the observation shares some striking similarities with the Rauch-Tung-Striebel smoothing procedure, which is at the heart of the FFBS procedure.
+The update of the smoothing distribution of the regime given the observations shares some striking similarities with the Rauch-Tung-Striebel smoothing procedure, which is at the heart of the FFBS procedure.
 The Rao-Blackwellisation requires to update backward in time the smoothing distribution of the states given the regimes and the observations, which is achieved by using a \emph{à la Kalman} backward update.
@@
 
-In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao-Blackwellisation steps allowing the sample "new" particles in the backward pass.
-This approach may be seen as an extension of the ideas of \cite{fearnhead:clifford:2003} for Rao-Blackwellized smoother.
+In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao-Blackwellisation steps allowing to sample "new" particles in the backward pass.
+This approach may be seen as an extension of the ideas of \cite{fearnhead:clifford:2003} for Rao-Blackwellized smoothers.
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
@@
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
-In Section~\ref{sec:crudeoil}, an application to commodity markets is presented; the performance of our procedure is illustrated on crude oil  data.
+In Section~\ref{sec:crudeoil}, an application to commodity markets is presented; the performance of our procedure is illustrated with crude oil  data.
 A detailed derivation of the algorithms is provided in the Appendix.
@@
 Each ancestral path has $J$ offsprings at time $i$, it is thus necessary to choose a given number of trajectories at time $i$ (for instance $N$) among the $NJ$ possible paths.
-Following \cite{fearnhead:clifford:2003,barembruch:garivier:moulines:2008}, this may be improved by considering all the offsprings of all ancestral trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$.
-Each ancestral path has $J$ offsprings at time $i$, it is thus necessary to choose a given number of trajectories at time $i$ (for instance $N$) among the $NJ$ possible paths.
 To obtain the weight associated with each offspring write the following approximation of  $p(a_{1:i}|y_{1:i})$ based on the weighted samples at time $i-1$:
@@
 \[
-p^N(a_{1:i}|a_{i+1:n},y_{1:n})\propto\sum_{k=1}^N \omega_{i|i-1}^k(a_i)Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{\eta^k_{i|i-1}(a_{i:n})/2\right\}\delta_{a^k_{1:i-1}}(a_{1:i-1})
+p^N(a_{1:i}|a_{i+1:n},y_{1:n})\propto\sum_{k=1}^N \omega_{i|i-1}^k(a_i)Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{\eta^k_{i|i-1}(a_{i:n})/2\right\}\delta_{a^k_{1:i-1}}(a_{1:i-1}).
 \]
@@
 \label{sec:numerical:experiments}
-This section highlights the performance improvement brought by  the additional Rao-Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$.
+This section highlights the improvements brought by  the additional Rao-Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$.
 In each setting the transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data;  see Section~\ref{sec:exp}.
@@
 The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles.
-Figure~\ref{fig:ffbs:posterior} illustrates that the algorithm with rejuvenation slightly improves the original algorithm in this simple case.
-\begin{figure}
-\centering
-\includegraphics[scale=.3]{FFBSplot}
-\caption{Estimated posterior probabilities by the two FFBS algorithms.}
-\label{fig:ffbs:posterior}
-\end{figure}
 Figure~\ref{fig:ffbs:err} displays the mean estimation error over $100$ independent Monte Carlo runs.
 The estimation error is defined as the difference between the the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
-\begin{figure}
-\centering
-\includegraphics[scale=.3]{err_est_ffbs}
-\caption{Posterior probabilities estimation error fo the two FFBS algorithms.}
-\label{fig:ffbs:err}
-\end{figure}
-%In addition, Table~\ref{fig:ffbs:RMSE} displays the RMSE for the estimation of the regimes as a function of time with $100$ independent batches of data and shows that the additional rejuvenation step improves the performance of the algorithm (at time $i$, the estimated label is set to $1$ if the estimation of $\mathbb{P}(a_i|Y_{1:n})$ is greater than $0.5$ and to $2$ otherwise).
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.3]{err_est_ffbs}
+%\caption{Posterior probabilities estimation error for the two FFBS algorithms.}
+%\label{fig:ffbs:err}
+%\end{figure}
+%In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for the two algorithms to illustrate that both estimators have similar variability. 
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.3]{var}
+%\caption{Empirical variances of the estimation for the two FFBS algorithms.}
+%\label{fig:ffbs:var}
+%\end{figure}
 
@@
 In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$.
-%and $B$ becomes:
-%\[
-%B = \begin{pmatrix}1 & 0 & 0.5 & 0 \\ 0 & 1 & 0.5 & 0\end{pmatrix}\eqsp.
-%\]
-%It becomes a situation more difficult because of the noisy level.
-The smoothing distributions are estimated with $N=15$ particles. Both methods are compared to the approximations provided by RBTF with $1000$ particles.
-%Figure~\ref{fig:exp2:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.
- %Figure~\ref{fig:exp1:label} and %\ref{fig:exp2:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.
-
-\medskip
-
-%Figure~\ref{fig:exp:artificial} displays the estimated posterior distributions of the labels $(a_k)_{1\le k\le 500}$ given by RBTF for the first $500$ time steps with the  artificial distributions:
-
-%In experiment~1, the three artificial laws behave similarly except that the first choice is unable to detect the last regime switching for more than 50 time steps.
-%In experiment 2, the second method is completely unstable which indicates that the third artificial leads to the best performance.
-Therefore, the third artificial law is used in the following numerical experiments.
-
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.5]{Artificial_Compare_Exp1_1_500.eps}
-%\includegraphics[scale=.5]{Artificial_Compare_Exp2_1_500.eps}
-%\caption{estimated probabilities by RBTF in experiment 1 [top] and experiment 2 [bottom] with the two artificial laws: yellow (the first one) and red (the third one).}
-%\label{fig:exp:artificial}
-%\end{figure}
+The smoothing distributions are estimated with $N=15$ particles.
+Both methods are compared to the approximations provided by RBTF with $1000$ particles.
 The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.
@@
 In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
-
-\begin{figure}
-\centering
-\includegraphics[scale=.5]{N10_1_500}
-\includegraphics[scale=.5]{N10_501_1000}
-\includegraphics[scale=.5]{N10_1001_1500}
-\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
-\label{fig:exp1:label}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.5]{N15_1_500}
-\includegraphics[scale=.5]{N15_501_1000}
-\includegraphics[scale=.5]{N15_1001_1500}
-\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
-\label{fig:exp2:label}
-\end{figure}
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.5]{N10_1_500}
+%\includegraphics[scale=.5]{N10_501_1000}
+%\includegraphics[scale=.5]{N10_1001_1500}
+%\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
+%\label{fig:exp1:label}
+%\end{figure}
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.5]{N15_1_500}
+%\includegraphics[scale=.5]{N15_501_1000}
+%\includegraphics[scale=.5]{N15_1001_1500}
+%\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
+%\label{fig:exp2:label}
+%\end{figure}
 
@@
 
-%Statistical estimation is performed using market observations and cannot be done directly under the risk-neutral probability but under the historical probability $\mathbb{P}$.
-It is assumed that each factor (spot, convenience yield) has its own constant market price of risk, $\lambda^S$ for $S_t$ and $\lambda^{\delta}$ for $\delta_t$.
-The historical dynamics of the Gibson-Schwartz model is then described as follows:
-%\begin{align*}
-%\rmd S_{t} & = (\mu -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t} \rmd \widetilde{W}_{t}^{1} \eqsp,\\
-%\rmd \delta_{t} & = \kappa_{a_t}(\widetilde{\alpha}_{a_t}-\delta_{t})\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd \langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho \rmd t\eqsp,
-%\end{align*}
-%where $( (\widetilde{W}_t^1,\widetilde{W}_t^2), t \geq 0)$ are standard Brownian motions under the historical probability $\mathbb{P}$ with  correlation $\rho_{a_t}$, $\mu = r - \lambda^S$ and $\widetilde{\alpha}_{a_t} = \alpha_{a_t} - \lambda^{\delta}/\kappa_{a_t} $.
 This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield.
 A two-regime model is already sufficient to produce stylized effects such as contango (increase of future prices) and backwardation (decrease of future prices).
-Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a Conditionally Linear Gaussian Model.
+Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a CLGM.
 The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:
@@
 \[
-\theta \eqdef \{\pi, Q, \mu_1, \Sigma_1, \kappa, (\alpha_j)_{1\leq j\leq J}, (\sigma_j)_{1\leq j\leq J}, (\eta_j)_{1\leq j\leq J}, (\rho_j)_{1\leq j\leq J}, \mu, (\lambda_j)_{1\leq j\leq J}, (g_{\ell})_{1\leq \ell\leq d} \}\eqsp .
+\theta \eqdef \{\pi, Q, \mu_1, \Sigma_1, \kappa, (\alpha_j)_{1\leq j\leq J}, (\sigma_j)_{1\leq j\leq J}, (\eta_j)_{1\leq j\leq J}, (\rho_j)_{1\leq j\leq J}, (g_{\ell})_{1\leq \ell\leq d} \}\eqsp .
 \]
@@
 	\centering
-	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|c|c|c|c|c||}
-		$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
+	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|c|c||}
+		$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
 		\hline
-		5.0 & 0.1 & -0.05 & 0.4 & 0.4 & 0.5 & 0.5 & 0.75 & 0.65 & 0.0 & 0.0 & 0.0 & 0.1 & 0.1 & 0.1 & 0.1
+		5.0 & 0.1 & -0.05 & 0.4 & 0.4 & 0.5 & 0.5 & 0.75 & 0.65 & 0.1 & 0.1 & 0.1 & 0.1
 	\end{tabular}
@@
 	\centering
-	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c||c|clc|c|c||}
-		Parameter&$\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
+	\begin{tabular}{||c|c|c|c|c|c|c|c|c||c|clc|c|c||}
 	\hline
-		Value&2.6378 & 0.0889 & -0.0281 & 0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
+		Parameter&  $\kappa$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$  \\
 		\hline
-		Std. Dev& 0.1999 & 0.004248 & 0.00149 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052   \\
+		Value & 2.6378 &0.3733 & 0.3485 & 0.5892 & 0.3814 & 0.8709 & 0.6761   \\
+		\hline
+		Std. Dev & 0.1999 & 0.005438 & 0.002884 & 0.0483 & 0.0349 & 0.0064 & 0.0052 \\  
+		\hline
 	\end{tabular}
-	\vspace{1.cm}
-	\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|||}
-		Parameter&$\mu$ & $\lambda_1$ & $\lambda_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
+
+	\vspace{.2cm}
+
+	\begin{tabular}{||c|c|c|c|c|c|c|c|c|c|c|||}
 	\hline
-		Value&0.1712 & 0.2422	& -0.1731 & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
+		Parameter & $\alpha_1$ & $\alpha_2$  & $g_1$ & $g_2$ & $g_3$ & $g_4$  & $Q(1,1)$ & $Q(2,2)$\\
 		\hline
-		Std. Dev& 0.004147 & 0.03243 & 0.01973 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4
+		Value & 0.0889 & -0.0281  & 2.3e-2 & 1.0e-4 & 3.0e-4 & 2.3e-2 & 0.9917 & 0.9880 \\
+		\hline
+		Std. Dev & 0.004248 & 0.00149 & 1.9e-4 & 2.6e-4 & 2.3e-4 & 2.1e-04 &  6.7e-4 & 9.6e-4\\
+		\hline
 	\end{tabular}
@@
 in contango).
-\begin{figure}
-\centering
-\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
-\caption{Log-price (red line) and slope of future curves (blue line)}
-\label{fig:LogPriceTermStructure:data}
-\end{figure}
-%The first regime (in backwardation) is identified by a higher mean-reversion of convenience yield $\alpha$ and a higher volatility in states variables ($\sigma$ and $\eta$).
-%And vice versa the second regime is characterized by a lower value of $\alpha, \sigma, \eta$.
-%\begin{verbatim}
-%TODO: Add reference: Explain why $rho$ must be high (at least positive).
-%TODO: Explain why the noises $g1$, $g4$ are greater than the intermediate contracts
-%\end{verbatim}
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
+%\caption{Log-price (red line) and slope of future curves (blue line)}
+%\label{fig:LogPriceTermStructure:data}
+%\end{figure}
 
@@
 
-%After convergence, the value obtained for $\kappa$ is slightly greater than the one obtained in \cite{almansour:2014} but we also obtain slightly lower values for $\eta_1$ and $\eta_2$.
-The result also shows that the future price of maturity $4W$ is such that $g_4$ is greater than all the other $g_i$'s.
-It may be explained by the fact that there is no observation of spot contract which is likely to impact mostly the front future contract.
-
 %\begin{figure}
-%\centering
-%\includegraphics[scale=.38]{alpha_M.png}\\
-%\includegraphics[scale=.38]{sigma_M.png}\\
-%\includegraphics[scale=.38]{eta_M.png}
-%\caption{Estimation of $\alpha$, $\sigma$ and $\eta$.}
+%	\centering
+%	\includegraphics[scale=.75,angle=90]{Posterior_Data}
+%	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
+%	\label{fig:posterior:data}
 %\end{figure}
-
-\begin{figure}
-	\centering
-	\includegraphics[scale=.75,angle=90]{Posterior_Data}
-	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
-	\label{fig:posterior:data}
-\end{figure}
-
-\begin{figure}
-	\centering
-	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
-	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
-	\label{fig:posterior:data:logprice}
-\end{figure}
-
-%\paragraph{Comparaison of estimated future prices with the observations}
-%It is interesting to assess the accuracy of the estimated parameters obtained with the EM algorithm.
-These estimated parameters are used to estimate the future prices and compare them with the observed ones for each maturity.
-The results of this estimation procedure is displayed in Figure~\ref{fig:density:4W3M} and Figure~\ref{fig:density:6M1Y}.
-
-%\begin{figure*}[htb]
+%
+%\begin{figure}
 %	\centering
-%		\begin{tabular}{||c|c|c|c|c||}
-%		Residuals & $F_{4W}$ & $F_{3M}$ & $F_{6M}$ & $F_{1Y}$\\
-%		\hline
-%		mean & -0.002654812 & 0.001178231 & 0.002130222 & 0.002391526 \\
-%		std & 0.01461169 & 0.00575261 & 0.01039791 & 0.01796323
-%		\end{tabular}
-%	\caption{Table of differences between estimated and observed future prices.}
-%	\label{tab:tableTest}
-%\end{figure*}
-%Mean value of the residuals between the estimated and observed log-future prices are close to $0$ with a small variance.
-The standard deviation of futures $4W$ has an important value in comparison to futures $3M$ or $6M$ which goes along with the remark for the estimation of $G$ in the equation \eqref{eq:model:obs} by the EM algorithm.
-
-%\FloatBarrier
-
-%Our algorithm is calibrated using future prices of maturities $4W, 3M, 6M$ and $1Y$ but may be used to estimate future prices of maturities $8W, 9M, 2Y, 3Y$.
-These estimations may be compared to the observations given in the market.
-Densities of the estimated prices for these others maturities $8W, 9M, 2Y, 3Y$ are shown in Figure~\ref{fig:densothers}.
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.4]{Density_Estim_Obs_OUTSIDE.png}
-%\caption{Densities of estimated and observed log-future prices $8W, 9M, 2Y, 3Y$.}
-%\label{fig:densothers}
+%	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
+%	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
+%	\label{fig:posterior:data:logprice}
 %\end{figure}
-%At first glance, mean values and standard deviations of the difference between observed and estimated future prices are greater than in the previous case since these maturities are not used to calibrate the model.
-Nevertheless, these errors remain close to $0$.
-In addition, standard deviations of futures $2Y$ and $3Y$ are greater than the ones of $8W$ and $9M$, it is reasonable since $2Y$ and $3Y$ futures are outside of the interval of maturities used to calibrate the model.
-%\FloatBarrier
+
+\section{Conclusions}
+In this paper, we proposed a survey of Rao Blackwellised Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces.
+These methods combine particle based approximations to sample the sequence of hidden regimes while the hidden linear states are explicitly integrated conditional on the sequence of regimes and observations.
+They  are based on the forward backward decomposition of the smoothing distributions using either two-filter based algorithms or forward filtering backward smoothing techniques mimicking the Rauch-Tung-Striebel smoother.
+The paper presents these algorithms in a common unifying framework and also provides a survey of different techniques that could be used in the forward filtering pass which may improve significantly the usual mixture Kalman filter.
+The filtering distributions are approximated at  each time step by considering all possible offsprings of all ancestral trajectories before discarding degenerated paths instead of resampling the ancestral paths before propagating them at the next time step.
+We also investigated the benefit of additional Rao Blackwellisation steps which allow to sample new regimes at each time step conditional on the forward and backward particles.
+This rejuvenation step uses explicit integration of the hidden linear states before merging the forward and backward filters for two-filter based algorithms or before sampling new states backward in time for FFBS based methods.
+The paper displays some Monte Carlo experiments with simulated data to illustrate that this additional rejuvenation step improves the performance of the smoothing algorithms with no additional computational costs or variance.
+They are also applied to commodity markets using WTI crude oil data.
 
@@
 
-\section{Technical lemmas}%~\ref{eq:technical:twofilter}-\ref{lem:integral:gammap}
+\section{Technical lemmas}
 \begin{lemma}
@@
 
-%\clearpage
-%\newpage
-%
-%\section{Algorithms}
-%The forward filter is described in Algorithm~\ref{algo:forward}.
-The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^j$ may be written:
-%\begin{align*}
-%\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
-%& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
-%\end{align*}
-%\begin{algorithm}
-%\DontPrintSemicolon
-%\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
-%\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
-%{\bf Initialization}\;
-%\For{$k \gets 1$ \textbf{to} $N$} {
-%Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
-%Compute $\mu_1^k$ and $P_1^k$:
-%\begin{align*}
-%K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
-%\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
-%P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
-%\end{align*}
-%Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
-%}
-%{\bf Iterations}\;
-%\For{$i \gets 2$ \textbf{to} $n$} {
-%Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
-%\For{$k \gets 1$ \textbf{to} $N$} {
-%For all $j \in \{1,\ldots, J\}$ compute
-%\[
-%\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
-%\]
-%Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
-%Update $\mu_i^k$ and $P_i^k$:
-%\begin{align*}
-%\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
-%P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
-%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
-%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
-%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
-%\end{align*}
-%Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
-%}
-%}
-%\caption{Forward filter, see \cite{chen:liu:2000}}
-%\label{algo:forward}
-%\end{algorithm}
-%
-%\begin{algorithm}
-%\DontPrintSemicolon
-%\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
-%\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
-%{\bf Initialization}\;
-%Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
-%{\bf Iterations}\;
-%\For{$i \gets n-1$ \textbf{to} $1$} {
-%\For{$k \gets 1$ \textbf{to} $N$} {
-%Compute
-%\[
-%\omega^k_{i|n} \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})\eqsp.
-%\]
-%}
-%Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$.
-%}
-%\caption{Smoothing algorithm based on the approximation of \cite{kim:1994}}
-%\label{algo:kim}
-%\end{algorithm}
-%
-%
-%\begin{algorithm}
-%\DontPrintSemicolon
-%\KwIn{Output of a forward filter: trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$, .}
-%\KwOut{Trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$.}
-%{\bf Initialization}\;
-%Set $\tilde{a}_n = a_n^k$ with probability $\omega_n^k$.\\
-%Compute $\hat{\Omega}_n$ and $\hat{\lambda}_n$:
-%\[
-%\hat{\Omega}_n = B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}B_{\tilde{a}_n}\quad\mbox{and}\quad \hat{\lambda}_n =B'_{\tilde{a}_n}\barG^{-1}_{\tilde{a}_n}(y_n-c_{\tilde{a}_n})\eqsp.
-%\]
-%{\bf Iterations}\;
-%\For{$i \gets n-1$ \textbf{to} $1$} {
-%Compute
-%\[
-%m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{\tilde{a}_{i+1}}\quad\mbox{and}\quad M_{i+1} = H_{\tilde{a}_{i+1}}'\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}} + I\eqsp.
-%\]
-%Set
-%\begin{align*}
-%\Omega_i &= T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})\hat{\Omega}_{i+1}T_{\tilde{a}_{i+1}}\eqsp,\\
-%\lambda_i &=T'_{\tilde{a}_{i+1}}(I-\hat{\Omega}_{i+1}H_{\tilde{a}_{i+1}}M^{-1}_{i+1}H'_{\tilde{a}_{i+1}})m_{i+1}\eqsp.
-%\end{align*}
-%\For{$k \gets 1$ \textbf{to} $N$} {
-%Compute
-%\[
-%\Lambda^k_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I \quad\mbox{and}\quad \eta^k_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}\eqsp.
-%\]
-%Set
-%\[
-%\tilde{\omega}_{i|n}^k \propto \omega_i^kQ(a_i^k,\tilde{a}_{i+1})|\Lambda^k_i|^{-1/2}\exp\left\{\eta^k_i/2\right\}\eqsp.
-%\]
-%}
-%Set $\tilde{a}_i = a_i^k$ with probability $\omega_{i|n}^k$ and
-%\[
-%\hat{\Omega}_i =  \Omega_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}B_{\tilde{a}_i} \quad\mbox{and}\quad \hat{\lambda}_i = \lambda_i + B'_{\tilde{a}_i}\barG^{-1}_{\tilde{a}_i}(y_i-c_{\tilde{a}_i})\eqsp.
-%\]
-%}
-%\caption{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
-%\label{algo:FFBS}
-%\end{algorithm}
-%
-%
-%\begin{algorithm}[b]
-%\DontPrintSemicolon
-%\KwIn{Artificial densities $(\gamma^{}_i)_{1\le i\le n}$ and proposal densities $(\tilde{q}_i)_{1\le i \le n}$.}
-%\KwOut{A set of weighted trajectories $(\tilde{a}^j_{i:n})_{1\le j\le N}$ with importance weights $(\tilde{\omega}^j_i)_{1\le j\le N}$, $1\le i \le n$.}
-%{\bf Initialization}\;
-%For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_n\sim\tilde{q}_n(\cdot)$.\;
-%For $1\le \ell \le N$, set:
-%\[
-%\tilde{\omega}^{\ell}_n \propto \frac{\int\gamma^{}_n(\tilde{a}^{\ell}_n,z')g_{}(\tilde{a}^{\ell}_n,z';y_n)\rmd z'}{\tilde{q}_n(\tilde{a}^{\ell}_n)}\eqsp.
-%\]
-%{\bf Iterations}\;
-%\For{$i \gets n-1$ \textbf{to} $1$} {
-%Resample the set $(\tilde{a}^{\ell}_{i+1:n})_{1\le j\le N}$ using the normalized weights $(\tilde{\omega}^{\ell}_{i+1})_{1\le j \le N}$.\;
-%For $1\le \ell \le N$, sample $\tilde{a}^{\ell}_{i}\sim\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot)$.\;
-%For $1\le \ell \le N$, set:
-%\[
-%\tilde{\omega}^{\ell}_i \propto \frac{q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
-%\]
-%}
-%\caption{Backward information filter, see \cite[Section~4.1.1]{briers:doucet:maskell:2010}}
-%\label{algo:backward}
-%\end{algorithm}
-
 \bibliographystyle{alpha}

### 2016-11-23 TexArxiv/clgm.tex
Final modifications before submission

@@
 \item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$, called \emph{regimes}.  
-%$Q$ denotes the transition matrix of this Markov chain and by $\pi$ its initial distribution.
 \item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
@@
 \end{enumerate}
-%\textcolor{red}{après on part de $k=1$}
 At each time step $k\ge 1$, the observation $Y_k$ is given by:
@@
 This is a specific instance of Rao-Blackwellised Monte Carlo filters, often referred to as the \emph{Mixture Kalman Filter}. 
-%These SMC methods based on sequential importance sampling and resampling techniques introduced in \cite{handschin:mayne:1969} and \cite{rubin:1987} allow to approximate any sequence of conditional distributions of the hidden states given the observations, see \cite{kitagawa:1987,gordon:salmond:smith:1993}.
 Improvements of these early filtering techniques  have been introduced in \cite{doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005}.
-%These algorithms sample only trajectories of indicator variables associated with importance weights, the corresponding state variables $(Z_1,\ldots,Z_k)$ are marginalized and estimated with a Kalman filter, which is the optimal filter in this case.
-%However, many maximum likelihood based estimation procedures such as  the Expectation Maximization  (EM)  algorithm introduced in \cite{dempster:laird:rubin:1977} require to approximate some marginal smoothing distributions of the states and indicator variables.
-%Several algorithms have been developed to solve the smoothing problem using SMC methods: for instance the Forward Filtering Backward Smoothing algorithm (FFBS), the Forward Filtering Backward Simulation algorithm (FFBSi) and the PArticle-based, Rapid Incremental Smoother (PaRIS)  developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000},  \cite{godsill:doucet:west:2004} and \cite{olsson:westerborn:2015}.
-%,lindsten:schon:2011
-%In the forward pass of these algorithms, a Rao Blackwellised particle filter such as \cite{chen:liu:2000} is used to approximate the filtering distributions and the corresponding particles, weights and associated trajectories of means and variances of the Kalman filters are stored.
 
@@
 The Rao-Blackwellisation requires to update backward in time the smoothing distribution of the states given the regimes and the observations, which is achieved by using a \emph{à la Kalman} backward update.
-%The backward pass samples trajectories of indicator variables backward in time using the particles and weights of the forward pass.
 
@@
 \]
-%Maximum likelihood estimation of the parameters using the Expectation Maximization algorithm or gradient ascent algorithms requires to compute the posterior distributions of the states $a_{i-1:i},z_{i-1:i}$  given the observations $Y_{1:n}$, written $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$, which cannot be computed in closed-form (the dependency of $\phi^n_{i}$ on $Y_{1:n}$ is dropped for better clarity).
-SMC methods provide a class of importance sampling and resampling procedures to approximate $\phi^n_{i}(a_{i-1:i},z_{i-1:i})$.
 All the algorithms considered in this paper are based on forward-backward or two-filter decompositions of the smoothing distributions and share the same forward filter presented in Section~\ref{sec:forward}.
@@
 
-%This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$.
 \noindent\textbf{Iterations}\\
-%For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
-The ancestral path of indicator variables at time $i-1$ is defined by:
-%\begin{equation}
-%\label{eq:ancestral:traj}
-%a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
-%\end{equation}
-%where the ancestral variables are defined as follows:
-%\[
-%\mathsf{A}^k_{i} = k \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
-%\]
-%Similarly, means and variances trajectories are defined by:
-%\[
-%\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
-%\]
-%The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
 Several procedures may be used to extend the trajectories $(a^k_{1:i-1})_{1\le k \le N}$ at time $i$.
@@
 \]
-The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ so that:% which means that $\gamma_i^{j,k}$ may be written:
+The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ so that:
 \[
@@
 Then, in both cases, all particles such that $\tilde{\Omega}^{j,k}_{i} = 0$ are discarded and for all the other trajectories defined as an ancestral path $(a^k_{1:i-1})$ extended by $a^k_i = j$, the new corresponding weight $\omega$ in \eqref{eq:forward:SMC} is given by the normalized weight $\tilde{\Omega}^{j,k}_{i}$.
-
-%$\mu_i^k$ and $P_i^k$ are updated according to:
-%\begin{align*}
-%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
-%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
-%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
-%\end{align*}
-%The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
-
-%\subsection{Path-space smoother of \cite{kitagawa:1996}}
-%As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step.
-In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
-%\[
-%p^{\mathsf{Kit}}_N(a_{1:n}|y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
-%\]
-%In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
-%\[
-%p^{\mathsf{Kit}}_N(a_{i}|y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
-%\]
-%This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps.
-See Section~\ref{sec:numerical:experiments} for a numerical illustration.
 
@@
 \]
-%\[
-%p^N_{}(a_{1:i-1},z_{i-1}|y_{1:i-1}) = \sum_{k=1}^N\delta_{a^k_{1:i-1}}(a_{1:i-1})\,\omega^k_{i-1}\,p_{}(z_{i-1}|a^k_{1:i-1},y_{1:i-1})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp.
-%\]
 On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the regime of the filtering pass $\{a_i^k\}_{1\le k \le N}$.
@@
 \label{sec:two-filer}
-%Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}.
 Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  regimes given all the observations).
@@
 Then, the forward and backward weighted samples are combined using \eqref{eq:dens:bwd_fwd_doucet} to approximate $p(a_{i},z_{i}|y_{1:n})$.
-%Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:
-%Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods.
 \textbf{Additional particle rejuvenation step}\\
@@
 \label{eq:dens:bwd_fwd}
-%\phi^n_{i}(a_{i-1:i},z_{i-1:i})  =  \sum_{a_{i-2}}\sum_{a_{i+1}} \int_{z_{i-2}} \int_{z_{i+1}} \psi^n_{i}(a_{i-2:i+1},z_{i-2:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp.
 p(a_{i},z_{i}|y_{1:n})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp,
@@
 Define
-%\begin{align*}
-%\Upsilon^k_{i-1}(a_{i}) &= (P^k_{i-1})^{-1} + T'_{a_{i}}\barH_{a_{i}}^{-1}T_{a_{i}}\eqsp,\\
-%\upsilon^k_{i-1}(a_{i})&= \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})(P^k_{i-1})^{-1}\mu^k_{i-1}\eqsp,\\
-%\Omega^k_{i-1}(a_{i}) &= \left(\barH_{a_{i}}^{-1} - \barH_{a_{i}}^{-1}T_{a_{i}}(\Upsilon^k_{i-1})^{-1}(a_{i})T'_{a_{i}}\barH_{a_{i}}^{-1}\right)^{-1}\eqsp,\\
-%r_{i-1}^k(a_{i}) &= \upsilon^k_{i-1}(a_{i})+(\Omega^k_{i-1}(a_{i}))^{-1}d_{a_{i}}\eqsp,\\
-%\omega_{\mathsf{f},i}^k(a_{i}) &= \omega^k_{i-1}Q(a^k_{i-1},a_{i})\left|P^k_{i-1}\right|^{-1/2}\left|\Upsilon^k_{i-1}(a_{i})\right|^{-1/2} \left|2\pi\barH_{a_{i}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{d_{a_{i}}}^2\right\}\\
-%&\hspace{1.5cm}\times\exp\left\{ -\frac{1}{2}\normMat{P_{i-1}^k}{\mu^k_{i-1}}^2+\frac{1}{2}\normMat{P_{i-1}^k\Upsilon^k_{i-1}(a_{i})P_{i-1}^k}{\mu^k_{i-1}}^2 - (\upsilon^k_{i-1})'(a_{i})d_{a_{i}} \right\}\eqsp.
-%\end{align*}
 \begin{align*}
@@
 r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
-%\times\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-2}(a_{i-1})}{z_{i-1}-d_{a_{i-1}}}^2\right\}\eqsp.%q(a^k_{i-2},a_{i-1})\omega^k_{i-2}\left|P^k_{i-2}\right|^{-1/2}\left|\Upsilon^k_{i-2}(a_{i-1})\right|^{-1/2}\exp\left\{ -\frac{1}{2}\normMat{P_{i-2}^k}{\mu^k_{i-2}}^2\right\}\\
-%\times \left|2\pi\barH_{a_{i-1}}\right|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{\barH_{a_{i-1}}}{z_{i-1}-d_{a_{i-1}}}^2\right\}\exp\left\{\frac{1}{2}(\upsilon^k_{i-2})'(a_{i-1})(\Upsilon^k_{i-2})^{-1}(a_{i-1})\upsilon^k_{i-2}(a_{i-1})\right\}\eqsp.
 \end{equation}
@@
 As $t^N_{i}(a_i,z_i,y_{i+1:n})$ is not normalized, the normalization of the SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
-%These combination and normalization steps are given in Appendix~\ref{sec:approx:smooth}.
 Our procedure allows to construct sequence of regimes with non-degenerated importance weights in the combination step.
@@
 
-%\subsection{Backward filter}
-%\label{sec:backward}
-%This section displays a SMC algorithm to approximate $t_{i}(a_i,z_i,y_{i:n})$ using backward recursions.
 As the function $(a_{i+1},z_{i+1})\mapsto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$ is not a probability density function, it is not possible to use SMC samples to approximate it. 
-%The backward filter relies on artificial densities to introduce a surrogate target density function which may be approximated using SMC methods.
 Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:
@@
 This choice differs from \cite{briers:doucet:maskell:2010} where it is advocated to set $\gamma^{}_i$ as the product of two independent densities $\gamma_i^{a}(a_i)$ and $\gamma_i^{z}(z_i)$.
-The accuracy of the algorithm relies heavily on a proper tuning of this artificial density.%: it is shown in Section~\ref{sec:exp} that fitting a mixture improves the performance of the algorithm.
+The accuracy of the algorithm relies heavily on a proper tuning of this artificial density.
 
@@
 The estimation error is defined as the difference between the the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.3]{err_est_ffbs}
-%\caption{Posterior probabilities estimation error for the two FFBS algorithms.}
-%\label{fig:ffbs:err}
-%\end{figure}
-%In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for the two algorithms to illustrate that both estimators have similar variability. 
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.3]{var}
-%\caption{Empirical variances of the estimation for the two FFBS algorithms.}
-%\label{fig:ffbs:var}
-%\end{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=.3]{err_est_ffbs}
+\caption{Posterior probabilities estimation error for the two FFBS algorithms.}
+\label{fig:ffbs:err}
+\end{figure}
+In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for the two algorithms to illustrate that both estimators have similar variability. 
+\begin{figure}
+\centering
+\includegraphics[scale=.3]{var}
+\caption{Empirical variances of the estimation for the two FFBS algorithms.}
+\label{fig:ffbs:var}
+\end{figure}
 
@@
 RBTF and the algorithm of \cite{briers:doucet:maskell:2010} are compared to the approximations provided by RBTF with $1000$ particles.
-%Figure~\ref{fig:exp1:label} display the estimated posterior distributions of the labels given by the two algorithms at different time instances.
 
@@
 In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.5]{N10_1_500}
-%\includegraphics[scale=.5]{N10_501_1000}
-%\includegraphics[scale=.5]{N10_1001_1500}
-%\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
-%\label{fig:exp1:label}
-%\end{figure}
-%
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.5]{N15_1_500}
-%\includegraphics[scale=.5]{N15_501_1000}
-%\includegraphics[scale=.5]{N15_1001_1500}
-%\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
-%\label{fig:exp2:label}
-%\end{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=.5]{N10_1_500}
+\includegraphics[scale=.5]{N10_501_1000}
+\includegraphics[scale=.5]{N10_1001_1500}
+\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
+\label{fig:exp1:label}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[scale=.5]{N15_1_500}
+\includegraphics[scale=.5]{N15_501_1000}
+\includegraphics[scale=.5]{N15_1001_1500}
+\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
+\label{fig:exp2:label}
+\end{figure}
 
@@
 \]
-%where (setting $\tau t_i = t_i-t_{i-1}=1$):
-%\[
-%d_{j} \eqdef
-%\begin{pmatrix} \left[r- \alpha_{j} - \sigma^2_{j}/2 \right]  + \alpha_{j}[1-\rme^{-\kappa}]/\kappa\\
-%\alpha_{j} [1-\rme^{-\kappa}] \end{pmatrix}\eqsp, \;\; T \eqdef
-%\begin{pmatrix} 1 & -[1-\rme^{-\kappa}]/\kappa \\ 0 & \rme^{-\kappa} \end{pmatrix} \eqsp,
-%\]
-%\begin{align*}
-%\barH_{j}(1,1) &= \sigma^2_{j} + \eta^2_{j}\left\{1 + (1-\rme^{-2\kappa})/(2\kappa) - 2(1-\rme^{-\kappa})/\kappa\right\}/\kappa^2
-%- 2\rho_{j} \eta_{j} \sigma_{j}\left\{ 1 - (1-\rme^{-\kappa})/\kappa \right\}/\kappa\eqsp, \\
-%\barH_{j}(1,2) & = \left(\rho_{j} \eta_{j} \sigma_{j}-\eta^2_{j}/\kappa\right)\left(1-\rme^{-\kappa} \right)/\kappa + \eta^2_{j}\left(1-\rme^{-2\kappa} \right)/(2\kappa^2)\eqsp,\\
-%\barH_{j}(2,1) &= \barH_{j}(1,2)\eqsp,\quad \barH_{j}(2,2) = \eta^2_{j}\left( 1-\rme^{-2\kappa} \right)/(2\kappa)\eqsp.
-%\end{align*}
 where (with $\barH_{a_{i-1}} \eqdef H_{a_{i-1}}H'_{a_{i-1}}$ and $\tau = t_i-t_{i-1}$):
@@
 \end{align*}
-%Remarks que the equations \eqref{eq:modele} under the pricing measure $\mathbb{Q}$ becomes:
-%\begin{equation}
-%\begin{cases}
-%\rmd S_{t} & = (r -\delta_{t})S_{t}\rmd t+\sigma_{a_t} S_{t}\rmd \widetilde{W}_{t}^{1} \eqsp,\\
-%\rmd\delta_{t} & = \left[\kappa(\alpha_{a_t} -\delta_{t}) - \lambda\right]\rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2} \eqsp, \quad \rmd\langle \widetilde{W}_t^1, \widetilde{W}_t^2\rangle = \rho_{a_t} \rmd t\eqsp.
-%\end{cases}
-%\end{equation}
-%where the constant $\lambda$ denotes the market price of convenience yield risk and $\widetilde{W}^1, \widetilde{W}^2$ are two $\mathbb{Q}-$Brownian motions.
-By introducing a new mean-reversion level for the convenience yeild process under $\mathbb{Q}$
-%\[
-%\widetilde{\alpha} = \alpha - \lambda/\kappa
-%\]
-%we obtain the dynamics:
-%\[
-%\rmd\delta_{t} = \kappa(\widetilde{\alpha}_{a_t} -\delta_{t}) \rmd t+\eta_{a_t} \rmd \widetilde{W}_{t}^{2}
-%\]
-%The behaviors of $Z_i$ under $\mathbb{Q}$ is $Z_i = \widetilde{d}_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i
-%$ with
-%\[
-%\widetilde{d}_{a_{i-1}} \eqdef
-%\begin{pmatrix} \left[r - \widetilde{\alpha}_{a_{i-1}} - \sigma^2_{a_{i-1}}/2 \right] \tau t_i + \widetilde{\alpha}_{a_{i-1}}[1-e^{-\kappa \tau t_i}]/\kappa\\
-%\widetilde{\alpha}_{a_{i-1}} [1-e^{-\kappa \tau t_i}] \end{pmatrix}\eqsp,
-%\]
-%The transition matrix $Q$ is assumed to be the same under the pricing measure.
 The observations are Wednesday future contracts of the West Texas Intermediate crude oil (WTI) traded in the Chicago Mercantile Exchange (CME) from $11$ January 1995 to $13$ November 2013.
@@
 At each time $t_i = i\tau$, with $\tau = 0.0192$, the observations of the $\dimy = 4$ future prices are $Y_i \eqdef (\ln(F^{(market)}_{i\tau t,m_1}), \ldots, \ln(F^{(market)}_{i\tau t,m_{\dimy}}))'$, where $F_{t_i,m}$ is the future price at $t_i$ for a maturity  $m$ weeks.
-%In addition, the backwardation is more likely to stay in backwardation than switch to contango and the same holds for the contango regime.
 A closed form solution for $F_{t_i,m}$  may be written:
@@
 \begin{equation*}
-%\label{eq:loglik}
 \ell_{\theta}^{n}(Y_{1:n}) \eqdef \log\left(\sum_{a_1=1}^J\ldots\sum_{a_n=1}^J\int p_{\theta}(a_{1:n},z_{1:n},Y_{1:n})\,\rmd z_{1:n}\right)\eqsp,
@@
 \begin{equation*}
-%\label{eq:completelik}
 p_{\theta}(a_{1:n},z_{1:n},Y_{1:n}) \eqdef \pi(a_1)\phi_{\mu_1,\Sigma_1}(z_1)g_{\theta}(a_1,z_1;Y_1)\prod^{n}_{i=2}Q(a_{i-1},a_i)m_{\theta}\left(a_{i},z_{i-1};z_i\right)g_{\theta}(a_i,,z_i;Y_i)\eqsp,
@@
 \end{enumerate}
-%\paragraph{E-step}
-%By \eqref{eq:completelik}, the intermediate quantity $Q(\theta,\theta_p)$ may be written (up to some additive constant which does not depend on $\theta$) $Q(\theta,\theta_p) = I^{n}_{1,\theta_p} + J^{n}_{1,\theta_p} + \sum_{i=2}^nI^{n}_{i,\theta_p} + \sum_{i=2}^nJ^{n}_{i,\theta_p}$,
-%where
-%\begin{align*}
-% I^{n}_{1,\theta_p} &\eqdef \mathbb{E}_{\theta_p}\left[\log\phi_{\mu_1,\Sigma_1}(Z_1)\middle|Y_{1:n}\right]\eqsp,\\
-% &= - \frac{1}{2}\log|\Sigma_1| -\frac{1}{2}\mathrm{Tr}\left(\Sigma_1^{-1}\mathbb{E}_{\theta_p}\left[Z_1Z'_1\middle|Y_{1:n}\right]\right) +\mu'_1\Sigma_1^{-1}\mathbb{E}_{\theta_p}\left[Z_1\middle|Y_{1:n}\right]- \frac{1}{2}\mu'_1\Sigma_1^{-1}\mu_1\eqsp,\\
-% J^{n}_{1,\theta_p} &\eqdef \mathbb{E}_{\theta_p}\left[\log g_{\theta}(a_1,z_1;Y_1)\middle|Y_{1:n}\right]\eqsp,\\
-% &= -\frac{1}{2}\log|\barG|-\frac{1}{2}\sum_{j=1}^J\mathrm{Tr}\left(\barG^{-1}\mathbb{E}_{\theta_p}\left[\left(Y_1-c_{j}-BZ_1\right)\left(Y_1-c_{j}-BZ_1\right)'\1_{a_1=j}\middle|Y_{1:n}\right]\right)
-%\end{align*}
-%and, for all $2\le i\le n$,
-%\begin{align*}
-% I^{n}_{i,\theta_p} &\eqdef\mathbb{E}_{\theta_p}\left[\log m_{\theta}\left(a_{i-1},z_{i-1};z_i\right)\middle|Y_{1:n}\right]\eqsp,\\
-% &= -\frac{1}{2}\sum_{j=1}^J\log |\barH_{j}|\mathbb{E}_{\theta_p}\left[\1_{a_{i-1}=j}\middle|Y_{1:n}\right]\\
-% &\hspace{.4cm}-\frac{1}{2}\sum_{j=1}^J \mathrm{Tr}\left(\barH^{-1}_j\mathbb{E}_{\theta_p}\left[\left(Z_i-d_{j}-TZ_{i-1}\right)\left(Z_i-d_{j}-TZ_{i-1}\right)'\1_{a_{i-1}=j}\middle|Y_{1:n}\right]\right)\eqsp.
-% \end{align*}
-% \begin{align*}
-% J^{n}_{i,\theta_p} &\eqdef\mathbb{E}_{\theta_p}\left[\log g_{\theta}(a_i,z_i;Y_i)\middle|Y_{1:n}\right]\eqsp,\\
-% &= -\frac{1}{2}\log|\barG|-\frac{1}{2}\sum_{j=1}^J\mathrm{Tr}\left(\barG^{-1}\mathbb{E}_{\theta_p}\left[\left(Y_i-c_{j}-BZ_i\right)\left(Y_i-c_{j}-BZ_i\right)'\1_{a_i=j}\middle|Y_{1:n}\right]\right)\eqsp.
-%\end{align*}
 All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our RBTF algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
-%\paragraph{M-step}
 As the function $\theta\mapsto Q^N(\theta,\theta_p)$ cannot be maximized analytically, the M-step is performed numerically using the Covariance Matrix Adaptation Evolution Strategy (CMA-ES) introduced in \cite{hansen:ostermeier:2001}.
@@
 
-%In details, among $510$ backwardation there are $485$ data whose regime staying in this regime, the same remarks on $417$ among $441$ contango doesn't change its regime.
-
-%The future expiration date usually expires on the third business day prior to the $25-th$ calendar day of the month preceeding the delivery month
-%\footnotetext{For WTI crude oil, trading in the current delivery month shall cease on the third business day prior to the $25$th calendar day of the month preceding the delivery month.
-If the $25$th calendar day of the month is a non-business day, trading shall cease on the third business dat prior to the last business day preceding the $25$th calendar day.
-See http://www.cmegroup.com for more details}
-
-%Contrary to \cite{almansour:2014}, we calculate the exact time to maturity for each day and each future contract.
-For example, on $19-05-2010$ the time to maturity of the front contract $F_1$ is only $1$ day (the earliest expiration date is $20-05-2010$) but it is considered with a time to maturity $1$ month in \cite{almansour:2014}.
-Our calculus takes into account exactly the real time to maturity.
-
 \subsection{Numerical results}
-%Following \cite{carmona:ludkovski:2004}, under the risk neutral probability $\mathbb{Q}$, we assume that the spot price $S_t$ of the Chicago Mercantile Exchange (CME) crude oil and its associated convenience yield $\delta_t$ are solutions to the SDE:
-%\begin{align*}
-%\rmd S_t &= \left(\tilde{\mu}_{a_{t}} - \delta_t\right) S_t \rmd t + \sigma_{a_t} S_t \rmd \tilde{W}^1_t \eqsp,\\
-%\rmd \delta_t &= \kappa \left( \tilde{\alpha}_{a_t} - \delta_t \right) \rmd t + \eta_{a_t} \rmd \tilde{W}^2_t \eqsp,
-%\end{align*}
-%where $\tilde{W}^1$ and $\tilde{W}^2$ are two correlated Brownian motions under the historical probability $\mathbb{P}$.
-%
-%In this case,  $\delta_t$ is an Orstein-Uhlenbeck process.
-%
-%Under the risk-neutral probability $\mathbb{Q}$, this dynamic becomes:
-%\begin{align*}
-%\rmd S_t &= \left(r - \delta_t\right)S_t \rmd t + \sigma_{a_t} S_t \rmd W^1_t \eqsp,\\
-%\rmd \delta_t &= \left[\kappa \left(\alpha_{a_t} - \delta_t \right) - \lambda_{a_t} \right] \rmd t + \eta_{a_t} \rmd  W^2_t \eqsp,
-%\end{align*}
-%Define $\alpha_{a_t} \eqdef \tilde{\alpha}_{a_t} - \lambda_{a_t}/\kappa$. Therefore, the dynamic of $Z_t = (\ln(S_t), \delta_t)$ under the risk-neutral mesure is:
-%\begin{align*}
-%\rmd \ln(S_t) &= \left(r - \delta_t- \sigma^2_{a_t}/2 \right) dt + \sigma_{a_t} \rmd W^1_t \eqsp,\\
-%\rmd \delta_t &= \kappa\left( \alpha_{a_t} - \delta_t\right)dt + \eta_{a_t} \rmd W^2_t \eqsp,
-%\end{align*}
-%where $(a_t)_{t\ge 0}$ is a Markov process taking values in  $\{1,\ldots,J\}$ indicating the market regime and $W^1$ and $W^2$ are Brownian motions under $\mathbb{Q}$ such that $\rmd<W^1, W^2>_t = \rho_{a_t}\rmd t$.
-The interest rate $r$ is constant and $\lambda_{a_t}$ is a risk associated with the convenience yield dynamic when the market regime is $a_t$.
-In the following, it is assumed that between two consecutive observations at times $t_{i-1}$ and $t_{i}$, the market regime is fixed so that the solutions to the previous SDE may be computed explicitly if $a_t$ is set to $a_{i-1} = a_{t_{i-1}}$.
-Therefore, it is assumed that the regime indicators are given at times $(t_i)_{i\ge 1}$ by a discrete time Markov chain $(a_{i})_{i\ge 1}$ and do not change between $t_{i-1}$ and $t_{i}$.
-%The explicit integration of \eqref{eq:modele} between $t_{i-1}$ and $t_i$ under the approximation that there is no regime change between these two successive time instants provides the following values of the parameter of the discrete time model:
-%for $Z_i \eqdef (\ln (S_{t_i}),\delta_{t_i})$:
-%\[
-%Z_i = d_{a_{i-1}} + T Z_{i-1} + H_{a_{i-1}} \varepsilon_i\eqsp,
-%\]
-%with $\varepsilon_i$ a standard $2$-dimensional Gaussian random variable and parameters defined as
-
-%For WTI crude oil, trading in the current delivery month shall cease on the third business day prior to the $25$th calendar day of the month preceding the delivery month.
-If the $25$th calendar day of the month is a non-business day, trading shall cease on the third business dat prior to the last business day preceding the $25$th calendar day.
-In the event that the official exchange holiday schedule changes subsequent to the listing of a crude oil futures, the originally listed expiration date shall remain in effect.
-In the event that the originally listed expiration day is declared a holiday, expiration will move to the business day immediately prior.
-For example, the $25$th Jan $2014$ is sunday, the business day immediately prior is $23$th Jan (friday) to which the $3$rd business day prior is $20th Jan$.
-Therefore the maturity dates of $Feb 2014$ Contract is $20th Jan 2014$.
-%For the CME WTI crude oil, the future prices are informed each trading day for each contract.
-Based on the expiration date, we may obtain an exact time to maturity for these futures.
-In our case, we are only interested in certain constant times to maturity: $4W$, $3M$, $6M$, $1Y$, $1,5Y$, $2Y$ and $3Y$.
-We chose the corresponding future contracts numbers: $1, 3, 7, 13, 19, 25$ and $37$ since this series is near the series $4W, 3M, 6M, 1Y, 1,5Y, 2Y, 3Y$.
-Futures prices for such fixed time to maturity may be estimated from the existing futures by smoothing splines or linear interpolation method.
-For our weekly (every Wednesday) future prices data, we used the linear interpolation from $1995$-$01$-$01$ to $2013$-$05$-$01$.
-The period $1995$-$01$-$01$ to $2013$-$05$-$01$ used for CME WTI Crude Oil future prices is the same as the one for the data used in \cite{almansour:2014} so that we may use the same weekly transition matrix $Q$: $Q(1,1) = 0.9807$ and $Q(2,2) = 0.9734$ (the authors used the New York Mercantile Exchange (NYMEX) WTI crude oil).
-The initials values for the EM algorithm are given below:
-
 The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi(1) = \pi(2) = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime.
@@
 \end{figure}\\
-The number of particles is set to $N = 100$, $\tau  = 1/52$. % and the backward filter is based on the  third artificial distribution introduced in Section~\ref{sec:artificial:distribution}.
+The number of particles is set to $N = 100$, $\tau  = 1/52$. 
 The interest rate is $r = 0.0296$ as in \cite{almansour:2016}. The initial guess for the mean and variance of the initial state are
@@
 The algorithm is stopped after $10000$ iterations.
-%and its bounded conditions are \\
-%\begin{tabular}{cccccccccccccc}
-%PAR & $\kappa$ & $\alpha_1$ & $\alpha_2$ & $\sigma_1$ & $\sigma_2$ & $\eta_1$ & $\eta_2$ & $\rho_1$ & $\rho_2$ & $g_1$ & $g_2$ & $g_3$ & $g_4$ \\
-%\hline
-%INF & 0.01 & -10.0 & -10.0 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 0.01 & 1e-10 & 1e-10 & 1e-10 & 1e-10 \\
-%SUP & 1e10 & 10.0 & 10.0 & 1e10 & 1e10 & 1e10 & 1e10 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 \\
-%\end{tabular}
+
 In Gibson-Schwartz model \cite{gibson:schwartz:1990}, a stronger backwardation effect implies a greater value for $\alpha$ for the same values of the other parameters.
@@
 
-%Remarks that the changement of $Q$ influences basically on the model parameters.
-A bad value of $Q$ can lead the parameters not to follow a good order in fonction of regimes.
-To prevent this trouble, our technique is firstly to fix the initial transition matrix $Q$ for the $100$ first iterations.
-That aids the model parameters to stabilize in a good order.
-More clearly in our experiment, fixing $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ up to 100 iterations drives CMAES to give a good order $\sigma_1 \geq \sigma_2$ $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$.
-When we no longer fix $Q$ after $100$ iterations, such the parameters in return conduct $\{Q(1,1), \; Q(2,2)\}$ to be in the good order.
-
 As expected, we obtain $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$ at convergence of the EM algorithm.
@@
 
-%Interestingly, we see that $\lambda^{\delta}_1 \geq \lambda^{\delta}_2$ \footnote{\cite{almansour:2016} obtain the same result in the opposite order, because of their definition of the historical mean-reverting value} and the two historical mean-reverting values are $\widetilde{\alpha}_1 = 0.0073$ and $\widetilde{\alpha}_2 = 0.0305$.
-This underlines that,
-
 Figure~\ref{fig:LogPriceTermStructure:data} compares the evolution of future $1M$ (the nearest contracts) to the \emph{term structure} observed from CME WTI crude oil, defined as the difference of future $13M$ and future $1M$ (to avoid seasonality).
@@
 in contango).
-%\begin{figure}
-%\centering
-%\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
-%\caption{Log-price (red line) and slope of future curves (blue line)}
-%\label{fig:LogPriceTermStructure:data}
-%\end{figure}
+\begin{figure}
+\centering
+\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
+\caption{Log-price (red line) and slope of future curves (blue line)}
+\label{fig:LogPriceTermStructure:data}
+\end{figure}
 
@@
 
-%\begin{figure}
-%	\centering
-%	\includegraphics[scale=.75,angle=90]{Posterior_Data}
-%	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
-%	\label{fig:posterior:data}
-%\end{figure}
-%
-%\begin{figure}
-%	\centering
-%	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
-%	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
-%	\label{fig:posterior:data:logprice}
-%\end{figure}
+\begin{figure}
+	\centering
+	\includegraphics[scale=.75,angle=90]{Posterior_Data}
+	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
+	\label{fig:posterior:data}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
+	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
+	\label{fig:posterior:data:logprice}
+\end{figure}
 

### 2017-05-13 TexArxiv/answer_referees.tex
Modifications for the revision

\include{clgm_def}
\begin{document}

\title{Particle rejuvenation of Rao-Blackwellized Sequential Monte Carlo smoothers for Conditionally Linear and Gaussian models}
\date{}

\author{Ngoc Minh Nguyen\footnote{LTCI, CNRS and T\'el\'ecom ParisTech, 46 rue Barrault 75634 Paris Cedex 13, France.}, Sylvain {L}e {C}orff\footnote{Laboratoire de Math\'ematiques d'Orsay, Univ. Paris-Sud, CNRS, Universit\'e Paris-Saclay, 91405 Orsay, France.} and Eric Moulines\footnote{Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique, France.}}

\lhead{Ngoc Minh Nguyen et al.}
\rhead{Conditionally Linear and Gaussian models}

\maketitle

The authors are grateful to the associate editor and the anonymous referees for their comments and remarks to improve the manuscript. These suggestions have been taken into consideration to propose a revision of the original paper. It has also been carefully proof read to remove remaining typos and minor mistakes. We provide below detailed answers for all comments.

\subsubsection*{Associate editor}

The paper surveys recent Rao-Blackwellized particle smoothers and presents a particle rejuvenation approach. The particle rejuvenation is interesting and seems useful, but it also shares some resemblance with the MCMC-based rejuvenation strategy proposed by [LBS+16, Section V.B] in the context of Rao-Blackwellized FFBS and with the strategy proposed by [FWT10] in the context of two-filter smoothing. Please relate the proposed method to these prior works. The detailed comments by the reviewers need to be addressed as well.

\vspace{.2cm}

{\em

It is true that our rejuvenation step shares some resemblance with the algorithms of [FWT10] and [LBS+16, Section V.B]. As noted in [FWT10, Secion~2.6], two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. The authors illustrate this issue in the case where the hidden state is an AR(2) process. To overcome the weakness of such standard two-filter approaches, [FWT10] samples new particles at time $i$ conditional on forward particles up to time $i-1$ and on backward particles from time $i+1$ which are appropriately weighted. This allows to produce new particles at time $i$ and to obtain a SMC approximation whose support is not restricted to the backward particles at time $i+1$ as in the original two-filter algorithm.  The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps.

On the other hand, another modification of the FFBS algorithm based on a Markov chain Monte Carlo sampling step has been introduced in [LBS+16]. Instead of sampling from  explicitly from (12), [LBS+16] proposed to draw a forward path $a_{1:i-1}$ in $(a^k_{1:i-1})_{1\le k \le N}$ and a sate $a_i$ in $\{1,\ldots,J\}$ according to:
\[
\widetilde{q}(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \widetilde{\vartheta}^k_{i-1}\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp,
\]
where $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ are adjustment multipliers and $\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})$ is a proposal kernel chosen by the user. This means that an ancestral path $a^{\star}_{1:i-1}$ is sampled in $(a^k_{1:i-1})_{1\le k \le N}$ with weights $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ and $a^{\star}_i$ is sampled from $\widetilde{q}(\cdot|a^{\star}_{1:i-1},a_{i+1:n},y_{1:n})$. Then the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio. The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability (11) for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  (11) for all combinations of sequences $(a^{k}_{1:i})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$. The method proposed in [LBS+16], is computationally more efficient as its complexity depends mainly on the number of MCMC samples produced by the algorithm. However it also requires a fair amount of tuning to choose efficient adjustment multipliers and proposal kernels. Our proposed method is computationally more intensive but allows to sample explicitly from (12) without additional tuning parameters. Note that both methods provide similar results for the same computational costs.
These remarks were added at the end of Section~2.2 and Section~2.3.1.
}

\subsubsection*{Referee 1}
A clear presentation of the state-of-the-art algorithms, e.g. what posterior pdf incl. linear states each algorithm (RB-FFBS and RB Two-filter smoother) tries to approximate, as well as a full description of these algorithms. In the current manuscript, the authors focus entirely on the SMC approximation part of the RB filters, and the modifications they intend to do (it is not really mentioned that the linear Gaussian part of the problem is not going to be presented, apologies if I missed it)

\vspace{.2cm}

{\em
\noindent The overall presentation of the paper has been modified to take this comment into account. First, the paper is not presented as a survey paper anymore as we want to focus on fully Rao-Blackwellized SMC approximations only and to highlight the improvements that can be made with simple rejuvenation steps. Therefore, we only mention the algorithm based on Kim's approximation in the introduction but detail only two-filter and FFBS based algorithms. In addition the Rao-Blackwellized two-filter of [BDM10] has been much more detailed in Section~2.3.1 before introducing the particle rejuvenation step in Section~2.3.2. For both algorithms a specific section is devoted to the original method and another section to the rejuvenation step. In addition, these rejuvenation steps are compared to other two-filter and FFBS improvements proposed in the literature.
}

\vspace{.5cm}

\noindent Some (more) intuition, about concepts and why certain steps are performed. Especially in the presentation of the RBTF in Section 2.4 I had problems to follow the overall idea. E.g. where do we integrate out the linear states in the backward pass of the original formulation presented by [BDM10]. I would expect this description before the modification of particle rejuvenation is presented. Also, some intuition about the explicit marginalization proposed in (9) would be helpful. Why is it chosen, what is the benefit compared to the original formulation? Currently, one gets easily lost in the derivations without maintaining the overall picture.

\vspace{.2cm}

{\em
\noindent  This comment helped to greatly improved the readability of the paper. As explained in the previous paragraph, Section~2.3.1 is now devoted to the original Rao-Blackwellized two-filter method and Section~2.3.2 to the rejuvenation step. Some additional motivations to introduce rejuvenation are given at the end of Sections~2.2.2 and~2.3.2 (see the answer to the associate editor). Two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. The particle rejuvenation proposed in our paper exploits the idea introduced in [FWT10] in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps. Most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities.
}

\vspace{.5cm}

\noindent Some comparisons in terms of performance, complexity, detailing the superiority (or not) of one method over the other. E.g. what are the pros and cons of RB-FFBS and RBTF. Also some numerical comparisons in the simulation section are expected.

\vspace{.2cm}

{\em
\noindent Some additional numerical experiments were added in Section~3. For each simulation study which was previously given in Section~3 for either FFBS or two-filter methods, now both algorithms are used and compared to their counterpart with rejuvenation. Our main objective is then to show the improvement of the particle rejuvenation steps for both methods.  This also confirms that FFBS algorithm perform better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of teach method.
}

\vspace{.5cm}

\noindent All minor comments were taken into consideration and typos were corrected. Some specific details are given below.

 \vspace{.5cm}

 \noindent The authors talk about two FFBS algorithms that they compare to the FFBS with
particle rejuvenation. But which two? Kim's and Lindsten's? Some more details
would be helpful, also which forward filter has been used.

\vspace{.2cm}

{\em 
\noindent The two FFBS algorithms are the original version of [LBS+16]  and the algorithm with rejuvenation.
A first (costly) run of the FFBS algorithm with rejuvenation with a great number of particles ($N=5000$) is used as a benchmark value. Then the posterior distributions are estimated by the original methods and the methods with rejuvenation with $N=50$ particles are compared to this  benchmark run.
}

\subsubsection*{Referee 2}
\noindent It is well known that SMC algorithms, in their most basic form, are numerically unstable
when used for approximating distributions defined on path spaces of increasing dimension. This is due to the particle path degeneracy phenomenon caused by the resampling mechanism of SMC algorithms. The degeneracy appears to be a potential problem for the estimators discussed in the paper, since the approximation of the
backward kernel (on, e.g., p. 8, line 52) involves quantities such as the mean and the
covariance matrix of the conditional distribution of $z_i$ given the states $a_{1:i}$ and the observations,
evaluated along the particle genealogical paths. I miss a discussion on this. If the effect exists, does the rejuvenation proposed by the authors counteract the same? (I do not think so).

\vspace{.2cm}
{\em
\noindent As explained to the associate editor and to  Referee~1, it is true that particle smoothers are prone to degenerate when they are based on  the genealogy of the forward or backward particles. In addition, in the specific case of two-filter based smoothers, [FWT10, Secion~2.6] noted that these algorithms are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$. 
The particle rejuvenation proposed in our paper produces an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$. In the specific case of linear and Gaussian models, Kalman filtering techniques allow to compute explicitly the approximation of the target distribution at time $i$ based on  $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$. The support of the approximated distribution is not restricted to $(\tilde{a}^k_{i})_{1\le k \le N}$ to overcome particle depletion. The marginalization over the linear states  builds new trajectories $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters. Although this does not overcome the overall degeneracy issue, numerical experiments highlight that this improves the performance of the algorithms.
}

\vspace{.5cm}

\noindent The rejuvenation approach proposed in Section 2.3 makes sense in that it allows the
production of backward draws that are not restricted to the discrete support formed by the particles generated in the forward  filtering pass. On the other hand, it is not obvious that there is a lot of variance to be gained in doing this (even though the simulation study in Section 3 indicates some improvement). Thus, I miss a heuristic
motivation for the rejuvenation step in Section 2.3. Are there specific models for which the rejuvenation step is supposed improve significantly the FFBS algorithm? As far as I understand, there is a price to pay for the rejuvenation in terms of some additional computational work, as the Kalman filter needs to be run for all the possible extensions
of each particle path. So, is the particle rejuvenation worthy of this additional effort? I guess it depends, e.g., on the number J of possible regimes?

\vspace{.2cm}
{\em
\noindent The intuition supporting particle rejuvenation is explained in the answers to the associate editor and to Referee~1. There is no theoretical result to prove that this approach provides a better approximation of the target distributions. However, our numerical experiments hint that this actually improves the accuracy and variance of the proposed estimators which would be even more interesting with additional regimes or additional rejuvenation steps. In addition, the additional computational cost is not prohibitive as Kalman filtering steps are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$ (but who is stil necessary when no rejuvenation step is introduced).
}

\vspace{.5cm}

\noindent It took a while before I understood the backward filter algorithm in Section 2.4. More
specifically, the expression of the importance weights on p. 11 seems to be based on
the identity
\[
\tilde{p}_{i}(a_{i:n}|y_{i:n}) \propto \left\{\prod_{u=i}^{n-1}Q(a_u,a_{u+1})\right\} \int \gamma^{}_{i}(a_i,z_i) p(y_{i:n}|z_i,a_{i:n})\rmd z_i\eqsp.
\]
which never appears explicitly in the description of the backward  filter. I would recommend
the authors to include a derivation of the backward filter on the basis of this identity.

\vspace{.2cm}
{\em
\noindent The identity was added in Section~2.3.1 in a more detailed derivation of the backward filter.
}

\vspace{.5cm}

\noindent Finally, I miss a discussion on the last sampling operation of the Rao-Blackwellized
two-filter smoother, i.e., the sampling of the mixture obtained when plugging the approximations into (9). This should yield a mixture with $N^2$ components, implying a not innocent computational complexity when N is large.

\vspace{.2cm}
{\em
\noindent The additional computational cost of the rejuvenation step comes from the Kalman filtering steps which are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$. However, this combination is also necessary in the original two-filter algorithm of [BDM10] with no rejuvenation.
}

\vspace{.5cm}

\noindent All minor comments were taken into consideration and typos were corrected.
\end{document}

### 2017-05-13 TexArxiv/clgm.tex
Modifications for the revision

@@
 
-\title{A survey of Rao-Blackwellized Sequential Monte Carlo smoothers for Conditionally Linear and Gaussian models}
+\title{Particle rejuvenation of Rao-Blackwellized Sequential Monte Carlo smoothers for Conditionally Linear and Gaussian models}
 \date{}
@@
 \begin{abstract}
-This paper focuses on Sequential Monte Carlo based approximations of smoothing distributions in conditionally linear and Gaussian state spaces.
-To reduce Monte Carlo variance of smoothers, it is typical in these models to use Rao-Blackwellisation: particle approximation is used to sample the sequence of hidden regimes while the Gaussian states are explicitly integrated conditional on the sequence of regimes and observations, using variants of the Kalman filter / smoother.
-The first successful attempt to use Rao-Blackwellisation for smoothing has been presented in  \cite{briers:doucet:maskell:2010} which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filters / smoothers.
-More recently, in a series of article, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed to use a  forward backward decomposition of the smoothing distribution, mimicking the Rauch-Tung-Striebel smoother for the regime combined with backward Kalman updates.
-In this paper, we provide a survey of these algorithms in a common unifying framework.
-We then investigate the benefit of introducing an additional rejuvenation step  which allows to sample at each time instant new  regimes conditional on the forward and backward particles.
-To support our claims, a Monte Carlo study is presented.
+This paper focuses on Sequential Monte Carlo approximations of smoothing distributions in conditionally linear and Gaussian state spaces.
+To reduce Monte Carlo variance of smoothers, it is typical in these models to use Rao-Blackwellization: particle approximation is used to sample sequences of hidden regimes while the Gaussian states are explicitly integrated conditional on the sequence of regimes and observations, using variants of the Kalman filter / smoother.
+The first successful attempt to use Rao-Blackwellization for smoothing has been presented in  \cite{briers:doucet:maskell:2010} which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filters / smoothers.
+More recently, \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015} proposed to use a  forward backward decomposition of smoothing distributions, mimicking the Rauch-Tung-Striebel smoother for the regimes combined with backward Kalman updates.
+This paper investigates the benefit of introducing additional rejuvenation steps in all these algorithms to sample at each time instant new  regimes conditional on the forward and backward particles.
+This defines particle based approximations of the smoothing distributions whose support is not restricted to the set of particles sampled in the forward or backward filter.
 These procedures are applied to commodity markets which are described using a two factor model based on the spot price and a convenience yield for crude oil data.
@@
 \section{Introduction} \label{sec:intro}
-State space models are bivariate stochastic processes $\{(Y_k,Z_k)\}_{k\ge 1}$ where the state sequence $\{Z_k\}_{k\ge 1}$ is a Markov chain which is only partially observed through the sequence $\{Y_k\}_{k\ge 1}$.
-Conditionally on the state sequence $\{Z_k\}_{k\ge 1}$ the observations are independent and for all $\ell\ge 1$ the conditional distribution of $Y_{\ell}$ given $\{Z_k\}_{k\ge 1}$ depends on $Z_{\ell}$ only.
+State space models are bivariate stochastic processes $\{(Y_i,Z_i)\}_{i\ge 1}$ where the state sequence $(Z_i)_{i\ge 1}$ is a Markov chain which is only partially observed through the sequence $(Y_i)_{i\ge 1}$.
+Conditionally on the state sequence $(Z_i)_{i\ge 1}$ the observations are independent and for all $\ell\ge 1$ the conditional distribution of $Y_{\ell}$ given $(Z_i)_{i\ge 1}$ depends on $Z_{\ell}$ only.
 These models are used in a large variety of disciplines such as financial econometrics, biology, signal processing, see \cite{delmoral:2013} and the references therein.
 In general state space models, bayesian filtering and smoothing problems, i.e.
-the computation of the posterior distributions of a sequence of states $(Z_{k},\ldots,Z_{p})$ for $1\le k\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks.
-Filtering refers to the estimation of the distributions of the hidden state $Z_k$ given the observations $(Y_1,\ldots,Y_k)$ up to time $k$, while fixed-interval smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le k\le p<\ell$.
+the computation of the posterior distributions of a sequence of states $(Z_{i},\ldots,Z_{p})$ for $1\le i\le p\le \ell$ given observations $(Y_{1},\ldots,Y_{\ell})$, are challenging tasks.
+Filtering refers to the estimation of the distributions of the hidden state $Z_i$ given the observations $(Y_1,\ldots,Y_i)$ up to time $i$, while fixed-interval smoothing stands for the estimation of the distribution of sequence of states $(Z_{i},\ldots,Z_{p})$ given observations $(Y_{1},\ldots,Y_{\ell})$ with $1\le i\le p<\ell$.
 
 When the state and observation models are linear and Gaussian, filtering can be solved explicitly using the Kalman filter \cite{kalman:1960}.
-Exact solutions of the fixed-horizon smoothing problem can be obtained using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier "two-filter" smoother \cite{bryson:frazier:1963}.
-This paper focuses on  Conditionally Linear and Gaussian Models (CLGM) given for $k\ge 1$ by:
+Exact solutions of the fixed-horizon smoothing problem can be obtained using either the Rauch-Tung-Striebel smoother \cite{rauch:striebel:tung:1965} or the Bryson-Frazier two-filter smoother \cite{bryson:frazier:1963}.
+This paper focuses on  Conditionally Linear and Gaussian Models (CLGM) given for $i\ge 2$ by:
 \begin{equation}
 \label{eq:model:state}
-Z_k = d_{a_{k}} + T_{a_{k}}Z_{k-1}+H_{a_{k}}\varepsilon_k\eqsp,
+Z_i = d_{a_{i}} + T_{a_{i}}Z_{i-1}+H_{a_{i}}\varepsilon_i\eqsp,
 \end{equation}
-where
+where:
 \begin{enumerate}[-]
-\item $(\varepsilon_k)_{k\ge 1}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance.
-\item $(a_k)_{k\ge 0}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$, called \emph{regimes}.  
-\item $(H_j)_{1\le j\le J}$ are positive-definite matrices.
-\item $Z_0$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_k)_{k\ge 1}$.
+\item $(\varepsilon_i)_{i\ge 2}$ is a sequence of independent and identically distributed (i.i.d.) $\dimz$-dimensional Gaussian vectors with zero mean and identity covariance.
+\item $(a_i)_{i\ge 1}$ is a homogeneous Markov chain taking values in a finite space $\{1,\ldots,J\}$, called \emph{regimes}, with initial distribution $\pi$ and transition matrix $Q$.  
+\item $(H_j)_{1\le j\le J}$ are $\dimz \times \dimz$ positive-definite matrices, $(d_j)_{1\le j\le J}$ $\dimz$-dimensional vectors and $(T_j)_{1\le j\le J}$ $\dimz \times \dimz$ positive-definite matrices.
+\item $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_i)_{i\ge 2}$.
 \end{enumerate}
-At each time step $k\ge 1$, the observation $Y_k$ is given by:
+Let $n$ be the number of observations. At each time step $1\le i \le n$, the observation $Y_i$ is given by:
 \begin{equation}
 \label{eq:model:obs}
-Y_k = c_{a_k} + B_{a_k}Z_k + G_{a_k}\eta_k\eqsp,
+Y_i = c_{a_i} + B_{a_i}Z_i + G_{a_i}\eta_i\eqsp,
 \end{equation}
-where
+where:
 \begin{enumerate}[-]
-\item $(\eta_k)_{k\ge 0}$ is a i.i.d. sequence of $\dimy$-dimensional Gaussian vectors, independent of $(\varepsilon_k)_{k\ge 1}$ and $Z_1$.
-\item $(G_j)_{1\le j \le J}$ are $\dimy \times \dimy$ positive-definite matrices and $(B_j)_{1\le j \le J}$ are $\dimy \times \dimz$ matrices.
+\item $(\eta_i)_{i\ge 1}$ is a i.i.d. sequence of $\dimy$-dimensional Gaussian vectors, independent of $(\varepsilon_i)_{i\ge 2}$ and $Z_1$.
+\item $(G_j)_{1\le j \le J}$ are $\dimy \times \dimy$ positive-definite matrices, $(c_j)_{1\le j \le J}$ $\dimy$-dimensional vectors and $(B_j)_{1\le j \le J}$ $\dimy \times \dimz$ matrices.
 \end{enumerate}
 CLGM play an important role in many applications; see \cite{sarkka:2013} and the references therein for an up-to-date account.
-A crucial feature of these models is that, conditional on the regime sequence $(a_1,\ldots,a_n)$ , both the state equation and the observation equation are linear and Gaussian, which implies that conditional on the sequence of regimes and the observations, the filtering and the smoothing distribution of the 'continuous' states $(Z_1,\dots,Z_n)$ can be computed explicitly.
+A crucial feature of these models is that, conditional on the regime sequence $(a_1,\ldots,a_n)$, both the state equation and the observation equation are linear and Gaussian, which implies that conditional on the sequence of regimes and on the observations, the filtering and the smoothing distributions of the continuous states $(Z_1,\dots,Z_n)$ can be computed explicitly.
 
-To exploit this specific structure, it has been suggested in the pioneering papers of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample the regimes with the Kalman filter to compute the conditional distribution of the states sequence $(Z_j)_{j=1}^n$ conditional on the regimes and the observations.
-This is a specific instance of Rao-Blackwellised Monte Carlo filters, often referred to as the \emph{Mixture Kalman Filter}. 
+To exploit this specific structure, it has been suggested in the pioneering works of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample the regimes with the Kalman filter to compute the conditional distribution of the states sequence $(Z_i)_{1\le i\le n}$ conditional on the regimes and on the observations.
+This is a specific instance of Rao-Blackwellized Monte Carlo filters, often referred to as the \emph{Mixture Kalman Filter}.
 Improvements of these early filtering techniques  have been introduced in \cite{doucet:gordon:krishnamurthy:2001,schon:gustafsson:nordlund:2005}.
 
-The use of Rao Blackwellisation to solve the smoothing problem has been proved to be more challenging and has received satisfactory solutions only recently.
-
-The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao Blackwellised as it required to sample the hidden linear states in the backward pass.
-
-An alternative approach, based on the so-called  'structural approximation' of the model suggested in an early paper by \cite{kim:1994}, was proposed by \cite{barber:2006} to avoid to sample a 'continuous' state in the backward pass.
+The use of Rao-Blackwellization to solve the smoothing problem has been proved to be more challenging and has received satisfactory solutions only recently.
+The first forward-backward smoother proposed in the literature \cite{fong:godsill:doucet:west:2002} was not fully Rao-Blackwellized as it required to sample the hidden linear states in the backward pass.
+An alternative approach, based on the so-called structural approximation of the model suggested in an early paper by \cite{kim:1994}, was proposed by \cite{barber:2006} to avoid to sample a continuous state in the backward pass.
 This approximation is rather ad-hoc and the resulting smoother is not consistent when the number of particles goes to infinity.
@@
 
-The first fully Rao Blackwellised SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao Blackwellisation steps for the forward and the backward filters.
+The first fully Rao-Blackwellized SMC smoother which should lead to consistent approximations when the number of particles grows to infinity was proposed by  \cite{briers:doucet:maskell:2010} and extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula with Rao-Blackwellization steps for the forward and the backward filters.
 This two-filter approach combines a forward filter with a backward information filter which are approximated numerically using SMC for the regime sequence and Kalman filtering techniques for the hidden linear states.
 
-More recently, \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} introduced a Rao Blackwellised smoother based on the forward-backward decomposition of the FFBS algorithm with Rao Blackwellisation steps both in the forward and backward time directions.
+More recently, \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015} introduced a Rao-Blackwellized smoother based on the forward-backward decomposition of the FFBS algorithm with Rao-Blackwellization steps both in the forward and backward time directions.
 The update of the smoothing distribution of the regime given the observations shares some striking similarities with the Rauch-Tung-Striebel smoothing procedure, which is at the heart of the FFBS procedure.
-The Rao-Blackwellisation requires to update backward in time the smoothing distribution of the states given the regimes and the observations, which is achieved by using a \emph{à la Kalman} backward update.
+The Rao-Blackwellization requires to update backward in time the smoothing distribution of the states given the regimes and the observations, which is achieved by using an \emph{\`a la Kalman} backward update.
 
-In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} by using additional Rao-Blackwellisation steps allowing to sample "new" particles in the backward pass.
+In this paper, we propose to improve the performance of the algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015} by using additional Rao-Blackwellization steps which allows to sample new particles in the backward pass.
 This approach may be seen as an extension of the ideas of \cite{fearnhead:clifford:2003} for Rao-Blackwellized smoothers.
 In \cite{briers:doucet:maskell:2010}, for all $1\le i\le n$, the sampled forward and backward sequences are merged to approximate the posterior distribution of $(a_i,z_i)$.
+This provides an approximation whose support is restricted to the particles produced at time $i$ by the backward particle filter.
+As noted in \cite[Secion~2.6]{fearnhead:wyncoll:tawn:2010}, these two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$.
 We propose to approximate the marginal smoothing distribution of $(a_i,z_i)$ by merging the sampled forward and backward trajectories at times $i-1$ and $i+1$ and integrating out all possible paths between time $i-1$ and time $i$ and between time $i$ and time $i+1$ instead of sampling random variables.
-Similarly, in the backward pass of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}, a regime $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$.
+Similarly, in the backward pass of \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015}, a regime $\tilde{a}_i$ is sampled at time $1\le i\le n-1$ using the particles produced by the forward filter at time $i$.
 In this case, particle rejuvenation may be introduced by using the forward weighted samples at time $i-1$ and extending these trajectories at time $i$ with a Kalman filter for all possible values of the regime.
@@
 The paper is organized as follows.
-The proposed modification of the fully Rao Blackwellised Monte Carlo smoothers are presented in Section~\ref{sec:RaoBlackwell}.
+The algorithms introduced in \cite{briers:doucet:maskell:2010} and in \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015} as long as the proposed rejuvenation associated with each method are presented in Section~\ref{sec:RaoBlackwell}.
 The performance of all these methods is illustrated in Section~\ref{sec:numerical:experiments} with simulated data.
@@
 
-\section{Rao Blackwellised smoothing algorithms}
+\section{Rao-Blackwellized smoothing algorithms}
 \label{sec:RaoBlackwell}
 This section details the Sequential Monte Carlo algorithms which can be used to approximate the conditional distribution of the states $(a_1,\ldots,a_n)$ or the marginal distributions of $(a_i,z_{i})$ given the observations $(Y_1,\ldots,Y_n)$.
-For all $m\,\times\,m$ matrix  let $|A|$ be the determinant of $A$.
+For all $\dimz\,\times\,\dimz$ matrix  let $|A|$ be the determinant of $A$.
 If $A$ is a positive-definite matrix, for all $z\in\mathbb{R}^{m}$ define
 \[
-\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp.
+\normMat{A}{z}^2 \eqdef z'A^{-1}z\eqsp,
 \]
+where for any vector or matrix $z$, $z'$ denotes the transpose matrix of $z$.
 Let $m_{}(a_{i},z_{i-1};z_i)$ be the probability density of the conditional distribution of $Z_i$ given $(a_{i},Z_{i-1})$ and $g_{}(a_{i},z_{i};y_i)$ be the probability density of the conditional distribution of $Y_i$ given $(a_{i},Z_{i})$:
@@
 \label{sec:forward}
-The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using a standard Rao Blackwellised  algorithm.
+The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using a standard Rao-Blackwellized  algorithm.
 The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with normalized importance weights $(\omega^k_{i})_{1\le k \le N}$ ($\sum_{k=1}^N \omega^k_i = 1$) used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
@@
 \label{eq:forward:SMC}
-p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
+p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\,\delta_{a^k_{1:i}}(a_{1:i})\eqsp,
 \end{equation}
-In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by using the Kalman filter's update.
+where $\delta$ is the Dirac delta function.
+In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by using the Kalman filter update.
 
@@
 \mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
-P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
+P^k_{1} &=\left(I_{\dimz}-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp,
 \end{align*}
-Each particle particle $a_1^k$ is associated with the importance weight $\omega^k_1 = 1/N$.
+where for all positive integer $p$, $I_p$ is the $p\times p$ identity matrix. Each particle particle $a_1^k$ is associated with the importance weight $\omega^k_1 = 1/N$.
 
@@
 Several procedures may be used to extend the trajectories $(a^k_{1:i-1})_{1\le k \le N}$ at time $i$.
-For all sampled trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$ and all $1\le j \le J$,  \cite{chen:liu:2000} used the incremental weights:
+For all sampled trajectories $(a_{1:i-1}^k)_{1\le k \le N}$ and all $1\le j \le J$,  \cite{chen:liu:2000} used the incremental weights:
 \[
@@
 \]
-The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ so that:
+The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}(a_i)$ and variance $B_{a_i}P^k_{i|i-1}(a_i)B'_{a_i} + \barG_{a_i}$ where
+\begin{align*}
+\mu^{k}_{i|i-1}(a_i) &= d_{a_i} + T_{a_i}\mu^k_{i-1}\eqsp,\\
+P^{k}_{i|i-1}(a_i) &= T_{a_i}P^k_{i-1}T'_{a_i} + \barH_{a_i}\eqsp.
+\end{align*} 
+Therefore,
 \[
@@
 \]
-where  $\mu^{j,k}_{i|i-1} = d_{j} + T_{j}\mu^k_{i-1}$ and $P^{j,k}_{i|i-1} = T_{j}P^k_{i-1}T'_{j} + \barH_{j}$.
+where:  
+\begin{align}
+\mu^{j,k}_{i|i-1} &= \mu^{k}_{i|i-1}(j) = d_{j} + T_{j}\mu^k_{i-1}\eqsp,\label{eq:mui|i-1}\\
+P^{j,k}_{i|i-1} &= P^{k}_{i|i-1}(j) = T_{j}P^k_{i-1}T'_{j} + \barH_{j}\eqsp.\label{eq:Pi|i-1}
+\end{align} 
 In \cite{chen:liu:2000}, for all $1\le k \le N$, an ancestral path is chosen with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.
@@
 Following \cite{barembruch:garivier:moulines:2008}, we might choose between the Kullback-Leibler Optimal Selection (KL-OS) or the Chi-Squared Optimal Selection (CS-OS) to associate a new weight to each of the $NJ$ trajectories.
-If the new weight is 0, then the corresponding particle can be removed.\\
-\textbf{KL-OS:} $\lambda$ is chosen as the solution of :
+If the new weight is 0, then the corresponding particle can be removed.
+
+\vspace{.2cm}
+
+\noindent\textbf{KL-OS:} $\lambda$ is chosen as the solution of :
 \[
@@
  \begin{array}{rl}
-  &\lambda \;\mbox{with probability } \tilde{\omega}^{j,k}_{i}/\lambda\eqsp,\\
-&0 \;\mbox{with probability } 1-\tilde{\omega}^{j,k}_{i}/\lambda\eqsp.
+  &\hspace{-.5cm}\lambda \;\mbox{with probability } \tilde{\omega}^{j,k}_{i}/\lambda\eqsp,\\
+&\hspace{-.5cm}0 \;\mbox{with probability } 1-\tilde{\omega}^{j,k}_{i}/\lambda\eqsp.
 \end{array}
@@
  \begin{array}{rl}
-  &\sqrt{\tilde{\omega}^{j,k}_{i}\lambda} \;\mbox{with probability } \sqrt{\tilde{\omega}^{j,k}_{i}/\lambda}\eqsp,\\
-&0 \;\mbox{with probability } 1-\sqrt{\tilde{\omega}^{j,k}_{i}/\lambda}\eqsp.
+  &\hspace{-.5cm}\sqrt{\tilde{\omega}^{j,k}_{i}\lambda} \;\mbox{with probability } \sqrt{\tilde{\omega}^{j,k}_{i}/\lambda}\eqsp,\\
+&\hspace{-.5cm}0 \;\mbox{with probability } 1-\sqrt{\tilde{\omega}^{j,k}_{i}/\lambda}\eqsp.
 \end{array}
@@
 Then, in both cases, all particles such that $\tilde{\Omega}^{j,k}_{i} = 0$ are discarded and for all the other trajectories defined as an ancestral path $(a^k_{1:i-1})$ extended by $a^k_i = j$, the new corresponding weight $\omega$ in \eqref{eq:forward:SMC} is given by the normalized weight $\tilde{\Omega}^{j,k}_{i}$.
+In the numerical sections of this paper, the Kullback-Leibler Optimal Selection (KL-OS) scheme is used.
 
-\subsection{Algorithm based on the structural approximation of \cite{kim:1994}, \cite{barber:2006}}
-The method to approximate the smoothing distributions of the regimes given all the observations proposed by \cite{barber:2006} relies on an approximation of $p(a_i|a_{i+1:n},y_{1:n})$, which is hardly justified and may introduce some bias.
-Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$ approximating the filtering distributions are available.
-Then, note that, for all $1\le i\le n$,
-\[
-p(a_{i:n}|y_{1:n}) = p(a_{i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp.
-\]
-Then, the approximation of \cite{kim:1994,barber:2006} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
-\[
-p(a_i|a_{i+1},y_{1:i}) \propto Q(a_i,a_{i+1})p(a_i|y_{1:i})\eqsp.
-\]
-Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$:
-\begin{enumerate}[-]
-\item $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$.
-\item Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$.
-\end{enumerate}
-If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the regimes is:
-\[
-p^{\mathsf{Kim}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
-\]
+%\subsection{Algorithm based on the structural approximation of \cite{kim:1994}, \cite{barber:2006}}
 
-\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
-\cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013} proposed a Rao-Blackwellised procedure to sample the regime backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} without using the ad-hoc approximation of \cite{kim:1994}.
+\subsection{FFBS based algorithms}% of \cite{lindsten:bunch:sarkka:schon:godsill:2015,sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013}}
+\subsubsection{FFBS algorithms of \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015}}
+%Forward filtering backward smoothing algorithms offer an alternative to the method proposed by \cite{barber:2006} which relies on a hardly justified and biased approximation of $p(a_{i:n}|y_{1:n})$, for all $1\le i\le n$.
+%Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$ approximating the filtering distributions are available.
+%
+%In \cite{kim:1994,barber:2006}, the decomposition:
+%\[
+%p(a_{i:n}|y_{1:n}) = p(a_{i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})
+%\]
+%is approximated replacing $p(a_i|a_{i+1:n},y_{1:n})$ by
+%\[
+%p(a_i|a_{i+1},y_{1:i}) \propto Q(a_i,a_{i+1})p(a_i|y_{1:i})\eqsp.
+%\]
+%Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$.
+%
+%\begin{enumerate}[-]
+%\item $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$.
+%\item Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$.
+%\end{enumerate}
+%If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the regimes is:
+%\[
+%p^{\mathsf{Kim}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
+%\]
+\cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellized procedure to sample the regime backward in time following the same steps as in the Forward Filtering Backward Smoothing algorithm \cite{huerzeler:kunsch:1998,doucet:godsill:andrieu:2000}.
 The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
 \[
-p(a_{1:n}|y_{1:n}) = p(a_{1:i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp,
+p(a_{1:n}|y_{1:n}) = p(a_{1:i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp.
 \]
@@
 The first factor on the right hand side of the previous equation is nevertheless more difficult to handle because it itself relies on all the observations.
-Nevertheless, as noted by \cite{sarkka:bunch:godsill:2012}, this term can be computed recursively by considering the following obvious decomposition:
-\[
+As noted by \cite{sarkka:bunch:godsill:2012}, this term can be computed recursively by considering the following decomposition:
+\begin{equation}
+\label{eq:FFBS:decomp}
 p(a_{1:i}|a_{i+1:n},y_{1:n}) \propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})p(a_{1:i}|y_{1:i})\eqsp.
-\]
+\end{equation}
 The second factor in the last equation may be approximated using the ancestral  trajectories $(a^k_{1:i})_{1\le k \le N}$ and the associated importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter.
@@
 \[
-p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a_{1:i})\eqsp, \quad \text{with} \quad \tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) \eqsp.
+p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a_{1:i}) \quad \text{with} \quad \tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) \eqsp.
 \]
@@
 This algorithm requires to compute the quantity  $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$.
-As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using  Kalman filtering techniques.
+This predictive quantity is available analytically using  Kalman filtering techniques.
 However, this has to be done for each trajectory $(a^k_{1:i})_{1\le k \le N}$, which leads to an algorithm with a prohibitive computational complexity.
@@
 \label{eq:FFBS:incremental}
-p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) = \int p(y_{i+1:n},a_{i+1:n}|z_i,a^k_{i})p(z_{i}|a^k_{1:i},y_{1:i})\rmd z_i\eqsp,
+p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) = \int p(y_{i+1:n},a_{i+1:n}|z_i,a^k_{i})p(z_{i}|a^k_{1:i},y_{1:i})\rmd z_i\eqsp.
 \end{equation}
@@
 \[
-p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\hat{\Omega}_i(a_{i:n})z_i-2\hat{\lambda}'_i(a_{i:n})z_i)/2\right\}\eqsp,
+p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\widehat{\Omega}_i(a_{i:n})z_i-2\widehat{\lambda}'_i(a_{i:n})z_i)/2\right\}\eqsp,
 \]
@@
 \begin{align*}
-\hat{\Omega}_n(a_n) &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
-\hat{\lambda}_n(a_n) &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
+\widehat{\Omega}_n(a_n) &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
+\widehat{\lambda}_n(a_n) &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
 \end{align*}
-Then, for $1\le i\le n-1$, define $m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\hat{\Omega}_{i+1}H_{a_{i+1}} + I$ and write
+Then, for $1\le i\le n-1$, define $m_{i+1} = \widehat{\lambda}_{i+1} - \widehat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\widehat{\Omega}_{i+1}H_{a_{i+1}} + I_{\dimz}$ and write
 \begin{align*}
-\Omega_i(a_{i+1:n}) &= T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}(a_{i+1:n})H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\hat{\Omega}_{i+1}(a_{i+1:n})T_{a_{i+1}}\eqsp,\\
-\lambda_i(a_{i+1:n}) &=T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}(a_{i+1:n})H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
+\Omega_i(a_{i+1:n}) &= T'_{a_{i+1}}(I_{\dimz}-\widehat{\Omega}_{i+1}(a_{i+1:n})H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\widehat{\Omega}_{i+1}(a_{i+1:n})T_{a_{i+1}}\eqsp,\\
+\lambda_i(a_{i+1:n}) &=T'_{a_{i+1}}(I_{\dimz}-\widehat{\Omega}_{i+1}(a_{i+1:n})H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
 \end{align*}
@@
 \begin{align*}
-\hat{\Omega}_i(a_{i:n}) &= \Omega_i(a_{i+1:n})+ B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
-\hat{\lambda}_i(a_{i:n}) & = \lambda_i(a_{i+1:n}) + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
+\widehat{\Omega}_i(a_{i:n}) &= \Omega_i(a_{i+1:n})+ B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
+\widehat{\lambda}_i(a_{i:n}) & = \lambda_i(a_{i+1:n}) + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
 \end{align*}
@@
 \label{eq:post:obs}
-p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})\propto Q(a_i^k,a_{i+1})|\Lambda_i|^{-1/2}\exp\left\{\eta_i/2\right\}\eqsp,
+p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})\propto Q(a_i^k,a_{i+1})|\Lambda^k_i(a_{i+1:n})|^{-1/2}\exp\left\{-\eta^k_i(a_{i+1:n})/2\right\}\eqsp,
 \end{equation}
@@
 \begin{align*}
-\Lambda^k_i(a_{i+1:n})&= (\Gamma_i^k)'\Omega_i(a_{i+1:n})\Gamma_i^k + I\eqsp,\\
+\Lambda^k_i(a_{i+1:n})&= (\Gamma_i^k)'\Omega_i(a_{i+1:n})\Gamma_i^k + I_{\dimz}\eqsp,\\
 \eta^k_i(a_{i+1:n}) &= \|\mu_i^k\|^2_{\Omega^{-1}_i(a_{i+1:n})} - 2\lambda'_i(a_{i+1:n})\mu_i^k-\|(\Gamma_i^k)'(\lambda_i(a_{i+1:n})-\Omega_i(a_{i+1:n})\mu_i^k)\|^2_{\Lambda_i(a_{i+1:n})}\eqsp,
@@
 \[
-\tilde{\omega}_{i|n} \propto \omega_i^kQ(a_i^k,a_{i+1})|\Lambda^k_i(a_{i+1:n})|^{-1/2}\exp\left\{\eta^k_i(a_{i+1:n})/2\right\}\eqsp.
+\tilde{\omega}_{i|n} \propto \omega_i^kQ(a_i^k,a_{i+1})|\Lambda^k_i(a_{i+1:n})|^{-1/2}\exp\left\{-\eta^k_i(a_{i+1:n})/2\right\}\eqsp.
 \]
@@
 \]
-\textbf{Additional particle rejuvenation step}\\
-An improved version of this FFBS algorithm may be defined for all $2\le i\le n-1$ by writing :
+
+\subsubsection{Particle rejuvenation of FFBS algorithms}
+The crucial step of the FFBS algorithm is the decomposition \eqref{eq:FFBS:decomp} which allows to extend a backward trajectory $\tilde{a}_{i+1:n}$ by choosing a particle in the set $(a_i^k)_{1\le k \le N}$ produced by the forward filter (and discarding the states $a^k_{1:i-1}$).
+An improved version of this FFBS algorithm which is not constrained to sample states in the support $(a_i^k)_{1\le k \le N}$ may be defined for all $2\le i\le n-1$ by writing:
 \begin{align*}
@@
 \end{align*}
-Replacing $p(a_{1:i-1},z_{i-1}|y_{1:i-1})$ in the integral by the particle approximation obtained during the forward pass and using Kalman filtering techniques for each trajectory $\{a^k_{1:i-1}\}_{1\le k\le N}$ and each $a_i\in\{1,\ldots,J\}$ yields:
+Replacing $p(a_{1:i-1},z_{i-1}|y_{1:i-1})$ in the integral by the particle approximation obtained during the forward pass and using Kalman filtering techniques for each trajectory $(a^k_{1:i-1})_{1\le k\le N}$ and each $a_i\in\{1,\ldots,J\}$ yields:
 \[
@@
 \[
-\omega_{i|i-1}^k(a_i) = \omega_{i-1}^k Q(a_{i-1}^k,a_i)|\Sigma^k_{i|i-1}(a_i)|^{-1/2}\mathrm{e}^{-\left(y_i - y^k_{i|i-1}(a_i)\right)'\left(\Sigma^k_{i|i-1}(a_i)\right)^{-1}\left(y_i - y^k_{i|i-1}(a_i)\right)/2}\eqsp,
+\omega_{i|i-1}^k(a_i) = \omega_{i-1}^k Q(a_{i-1}^k,a_i)|\Sigma^k_{i|i-1}(a_i)|^{-1/2}\mathrm{exp}\left\{-\frac{1}{2}\|y_i - y^k_{i|i-1}(a_i)\|_{\Sigma^k_{i|i-1}(a_i)}\right\}\eqsp,%\mathrm{e}^{-\left(y_i - y^k_{i|i-1}(a_i)\right)'\left(\Sigma^k_{i|i-1}(a_i)\right)^{-1}\left(y_i - y^k_{i|i-1}(a_i)\right)/2}\eqsp,
 \]
@@
 \]
-On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the regime of the filtering pass $\{a_i^k\}_{1\le k \le N}$.
+On the other hand, for all $1\le k \le N$, $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i})$ is computed as in \eqref{eq:post:obs} with all possible values $a_i\in\{1,\ldots,J\}$ and not only the regime of the filtering pass $(a_i^k)_{1\le k \le N}$.
 This means that a Kalman filter must be used for each trajectory $a^k_{1:i-1}$ which may be extended by $a_i\in\{1,\ldots,J\}$.
-Denote by $\mu_{i|i-1}^k(a_i)$ and $P_{i|i-1}^k(a_i)$ the mean and covariance matrix of the law of $z_i$ given $(a^k_{1:i-1},a_i)$.
+Denote by $\mu_{i|i-1}^k(a_i)$ and $P_{i|i-1}^k(a_i)$ the mean and covariance matrix of the law of $z_i$ given $(a^k_{1:i-1},a_i)$ obtained as in \eqref{eq:mui|i-1} and \eqref{eq:Pi|i-1}.
 Then,
-\[
-p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i}) = Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{\eta^k_{i|i-1}(a_{i:n})/2\right\}\eqsp,
-\]
+\begin{equation}
+\label{eq:posterior:FFBS}
+p(y_{i+1:n},a_{i+1:n}|a^k_{1:i-1},a_i,y_{1:i}) = Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{-\eta^k_{i|i-1}(a_{i:n})/2\right\}\eqsp,
+\end{equation}
 where the proportionality is with respect to $(a^k_{1:i-1},a_i)$ and
 \begin{align*}
-\Lambda^k_{i|i-1}(a_{i:n})&= (\Gamma_{i|i-1}^k(a_i))'\Omega_i(a_{i+1:n})\Gamma_{i|i-1}^k(a_i) + I\eqsp,\\
+\Lambda^k_{i|i-1}(a_{i:n})&= (\Gamma_{i|i-1}^k(a_i))'\Omega_i(a_{i+1:n})\Gamma_{i|i-1}^k(a_i) + I_{\dimz}\eqsp,\\
 \eta^k_{i|i-1}(a_{i:n}) &= \|\mu_{i|i-1}^k(a_i)\|^2_{\Omega^{-1}_i(a_{i+1:n})} - 2\lambda'_i(a_{i+1:n})\mu_{i|i-1}^k(a_i)\\
@@
 \end{align*}
-where $P_{i|i-1}^k(a_i) = \Gamma_{i|i-1}^k(a_i)(\Gamma_{i|i-1}^k(a_i))'$. $p(a_{1:i}|a_{i+1:n},y_{1:n})$ is then approximated by :
+where $\Gamma_{i|i-1}^k(a_i)$ is defined as $P_{i|i-1}^k(a_i) = \Gamma_{i|i-1}^k(a_i)(\Gamma_{i|i-1}^k(a_i))'$. The distribution $p(a_{1:i}|a_{i+1:n},y_{1:n})$ is then approximated by :
+\begin{multline}
+\label{eq:FFBS:rejuv}
+p^N(a_{1:i}|a_{i+1:n},y_{1:n})\\
+\propto\sum_{k=1}^N \omega_{i|i-1}^k(a_i)Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{-\eta^k_{i|i-1}(a_{i:n})/2\right\}\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp.
+\end{multline}
+By integrating over all possible paths $a_{1:i-1}$, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$.
+This particle rejuvenation step allows to explore states which are not in the support of the particles produced by the forward filter  and improves the accuracy and the variance of the original FFBS algorithm, see Section~\ref{sec:numerical:experiments} for numerical illustrations.
+
+Another modification of the FFBS algorithm based on a Markov chain Monte Carlo (MCMC) sampling step was introduced in \cite[Section~5.2]{lindsten:bunch:sarkka:schon:godsill:2015}.
+Instead of sampling from \eqref{eq:FFBS:rejuv}, \cite[Section~5.2]{lindsten:bunch:sarkka:schon:godsill:2015} proposed to draw a forward path $a_{1:i-1}$ in $(a^k_{1:i-1})_{1\le k \le N}$ and a sate $a_i$ in $\{1,\ldots,J\}$ according to:
 \[
-p^N(a_{1:i}|a_{i+1:n},y_{1:n})\propto\sum_{k=1}^N \omega_{i|i-1}^k(a_i)Q(a_i,a_{i+1})|\Lambda^k_{i|i-1}(a_{i:n})|^{-1/2}\exp\left\{\eta^k_{i|i-1}(a_{i:n})/2\right\}\delta_{a^k_{1:i-1}}(a_{1:i-1}).
+\widetilde{q}(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \widetilde{\vartheta}^k_{i-1}\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})\delta_{a^k_{1:i-1}}(a_{1:i-1})\eqsp,
 \]
-By integrating over all possible paths $a_{1:i-1}$, $\tilde{a}_i$ is sampled in $\{1,\ldots,J\}$.
-\subsection{Rao Blackwellised Two-filter Smoother (RBTF)}
+where $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ are adjustment multipliers and $\widetilde{q}(a_{i}|a^k_{1:i-1},a_{i+1:n},y_{1:n})$ is a proposal kernel chosen by the user.
+This means that an ancestral path $a^{\star}_{1:i-1}$ is sampled in $(a^k_{1:i-1})_{1\le k \le N}$ with weights $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ and $a^{\star}_i$ is sampled from $\widetilde{q}(\cdot|a^{\star}_{1:i-1},a_{i+1:n},y_{1:n})$.
+Then the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio.
+The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability \eqref{eq:posterior:FFBS} for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  \eqref{eq:posterior:FFBS} for all combinations of sequences $(a^{k}_{1:i})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$.
+Sampling from \eqref{eq:FFBS:rejuv} is computationally more intensive, especially when $N$ is large, but our method is based on a direct approximation of $p(a_{1:i}|a_{i+1:n},y_{1:n})$ based on  $(a^k_{1:i-1})_{1\le k \le N}$ and $a_{i+1:n}$ instead of approximate MCMC draws.
+
+\subsection{Rao-Blackwellized Two-filter Smoother}
 \label{sec:two-filer}
+\subsubsection{Rao-Blackwellized Two-filter Smoother of \cite{briers:doucet:maskell:2010}}
+\label{eq:rbtf}
 Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  regimes given all the observations).
@@
 \end{equation}
-The first term in the right hand side may be approximated using the forward filter, however, as the function $(a_{i},z_{i})\mapsto p_{}(y_{i:n}|a_{i},z_{i})$ is not a probability density function, it is not possible to use SMC samples to approximate it.
-The backward filter uses artificial densities to introduce a surrogate target density function which may be approximated recursively using SMC methods.
-Then, the forward and backward weighted samples are combined using \eqref{eq:dens:bwd_fwd_doucet} to approximate $p(a_{i},z_{i}|y_{1:n})$.
-\textbf{Additional particle rejuvenation step}\\
-for $2\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
-\begin{equation}
-\label{eq:dens:bwd_fwd}
-p(a_{i},z_{i}|y_{1:n})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-2}\rmd z_{i+1}\eqsp,
-\end{equation}
-where $\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-1$ and $i+1$.
-Note that the EM algorithm requires the approximation of $p(a_{i-1},z_{i-1},a_{i-1},z_{i-1}|y_{1:n})$ in the E-step, this may be obtained following the same steps by marginalizing explicitly the linear states at time $i-2$ and $i+1$, see Section~\ref{sec:numerical:experiments}.
-Intermediate computations follow the same steps as  for the approximation of $p(a_{i},z_{i}|y_{1:n})$.
-$\psi^n_{i}$ may be decomposed as follows:
-\begin{multline*}
-\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1}) \propto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})p_{}(a_{i-1},z_{i-1}|y_{1:i-1})Q(a_{i-1},a_{i})m_{}(a_{i},z_{i-1};z_{i})\\
-\times g_{}(a_{i},z_{i};y_{i})Q(a_{i},a_{i+1})m_{}(a_{i+1},z_{i};z_{i+1})\eqsp,
-\end{multline*}
-where the proportionality is with respect to $(a_{i-1:i+1},z_{i-1:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $p(a_{i},z_{i}|y_{1:n}) $ may be written as
-\begin{equation}
-\label{eq:smoothing}
-p(a_{i},z_{i}|y_{1:n})   \propto r_{i}(a_{i},z_{i},y_{1:i-1})g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
-\end{equation}
-where $m$ and $g$ are defined in \eqref{eq:definition-m} and \eqref{eq:definition-g} and
-\begin{align}
-r_{i}(a_{i},z_{i},y_{1:i-1}) &= p_{}(a_{i},z_{i}|y_{1:i-1})\eqsp,\\
-&=\sum_{a_{i-1}}\int_{z_{i-1}}p_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-1}\eqsp,\label{eq:smooth:r}\\
-t_{i}(a_i,z_i,y_{i+1:n}) &= \sum_{a_{i+1}}\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp.\label{eq:smooth:t}
-\end{align}
+The first term on the right hand side may be approximated using the forward filter by noting that:
+\[
+ p_{}(a_{i},z_{i}|y_{1:i-1}) =\sum_{a_{i-1}}\int_{z_{i-1}}p_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-1}\eqsp.
+\]
 In the forward pass described in Section~\ref{sec:forward}, a set of possible sequences of regimes $a_{1:i-1}^k$ associated with importance weights $\omega_{i-1}^k$, $1\le k\le N$ are sampled to approximate $p_{}(a_{i-1},z_{i-1}|y_{1:i-1})$.
-Plugging this approximation into \eqref{eq:smooth:r} provides a normalized approximation $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ of $r_{i}(a_{i},z_{i},y_{1:i-1})$ integrating over all possible choices of $(a_{i-1},z_{i-1})$:
-\[
-r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{a_{i-1}}\int_{z_{i-1}}p^N_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-1}\eqsp.
-\]
+This provides a normalized approximation $p^N(a_{i},z_{i}|y_{1:i-1})$ of $p(a_{i},z_{i}|y_{1:i-1})$.
 Define
@@
 \label{eq:rN}
-r^N_{i}(a_{i},z_{i},y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
+p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
 \end{equation}
 
-The backward pass described below produces a sequence of states $\tilde{a}_{i+1:n}^{\ell}$ associated with importance weights $\tilde{\omega}_{i+1}^{\ell}$, $1\le \ell\le N$ which are used to approximate $p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$.
-Plugging this approximation into \eqref{eq:smooth:t} provides a non normalized approximation $t^N_{i}(a_i,z_i,y_{i+1:n})$ of $t_{i}(a_i,z_i,y_{i+1:n})$ integrating over all possible choices $(a_{i+1},z_{i+1})$.
-
-These steps are then combined to form a non normalized SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ using \eqref{eq:smoothing}.
-As $t^N_{i}(a_i,z_i,y_{i+1:n})$ is not normalized, the normalization of the SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ is obtained by integrating over the states $a_{i},z_{i}$, when $r_{i}(a_{i},z_{i},y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $r^N_{i}(a_{i},z_{i},y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
-Our procedure allows to construct sequence of regimes with non-degenerated importance weights in the combination step.
-This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $p(a_{i},z_{i}|y_{1:n})$ over the states at times $i-1$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged.
-This method often leads to importance weights which are close to be numerically degenerated.
-
-As the function $(a_{i+1},z_{i+1})\mapsto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$ is not a probability density function, it is not possible to use SMC samples to approximate it. 
+As the function $(a_{i},z_{i})\mapsto p_{}(y_{i:n}|a_{i},z_{i})$ is not a probability density function, approximating the second term of \eqref{eq:dens:bwd_fwd_doucet}  using SMC samples is not straightforward.
+The backward filter uses artificial densities to introduce a surrogate target density function which may be approximated recursively using SMC methods.
+Then, the forward and backward weighted samples are combined using \eqref{eq:dens:bwd_fwd_doucet} to approximate $p(a_{i},z_{i}|y_{1:n})$.
 Following \cite{briers:doucet:maskell:2010}, for any probability densities $(\gamma^{}_{i})_{1\le i \le n}$, define the following joint probability densities:
@@
 \end{align*}
-By Lemma~\ref{eq:technical:twofilter}, these probability densities may be used to approximate the quantities $p_{}(y_{i:n}|a_{i},z_{i})$, $1\le i \le n$, in \eqref{eq:smooth:t}.
+Note that this choice differs slightly from \cite{briers:doucet:maskell:2010} where it is advocated to set $\gamma^{}_i$ as the product of two independent densities $\gamma_i^{a}(a_i)$ and $\gamma_i^{z}(z_i)$.
+As the accuracy of the algorithm relies heavily on a proper tuning of this artificial density, a more general choice of $\gamma^{}_i$ is considered in this paper.
+By Lemma~\ref{eq:technical:twofilter}, these probability densities may be used to approximate the quantities $p_{}(y_{i:n}|a_{i},z_{i})$, $1\le i \le n$, in \eqref{eq:dens:bwd_fwd_doucet}.
 \begin{lemma}
@@
 \begin{proof}
-The proof is postponed to Appendix~\ref{sec:app:proofs}.
+The proof is postponed to Appendix~\ref{sec:app}.
 \end{proof}
-A set of weighted trajectories $(\tilde{a}^\ell_{i:n})_{1\le \ell\le N}$ with importance weights $(\tilde{\omega}^\ell_i)_{1\le \ell\le N}$, $1\le i \le n$, may be sampled recursively backward in time to produce a SMC approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$.
+By definition of $\tilde{p}_i$ for all $1\le i \le n$,
+\begin{align*}
+\tilde{p}_{i}(a_{i:n},z_i|y_{i:n}) &\propto \gamma^{}_{i}(a_i,z_i) \int p_{}(y_{i:n}|a_{i:n},z_{i:n})p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)\rmd z_{i+1:n}\eqsp,\\
+&\propto\gamma^{}_{i}(a_i,z_i)\left\{\prod_{u=i}^{n-1}Q(a_u,a_{u+1})\right\}p(y_{i:n}|z_i,a_{i:n})\eqsp.
+\end{align*}
+This yields:
+\[
+\tilde{p}_{i}(a_{i:n}|y_{i:n}) \propto \left\{\prod_{u=i}^{n-1}Q(a_u,a_{u+1})\right\} \int \gamma^{}_{i}(a_i,z_i) p(y_{i:n}|z_i,a_{i:n})\rmd z_i\eqsp.
+\]
+A set of weighted trajectories $(\tilde{a}^\ell_{i:n})_{1\le \ell\le N}$ with importance weights $(\tilde{\omega}^\ell_i)_{1\le \ell\le N}$, $1\le i \le n$, may then be sampled recursively backward in time to produce a SMC approximation of $\tilde{p}_{}(a_{i:n}|y_{i:n})$ as follows.
 \begin{enumerate}[-]
@@
 \[
-\tilde{\omega}^{\ell}_i \propto \frac{q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
+\tilde{\omega}^{\ell}_i \propto \frac{Q(\tilde{a}^{\ell}_i,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}{\tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\tilde{a}^{\ell}_i)\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
 \]
@@
 \[
-\tilde{q}_n(\cdot) = \int\gamma^{}_n(\cdot,z')g_{}(\cdot,z';y_n)\rmd z'\quad\mbox{and}\quad \tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot) = \frac{q(\cdot,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\cdot,z')p_{}(y_{i:n}|(\cdot,\tilde{a}^{\ell}_{i+1:n}),z')\rmd z'}{\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
+\tilde{q}_n(\cdot) = \int\gamma^{}_n(\cdot,z')g_{}(\cdot,z';y_n)\rmd z'\quad\mbox{and}\quad \tilde{q}_i(\tilde{a}^{\ell}_{i+1:n},\cdot) = \frac{Q(\cdot,\tilde{a}^{\ell}_{i+1})\int \gamma^{}_i(\cdot,z')p_{}(y_{i:n}|(\cdot,\tilde{a}^{\ell}_{i+1:n}),z')\rmd z'}{\int \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1},z')p_{}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z')\rmd z'}\eqsp.
 \]
-By \eqref{eq:p:y1:n} and \eqref{eq:p:ai:zi}, we get
+By \eqref{eq:p:y1:n} and \eqref{eq:p:ai:zi}, 
 \begin{align*}
 p_{}(y_{i:n}|a_i,z_i)
-&= \frac{\tilde{p}_{i}(y_{i:n})  \tilde{p}_{i}(a_i,z_i|y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\eqsp, \\
-&= \tilde{p}_{i}(y_{i:n})  \sum_{a_{i+1:n}=1}^J\frac{\tilde{p}_{i}(a_{i:n}|y_{i:n})p_{}(y_{i:n}|a_{i:n},z_{i})}{\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'}\label{eq:p:ai:zi}\eqsp,
+&= \frac{\tilde{p}_{i}(y_{i:n})  \tilde{p}_{i}(a_i,z_i|y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}= \tilde{p}_{i}(y_{i:n})  \sum_{a_{i+1:n}=1}^J\frac{\tilde{p}_{i}(a_{i:n}|y_{i:n})p_{}(y_{i:n}|a_{i:n},z_{i})}{\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'}\label{eq:p:ai:zi}\eqsp,
 \end{align*}
-which suggests the following particle approximation $p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$:
-\[
-p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_{i}(y_{i:n}) \sum_{\ell=1}^N\frac{\tilde{\omega}^{\ell}_i p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z_{i})}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}\delta_{\tilde{a}^{\ell}_{i}}(a_{i})\eqsp.
-\]
+which suggests the following particle approximation $p^{N}_{}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$:
+\begin{equation}
+\label{eq:likelihoodTF}
+p^{N}_{}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_{i}(y_{i:n}) \sum_{\ell=1}^N\frac{\tilde{\omega}^{\ell}_i p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z_{i})}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'}\delta_{\tilde{a}^{\ell}_{i}}(a_{i})\eqsp.
+\end{equation}
 The conditional likelihood of the observations given the sequence of states $p_{}(y_{i:n}|a_{i:n},z_{i})$ can be computed explicitly using a Gaussian backward smoother; these computations are summarized in Lemma~\ref{lem:py}.
-\begin{lemma}
-\label{lem:py}
-For all $1\le i \le n$,
+In the numerical experiments, $\gamma^{}_i(a_i,z_i)$ is set as a mixture of Gaussian distributions.
+Note that for such a choice, the integral $\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'$ may be computed explicitly, see Lemma~\ref{lem:integral:gammap}.
+Then, combining \eqref{eq:likelihoodTF} and \eqref{eq:rN} with \eqref{eq:dens:bwd_fwd_doucet} provides an approximation of $p(a_{i},z_{i}|y_{1:n})$ by merging the forward particles $(a^k_{i-1})_{1\le k \le N}$ with the backward particles $(\tilde{a}^k_{i+1})_{1\le k \le N}$, the support of this SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ being $(\tilde{a}^k_{i+1})_{1\le k \le N}$.
+
+As noted in \cite[Secion~2.6]{fearnhead:wyncoll:tawn:2010}, two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$.
+The authors illustrate this issue in the case where the hidden state is an AR(2) process.
+To overcome the weakness of such standard two-filter approaches the particle rejuvenation proposed in Section~\ref{sec:tf:rejuv} follows the idea introduced in \cite{fearnhead:wyncoll:tawn:2010} where new particles at time $i$ are sampled conditional on $(a^k_{1:i-1})_{1\le k \le N}$ and on $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ and appropriately weighted.
+This allows to produce new particles at time $i$ and to obtain a SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ whose support is not restricted to $(\tilde{a}^k_{i+1})_{1\le k \le N}$.
+Section~\ref{sec:tf:rejuv} exploits this idea in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps.
+
+\subsubsection{Particle rejuvenation of two-filter based algorithms}
+\label{sec:tf:rejuv}
+For $2\le i \le n-1$, particle rejuvenation relies on the explicit marginalization:
 \begin{equation}
-\label{eq:py}
-p_{}(y_{i:n}|a_{i:n},z_i) =  \exp\left\{-\frac{1}{2}\tilde{c}_{i}(a_{i:n}) - \frac{1}{2}\normMat{\tilde{P}_{i}(a_{i:n})}{z_i}^2 + z'_i\tilde{\nu}_{i}(a_{i:n})\right\}\eqsp,
+\label{eq:dens:bwd_fwd}
+p(a_{i},z_{i}|y_{1:n})  =  \sum_{a_{i-1}}\sum_{a_{i+1}} \int_{z_{i-1}} \int_{z_{i+1}} \psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})\rmd z_{i-1}\rmd z_{i+1}\eqsp,
 \end{equation}
-where
-\begin{align}
-\tilde{c}_{n}(a_{n}) &= \dimy\log(2\pi) + \log \left|\barG_{a_n}\right| + \normMat{\barG_{a_n}}{y_n-c_{a_n}}^2\eqsp,\label{eq:cn}\\
-\tilde{P}_{n}^{-1}(a_{n}) &=B'_{a_n}\barG_{a_n}^{-1}B_{a_n}\eqsp,\label{eq:Pn}\\
-\tilde{\nu}_{n}(a_{n})&=B'_{a_n}\barG_{a_n}^{-1}(y_n-c_{a_n})\label{eq:nun}
-\end{align}
-and, for all $1\le i\le n-1$,
-\begin{align}
-\tilde{c}_{i}(a_{i:n}) &= \tilde{c}_{i|i+1}(a_{i:n}) + \dimy\log(2\pi)  + \log|\barG_{a_i}| + \normMat{\barG_{a_i}}{y_i-c_{a_i}}^2\eqsp,\label{eq:ck}\\
-\tilde{P}_{i}^{-1}(a_{i:n}) &= \tilde{P}_{i|i+1}^{-1}(a_{i:n}) + B'_{a_i}\barG_{a_i}^{-1}B_{a_i}\eqsp,\label{eq:Pk}\\
-\tilde{\nu}_{i}(a_{i:n})&=\tilde{\nu}_{i|i+1}(a_{i:n}) + B'_{a_i}\barG_{a_i}^{-1}(y_{i}-c_{a_{i}})\eqsp,\label{eq:nuk}
-\end{align}
-with
-\begin{align*}
-\Delta_{i}(a_{i+1:n}) &= \left(I_{\dimz} + H'_{a_{i+1}}\tilde{P}_{i+1}^{-1}(a_{i+1:n})H_{a_{i+1}}\right)^{-1}\eqsp,\\
-\tilde{r}_{i|i+1}(a_{i+1:n})&= \tilde{\nu}_{i+1}(a_{i+1:n}) + \barH_{a_{i+1}}^{-1}d_{a_{i+1}}\eqsp,\\
-\tilde{c}_{i|i+1}(a_{i+1:n}) &= \tilde{c}_{i+1}(a_{i+1:n}) + \log|\barH_{a_{i+1}}| + d'_{a_{i+1}}\barH_{a_{i+1}}^{-1}d_{a_{i+1}}-\log|H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}|\\
-&\hspace{4.0cm} - \tilde{r}'_{i|i+1}(a_{i+1:n})H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\tilde{r}_{i|i+1}(a_{i+1:n})\eqsp,\\
-\tilde{P}_{i|i+1}^{-1}(a_{i+1:n}) &=T'_{a_{i+1}}\left(I_{\dimz}-\barH_{a_{i+1}}^{-1}H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\right)\barH_{a_{i+1}}^{-1}T_{a_{i+1}}\eqsp,\\
-\tilde{\nu}_{i|i+1}(a_{i+1:n})&=T'_{a_{i+1}}\barH_{a_{i+1}}^{-1}\left[-d_{a_{i+1}}+H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\left(\tilde{\nu}_{i+1}(a_{i+1:n})+\barH_{a_{i+1}}^{-1}d_{a_{i+1}}\right)\right]\eqsp.
-\end{align*}
-\end{lemma}
-\begin{proof}
-The proof is postponed to Appendix~\ref{sec:app:proofs}.
-\end{proof}
-By Lemma~\ref{lem:py}, the SMC approximation $p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$ is then given by
+where $\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1})$ is the smoothing distribution of the hidden regimes and states between time indices $i-1$ and $i+1$.
+Note that the EM algorithm requires the approximation of $p(a_{i-1},z_{i-1},a_{i-1},z_{i-1}|y_{1:n})$ in the E-step, this may be obtained following the same steps by marginalizing explicitly the linear states at time $i-2$ and $i+1$.
+Intermediate computations follow the same steps as  for the approximation of $p(a_{i},z_{i}|y_{1:n})$.
+First, $\psi^n_{i}$ may be decomposed as follows:
+\begin{multline*}
+\psi^n_{i}(a_{i-1:i+1},z_{i-1:i+1}) \propto p_{}(y_{i+1:n}|a_{i+1},z_{i+1})p_{}(a_{i-1},z_{i-1}|y_{1:i-1})Q(a_{i-1},a_{i})m_{}(a_{i},z_{i-1};z_{i})\\
+\times g_{}(a_{i},z_{i};y_{i})Q(a_{i},a_{i+1})m_{}(a_{i+1},z_{i};z_{i+1})\eqsp,
+\end{multline*}
+where the proportionality is with respect to $(a_{i-1:i+1},z_{i-1:i+1})$. Then, by \eqref{eq:dens:bwd_fwd}, the smoothing distribution $p(a_{i},z_{i}|y_{1:n}) $ may be written as
+\begin{equation}
+\label{eq:smoothing}
+p(a_{i},z_{i}|y_{1:n})   \propto p(a_{i},z_{i}|y_{1:i-1})g_{}(a_{i},z_{i};y_{i})t_{i}(a_i,z_i,y_{i+1:n})\eqsp,
+\end{equation}
+where $m$ and $g$ are defined in \eqref{eq:definition-m} and \eqref{eq:definition-g} and
+\begin{equation}
+%r_{i}(a_{i},z_{i},y_{1:i-1}) &= p_{}(a_{i},z_{i}|y_{1:i-1})\eqsp,\\
+%&=\sum_{a_{i-1}}\int_{z_{i-1}}p_{}(a_{i-1},z_{i-1}|y_{1:i-1})m_{}(a_{i},z_{i-1};z_{i})Q(a_{i-1},a_{i}) \rmd z_{i-1}\eqsp,\label{eq:smooth:r}
+t_{i}(a_i,z_i,y_{i+1:n}) = \sum_{a_{i+1}}\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp.\label{eq:smooth:t}
+\end{equation}
+The backward pass described in Section~\ref{eq:rbtf} produces a sequence of states $\tilde{a}_{i+1:n}^{\ell}$ associated with importance weights $\tilde{\omega}_{i+1}^{\ell}$, $1\le \ell\le N$ which are used to approximate $p_{}(y_{i+1:n}|a_{i+1},z_{i+1})$.
+Plugging this approximation into \eqref{eq:smooth:t} provides an approximation $t^N_{i}(a_i,z_i,y_{i+1:n})$ of $t_{i}(a_i,z_i,y_{i+1:n})$ integrating over all possible choices $(a_{i+1},z_{i+1})$.
+These steps are then combined to form a non normalized SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ using \eqref{eq:smoothing}.
+The normalization of the SMC approximation of $p(a_{i},z_{i}|y_{1:n})$ is obtained by integrating over the states $a_{i},z_{i}$, when $p(a_{i},z_{i}|y_{1:i-1})$ and $t_{i}(a_i,z_i,y_{i+1:n})$ are replaced by $p^N(a_{i},z_{i}|y_{1:i-1})$ and $t^N_{i}(a_i,z_i,y_{i+1:n})$ in \eqref{eq:smoothing}.
+Our procedure allows to construct sequence of regimes with non-degenerated importance weights in the combination step.
+This procedure improves significantly \cite{briers:doucet:maskell:2010} where no marginalization of $p(a_{i},z_{i}|y_{1:n})$ over the states at times $i-1$ and $i+1$ is performed and where the proposed forward and backward paths are directly merged.
+This method often leads to importance weights which are close to be numerically degenerated.
+By Lemma~\ref{lem:py}, the SMC approximation $p^{N}_{}(y_{i:n}|a_{i},z_{i})$ of $p_{}(y_{i:n}|a_{i},z_{i})$ is then given by:
 \begin{equation}
 \label{eq:pN:yi:n}
-p^{\mathsf{two-filt}}_{N}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_i(y_{i:n}) \sum_{\ell=1}^N  \frac{\delta_{\tilde{a}^{\ell}_i}(a_i)\tilde{\omega}_{i}^{\ell}}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'} \exp\left\{- \frac{1}{2}\normMat{\tilde{P}^{\ell}_{i}}{z_i}^2 + z'_i\tilde{\nu}^{\ell}_{i} - \frac{1}{2} \tilde{c}^{\ell}_i \right\}\eqsp,
+p^{N}_{}(y_{i:n}|a_{i},z_{i}) = \tilde{p}_i(y_{i:n}) \sum_{\ell=1}^N  \frac{\delta_{\tilde{a}^{\ell}_i}(a_i)\tilde{\omega}_{i}^{\ell}}{\int \gamma^{}_{i}(\tilde{a}^{\ell}_i,z')p_{}(y_{i:n}|\tilde{a}^{\ell}_{i:n},z')\rmd z'} \exp\left\{- \frac{1}{2}\normMat{\tilde{P}^{\ell}_{i}}{z_i}^2 + z'_i\tilde{\nu}^{\ell}_{i} - \frac{1}{2} \tilde{c}^{\ell}_i \right\}\eqsp,
 \end{equation}
-where $(\tilde{P}^{\ell}_{i})^{-1} \eqdef \tilde{P}_{i}^{-1}(\tilde{a}^{\ell}_{i:n})$, $\tilde{\nu}^{\ell}_{i}\eqdef \tilde{\nu}_{i}(\tilde{a}^{\ell}_{i:n})$, $\tilde{c}^{\ell}_{i} \eqdef \tilde{c}^{\ell}_{i}(\tilde{a}^{\ell}_{i:n})$.
+where $(\tilde{P}^{\ell}_{i})^{-1} \eqdef \tilde{P}_{i}^{-1}(\tilde{a}^{\ell}_{i:n})$, $\tilde{\nu}^{\ell}_{i}\eqdef \tilde{\nu}_{i}(\tilde{a}^{\ell}_{i:n})$ and $\tilde{c}^{\ell}_{i} \eqdef \tilde{c}^{\ell}_{i}(\tilde{a}^{\ell}_{i:n})$ are defined in Lemma~\ref{lem:py}.
 Define
 \begin{align*}
-\Delta^{\ell}_{i} &\eqdef \left(I + H'_{\tilde{a}^\ell_{i+1}}(\tilde{P}^{\ell}_{i+1})^{-1}H_{\tilde{a}^\ell_{i+1}}\right)^{-1}\eqsp,\\
-\delta^{\ell}_{i}&\eqdef\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}(d_{\tilde{a}^\ell_{i+1}}+T_{\tilde{a}^\ell_{i+1}}z_i)\eqsp.
+\Delta^{\ell}_{i+1} &\eqdef \left(I_{\dimz} + H'_{\tilde{a}^\ell_{i+1}}(\tilde{P}^{\ell}_{i+1})^{-1}H_{\tilde{a}^\ell_{i+1}}\right)^{-1}\eqsp,\\
+\delta^{\ell}_{i+1}&\eqdef\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}(d_{\tilde{a}^\ell_{i+1}}+T_{\tilde{a}^\ell_{i+1}}z_i)\eqsp.
 \end{align*}
@@
 t^N_{i}(a_i,z_i,y_{i+1:n}) &=\sum_{a_{i+1}=1}^J\int_{z_{i+1}}m_{}(a_{i+1},z_{i};z_{i+1})Q(a_{i},a_{i+1})p^N_{}(y_{i+1:n}|a_{i+1},z_{i+1}) \rmd z_{i+1}\eqsp,\nonumber\\
-&=  \tilde{p}_{i+1}(y_{i+1:n}) \sum_{\ell=1}^N C_i^{-1}(\tilde{a}^{\ell}_{i+1:n}) Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_i H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2} \eqsp\nonumber \\
-&\hspace{2.3cm} \times \exp\left\{\frac{1}{2} (\delta^{\ell}_i)'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}} \delta^{\ell}_i -\frac{1}{2}\normMat{\barH_{\tilde{a}^{\ell}_{i+1}}}{d_{\tilde{a}^{\ell}_{i+1}}+T_{\tilde{a}^{\ell}_{i+1}}z_i}^2\right\} \eqsp,\\
+&=  \tilde{p}_{i+1}(y_{i+1:n}) \sum_{\ell=1}^N C_i^{-1}(\tilde{a}^{\ell}_{i+1:n}) Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_{i+1} H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2} \eqsp\nonumber \\
+&\hspace{2cm} \times \exp\left\{\frac{1}{2} (\delta^{\ell}_{i+1})'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_{i+1}H'_{\tilde{a}^{\ell}_{i+1}} \delta^{\ell}_{i+1} -\frac{1}{2}\normMat{\barH_{\tilde{a}^{\ell}_{i+1}}}{d_{\tilde{a}^{\ell}_{i+1}}+T_{\tilde{a}^{\ell}_{i+1}}z_i}^2\right\} \eqsp,\nonumber\\
 &= \sum_{\ell=1}^N\tilde{\omega}_{\mathsf{b},i}^\ell(a_i)\exp\left\{-\frac{1}{2}\normMat{\tilde{S}_{i+1}^\ell}{z_i}^2+z_i'\tilde{s}_{i+1}^\ell\right\}\eqsp,\label{eq:tN}
@@
 C_i(\tilde{a}^{\ell}_{i+1:n}) &\eqdef \exp \left\{\tilde{c}^{\ell}_{i+1}/2\right\}  \int_{z_{i+1}} \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1}, z)\tilde{p}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z) \rmd z\eqsp,\\
-\tilde{\omega}_{\mathsf{b},i}^\ell(a_i) &= \tilde{p}_{i+1}(y_{i+1:n})C_i(\tilde{a}^{\ell}_{i+1:n})^{-1}Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_i H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2}\\
-&\times\exp\{-d'_{\tilde{a}^\ell_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}}/2\}\exp\{(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}}(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})/2\} \eqsp,\\
-(\tilde{S}_{i+1}^\ell)^{-1} &= T'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}(T_{\tilde{a}^{\ell}_{i+1}}-H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}T_{\tilde{a}^{\ell}_{i+1}})\eqsp,\\
-\tilde{s}_{i+1}^\ell &=T'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}(H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_iH'_{\tilde{a}^{\ell}_{i+1}}(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})-d_{\tilde{a}^\ell_{i+1}})\eqsp.
+\tilde{\omega}_{\mathsf{b},i}^\ell(a_i) &= \tilde{p}_{i+1}(y_{i+1:n})C_i(\tilde{a}^{\ell}_{i+1:n})^{-1}Q(a_i, \tilde{a}^{\ell}_{i+1}) \tilde{\omega}^{\ell}_{i+1} |\barH_{\tilde{a}^{\ell}_{i+1}}|^{-1/2} |H_{\tilde{a}^{\ell}_{i+1}} \Delta^{\ell}_{i+1} H'_{\tilde{a}^{\ell}_{i+1}}|^{1/2}\\
+&\times\exp\{-d'_{\tilde{a}^\ell_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}}/2\}\exp\{(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})'H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_{i+1}H'_{\tilde{a}^{\ell}_{i+1}}(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})/2\} \eqsp,\\
+(\tilde{S}_{i+1}^\ell)^{-1} &= T'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}(T_{\tilde{a}^{\ell}_{i+1}}-H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_{i+1}H'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}T_{\tilde{a}^{\ell}_{i+1}})\eqsp,\\
+\tilde{s}_{i+1}^\ell &=T'_{\tilde{a}^{\ell}_{i+1}}\barH_{\tilde{a}^\ell_{i+1}}^{-1}(H_{\tilde{a}^{\ell}_{i+1}}\Delta^{\ell}_{i+1}H'_{\tilde{a}^{\ell}_{i+1}}(\tilde{\nu}^{\ell}_{i+1} + \barH_{\tilde{a}^\ell_{i+1}}^{-1}d_{\tilde{a}^\ell_{i+1}})-d_{\tilde{a}^\ell_{i+1}})\eqsp.
 \end{align*}
-In the experiments (see section~\ref{sec:exp}), $\gamma^{}_i(a_i,z_i)$ is set as a mixture of Gaussian distributions. Note that for such a choice,
- the integral $\int \gamma^{}_{i}(a_i,z')p_{}(y_{i:n}|a_{i:n},z')\rmd z'$ may be computed explicitly; see Lemma~\ref{lem:integral:gammap}.
-This choice differs from \cite{briers:doucet:maskell:2010} where it is advocated to set $\gamma^{}_i$ as the product of two independent densities $\gamma_i^{a}(a_i)$ and $\gamma_i^{z}(z_i)$.
-The accuracy of the algorithm relies heavily on a proper tuning of this artificial density.
-
-\begin{lemma}
-\label{lem:integral:gammap}
-For all $1\le i\le n$,
-\begin{multline*}
-\int \phi_{\mu_i,\Sigma_i}(z_i)p_{}(y_{i:n}|a_{i:n},z_i)\rmd z_{i} = \exp\left\{-\frac{1}{2}\log |\Sigma_i| - \frac{1}{2}\mu_i'\Sigma^{-1}_i\mu_i\right\}\\
-\times\exp\left\{-\frac{1}{2}\tilde{c}_{i}(a_{i:n})+\frac{1}{2}\log|\tilde{\Omega}_{i}(a_{i:n})| +\frac{1}{2}\tilde{z}'_i(a_{i:n})\tilde{\Omega}_{i}(a_{i:n})\tilde{z}_i(a_{i:n})\right\}\eqsp,
-\end{multline*}
-where $\phi_{\mu,\Sigma}$ is the probability density function of a $\dimz$ dimensional Gaussian random variable with mean $\mu$ and variance matrix $\Sigma$ and
-\[
-\tilde{\Omega}_{i}(a_{i:n})\eqdef \left(\Sigma_i^{-1} + \tilde{P}_{i}^{-1}(a_{i:n})\right)^{-1}\quad\mbox{and}\quad
-\tilde{z}_i(a_{i:n})\eqdef \Sigma_i^{-1}\mu_i+\tilde{\nu}_{i}(a_{i:n})
-\]
-and where $c_{i}$, $\tilde{P}_{i}$ and $\nu_{i}$ are given in Lemma~\ref{lem:py}.
-\end{lemma}
-\begin{proof}
-The proof is postponed to Appendix~\ref{sec:app:proofs}.
-\end{proof}
+In the numerical experiments, $\gamma^{}_i(a_i,z_i)$ is set as a mixture of Gaussian distributions.
+As explained in Section~\ref{eq:rbtf}, the integral $\int_{z_{i+1}} \gamma^{}_{i+1}(\tilde{a}^{\ell}_{i+1}, z)\tilde{p}(y_{i+1:n}|\tilde{a}^{\ell}_{i+1:n},z) \rmd z$ may be computed explicitly, see Lemma~\ref{lem:integral:gammap}.
 
@@
 \label{sec:numerical:experiments}
-This section highlights the improvements brought by  the additional Rao-Blackwellisation steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$.
-In each setting the transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data;  see Section~\ref{sec:exp}.
-First, the FFBS algorithm is applied to a simple one-dimensional model with:
+This section highlights the improvements brought by  the additional Rao-Blackwellization steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$.
+In each setting the transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data, see Section~\ref{sec:exp}.
+First, the algorithms are applied to a simple one-dimensional model with:
 \[
-\pi(1)=\pi(2) = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
+\pi_1=\pi_2 = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
 \]
@@
 \]
-The two FFBS algorithms are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles.
+The original FFBS algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015} and the FFBS algorithm with rejuvenation proposed in this paper are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $5000$ particles considered as a benchmark value.
 Figure~\ref{fig:ffbs:err} displays the mean estimation error over $100$ independent Monte Carlo runs.
-The estimation error is defined as the difference between the the estimation obtained with the proposed FFBS algorithm with rejuvenation and $2000$ particles and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
+The estimation error is defined as the difference between the benchmark value and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
 \begin{figure}
@@
 \end{figure}
-
 Then, the performance of the two-filter based algorithms are compared in two different settings with $n=1500$ observations.
@@
 \[
-\pi(1)=\pi(2) = 1/2\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
+\pi_1=\pi_2 = 1/2\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
 \]
@@
 \]
-The approximations of the smoothing distributions are computed with $N=10$.
-RBTF and the algorithm of \cite{briers:doucet:maskell:2010} are compared to the approximations provided by RBTF with $1000$ particles.
+In this case, the benchmark value is given by the two-filter algorithm with rejuvenation and $N=1000$ particles.
+The two-filter approximations of the smoothing distributions are computed with $N=10$.
 
@@
 The smoothing distributions are estimated with $N=15$ particles.
-Both methods are compared to the approximations provided by RBTF with $1000$ particles.
 The specific form of $B_1$ (first row with entries equal to one) and the fact that $T_1 = T_2$, $B_1 = B_2 $ and $\barG_1 = \barG_2$ are also motivated by the application to WTI crude oil data.
 It is assumed that the differences between the two regimes are given by the state's dynamics ($d_1$, $d_2$, $\barH_1$ and $\barH_2$).
-Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using:
-\[
-\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
-\]
-In both cases, RBTF behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
+Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using $\gamma^{}_i (a_i, z_i) = p^N(a_{i},z_{i}|y_{1:i-1})$ where $p^N(a_{i},z_{i}|y_{1:i-1})$ is defined by \eqref{eq:rN}.
+%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
+%\end{equation}:
+%\[
+%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
+%\]
+In both cases, the two-filter algorithm with rejuvenation behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
 \begin{figure}
@@
 \]
-where $\xi_i$ is a standard multivariate Gaussian random variable and:
+where $\eta_i$ is a standard multivariate Gaussian random variable and:
 \[
@@
 \begin{equation*}
-p_{\theta}(a_{1:n},z_{1:n},Y_{1:n}) \eqdef \pi(a_1)\phi_{\mu_1,\Sigma_1}(z_1)g_{\theta}(a_1,z_1;Y_1)\prod^{n}_{i=2}Q(a_{i-1},a_i)m_{\theta}\left(a_{i},z_{i-1};z_i\right)g_{\theta}(a_i,,z_i;Y_i)\eqsp,
+p_{\theta}(a_{1:n},z_{1:n},Y_{1:n}) \eqdef \pi(a_1)\phi_{\mu_1,\Sigma_1}(z_1)g_{\theta}(a_1,z_1;Y_1)\prod^{n}_{i=2}Q(a_{i-1},a_i)m_{\theta}\left(a_{i},z_{i-1};z_i\right)g_{\theta}(a_i,,z_i;Y_i)\eqsp.
 \end{equation*}
-where $\phi_{\mu_1,\Sigma_1}$ is the Gaussian probability density function with mean $\mu_1$ and variance $\Sigma_1$.
+%where $\phi_{\mu_1,\Sigma_1}$ is the Gaussian probability density function with mean $\mu_1$ and variance $\Sigma_1$.
 Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{1:n}\right]$  the conditional expectation given $Y_{1:n}$ when the parameter value is set to $\theta$.
@@
 \begin{enumerate}
-	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p})\eqdef \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{1:n},Y_{1:n})\middle|Y_{1:n}\right]$ ;
+	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p})\eqdef \mathbb{E}_{\theta_p}\left[\log p_{\theta}(a_{1:n},Z_{1:n},Y_{1:n})\middle|Y_{1:n}\right]$ ;
 	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
@@
 \subsection{Numerical results}
-The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi(1) = \pi(2) = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime.
+The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi_1 = \pi_2 = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime.
 The other parameters are initialized as shown in the following table:
@@
 Therefore, this condition is imposed for all simulations in the CMA-ES algorithm.
-The results after $2500$ iterations of the EM algorithm are given in Figure~\ref{fig:CMEfinalestimates}.
+The results after $2500$ iterations of the EM algorithm are given in Table~\ref{fig:CMEfinalestimates}.
 \begin{table}[h!]
@@
 	\end{tabular}
-	\caption{Final estimates after 3500 iterations.}
+	\caption{Final estimates after 2500 iterations.}
 	\label{fig:CMEfinalestimates}
 \end{table}
-
 As expected, we obtain $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$ at convergence of the EM algorithm.
@@
 
-Figure~\ref{fig:posterior:data} and~\ref{fig:posterior:data:logprice} displays the the estimated posterior probabilities of the regimes and the observed future slope.
+Figure~\ref{fig:posterior:data} and~\ref{fig:posterior:data:logprice} display the the estimated posterior probabilities of the regimes and the observed future slope.
 When the future curve is in backwardation (resp.
@@
 \section{Conclusions}
-In this paper, we proposed a survey of Rao Blackwellised Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces.
-These methods combine particle based approximations to sample the sequence of hidden regimes while the hidden linear states are explicitly integrated conditional on the sequence of regimes and observations.
-They  are based on the forward backward decomposition of the smoothing distributions using either two-filter based algorithms or forward filtering backward smoothing techniques mimicking the Rauch-Tung-Striebel smoother.
-The paper presents these algorithms in a common unifying framework and also provides a survey of different techniques that could be used in the forward filtering pass which may improve significantly the usual mixture Kalman filter.
+This paper presents Rao-Blackwellized Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces in a common unifying framework and also provides a survey of different techniques that could be used in the forward filtering pass to improve significantly the usual mixture Kalman filter.
 The filtering distributions are approximated at  each time step by considering all possible offsprings of all ancestral trajectories before discarding degenerated paths instead of resampling the ancestral paths before propagating them at the next time step.
-We also investigated the benefit of additional Rao Blackwellisation steps which allow to sample new regimes at each time step conditional on the forward and backward particles.
+We also investigated the benefit of additional Rao-Blackwellization steps to sample new regimes at each time step conditional on the forward and backward particles.
 This rejuvenation step uses explicit integration of the hidden linear states before merging the forward and backward filters for two-filter based algorithms or before sampling new states backward in time for FFBS based methods.
@@
 \section{Technical lemmas}
+\label{sec:app}
+Lemmas~\ref{eq:technical:twofilter}, \ref{lem:py} and \ref{lem:integral:gammap} are close to \cite[Proposition 5, Proposition 6]{briers:doucet:maskell:2010}.
+The proofs are detailed in this appendix for completeness.
+\begin{proof}[Proof of Lemma~\ref{eq:technical:twofilter}]
+For all $1\le i \le n-1$,
+\begin{align*}
+p_{}(y_{i:n}|a_i,z_i) &= \sum_{a_{i+1:n}}\int p_{}(y_{i:n},a_{i+1:n},z_{i+1:n}|a_i,z_i)\rmd z_{i+1:n}\eqsp,\\
+&=  \sum_{a_{i+1:n}}\int p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})\rmd z_{i+1:n}\eqsp,\\
+&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\sum_{a_{i+1:n}}\int \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})\rmd z_{i+1:n}\eqsp,\\
+&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\sum_{a_{i+1:n}}\int \tilde{p}_{i}(a_{i:n},z_{i:n}|y_{i:n})\rmd z_{i+1:n}\eqsp,\\
+&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\tilde{p}_{i}(a_{i},z_{i}|y_{i:n})\eqsp,
+\end{align*}
+which concludes the proof of \eqref{eq:p:y1:n}. To prove \eqref{eq:p:ai:zi} write,
+\begin{align*}
+\tilde{p}_{i}(a_{i:n},z_{i}|y_{i:n}) &= \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\int p_{}(y_{i:n}|a_{i:n},z_{i:n})p_{}(a_{i+1:n},z_{i+1:n}|a_{i},z_{i})\rmd z_{i+1:n}\eqsp,\\
+&= \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\int \frac{p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(z_{i+1:n}|y_{i:n},a_{i:n},z_i)}{p_{}(z_{i+1:n}|a_{i:n},z_i)}p_{}(a_{i+1:n},z_{i+1:n}|a_{i},z_{i})\rmd z_{i+1:n}\eqsp,\\
+&=\frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(a_{i+1:n}|a_i)\eqsp.
+\end{align*}
+Therefore,
+\[
+\tilde{p}_{i}(a_{i},z_{i}|y_{i:n}) = \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\sum_{a_{i+1:n}}p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(a_{i+1:n}|a_i)
+\]
+and the proof is completed upon noting that
+\[
+\tilde{p}_{i}(a_{i:n}|y_{i:n}) = \frac{p_{}(a_{i+1:n}|a_i)}{\tilde{p}_{i}(y_{i:n})} \int \gamma^{}_{i}(a_i,z)p_{}(y_{i:n}|a_{i:n},z)\rmd z\eqsp.
+\]
+\end{proof}
+
+\begin{lemma}
+\label{lem:py}
+For all $1\le i \le n$,
+\begin{equation}
+\label{eq:py}
+p_{}(y_{i:n}|a_{i:n},z_i) =  \exp\left\{-\frac{1}{2}\tilde{c}_{i}(a_{i:n}) - \frac{1}{2}\normMat{\tilde{P}_{i}(a_{i:n})}{z_i}^2 + z'_i\tilde{\nu}_{i}(a_{i:n})\right\}\eqsp,
+\end{equation}
+where
+\begin{align}
+\tilde{c}_{n}(a_{n}) &= \dimy\log(2\pi) + \log \left|\barG_{a_n}\right| + \normMat{\barG_{a_n}}{y_n-c_{a_n}}^2\eqsp,\label{eq:cn}\\
+\tilde{P}_{n}^{-1}(a_{n}) &=B'_{a_n}\barG_{a_n}^{-1}B_{a_n}\eqsp,\label{eq:Pn}\\
+\tilde{\nu}_{n}(a_{n})&=B'_{a_n}\barG_{a_n}^{-1}(y_n-c_{a_n})\label{eq:nun}
+\end{align}
+and, for all $1\le i\le n-1$,
+\begin{align}
+\tilde{c}_{i}(a_{i:n}) &= \tilde{c}_{i|i+1}(a_{i+1:n}) + \dimy\log(2\pi)  + \log|\barG_{a_i}| + \normMat{\barG_{a_i}}{y_i-c_{a_i}}^2\eqsp,\label{eq:ck}\\
+\tilde{P}_{i}^{-1}(a_{i:n}) &= \tilde{P}_{i|i+1}^{-1}(a_{i+1:n}) + B'_{a_i}\barG_{a_i}^{-1}B_{a_i}\eqsp,\label{eq:Pk}\\
+\tilde{\nu}_{i}(a_{i:n})&=\tilde{\nu}_{i|i+1}(a_{i+1:n}) + B'_{a_i}\barG_{a_i}^{-1}(y_{i}-c_{a_{i}})\eqsp,\label{eq:nuk}
+\end{align}
+with
+\begin{align*}
+\Delta_{i+1}(a_{i+1:n}) &= \left(I_{\dimz} + H'_{a_{i+1}}\tilde{P}_{i+1}^{-1}(a_{i+1:n})H_{a_{i+1}}\right)^{-1}\eqsp,\\
+\tilde{r}_{i|i+1}(a_{i+1:n})&= \tilde{\nu}_{i+1}(a_{i+1:n}) + \barH_{a_{i+1}}^{-1}d_{a_{i+1}}\eqsp,\\
+\tilde{c}_{i|i+1}(a_{i+1:n}) &= \tilde{c}_{i+1}(a_{i+1:n}) + \log|\barH_{a_{i+1}}| + d'_{a_{i+1}}\barH_{a_{i+1}}^{-1}d_{a_{i+1}}-\log|H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}|\\
+&\hspace{4.0cm} - \tilde{r}'_{i|i+1}(a_{i+1:n})H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\tilde{r}_{i|i+1}(a_{i+1:n})\eqsp,\\
+\tilde{P}_{i|i+1}^{-1}(a_{i+1:n}) &=T'_{a_{i+1}}\left(I_{\dimz}-\barH_{a_{i+1}}^{-1}H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\right)\barH_{a_{i+1}}^{-1}T_{a_{i+1}}\eqsp,\\
+\tilde{\nu}_{i|i+1}(a_{i+1:n})&=T'_{a_{i+1}}\barH_{a_{i+1}}^{-1}\left[-d_{a_{i+1}}+H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\left(\tilde{\nu}_{i+1}(a_{i+1:n})+\barH_{a_{i+1}}^{-1}d_{a_{i+1}}\right)\right]\eqsp.
+\end{align*}
+\end{lemma}
+
+\begin{proof}
+The result is proved by backward induction.
+\eqref{eq:cn}, \eqref{eq:Pn} and \eqref{eq:nun} follow directly from \eqref{eq:model:obs}.
+Assume that for a given $1\le i\le n-1$, $p(y_{i+1:n}|a_{i+1:n},z_{i+1})$ is given by \eqref{eq:py}.
+Write
+\[
+p(y_{i:n}|a_{i:n},z_i) = \int m(a_{i+1},z_i;z_{i+1})g(a_i,z_i;y_i)p(y_{i+1:n}|a_{i+1:n},z_{i+1})\rmd z_{i+1}\eqsp,
+\]
+with
+\begin{align*}
+m(a_{i+1},z_i;z_{i+1}) &= \exp\left\{-\frac{\dimz}{2}\log(2\pi)-\frac{1}{2}\log|\barH_{a_{i+1}}|-\frac{1}{2}\normMat{\barH_{a_{i+1}}}{z_{i+1}-d_{a_{i+1}}-T_{a_{i+1}}z_i}^2\right\}\eqsp,\\
+g(a_i,z_i;y_i)& = \exp\left\{-\frac{\dimy}{2}\log(2\pi)-\frac{1}{2}\log|\barG_{a_i}|-\frac{1}{2}\normMat{\barG_{a_i}}{y_{i}-c_{a_{i}}-B_{a_i}z_i}^2\right\}\eqsp,\\
+p(y_{i+1:n}|a_{i+1:n},z_{i+1})& = \exp\left\{-\frac{1}{2}c_{i+1}(a_{i+1:n}) - \frac{1}{2}\normMat{\tilde{P}_{i+1}(a_{i+1:n})}{z_{i+1}}^2 + z'_{i+1}\tilde{\nu}_{i+1}(a_{i+1:n})\right\}\eqsp.
+\end{align*}
+Let $\Delta_{i+1}$ and  $\delta_{i+1}$ be given by:
+\begin{align*}
+\Delta_{i+1}(a_{i+1:n}) &\eqdef \left(I_{\dimz} + H'_{a_{i+1}}\tilde{P}_{i+1}^{-1}(a_{i+1:n})H_{a_{i+1}}\right)^{-1}\eqsp,\\
+\delta_{i+1}(a_{i+1:n}) &\eqdef \nu_{i+1}(a_{i+1:n}) + \barH_{a_{i+1}}^{-1}(d_{a_{i+1}}+T_{a_{i+1}}z_i)\eqsp.
+\end{align*}
+Then, $\barH_{a_{i+1}}^{-1} + \tilde{P}_{i+1}^{-1}(a_{i+1:n}) = \left(H_{a_{i+1}}\Delta_{i+1}(a_{i+1:n})H'_{a_{i+1}}\right)^{-1}$ and \eqref{eq:ck}, \eqref{eq:Pk} and \eqref{eq:nuk} follows from
+\begin{multline*}
+\int \exp\left\{-\frac{1}{2}\normMat{H_{a_{i+1}}\Delta_{i+1}(a_{i+1:n})H'_{a_{i+1}}}{z_{i+1}}^2 + z'_{i+1}\delta_{i+1}(a_{i+1:n})\right\}\rmd z_{i+1} \\
+=\exp\left\{\frac{1}{2}\log(2\pi) + \frac{1}{2}\log|H_{a_{i+1}}\Delta_{i+1}(a_{i+1:n})H'_{a_{i+1}}|\right\}\\
+\times\exp\left\{\frac{1}{2}\delta'_{i+1}(a_{i+1:n})'H_{a_{i+1}}\Delta_{i+1}(a_{i+1:n})H'_{a_{i+1}}\delta_{i+1}(a_{i+1:n})\right\}\eqsp.
+\end{multline*}
+\end{proof}
+
+\begin{lemma}
+\label{lem:integral:gammap}
+For all $1\le i\le n$,
+\begin{multline*}
+\int \phi_{\mu_i,\Sigma_i}(z_i)p_{}(y_{i:n}|a_{i:n},z_i)\rmd z_{i} = \exp\left\{-\frac{1}{2}\log |\Sigma_i| - \frac{1}{2}\mu_i'\Sigma^{-1}_i\mu_i\right\}\\
+\times\exp\left\{-\frac{1}{2}\tilde{c}_{i}(a_{i:n})+\frac{1}{2}\log|\tilde{\Omega}_{i}(a_{i:n})| +\frac{1}{2}\tilde{z}'_i(a_{i:n})\tilde{\Omega}_{i}(a_{i:n})\tilde{z}_i(a_{i:n})\right\}\eqsp,
+\end{multline*}
+where $\phi_{\mu,\Sigma}$ is the probability density function of a $\dimz$ dimensional Gaussian random variable with mean $\mu$ and variance matrix $\Sigma$ and
+\[
+\tilde{\Omega}_{i}(a_{i:n})\eqdef \left(\Sigma_i^{-1} + \tilde{P}_{i}^{-1}(a_{i:n})\right)^{-1}\quad\mbox{and}\quad
+\tilde{z}_i(a_{i:n})\eqdef \Sigma_i^{-1}\mu_i+\tilde{\nu}_{i}(a_{i:n})
+\]
+and where $c_{i}$, $\tilde{P}_{i}$ and $\nu_{i}$ are given in Lemma~\ref{lem:py}.
+\end{lemma}
+\begin{proof}
+By Lemma~\ref{lem:py},
+\begin{multline*}
+\phi_{\mu_i,\Sigma_i}(z_i)p(y_{i:n}|a_{i:n},z_i) = \exp\left\{-\frac{\dimz}{2}\log(2\pi) -\frac{1}{2}\log|\Sigma_i|-\frac{1}{2}\normMat{\Sigma_i}{z_i-\mu_i}^2\right\}\\
+\times\exp\left\{-\frac{1}{2}c_{i}(a_{i:n})-\frac{1}{2}\normMat{\tilde{P}_{i}(a_{i:n})}{z_i}^2+z'_i\nu_{i}(a_{i:n})\right\}\eqsp.
+\end{multline*}
+The proof is completed noting that
+\begin{multline*}
+\int \exp\left\{-\frac{1}{2}z'_i\tilde{\Omega}_{i}^{-1}(a_{i:n})z_i + z'_i(\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n}))\right\}\rmd z_{i}\\
+= \exp\left\{\frac{\dimz}{2}\log(2\pi) + \frac{1}{2}\log|\tilde{\Omega}_{i}(a_{i:n})| +\frac{1}{2}\left[\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n})\right]'\tilde{\Omega}_{i}(a_{i:n})\left[\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n})\right]\right\}\eqsp.
+\end{multline*}\end{proof}
+
 \begin{lemma}
@@
 The proof is concluded upon noting that $\tilde{W}^1_t$ and $\tilde{W}^2_t$ are centered Gaussian random variables such that:
-\begin{itemize}
+\begin{enumerate}[-]
 \item $\!\!\mathrm{Var}\left[\tilde{W}^1_t\right] = \sigma^2  t + \eta^2\left\{ t + (1-e^{-2\kappa t})/(2\kappa) - 2(1-e^{-\kappa  t})/\kappa\right\}/\kappa^2 - 2\rho \eta \sigma\left\{ t - (1-e^{-\kappa  t})/\kappa \right\}/\kappa\eqsp,$
@@
 \item $\!\!\mathrm{Cov}\left[\tilde{W}^1_t,\tilde{W}^2_t\right] = \left(\rho \eta\sigma-\eta^2/\kappa\right)\left(1-e^{-\kappa t} \right)/\kappa + \eta^2 \left(1-e^{-2\kappa t} \right)/(2\kappa^2)\eqsp.$
-\end{itemize}
-\end{proof}
-
-\label{sec:app:proofs}
-Lemmas~\ref{eq:technical:twofilter}, \ref{lem:py} and \ref{lem:integral:gammap} are close to \cite[Proposition 5, Proposition 6]{briers:doucet:maskell:2010}.
-The proofs are detailed in this appendix for completeness.
-\begin{proof}[Proof of Lemma~\ref{eq:technical:twofilter}]
-For all $1\le i \le n-1$,
-\begin{align*}
-p_{}(y_{i:n}|a_i,z_i) &= \sum_{a_{i+1:n}}\int p_{}(y_{i:n},a_{i+1:n},z_{i+1:n}|a_i,z_i)\rmd z_{i+1:n}\eqsp,\\
-&=  \sum_{a_{i+1:n}}\int p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})\rmd z_{i+1:n}\eqsp,\\
-&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\sum_{a_{i+1:n}}\int \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}p_{}(a_{i+1:n},z_{i+1:n}|a_i,z_i)p_{}(y_{i:n}|a_{i:n},z_{i:n})\rmd z_{i+1:n}\eqsp,\\
-&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\sum_{a_{i+1:n}}\int \tilde{p}_{i}(a_{i:n},z_{i:n}|y_{i:n})\rmd z_{i+1:n}\eqsp,\\
-&=  \frac{\tilde{p}_{i}(y_{i:n})}{\gamma^{}_{i}(a_i,z_i)}\tilde{p}_{i}(a_{i},z_{i}|y_{i:n})\eqsp,
-\end{align*}
-which concludes the proof of \eqref{eq:p:y1:n}. To prove \eqref{eq:p:ai:zi} write,
-\begin{align*}
-\tilde{p}_{i}(a_{i:n},z_{i}|y_{i:n}) &= \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\int p_{}(y_{i:n}|a_{i:n},z_{i:n})p_{}(a_{i+1:n},z_{i+1:n}|a_{i},z_{i})\rmd z_{i+1:n}\eqsp,\\
-&= \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\int \frac{p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(z_{i+1:n}|y_{i:n},a_{i:n},z_i)}{p_{}(z_{i+1:n}|a_{i:n},z_i)}p_{}(a_{i+1:n},z_{i+1:n}|a_{i},z_{i})\rmd z_{i+1:n}\eqsp,\\
-&=\frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(a_{i+1:n}|a_i)\eqsp.
-\end{align*}
-Therefore,
-\[
-\tilde{p}_{i}(a_{i},z_{i}|y_{i:n}) = \frac{\gamma^{}_{i}(a_i,z_i)}{\tilde{p}_{i}(y_{i:n})}\sum_{a_{i+1:n}}p_{}(y_{i:n}|a_{i:n},z_{i})p_{}(a_{i+1:n}|a_i)
-\]
-and the proof is completed upon noting that
-\[
-\tilde{p}_{i}(a_{i:n}|y_{i:n}) = \frac{p_{}(a_{i+1:n}|a_i)}{\tilde{p}_{i}(y_{i:n})} \int \gamma^{}_{i}(a_i,z)p_{}(y_{i:n}|a_{i:n},z)\rmd z\eqsp.
-\]
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lem:py}]
-The result is proved by backward induction.
-\eqref{eq:cn}, \eqref{eq:Pn} and \eqref{eq:nun} follow directly from \eqref{eq:model:obs}.
-Assume that for a given $1\le i\le n-1$, $p(y_{i+1:n}|a_{i+1:n},z_{i+1})$ is given by \eqref{eq:py}.
-Write
-\[
-p(y_{i:n}|a_{i:n},z_i) = \int p(z_{i+1}|z_i,a_{i+1})p(y_i|z_i,a_i)p(y_{i+1:n}|a_{i+1:n},z_{i+1})\rmd z_{i+1}\eqsp,
-\]
-with
-\begin{align*}
-p(z_{i+1}|z_i,a_{i+1}) &= \exp\left\{-\frac{\dimz}{2}\log(2\pi)-\frac{1}{2}\log|\barH_{a_{i+1}}|-\frac{1}{2}\normMat{\barH_{a_{i+1}}}{z_{i+1}-d_{a_{i+1}}-T_{a_{i+1}}z_i}^2\right\}\eqsp,\\
-p(y_i|z_i,a_i)& = \exp\left\{-\frac{\dimy}{2}\log(2\pi)-\frac{1}{2}\log|\barG_{a_i}|-\frac{1}{2}\normMat{\barG_{a_i}}{y_{i}-c_{a_{i}}-B_{a_i}z_i}^2\right\}\eqsp,\\
-p(y_{i+1:n}|a_{i+1:n},z_{i+1})& = \exp\left\{-\frac{1}{2}c_{i+1}(a_{i+1:n}) - \frac{1}{2}\normMat{\tilde{P}_{i+1}(a_{i+1:n})}{z_{i+1}}^2 + z'_{i+1}\tilde{\nu}_{i+1}(a_{i+1:n})\right\}\eqsp.
-\end{align*}
-Let $\Delta_{i}$ and  $\delta_{i}$ be given by:
-\begin{align*}
-\Delta_{i}(a_{i+1:n}) &\eqdef \left(I_{\dimz} + H'_{a_{i+1}}\tilde{P}_{i+1}^{-1}(a_{i+1:n})H_{a_{i+1}}\right)^{-1}\eqsp,\\
-\delta_{i}(a_{i+1:n}) &\eqdef \nu_{i+1}(a_{i+1:n}) + \barH_{a_{i+1}}^{-1}(d_{a_{i+1}}+T_{a_{i+1}}z_i)\eqsp.
-\end{align*}
-Then, $\barH_{a_{i+1}}^{-1} + \tilde{P}_{i+1}^{-1}(a_{i+1:n}) = \left(H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\right)^{-1}$ and \eqref{eq:ck}, \eqref{eq:Pk} and \eqref{eq:nuk} follows from
-\begin{multline*}
-\int \exp\left\{-\frac{1}{2}\normMat{H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}}{z_{i+1}}^2 + z'_{i+1}\delta_{i}(a_{i+1:n})\right\}\rmd z_{i+1}\\
- =\exp\left\{\frac{1}{2}\log(2\pi) + \frac{1}{2}\log|H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}|+\frac{1}{2}\delta'_{i}(a_{i+1:n})'H_{a_{i+1}}\Delta_{i}(a_{i+1:n})H'_{a_{i+1}}\delta_{i}(a_{i+1:n})\right\}\eqsp.
-\end{multline*}
-\end{proof}
-
-\begin{proof}[Proof of Lemma~\ref{lem:integral:gammap}]
-By Lemma~\ref{lem:py},
-\begin{multline*}
-\phi_{\mu_i,\Sigma_i}(z_i)p(y_{i:n}|a_{i:n},z_i) = \exp\left\{-\frac{\dimz}{2}\log(2\pi) -\frac{1}{2}\log|\Sigma_i|-\frac{1}{2}\normMat{\Sigma_i}{z_i-\mu_i}^2\right\}\\
-\times\exp\left\{-\frac{1}{2}c_{i}(a_{i:n})-\frac{1}{2}\normMat{\tilde{P}_{i}(a_{i:n})}{z_i}^2+z'_i\nu_{i}(a_{i:n})\right\}\eqsp.
-\end{multline*}
-The proof is completed noting that
-\begin{multline*}
-\int \exp\left\{-\frac{1}{2}z'_i\tilde{\Omega}_{i}^{-1}(a_{i:n})z_i + z'_i(\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n}))\right\}\rmd z_{i}\\
-= \exp\left\{\frac{\dimz}{2}\log(2\pi) + \frac{1}{2}\log|\tilde{\Omega}_{i}(a_{i:n})| +\frac{1}{2}\left[\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n})\right]'\tilde{\Omega}_{i}(a_{i:n})\left[\Sigma_i^{-1}\mu_i+\nu_{i}(a_{i:n})\right]\right\}\eqsp.
-\end{multline*}
+\end{enumerate}
 \end{proof}

### 2017-05-14 TexArxiv/clgm.tex
New simulations for revision

@@
 \[
-\pi_1=\pi_2 = 1/2\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= c_2 = 0 \eqsp,
+\pi_1=\pi_2 = 0.5\\;\;d_1 = 0.5\;\; d_2 = 0;\;c_1= 0.1\;\; c_2 = 0 \eqsp,
 \]
 \[
-Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T_1 = T_2 = 1\;\;\barH_1 = \barH_2= 0.05\eqsp,
+Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T_1 = T_2 = 1\;\;\barH_1 = \barH_2= 0.1\eqsp,
 \]
@@
 \]
-The original FFBS algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015} and the FFBS algorithm with rejuvenation proposed in this paper are used with $N = \tilde{N} = 50$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$ are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $5000$ particles considered as a benchmark value.
+The original FFBS algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015} and the FFBS algorithm with rejuvenation proposed in this paper are used with $N = \tilde{N} = 25$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$.
+For comparable computational costs, the two-filter method of \cite{briers:doucet:maskell:2010} and the method with rejuvenation are run with $N=100$.
+All these algorithms are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $5000$ particles considered as a benchmark value.
 Figure~\ref{fig:ffbs:err} displays the mean estimation error over $100$ independent Monte Carlo runs.
-The estimation error is defined as the difference between the benchmark value and the estimation given by the two FFBS algorithms with $N = \tilde{N} = 50$.
+The estimation error is defined as the absolute difference between the benchmark value and the estimations given by all algorithms.
 \begin{figure}
 \centering
-\includegraphics[scale=.3]{err_est_ffbs}
-\caption{Posterior probabilities estimation error for the two FFBS algorithms.}
+\includegraphics[scale=.3]{err_est}
+\caption{Posterior probabilities estimation error for all algorithms.}
 \label{fig:ffbs:err}
 \end{figure}
-In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for the two algorithms to illustrate that both estimators have similar variability. 
+In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for each algorithm to illustrate that both estimators have similar variability. 
 \begin{figure}
 \centering
-\includegraphics[scale=.3]{var}
-\caption{Empirical variances of the estimation for the two FFBS algorithms.}
+\includegraphics[scale=.3]{var2}
+\caption{Empirical variances of the estimation of $\mathbb{P}(a_k=1|Y_{1:n})$ for all algorithms.}
 \label{fig:ffbs:var}
 \end{figure}
-Then, the performance of the two-filter based algorithms are compared in two different settings with $n=1500$ observations.
-\paragraph{Experiment 1:}
-\[
-\pi_1=\pi_2 = 1/2\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
-\]
-\[
-Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T = \begin{pmatrix}1 & -0.02\\ 0 & 0.96\end{pmatrix}\;\;\barH_1 = \begin{pmatrix}0.25 & 0.01\\ 0.01 & 0.01\end{pmatrix}\;\;\barH_2 = \begin{pmatrix}0.3 & 0\\ 0 & 0.3\end{pmatrix}\eqsp,
-\]
-\[
-B = \begin{pmatrix}1 & 1 & 1 & 1\\ 0 & 0.5 & 0.5 & 0.1 \end{pmatrix}'\;\; \barG = \mathrm{diag}\left(0.1,0.1,0.1,0.1\right)\;\; \mu_1 = \begin{pmatrix}1 \\ 0\end{pmatrix}\;\; \Sigma_1 = \begin{pmatrix} 5 & 0 \\ 0  & 5\end{pmatrix}\eqsp.
-\]
-In this case, the benchmark value is given by the two-filter algorithm with rejuvenation and $N=1000$ particles.
-The two-filter approximations of the smoothing distributions are computed with $N=10$.
-
-\paragraph{Experiment 2:}
-In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$.
+%Then, the performance of the two-filter based algorithms are compared in two different settings with $n=1500$ observations.
+%\paragraph{Experiment 1:}
+%\[
+%\pi_1=\pi_2 = 0.5\\;\;d_1 = \begin{pmatrix}0 \\ 0.5 \end{pmatrix}\;\; d_2 = \begin{pmatrix}0 \\ 0\end{pmatrix}\\;\;c_1= \begin{pmatrix}0 & 0& 0 & 0 \end{pmatrix}'\;\;c_2 =\begin{pmatrix}0 & 0 & 0 & 0 \end{pmatrix}'\eqsp,
+%\]
+%\[
+%Q = \begin{pmatrix}0.99 & 0.01\\ 0.03 & 0.97\end{pmatrix}\;\;T = \begin{pmatrix}1 & -0.02\\ 0 & 0.96\end{pmatrix}\;\;\barH_1 = \begin{pmatrix}0.25 & 0.01\\ 0.01 & 0.01\end{pmatrix}\;\;\barH_2 = \begin{pmatrix}0.3 & 0\\ 0 & 0.3\end{pmatrix}\eqsp,
+%\]
+%\[
+%B = \begin{pmatrix}1 & 1 & 1 & 1\\ 0 & 0.5 & 0.5 & 0.1 \end{pmatrix}'\;\; \barG = \mathrm{diag}\left(0.1,0.1,0.1,0.1\right)\;\; \mu_1 = \begin{pmatrix}1 \\ 0\end{pmatrix}\;\; \Sigma_1 = \begin{pmatrix} 5 & 0 \\ 0  & 5\end{pmatrix}\eqsp.
+%\]
+%In this case, the benchmark value is given by the two-filter algorithm with rejuvenation and $N=1000$ particles.
+%The two-filter approximations of the smoothing distributions are computed with $N=10$. 
+%
+%\paragraph{Experiment 2:}
+%In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$.
 The smoothing distributions are estimated with $N=15$ particles.
@@
 Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using $\gamma^{}_i (a_i, z_i) = p^N(a_{i},z_{i}|y_{1:i-1})$ where $p^N(a_{i},z_{i}|y_{1:i-1})$ is defined by \eqref{eq:rN}.
-%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
-%\end{equation}:
-%\[
-%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
-%\]
-In both cases, the two-filter algorithm with rejuvenation behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
-\begin{figure}
-\centering
-\includegraphics[scale=.5]{N10_1_500}
-\includegraphics[scale=.5]{N10_501_1000}
-\includegraphics[scale=.5]{N10_1001_1500}
-\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
-\label{fig:exp1:label}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.5]{N15_1_500}
-\includegraphics[scale=.5]{N15_501_1000}
-\includegraphics[scale=.5]{N15_1001_1500}
-\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
-\label{fig:exp2:label}
-\end{figure}
+%%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
+%%\end{equation}:
+%%\[
+%%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
+%%\]
+%In both cases, the two-filter algorithm with rejuvenation behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.5]{N10_1_500}
+%\includegraphics[scale=.5]{N10_501_1000}
+%\includegraphics[scale=.5]{N10_1001_1500}
+%\caption{Experiment 1 - estimated probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=10$ particles,  RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
+%\label{fig:exp1:label}
+%\end{figure}
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.5]{N15_1_500}
+%\includegraphics[scale=.5]{N15_501_1000}
+%\includegraphics[scale=.5]{N15_1001_1500}
+%\caption{Experiment 2 - estimated smoothing probabilities by \cite{briers:doucet:maskell:2010} (blue) and RBTF (red) with $N=15$ particles, RBTF with $N=1000$ particles (black line) and true regimes (squared black).}
+%\label{fig:exp2:label}
+%\end{figure}
 

### 2017-05-15 TexArxiv/answer_referees.tex
Final modifications for revision

@@
 This allows to produce new particles at time $i$ and to obtain a SMC approximation whose support is not restricted to the backward particles at time $i+1$ as in the original two-filter algorithm.
-The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps.
+The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$.
+In addition, it produces a direct approximation of the target distribution based on $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$  without any additional importance sampling steps.
 
-On the other hand, another modification of the FFBS algorithm based on a Markov chain Monte Carlo sampling step has been introduced in [LBS+16].
+On the other hand, another modification of the FFBS algorithm based on a Markov chain Monte Carlo sampling step was introduced in [LBS+16].
 Instead of sampling from  explicitly from (12), [LBS+16] proposed to draw a forward path $a_{1:i-1}$ in $(a^k_{1:i-1})_{1\le k \le N}$ and a sate $a_i$ in $\{1,\ldots,J\}$ according to:
@@
 The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability (11) for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  (11) for all combinations of sequences $(a^{k}_{1:i})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$.
-The method proposed in [LBS+16], is computationally more efficient as its complexity depends mainly on the number of MCMC samples produced by the algorithm.
+The method proposed in [LBS+16] is computationally less costly as its complexity depends mainly on the number of MCMC samples produced by the algorithm.
 However it also requires a fair amount of tuning to choose efficient adjustment multipliers and proposal kernels.
@@
 As explained in the previous paragraph, Section~2.3.1 is now devoted to the original Rao-Blackwellized two-filter method and Section~2.3.2 to the rejuvenation step.
+The overall idea supporting two-filter based algorithms is clearly presented in Section~2.3.1 and most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities.
 Some additional motivations to introduce rejuvenation are given at the end of Sections~2.2.2 and~2.3.2 (see the answer to the associate editor).
@@
 The particle rejuvenation proposed in our paper exploits the idea introduced in [FWT10] in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$ and without any additional sampling steps.
-Most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities.
 }
@@
 {\em
-\noindent Some additional numerical experiments were added in Section~3.
-For each simulation study which was previously given in Section~3 for either FFBS or two-filter methods, now both algorithms are used and compared to their counterpart with rejuvenation.
+\noindent Even is the paper is not a survey of all Rao-Blackwellized methods for CLGM anymore, some additional numerical experiments were added in Section~3.
+Each simulation study which was previously given in Section~3 was devoted either to FFBS or two-filter based methods.
+Now, both algorithms are used and compared to their counterpart with rejuvenation.
 Our main objective is then to show the improvement of the particle rejuvenation steps for both methods.
-This also confirms that FFBS algorithm perform better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of teach method.
+This also confirms that FFBS algorithm perform better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of each method.
 }
@@
 
-\noindent All minor comments were taken into consideration and typos were corrected. Some specific details are given below.
+\noindent The minor comments were taken into consideration and typos were corrected. Some specific details are given below.
 
@@
 A first (costly) run of the FFBS algorithm with rejuvenation with a great number of particles ($N=5000$) is used as a benchmark value.
-Then the posterior distributions are estimated by the original methods and the methods with rejuvenation with $N=50$ particles are compared to this  benchmark run.
+Then the posterior distributions are estimated by the original methods and the methods with rejuvenation with fewer particles and compared to this  benchmark run.
+}
+
+\vspace{.5cm}
+
+\noindent Figure 1 shall be Table 1? In the text it is 2500 iterations, in the Table
+caption it is 3500? How is Std. Dev obtained? I mean is this simulated data or real data where
+the true parameters $\theta$ are usually unknown?
+
+\vspace{.2cm}
+
+\noindent {\em 
+The estimated values were indeed obtained after 2500 iterations of the algorithm.
+The standard deviations are obtained with 50 independent Monte Carlo runs of the estimation algorithm (and the estimated value is set as the mean over all runs).
+This is now stated clearly in the paper.
 }
@@
 \noindent As explained to the associate editor and to  Referee~1, it is true that particle smoothers are prone to degenerate when they are based on  the genealogy of the forward or backward particles.
-In addition, in the specific case of two-filter based smoothers, [FWT10, Secion~2.6] noted that these algorithms are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$.
+In addition, in the specific case of two-filter based smoothers, [FWT10, Section~2.6] noted that these algorithms are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$.
 The particle rejuvenation proposed in our paper produces an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$.
 In the specific case of linear and Gaussian models, Kalman filtering techniques allow to compute explicitly the approximation of the target distribution at time $i$ based on  $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$.
-The support of the approximated distribution is not restricted to $(\tilde{a}^k_{i})_{1\le k \le N}$ to overcome particle depletion.
-The marginalization over the linear states  builds new trajectories $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters.
-Although this does not overcome the overall degeneracy issue, numerical experiments highlight that this improves the performance of the algorithms.
+
+Therefore, the support of the approximated distribution is not restricted to $(\tilde{a}^k_{i})_{1\le k \le N}$ to overcome particle depletion.
+The marginalization over the linear states  builds new trajectories of the form $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters.
+Although this does not overcome the overall degeneracy issue, numerical experiments highlight that this improves the performance of all Rao-Blackwellized SMC smoothers.
+Additional comments were added at the end of Section~2.2 and Section~2.3.1.
 }
@@
 However, our numerical experiments hint that this actually improves the accuracy and variance of the proposed estimators which would be even more interesting with additional regimes or additional rejuvenation steps.
+[FWT10, Section~2.6] illustrated the degeneracy issue of two-filter based smoothers with the example of an AR(2) process.
+Our rejuvenation exploits the idea introduced in [FWT10, Section~2.6] to overcome degeneracy issues in the context of CLGM.
 In addition, the additional computational cost is not prohibitive as Kalman filtering steps are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$ (but who is stil necessary when no rejuvenation step is introduced).
@@
 {\em
-\noindent The identity was added in Section~2.3.1 in a more detailed derivation of the backward filter.
+\noindent The identity was added in Section~2.3.1 in a more detailed derivation of the backward filter. We thank Referee~2 for this help to improve the presentation of the backward filter.
 }
@@
 {\em
-\noindent The additional computational cost of the rejuvenation step comes from the Kalman filtering steps which are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$.
+\noindent The additional computational cost of the rejuvenation step comes from the Kalman filtering steps needed to extend all backward trajectories (built with particles).
+These steps are much faster than the combination of the forward and backward particles whose complexity grows with $N^2$.
 However, this combination is also necessary in the original two-filter algorithm of [BDM10] with no rejuvenation.
+A comment was added in the numerical section.
 }

### 2017-05-15 TexArxiv/clgm.tex
Final modifications for revision

@@
 %The two-filter approximations of the smoothing distributions are computed with $N=10$. 
-%
-%\paragraph{Experiment 2:}
-%In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$.
+%%
+%%\paragraph{Experiment 2:}
+%%In this second experiment, we use the same parameters except that we increase by a factor $10$ the values in $\barG$.
 The smoothing distributions are estimated with $N=15$ particles.
@@
 Figures~\ref{fig:exp1:label} and~\ref{fig:exp2:label} displays the approximations of the  posterior distributions of the regimes given by the algorithms using $\gamma^{}_i (a_i, z_i) = p^N(a_{i},z_{i}|y_{1:i-1})$ where $p^N(a_{i},z_{i}|y_{1:i-1})$ is defined by \eqref{eq:rN}.
-%%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
-%%\end{equation}:
-%%\[
-%%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
-%%\]
+%%%p^N(a_{i},z_{i}|y_{1:i-1}) = \sum_{k=1}^N \omega_{\mathsf{f},i}^k(a_{i})\exp\left\{-\frac{1}{2}\normMat{\Omega^k_{i-1}(a_{i})}{z_{i}}^2+z'_{i}r_{i-1}^k(a_{i})\right\}\eqsp.
+%%%\end{equation}:
+%%%\[
+%%%\gamma^{}_i (a_i, z_i) = \sum_{k=1}^{N} Q(a^{k}_{i-1},a_i) \omega^{k}_{i-1} \exp\left\{ -\frac{1}{2} \log \left|\barH_{a^k_{i-1}} \right| -\frac{1}{2}\normMat{\barH_{a^k_{i-1}}}{z_i - d_{a^k_{i-1}} -T \mu^k_{i-1} }^2\right\} \eqsp.
+%%%\]
 %In both cases, the two-filter algorithm with rejuvenation behaves slightly better than \cite{briers:doucet:maskell:2010} which does not detect some regime switchings (for instance  at time 780 in Figure~\ref{fig:exp1:label} and at time 300 in Figure~\ref{fig:exp2:label}) or gives false detections (for instance at time 10 in Figure~\ref{fig:exp2:label}).
@@
 where the parameter $\left(r, \sigma, \kappa, \alpha, \eta, \rho \right)$ are constant and $( (W_t^1,W_t^2), t \geq 0)$ are standard Brownian motions.
-This model appear to be too restrictive as energy markets are not likely to revert to a single equilibrium value.
+This model appears to be too restrictive as energy markets are not likely to revert to a single equilibrium value.
 This assumption is relaxed using Markov switching models to allow several possible regimes for the spot price and the convenience yield.
@@
 where $(a_t)_{t\ge 0}$ is a finite state space Markov process.
-
 This model allows to exhibit fundamental features of commodity future prices, which typically display different regimes of volatility and/or convenience yield.
@@
 Assuming that the switching rate between regimes is negligible compared to the inverse of the discretization period, the discretized version of the spot price and convenience yield $Z_i= (\ln S_i, \delta_i)$ (the sampling period is taken to be 1) is modeled as a CLGM.
-The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 0}$:
+The explicit integration of this SDE detailed in Lemma~\ref{lem:integratedSDE} yields the following discrete time model for $(Z_i)_{i\ge 2}$:
 \[
@@
 \end{enumerate}
-All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our RBTF algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
+All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our two-filter algorithm with rejuvenation algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
 As the function $\theta\mapsto Q^N(\theta,\theta_p)$ cannot be maximized analytically, the M-step is performed numerically using the Covariance Matrix Adaptation Evolution Strategy (CMA-ES) introduced in \cite{hansen:ostermeier:2001}.
@@
 The results after $2500$ iterations of the EM algorithm are given in Table~\ref{fig:CMEfinalestimates}.
+The estimated values and standard deviations are obtained with $50$ independent runs of the algorithm.
 \begin{table}[h!]
@@
 As expected, we obtain $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$, $\eta_1 \geq \eta_2$ and $\rho_1 \geq \rho_2$ at convergence of the EM algorithm.
-Moreover, $Q(1,1) > Q(2,2)$ is corresponding to the prediction that we did from the data description.
-From the results, $\sigma_1 \geq \sigma_2$, $\alpha_1 \geq \alpha_2$ indicates the first regime (backwardation) characterized by a higher value in both volatility and equilibrium level of convenience yield, and the second regime (contango) characterized by a lower value in both volatility and equilibrium convenience yield level.
+Moreover, $Q(1,1) > Q(2,2)$ corresponds to the prediction that we did from the data description.
+The fact that $\sigma_1 \geq \sigma_2$ and $\alpha_1 \geq \alpha_2$ indicates the first regime (backwardation) characterized by a higher value in both volatility and equilibrium level of convenience yield, and the second regime (contango) characterized by a lower value in both volatility and equilibrium convenience yield level.
 This in accordance with the theory of storage that the volatility of the commodity spot price is high when the inventory is low, and the convenience yield is all the higher as inventory is low.
@@
 \includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
-\caption{Log-price (red line) and slope of future curves (blue line)}
+\caption{Log-price (red line) and slope of future curves (blue line).}
 \label{fig:LogPriceTermStructure:data}
@@
 The correlation between the spot price and the convenience yield is positive and high in both two regimes.
-This is an accordance to what as been observed in most commodity market; see \cite{gibson:schwartz:1990}.
+This is an accordance to what as been observed in most commodity market, see \cite{gibson:schwartz:1990}.
 The slope of future curve decreases in function of maturity.
@@
 	\includegraphics[scale=.75,angle=90]{Posterior_Data}
-	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
+	\caption{Posterior probability (triangle black) and slope of future curves (blue line).}
 	\label{fig:posterior:data}
@@
 	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
-	\caption{Posterior probability (triangle black) and slope of future curves (blue line)}
+	\caption{Posterior probability (triangle black) and slope of future curves (blue line).}
 	\label{fig:posterior:data:logprice}
@@
 \section{Conclusions}
-This paper presents Rao-Blackwellized Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces in a common unifying framework and also provides a survey of different techniques that could be used in the forward filtering pass to improve significantly the usual mixture Kalman filter.
+This paper presents Rao-Blackwellized Sequential Monte Carlo methods to approximate smoothing distributions in conditionally linear and Gaussian state spaces in a common unifying framework.
+It also provides different techniques that could be used in the forward filtering pass to improve significantly the usual mixture Kalman filter.
 The filtering distributions are approximated at  each time step by considering all possible offsprings of all ancestral trajectories before discarding degenerated paths instead of resampling the ancestral paths before propagating them at the next time step.
-We also investigated the benefit of additional Rao-Blackwellization steps to sample new regimes at each time step conditional on the forward and backward particles.
+The paper investigates the benefit of additional Rao-Blackwellization steps to sample new regimes at each time step conditional on the forward and backward particles.
 This rejuvenation step uses explicit integration of the hidden linear states before merging the forward and backward filters for two-filter based algorithms or before sampling new states backward in time for FFBS based methods.
-The paper displays some Monte Carlo experiments with simulated data to illustrate that this additional rejuvenation step improves the performance of the smoothing algorithms with no additional computational costs or variance.
+The paper displays some Monte Carlo experiments with simulated data to illustrate that this additional rejuvenation step improves the performance of the smoothing algorithms with no substantial additional computational costs.
 They are also applied to commodity markets using WTI crude oil data.

### 2017-05-15 TexArxiv/answer_referees.tex
Final modifications for revision

@@
 This allows to produce new particles at time $i$ and to obtain a SMC approximation whose support is not restricted to the backward particles at time $i+1$ as in the original two-filter algorithm.
-The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allows to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$.
+The particle rejuvenation proposed in our paper exploits this idea in the specific case of linear and Gaussian models where explicit computations allow to produce an approximation using $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$ with support $\{1,\ldots,J\}$.
 In addition, it produces a direct approximation of the target distribution based on $(a^k_{1:i-1})_{1\le k \le N}$ and $(\tilde{a}^k_{i+1:n})_{1\le k \le N}$  without any additional importance sampling steps.
@@
 This means that an ancestral path $a^{\star}_{1:i-1}$ is sampled in $(a^k_{1:i-1})_{1\le k \le N}$ with weights $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ and $a^{\star}_i$ is sampled from $\widetilde{q}(\cdot|a^{\star}_{1:i-1},a_{i+1:n},y_{1:n})$.
-Then the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio.
-The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability (11) for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  (11) for all combinations of sequences $(a^{k}_{1:i})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$.
+Then, the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio.
+The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability (11) for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  (11) for all combinations of sequences $(a^{k}_{1:i-1})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$.
 The method proposed in [LBS+16] is computationally less costly as its complexity depends mainly on the number of MCMC samples produced by the algorithm.
@@
 As explained in the previous paragraph, Section~2.3.1 is now devoted to the original Rao-Blackwellized two-filter method and Section~2.3.2 to the rejuvenation step.
-The overall idea supporting two-filter based algorithms is clearly presented in Section~2.3.1 and most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities.
-Some additional motivations to introduce rejuvenation are given at the end of Sections~2.2.2 and~2.3.2 (see the answer to the associate editor).
+The idea supporting two-filter based algorithms is clearly presented in Section~2.3.1 and most of the computations have also been postponed to the appendix in order to highlight the overall picture and avoid technicalities.
+Some additional motivations to introduce rejuvenation are given at the end of Sections~2.2 and~2.3.2 (see the answer to the associate editor).
 Two-filter smoothers are prone to suffer from degeneracy issues when the algorithm associates forward particles at time $i-1$ with backward particles at time $i$.
@@
 Our main objective is then to show the improvement of the particle rejuvenation steps for both methods.
-This also confirms that FFBS algorithm perform better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of each method.
+This also confirms that FFBS algorithm performs better even with a sharp choice for the  $\gamma_i$'s but also that rejuvenation steps improve both the accuracy and the variance of each method.
 }
@@
 
-\noindent The minor comments were taken into consideration and typos were corrected. Some specific details are given below.
+\noindent {\em The minor comments were taken into consideration and typos were corrected. Some specific details are given below.}
 
@@
 Therefore, the support of the approximated distribution is not restricted to $(\tilde{a}^k_{i})_{1\le k \le N}$ to overcome particle depletion.
-The marginalization over the linear states  builds new trajectories of the form $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters.
+Marginalization over the linear states  builds new trajectories of the form $(a^k_{1:i-1},j,\tilde{a}^\ell_{i+1:n})_{1\le k,\ell \le N}$ for all $1\le j \le J$ which have not been sampled by the forward and backward filters.
 Although this does not overcome the overall degeneracy issue, numerical experiments highlight that this improves the performance of all Rao-Blackwellized SMC smoothers.
@@
 
-\noindent All minor comments were taken into consideration and typos were corrected.
+\noindent {\em All minor comments were taken into consideration and typos were corrected.}
 \end{document}

### 2017-05-15 TexArxiv/clgm.tex
Final modifications for revision

@@
 \end{figure}
-In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for each algorithm to illustrate that both estimators have similar variability. 
+In addition, Figure~\ref{fig:ffbs:var} displays the empirical variance of the estimation for each algorithm.
+Figure~\ref{fig:ffbs:err} and  Figure~\ref{fig:ffbs:var} illustrate that in both cases the additional rejuvenation step improves the accuracy and the variability of SMC smoothers.
+In addition, even with a sharp choice for the artificial distributions $\gamma_i$, $1\le i\le n$,  FFBS based methods outperform two-filter based smoothers.
 \begin{figure}

### 2017-05-17 TexArxiv/answer_referees.tex
Final modifications #3

@@
 \noindent Even is the paper is not a survey of all Rao-Blackwellized methods for CLGM anymore, some additional numerical experiments were added in Section~3.
-Each simulation study which was previously given in Section~3 was devoted either to FFBS or two-filter based methods.
+The simulation study which was previously given in Section~3 was devoted either to FFBS or two-filter based methods.
 Now, both algorithms are used and compared to their counterpart with rejuvenation.
@@
 This is now stated clearly in the paper.
+}
+
+\vspace{.5cm}
+
+\noindent p.7, L.51: Is $\Omega_i^{-1}$, $\Lambda_i$ correct? In the paper by Lindsten it is
+$\Omega_i$, $\Lambda_i^{-1}$.
+
+\vspace{.2cm}
+
+\noindent {\em 
+The equation is correct. The difference comes from the fact that $\|z\|_{A} = z'Az$ in the paper by Linddsten and $\|z\|_{A} = z'A^{-1}z$ in our paper.
 }

### 2017-05-17 TexArxiv/clgm.tex
Final modifications #3

@@
 To reduce Monte Carlo variance of smoothers, it is typical in these models to use Rao-Blackwellization: particle approximation is used to sample sequences of hidden regimes while the Gaussian states are explicitly integrated conditional on the sequence of regimes and observations, using variants of the Kalman filter / smoother.
-The first successful attempt to use Rao-Blackwellization for smoothing has been presented in  \cite{briers:doucet:maskell:2010} which extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filters / smoothers.
-More recently, \cite{sarkka:bunch:godsill:2012,lindsten:bunch:godsill:schon:2013,lindsten:bunch:sarkka:schon:godsill:2015} proposed to use a  forward backward decomposition of smoothing distributions, mimicking the Rauch-Tung-Striebel smoother for the regimes combined with backward Kalman updates.
+The first successful attempt to use Rao-Blackwellization for smoothing extends the Bryson-Frazier smoother for Gaussian linear state space models using the generalized two-filter formula together with Kalman filters / smoothers.
+More recently, a  forward backward decomposition of smoothing distributions mimicking the Rauch-Tung-Striebel smoother for the regimes combined with backward Kalman updates has been introduced.
 This paper investigates the benefit of introducing additional rejuvenation steps in all these algorithms to sample at each time instant new  regimes conditional on the forward and backward particles.
@@
 \item $(H_j)_{1\le j\le J}$ are $\dimz \times \dimz$ positive-definite matrices, $(d_j)_{1\le j\le J}$ $\dimz$-dimensional vectors and $(T_j)_{1\le j\le J}$ $\dimz \times \dimz$ positive-definite matrices.
-\item $Z_1$ is a $\dimz$-dimensional Gaussian random with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_i)_{i\ge 2}$.
+\item $Z_1$ is a $\dimz$-dimensional Gaussian random variable with mean $\mu_1$ and variance $\Sigma_1$ independent of $(\varepsilon_i)_{i\ge 2}$.
 \end{enumerate}
@@
 
-To exploit this specific structure, it has been suggested in the pioneering works of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) to sample the regimes with the Kalman filter to compute the conditional distribution of the states sequence $(Z_i)_{1\le i\le n}$ conditional on the regimes and on the observations.
+To exploit this specific structure, it has been suggested in the pioneering works of \cite{chen:liu:2000,doucet:godsill:andrieu:2000} to solve the filtering problem by combining Sequential Monte Carlo (SMC) methods to sample the regimes with the Kalman filter to compute the conditional distribution of the states sequence $(Z_i)_{1\le i\le n}$ conditional on the regimes and on the observations.
 This is a specific instance of Rao-Blackwellized Monte Carlo filters, often referred to as the \emph{Mixture Kalman Filter}.
@@
 A drawback of this method is that only ancestral paths that have been selected using the importance weights $(\omega^k_{i-1})_{1\le k \le N}$ are extended at time $i$.
-Following \cite{barembruch:garivier:moulines:2008}, this may be improved by considering all the offsprings of all ancestral trajectories $\{a_{1:i-1}^k\}_{1\le k \le N}$.
+Following \cite{barembruch:garivier:moulines:2008}, this may be improved by considering all the offsprings of all ancestral trajectories $(a_{1:i-1}^k)_{1\le k \le N}$.
 Each ancestral path has $J$ offsprings at time $i$, it is thus necessary to choose a given number of trajectories at time $i$ (for instance $N$) among the $NJ$ possible paths.
@@
 This means that an ancestral path $a^{\star}_{1:i-1}$ is sampled in $(a^k_{1:i-1})_{1\le k \le N}$ with weights $(\widetilde{\vartheta}^k_{i-1})_{1\le k \le N}$ and $a^{\star}_i$ is sampled from $\widetilde{q}(\cdot|a^{\star}_{1:i-1},a_{i+1:n},y_{1:n})$.
-Then the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio.
+Then, the proposed sequence $a^{\star}_{1:i}$ is accepted or rejected using the usual Metropolis-Hastings acceptance ratio.
 The choice of MCMC rejuvenation has interesting practical consequences as the computation of the acceptance ratio only requires to compute the posterior probability \eqref{eq:posterior:FFBS} for the proposed sequence $a^{\star}_{1:i}$ while our technique is based on the computation of  \eqref{eq:posterior:FFBS} for all combinations of sequences $(a^{k}_{1:i})_{1\le k \le N}$ and states $a_i\in\{1,\ldots,J\}$.
@@
 This section highlights the improvements brought by  the additional Rao-Blackwellization steps for the two-filter and the FFBS approximations of the marginal smoothing distributions in the case where the number of states is $J=2$.
-In each setting the transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data, see Section~\ref{sec:exp}.
+The transition matrix $Q$ is such that the probability of switching from one regime to the other is small, as expected for the WTI crude oil data, see Section~\ref{sec:exp}.
 First, the algorithms are applied to a simple one-dimensional model with:
@@
 \]
-The original FFBS algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015} and the FFBS algorithm with rejuvenation proposed in this paper are used with $N = \tilde{N} = 25$ and the estimated posterior probabilities $\mathbb{P}(a_k=1|Y_{1:n})$.
+The original FFBS algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015} and the FFBS algorithm with rejuvenation proposed in this paper are used with $N = \tilde{N} = 25$.
 For comparable computational costs, the two-filter method of \cite{briers:doucet:maskell:2010} and the method with rejuvenation are run with $N=100$.
+The artificial distributions are chosen as $\gamma^{}_i (a_i, z_i) = p^N(a_{i},z_{i}|y_{1:i-1})$ where $p^N(a_{i},z_{i}|y_{1:i-1})$ is defined by \eqref{eq:rN}.
 All these algorithms are compared to the estimation obtained with the proposed FFBS algorithm with rejuvenation and $5000$ particles considered as a benchmark value.
@@
 Figure~\ref{fig:ffbs:err} and  Figure~\ref{fig:ffbs:var} illustrate that in both cases the additional rejuvenation step improves the accuracy and the variability of SMC smoothers.
-In addition, even with a sharp choice for the artificial distributions $\gamma_i$, $1\le i\le n$,  FFBS based methods outperform two-filter based smoothers.
+In addition, even with a sharp choice for the artificial distributions $\gamma_i$, $1\le i\le n$,  FFBS based methods outperform two-filter based smoothers for this model.
 \begin{figure}
@@
 \end{enumerate}
-All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our two-filter algorithm with rejuvenation algorithm to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
+All the conditional expectations involved in $Q(\theta,\theta_p)$ are approximated using our two-filter algorithm with rejuvenation to define the SMC approximation  $\theta\mapsto Q^N(\theta,\theta_p)$ of $\theta\mapsto Q(\theta,\theta_p)$.
 As the function $\theta\mapsto Q^N(\theta,\theta_p)$ cannot be maximized analytically, the M-step is performed numerically using the Covariance Matrix Adaptation Evolution Strategy (CMA-ES) introduced in \cite{hansen:ostermeier:2001}.

### 2017-07-03 TexSubmission/clgm.tex
Submitted version

--- 2017-05-17 TexArxiv/clgm.tex
@@
-\include{clgm_def}
+\documentclass[nolayout]{article}
+
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+\usepackage{amsmath,amsthm}
+\usepackage{geometry}
+\usepackage[textwidth=4cm,textsize=footnotesize]{todonotes}
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\usepackage{amssymb,color,bbm,xargs}
+\usepackage{graphicx}
+\usepackage[active]{srcltx}
+\usepackage{ifthen}
+\usepackage{enumerate}
+\usepackage{color}
+\usepackage{accents}
+\usepackage{dsfont}
+\usepackage[ruled,vlined]{algorithm2e}
+\usepackage{subfig}
+\usepackage{multirow}
+\usepackage{placeins}
+
+\usepackage{aliascnt}
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}
+\newtheorem{assumption}{H\hspace{-3pt}}
+\newtheorem*{assumption*}{A}
+\newtheorem{assumptionDelta}{H\hspace{-3pt}(\Delta)}
+\newtheorem{assumptionsmallSetD}{H\hspace{-3pt}(\smallSetD \times \smallSetD)}
+\newtheorem{assumptionAR}{AR\hspace{-3pt}}
+\newtheorem{assumptionCN}{CN\hspace{-3pt}}
+
+\newaliascnt{proposition}{theorem}
+\newtheorem{proposition}[proposition]{Proposition}
+\aliascntresetthe{proposition}
+
+\newaliascnt{lemma}{theorem}
+\newtheorem{lemma}[lemma]{Lemma}
+\aliascntresetthe{lemma}
+\newaliascnt{corollary}{theorem}
+\newtheorem{corollary}[corollary]{Corollary}
+\aliascntresetthe{corollary}
+
+\newtheorem{hypothese}{Hypothesis}
+\newtheorem{fact}[theorem]{Fact}
+
+\theoremstyle{definition}
+\newaliascnt{definition}{theorem}
+\newtheorem{definition}[definition]{Definition}
+\aliascntresetthe{definition}
+
+\newaliascnt{remark}{theorem}
+\newtheorem{remark}[remark] {Remark}
+\aliascntresetthe{remark}
+
+\usepackage{hyperref}
+\usepackage[nameinlink,noabbrev]{cleveref}
+
+\providecommand*{\definitionautorefname}{Definition}
+\providecommand*{\lemmaautorefname}{Lemma}
+\providecommand*{\exampleautorefname}{Example}
+\providecommand*{\propositionautorefname}{Proposition}
+\providecommand*{\procautorefname}{Procedure}
+\providecommand*{\exerciseautorefname}{Exercise}
+\providecommand*{\corollaryautorefname}{Corollary}
+\providecommand*{\remarkautorefname}{Remark}
+\providecommand*{\algorithmautorefname}{Algorithm}
+\providecommand*{\assumptionautorefname}{H\hspace*{-3pt}}  
+\providecommand*{\assumptionARautorefname}{AR\hspace*{-3pt}}
+\providecommand*{\assumptionCNautorefname}{CN\hspace*{-3pt}}
+
+\newcommand*{\largedot}{\cdot}
+
+\def\Rset{\mathbb{R}}
+\def\Nset{\mathbb{N}}
+\def\nset{\mathbb{N}}
+\def\rset{\mathbb{R}}
+\def\varb{b}
+\newcommand{\1}{\mathbbm{1}}
+
+\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
+\newcommand{\ceil}[1]{\left\lceil #1 \right\rceil}
+
+\def\ie{\textit{i.e.}}
+\newcommand{\coint}[1]{\left[#1\right)}
+\newcommand{\ocint}[1]{\left(#1\right]}
+\newcommand{\ooint}[1]{\left(#1\right)}
+\newcommand{\ccint}[1]{\left[#1\right]}
+\newcommand{\Wienerspace}{\mathbf{W}}
+
+\def\PE{\mathbb{E}}
+\newcommandx{\PEt}[2][1=]{\mathbb{E}_{#1}\left[#2\right]}
+\def\rmd{\mathrm{d}}
+\def\rme{\mathrm{e}}
+\def\target{\pi}
+\def\dim{d}
+\def\score{\accentset{\largedot}{\ell}}
+\def\Vdot{\dot{V}}
+\def\unitarget{f}
+\def\rmL{\mathrm{L}}
+\def\dimy{\mathsf{p}}
+\def\dimz{\mathsf{m}}
+\def\bfX{\boldsymbol{X}}
+\def\xbf{\boldsymbol{x}}
+\def\ybf{\boldsymbol{y}}
+\def\Xbf{\boldsymbol{X}}
+\def\G{\mathcal{G}}
+\def\barG{\overline{G}}
+\def\barH{\overline{H}}
+\def\Gammabf{\mathbf{\Gamma}}
+\def\Ybf{\boldsymbol{Y}}
+\newcommand{\norm}[1]{\left\Vert #1 \right \Vert}
+\newcommand{\plusinfty}{+ \infty}
+\newcommand{\defEns}[1]{\left \{ #1 \right \}}
+\newcommand{\abs}[1]{\left\vert #1 \right\vert}
+\def\Zbf{\boldsymbol{Z}}
+\def\eqsp{\,}
+\def\leb{\lambda}
+\def\limd{\underset{d\to +\infty}{\longrightarrow}}
+\newcommand{\normMat}[2]{\left\|#2\right\|_{#1}}
+
+\newcommand{\parenthese}[1]{\left( #1 \right)}
+\newcounter{hypH}
+\newenvironment{hypH}{\refstepcounter{hypH}\begin{itemize}
+\item[{\bf H\arabic{hypH}}]}{\end{itemize}}
+\newcommand{\eqdef}{\ensuremath{:=}}
+\newcommand{\setAccept}{\mathcal{A}}
+\def\iid{\operatorname{i.i.d.}}
+\newcommand{\sachant}[1]{\left| #1 \right.}
+\newcommand{\setDisconDotV}{\mathcal{D}_{\dot{V}}}
+
+\def\restriction#1#2{\mathchoice
+              {\setbox1\hbox{${\displaystyle #1}_{\scriptstyle #2}$}
+              \restrictionaux{#1}{#2}}
+              {\setbox1\hbox{${\textstyle #1}_{\scriptstyle #2}$}
+              \restrictionaux{#1}{#2}}
+              {\setbox1\hbox{${\scriptstyle #1}_{\scriptscriptstyle #2}$}
+              \restrictionaux{#1}{#2}}
+              {\setbox1\hbox{${\scriptscriptstyle #1}_{\scriptscriptstyle #2}$}
+              \restrictionaux{#1}{#2}}}
+\def\restrictionaux#1#2{{#1\,\smash{\vrule height .8\ht1 depth .85\dp1}}_{\,#2}}
+
+\newcommandx\sequence[3][2=,3=]
+{\ifthenelse{\equal{#3}{}}{\ensuremath{\{ #1_{#2}\}}}{\ensuremath{( #1_{#2}, \eqsp #2 \in #3 )}}}
+\newcommandx{\sequencen}[2][2=n\in\nset]{\ensuremath{( #1, \eqsp #2 )}}
+\newcommandx\dsequence[4][3=k,4=\zset]{\ensuremath{( (#1_{#3}, #2_{#3}), \eqsp #3 \in #4 )}}
+\newcommandx{\as}[1][1=\PP]{\ensuremath{#1\--\mathrm{a.s.}}}
+\newcommand{\eqspp}{\ \ }
+
+\def\canonicalFiltration{\mathscr{B}}
+
+\def\tildeX{\widetilde{X}}
+\def\tildeY{\widetilde{Y}}
+\def\barX{\bar{X}}
+\def\barY{\bar{Y}}
+\def\barS{\bar{S}}
+\newcommand{\coupling}[2]{\mathcal{C}(#1,#2)}
+\newcommand\proba[1]{\mathbb{P}\left[ #1 \right]}
+\newcommand\expe[1]{\mathbb{E}\left[ #1 \right]}
+\def\osc{\operatorname{osc}}
+\def\borel{\mathcal{B}}
+\def\mcf{\mathcal{F}}
+
 \begin{document}
@@
 The initial transition probability in CMA-ES is chosen as $Q(1,1) = 0.98, \; Q(2,2) = 0.97$ and $\pi_1 = \pi_2 = 0.5$ where the number $1$ represents the backwardation regime and $2$ represents the contango regime.
-The other parameters are initialized as shown in the following table:
-\begin{figure}[h!]
+The other parameters are initialized as shown in Table~\ref{fig:CMEinitialestimates}.
+\begin{table}[h!]
 	\centering
@@
 	\caption{Initial values for the EM algorithm.}
-\end{figure}\\
+	\label{fig:CMEinitialestimates}
+\end{table}
 The number of particles is set to $N = 100$, $\tau  = 1/52$. 
@@
 
+\subsection*{Competing interests}
+The authors have no competing interests.
+
+\subsection*{Funding}
+This work has been developed during a three year Ph.D. at T\'el\'ecom ParisTech and Lunalogic.
+
+\subsection*{Authors' contributions}
+All the authors have contributed to the conception of the algorithms and to the redaction of the manuscript.
+
 \appendix
@@
 
-\bibliography{./clgm_bib}
+\newcommand{\etalchar}[1]{$^{#1}$}
+\begin{thebibliography}{FGDW02}
+
+\bibitem[Alm16]{almansour:2016}
+A.~Almansour.
+\newblock Convenience yield in commodity price modeling: A regime switching
+  approach.
+\newblock {\em Energy Economics}, 53:238--247, 2016.
+
+\bibitem[Bar06]{barber:2006}
+D.~Barber.
+\newblock {Expectation correction for smoothed inference in switching linear
+  dynamical systems}.
+\newblock {\em Journal of {M}achine {L}earning {R}esearch}, 7:2515--2540, 2006.
+
+\bibitem[BDM10]{briers:doucet:maskell:2010}
+M.~Briers, A.~Doucet, and S.~Maskell.
+\newblock Smoothing algorithms for state-space models.
+\newblock {\em {A}nn. {I}nst. {S}tat. {M}ath}, 62(1):61--89, 2010.
+
+\bibitem[BF63]{bryson:frazier:1963}
+A.E. Bryson and M.~Frazier.
+\newblock {Smoothing for linear and nonlinear dynamic systems}.
+\newblock {\em {Proc. Optimum Sys. Synthesis Conf.}}, 1963.
+
+\bibitem[BGM08]{barembruch:garivier:moulines:2008}
+S.~Barembruch, A.~Garivier, and E.~Moulines.
+\newblock {On optimal sampling for particle filtering in digital
+  communication}.
+\newblock {\em IEEE 9th Workshop on Signal Processing Advances in Wireless},
+  pages 634--638, 2008.
+
+\bibitem[CL00]{chen:liu:2000}
+R.~Chen and J.S. Liu.
+\newblock Mixture kalman filters.
+\newblock {\em Journal of the Royal Statistical Society B}, 62:493--508, 2000.
+
+\bibitem[DGA00]{doucet:godsill:andrieu:2000}
+A.~Doucet, S.~Godsill, and C.~Andrieu.
+\newblock On sequential monte carlo sampling methods for bayesian filtering.
+\newblock {\em Statistics and computing}, 10:197--208, 2000.
+
+\bibitem[DGK01]{doucet:gordon:krishnamurthy:2001}
+A.~Doucet, N.~Gordon, and V.~Krishnamurthy.
+\newblock Particle filters for state estimation of jump {M}arkov linear
+  systems.
+\newblock {\em {IEEE T}ransactions on {S}ignal {P}rocessing}, 49(3):613--624,
+  2001.
+
+\bibitem[DLR77]{dempster:laird:rubin:1977}
+A.~P. Dempster, N.~M. Laird, and D.~B. Rubin.
+\newblock Maximum likelihood from incomplete data via the {EM} algorithm.
+\newblock {\em Journal of the Royal Statistical Society B}, 39(1):1--38 (with
+  discussion), 1977.
+
+\bibitem[DM13]{delmoral:2013}
+P.~Del~Moral.
+\newblock {\em Mean field simulation for Monte Carlo integration}.
+\newblock Chapman \& Hall / CRC Monographs on Statistics \& Applied
+  Probability, 2013.
+
+\bibitem[FC03]{fearnhead:clifford:2003}
+P.~Fearnhead and P.~Clifford.
+\newblock {On-line inference for hidden {M}arkov models via particle filters}.
+\newblock {\em {J.R. Statist. Soc. B}}, 65:887--899, 2003.
+
+\bibitem[FGDW02]{fong:godsill:doucet:west:2002}
+W.~Fong, S.J. Godsill, A.~Doucet, and M.~West.
+\newblock {Monte Carlo smoothing with application to audio signal enhancement}.
+\newblock {\em IEEE Transactions on {S}ignal {P}rocessing}, 50(2):438--449,
+  2002.
+
+\bibitem[FWT10]{fearnhead:wyncoll:tawn:2010}
+P.~Fearnhead, D.~Wyncoll, and J.~Tawn.
+\newblock {A sequential smoothing algorithm with linear computational cost}.
+\newblock {\em {Biometrika}}, 97:447--464, 2010.
+
+\bibitem[GE90]{gibson:schwartz:1990}
+R.~Gibson and Schwartz E.S.
+\newblock Stochastic convenience yield and the pricing of oil contingent
+  claims.
+\newblock {\em Journal of finance}, 45(3):959--976, 1990.
+
+\bibitem[HK98]{huerzeler:kunsch:1998}
+M.~H{\"u}rzeler and H.~R. K{\"u}nsch.
+\newblock {M}onte {C}arlo approximations for general state-space models.
+\newblock {\em J. Comput. Graph. Statist.}, 7:175--193, 1998.
+
+\bibitem[HK04]{hansen:kern:2004}
+N.~Hansen and S.~Kern.
+\newblock {Evaluating the CMA Evolution Strategy on Multimodal Test Functions}.
+\newblock {\em Eighth International Conference on Parallel Problem Solving from
+  Nature}, 72:337--354, 2004.
+
+\bibitem[HO01]{hansen:ostermeier:2001}
+N.~Hansen and A.~Ostermeier.
+\newblock {Completely derandomized self-adaptation in evolution strategies}.
+\newblock {\em Evolutionary {C}omputation}, 9(2):159--195, 2001.
+
+\bibitem[Kal60]{kalman:1960}
+R.E. Kalman.
+\newblock {A new approach to linear filtering and prediction problems}.
+\newblock {\em J. {B}asic {E}ngng}, 82:35--45, 1960.
+
+\bibitem[Kim94]{kim:1994}
+CJ. Kim.
+\newblock Dynamic linear models with markov-switching.
+\newblock {\em Journal of Econometrics}, 60(1-2):1--22, 1994.
+
+\bibitem[LBGS13]{lindsten:bunch:godsill:schon:2013}
+F.~Lindsten, P.~Bunch, S.J. Godsill, and T.B. Schon.
+\newblock Rao-{B}lackwellized particle smoothers for mixed linear/nonlinear
+  state-space models.
+\newblock {\em Proceedings of the 38th {IEEE I}nternational {C}onference on
+  {A}coustics, {S}peech and {S}ignal {P}rocessing ({ICASSP})}, 2013.
+
+\bibitem[LBS{\etalchar{+}}16]{lindsten:bunch:sarkka:schon:godsill:2015}
+F.~Lindsten, P.~Bunch, S.~Sarkka, T.B. Schon, and S.J. Godsill.
+\newblock Rao-{B}lackwellized particle smoothers for conditionally linear
+  {G}aussian models.
+\newblock {\em IEEE {J}ournal of {S}elected {T}opics in {S}ignal {P}rocessing},
+  10(2):353--365, 2016.
+
+\bibitem[RST65]{rauch:striebel:tung:1965}
+A.E. Rauch, C.T. Striebel, and F.~Tung.
+\newblock {Maximum likelihood estimates of linear dynamic systems}.
+\newblock {\em American Institute of Aeronautics and Astronautics journal},
+  3(8):1445--1450, 1965.
+
+\bibitem[Sar13]{sarkka:2013}
+S.~Sarkka.
+\newblock {\em Bayesian filtering and smoothing}.
+\newblock Cambridge {U}niversity {P}ress, 2013.
+
+\bibitem[SBG12]{sarkka:bunch:godsill:2012}
+S.~Sarkka, P.~Bunch, and S.J. Godsill.
+\newblock A backward-simulation based {R}ao-{B}lackwellized particle smoother
+  for conditionally linear {G}aussian models.
+\newblock {\em Proceedings of the 16th {IFAC S}ymposium on {S}ystem
+  {I}dentification {(SYSID)}}, 2012.
+
+\bibitem[Sch97]{schwartz:1997}
+E.S. Schwartz.
+\newblock The stochastic behaviour of commodity prices: Implications for
+  pricing and hedging.
+\newblock {\em The {J}ournal of {F}inance}, 3:923--973, 1997.
+
+\bibitem[SGN05]{schon:gustafsson:nordlund:2005}
+T.~Schon, F.~Gustafsson, and P.-J. Nordlund.
+\newblock Marginalized particle filters for mixed linear/nonlinear state-space
+  models.
+\newblock {\em {IEEE T}ransactions on {S}ignal {P}rocessing}, 53(7):2279--2289,
+  2005.
+
+\end{thebibliography}
+
+%\bibliography{./clgm_bib}
 \end{document}

### 2017-07-03 TexArxiv/clgm.tex
ArXiv standards

@@
 
-\author{Ngoc Minh Nguyen\footnote{LTCI, CNRS and T\'el\'ecom ParisTech, 46 rue Barrault 75634 Paris Cedex 13, France.}, Sylvain {L}e {C}orff\footnote{Laboratoire de Math\'ematiques d'Orsay, Univ.
-Paris-Sud, CNRS, Universit\'e Paris-Saclay, 91405 Orsay, France.} and Eric Moulines\footnote{Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique, France.}}
+\author{Ngoc Minh Nguyen\footnote{LTCI, CNRS and T\'el\'ecom ParisTech, 46 rue Barrault 75634 Paris Cedex 13, France.}\and Sylvain {L}e {C}orff\footnote{Laboratoire de Math\'ematiques d'Orsay, Univ.
+Paris-Sud, CNRS, Universit\'e Paris-Saclay, 91405 Orsay, France.}\and Eric Moulines\footnote{Centre de Math\'ematiques Appliqu\'ees, UMR 7641, Ecole Polytechnique, France.}}
 

### 2017-07-03 TexArxiv/clgm.tex
Minor modifications for ArXiv

@@
 \centering
-\includegraphics[scale=.3]{err_est}
+\includegraphics[scale=.3]{err_est.eps}
 \caption{Posterior probabilities estimation error for all algorithms.}
@@
 \centering
-\includegraphics[scale=.3]{var2}
+\includegraphics[scale=.3]{var2.eps}
 \caption{Empirical variances of the estimation of $\mathbb{P}(a_k=1|Y_{1:n})$ for all algorithms.}
@@
 \centering
-\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure}
+\includegraphics[scale=.6,angle=90]{LogPrice_TermStructure.eps}
 \caption{Log-price (red line) and slope of future curves (blue line).}
@@
 	\centering
-	\includegraphics[scale=.75,angle=90]{Posterior_Data}
+	\includegraphics[scale=.75,angle=90]{Posterior_Data.eps}
 	\caption{Posterior probability (triangle black) and slope of future curves (blue line).}
@@
 	\centering
-	\includegraphics[scale=.75]{Posterior_Data_LogPrice}
+	\includegraphics[scale=.75]{Posterior_Data_LogPrice.eps}
 	\caption{Posterior probability (triangle black) and slope of future curves (blue line).}
@@
 \newblock Smoothing algorithms for state-space models.
-\newblock {\em {A}nn. {I}nst. {S}tat. {M}ath}, 62(1):61--89, 2010.
+\newblock {\em {A}nnals of the {I}nstitute of {S}tatistical {M}athematics}, 62(1):61--89, 2010.
 
@@
 \newblock {Smoothing for linear and nonlinear dynamic systems}.
-\newblock {\em {Proc. Optimum Sys. Synthesis Conf.}}, 1963.
+\newblock {\em {Proceedings of the optimum system synthesis conference}}, 1963.
 
@@
 \newblock Maximum likelihood from incomplete data via the {EM} algorithm.
-\newblock {\em Journal of the Royal Statistical Society B}, 39(1):1--38 (with
+\newblock {\em Journal of the Royal Statistical Society: Series B}, 39(1):1--38 (with
   discussion), 1977.
@@
 \newblock {On-line inference for hidden {M}arkov models via particle filters}.
-\newblock {\em {J.R. Statist. Soc. B}}, 65:887--899, 2003.
+\newblock {\em {Journal of the Royal Statistical Society: Series B}}, 65:887--899, 2003.
 
@@
 \newblock {M}onte {C}arlo approximations for general state-space models.
-\newblock {\em J. Comput. Graph. Statist.}, 7:175--193, 1998.
+\newblock {\em Journal of {C}omputational and {G}raphical {S}tatistics}, 7:175--193, 1998.
 
@@
 \newblock {A new approach to linear filtering and prediction problems}.
-\newblock {\em J. {B}asic {E}ngng}, 82:35--45, 1960.
+\newblock {\em Journal of {B}asic {E}ngineering}, 82:35--45, 1960.
 
